### 2024-04-17 manuscript/manuscript.qmd
add manuscript template

---
title: Short Paper
subtitle: A Short Subtitle
author:
  - name: Anders Lorentzen Kolstad
    email: anders.kolstad@nina.no
    orchid: https://orcid.org/0000-0002-9623-9491
    affiliations: 
        - id: nina
          name: Norwegian Institute for Nature Research
          department: Department of Terrestrial Ecology
          address: Pb 5685 Torgarden
          city: Trondheim
          postal-code: 7485
    attributes:
        corresponding: true
  - name: Hanno Sandvik
    email: hanno.sandvik@nina.no
    orchid: https://orcid.org/0000-0002-5889-1606
    affiliations:
      - id: nina
        name: Norwegian Institute for Nature Research
        department: Department of Aquatic Biodiversity
        address: Pb 5685 Torgarden
        city: Trondheim
        postal-code: 7485
  - name: Chloé R. Nater
    email: chloe.nater@nina.no
    orchid: https://orcid.org/0000-0002-7975-0108
    affiliations:
        - ref: nina
abstract: |
  This is the abstract. Lorem ipsum dolor sit amet, consectetur adipiscing elit. Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit. Nam placerat turpis purus, eu tristique ex tincidunt et. Mauris sed augue eget turpis ultrices tincidunt. Sed et mi in leo porta egestas. Aliquam non laoreet velit. Nunc quis ex vitae eros aliquet auctor nec ac libero. Duis laoreet sapien eu mi luctus, in bibendum leo molestie. Sed hendrerit diam diam, ac dapibus nisl volutpat vitae. Aliquam bibendum varius libero, eu efficitur justo rutrum at. Sed at tempus elit.
keywords: 
  - keyword1
  - keyword2
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Ecological Indicators
      formatting: preprint
      model: 3p
      cite-style: super
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
execute:
  warning: false
  message: false
---

# Introduction

# Methods

# Results

# Discussion

# References {-}

### 2024-04-17 R/example1.R
add example flowcharts

@@ -1,0 +1,147 @@
+
+
+
+library(DiagrammeR)
+#library(DiagrammeRsvg)
+#library(magrittr)
+#library(rsvg)
+
+example1 <- grViz("digraph{
+  
+   
+   graph [
+            compound = true, 
+            nodesep = .8, 
+            ranksep = .8,
+            color = crimson, 
+            rankdir = LR
+        ]
+
+  
+    node [
+            fontname = Helvetica, 
+            fontcolor = darkslategray,
+            shape = rectangle, 
+            fixedsize = true, 
+            width = 1,
+            color = darkslategray
+        ]
+        
+        
+    edge [
+            color = black
+        ]
+        
+        
+   subgraph clusterTop {
+      graph[color = black]
+      
+    subgraph cluster0 {
+      peripheries=0;
+       
+       subgraph cluster0_1 {
+        label='EA-1';
+        labeljust=l;
+        Area1[label = 'Value = 5\nRef. = 10']
+       }
+       
+       subgraph cluster0_2 {
+        label='EA-2';
+        labeljust=l;
+        Area2[label = 'Value = 1\nRef. = 4']
+       }
+    }
+    
+    subgraph clusterScaledUpper {
+      graph[style=dashed]
+      style=filled; color=grey90;
+      
+    subgraph cluster1 {
+      peripheries=0;
+        
+        subgraph cluster1_1 {
+          label='EA-1';
+          labeljust=l;
+          Area1_2[label = '.5']
+        }
+        
+        subgraph cluster1_2 {
+          label='EA-2';
+          labeljust=l;
+          Area2_2[label = '.25']
+        }
+    }   
+    
+    
+    subgraph cluster2 {
+      peripheries=0;
+      label = EAA;
+      labeljust=l;
+      Norge[label = '.375']
+    }
+    }
+    
+    Area1 -> Area1_2 -> Norge
+    Area2 -> Area2_2 -> Norge
+  
+   }  
+   
+   
+   
+   
+   subgraph clusterBottom {
+    graph[color = black];
+
+    
+    subgraph cluster0b {
+       peripheries=0;
+       
+       
+       subgraph cluster0_1b {
+        label='EA-1';
+        labeljust=l;
+        Area1b[label = 'Value = 5\nRef = 10']
+       }
+       
+       subgraph cluster0_2b {
+        label='EA-2';
+        labeljust=l;
[…]

### 2024-04-17 R/example2.R

@@ -1,0 +1,117 @@
+
+
+
+library(DiagrammeR)
+#library(DiagrammeRsvg)
+#library(magrittr)
+#library(rsvg)
+
+example2 <- 
+  grViz("digraph{
+  
+   
+   graph [
+            compound = true, 
+            nodesep = .8, 
+            ranksep = .8,
+            color = crimson, 
+            rankdir = LR
+        ]
+
+  
+    node [
+            fontname = Helvetica, 
+            fontcolor = darkslategray,
+            shape = rectangle, 
+            fixedsize = false, 
+            width = 1.5,
+            heigth=3,
+            color = darkslategray
+        ]
+        
+        
+    edge [
+            color = black
+        ]
+        
+        
+   subgraph clusterAll1 {
+      graph[color = black]
+      peripheries=0;
+      
+    subgraph cluster0 {
+      peripheries=0;
+       
+       subgraph clusterSubReg1 {
+        peripheries=1;
+        label='EA-1';
+        labeljust=l;
+        Area1A[label = 'Indicator = A\nValue = 5\nRef. = 10', color=blue]
+        Area1B[label = 'Indicator = B\nValue = 5\nRef. = 25']
+        Area1_2A[label = 'Indicator A = .5',  style=filled, color=grey90]
+        Area1_2B[label = 'Indicator B = .2', style=filled, color=grey90]
+        Area1A -> Area1_2A
+        Area1B -> Area1_2B
+        
+        SR1[label='EA-1 = 0.35', style=filled, color=grey90]
+        Area1_2A -> SR1
+        Area1_2B -> SR1
+
+       }
+       
+       subgraph clusterSubReg2 {
+       peripheries=1;
+        label='EA-2';
+        labeljust=l;
+        Area2A[label = 'Indicator= A\nValue = 1\nRef. = 4', color=blue]
+        Area2B[label = 'Indicator= B\nValue = 1\nRef. = 10']
+        Area2_2A[label = 'Indicator A = .25', style=filled, color=grey90]
+        Area2_2B[label = 'Indicator B = .1', style=filled, color=grey90]
+        Area2A -> Area2_2A
+        Area2B -> Area2_2B
+        
+        SR2[label='EA-2 = 0.18', style=filled, color=grey90]
+        Area2_2A -> SR2
+        Area2_2B -> SR2
+    
+       }
+    }
+    
+   
+ 
[…]

### 2024-04-17 R/example3.R

@@ -1,0 +1,88 @@
+
+
+
+library(DiagrammeR)
+
+# Example with two indicator and two EAs
+# where we scale variables at the EA level
+# and calculate an index at the EA level
+# and then aggregate the indeces for the EAA level
+
+example3 <- grViz("digraph{
+  
+   
+   graph [
+            compound = true, 
+            nodesep = .8, 
+            ranksep = .8,
+            color = crimson, 
+            rankdir = LR
+        ]
+
+  
+    node [
+            fontname = Helvetica, 
+            fontcolor = darkslategray,
+            shape = rectangle, 
+            fixedsize = true, 
+            width = 1,
+            color = darkslategray
+        ]
+        
+        
+    edge [
+            color = black
+        ]
+        
+# Orange part --------------------------
+
+   subgraph clusterTop {
+      graph[color = orange]
+      
+    subgraph cluster0 {
+      graph[color = green]
+      peripheries=0;
+       
+       subgraph cluster0_1 {
+        label='EA-1';
+        labeljust=l;
+        peripheries=1;
+        V2_EA1[label = 'Value = 5\nRef. = 25', color = orange, style = filled, fontcolor = white]
+        V1_EA1[label = 'Value = 5\nRef. = 10', color = blue, style = filled, fontcolor = white]
+        I1_EA1[label = ' .5', color = blue, style = filled, fontcolor = white]
+        I2_EA1[label = ' .2', color = orange, style = filled, fontcolor = white]
+        X_EA1[label = ' .35', color = grey, style = filled]
+        V1_EA1 -> I1_EA1 -> X_EA1
+        V2_EA1 -> I2_EA1 -> X_EA1
+        
+       }
+       
+       subgraph cluster0_2 {
+        label='EA-2';
+        labeljust=l;
+        peripheries=1;
+        V2_EA2[label = 'Value = 1\nRef. = 10', color = orange, style = filled, fontcolor = white]
+        V1_EA2[label = 'Value = 1\nRef. = 4', color = blue, style = filled, fontcolor = white]
+        I1_EA2[label = ' .25', color = blue, style = filled, fontcolor = white]
+        I2_EA2[label = ' .1', color = orange, style = filled, fontcolor = white]
+        X_EA2[label = ' .175', color = grey, style = filled]
+        V1_EA2 -> I1_EA2 -> X_EA2
[…]

### 2024-04-17 R/example4.R

@@ -1,0 +1,92 @@
+
+
+
+library(DiagrammeR)
+
+# Example with two indicator and two EAs
+# where we scale variables at the EA level
+# and calculate an index at the EA level
+# and then aggregate the indeces for the EAA level
+# giving equal weight to EAs
+
+example4 <-  grViz("digraph{
+  
+   
+   graph [
+            compound = true, 
+            nodesep = .8, 
+            ranksep = .8,
+            color = crimson, 
+            rankdir = LR
+        ]
+
+  
+    node [
+            fontname = Helvetica, 
+            fontcolor = darkslategray,
+            shape = rectangle, 
+            fixedsize = true, 
+            width = 1,
+            color = darkslategray
+        ]
+        
+        
+    edge [
+            color = black
+        ]
+        
+# Orange part --------------------------
+
+   subgraph clusterTop {
+      
+      graph[color = orange]
+       
+    subgraph cluster0 {
+      graph[color = green]
+      peripheries=0;
+       
+       subgraph cluster0_1 {
+        label='EA-1';
+        labeljust=l;
+        peripheries=1;
+        V1_EA1[label = 'Value = 5\nRef. = 10', color = blue, style = filled, fontcolor = white]
+        V2_EA1[label = 'Value = 5\nRef. = 25', color = orange, style = filled, fontcolor = white]
+        I1_EA1[label = ' .5', color = blue, style = filled, fontcolor = white]
+        I2_EA1[label = ' .2', color = orange, style = filled, fontcolor = white]
+        V1_EA1 -> I1_EA1
+        V2_EA1 -> I2_EA1
+        
+       }
+       
+       subgraph cluster0_2 {
+        label='EA-2';
+        labeljust=l;
+        peripheries=1;
+        V1_EA2[label = 'Value = 1\nRef. = 4', color = blue, style = filled, fontcolor = white]
+        V2_EA2[label = 'Value = 1\nRef. = 10', color = orange, style = filled, fontcolor = white]
+        I1_EA2[label = ' .25', color = blue, style = filled, fontcolor = white]
+        I2_EA2[label = ' .1', color = orange, style = filled, fontcolor = white]
+        V1_EA2 -> I1_EA2
+        V2_EA2 -> I2_EA2
+       }
+    }
+
[…]

### 2024-04-17 R/example5.R

@@ -1,0 +1,92 @@
+
+
+
+library(DiagrammeR)
+
+# Example with two indicator and two EAs
+# where we scale variables at the EA level
+# and calculate an index at the EA level
+# and then aggregate the indeces for the EAA level
+# giving UNequal weight to EAs
+
+example5 <-  grViz("digraph{
+  
+   
+   graph [
+            compound = true, 
+            nodesep = .8, 
+            ranksep = .8,
+            color = crimson, 
+            rankdir = LR
+        ]
+
+  
+    node [
+            fontname = Helvetica, 
+            fontcolor = darkslategray,
+            shape = rectangle, 
+            fixedsize = true, 
+            width = 1,
+            color = darkslategray
+        ]
+        
+        
+    edge [
+            color = black
+        ]
+        
+# Orange part --------------------------
+
+   subgraph clusterTop {
+      
+      graph[color = orange]
+       
+    subgraph cluster0 {
+      graph[color = green]
+      peripheries=0;
+       
+       subgraph cluster0_1 {
+        label='EA-1, 1 hectar';
+        labeljust=l;
+        peripheries=1;
+        V1_EA1[label = 'Value = 5\nRef. = 10', color = blue, style = filled, fontcolor = white]
+        V2_EA1[label = 'Value = 5\nRef. = 25', color = orange, style = filled, fontcolor = white]
+        I1_EA1[label = ' .5', color = blue, style = filled, fontcolor = white]
+        I2_EA1[label = ' .2', color = orange, style = filled, fontcolor = white]
+        V1_EA1 -> I1_EA1
+        V2_EA1 -> I2_EA1
+        
+       }
+       
+       subgraph cluster0_2 {
+        label='EA-2, 2 hectar';
+        labeljust=l;
+        peripheries=1;
+        V1_EA2[label = 'Value = 1\nRef. = 4', color = blue, style = filled, fontcolor = white]
+        V2_EA2[label = 'Value = 1\nRef. = 10', color = orange, style = filled, fontcolor = white]
+        I1_EA2[label = ' .25', color = blue, style = filled, fontcolor = white]
+        I2_EA2[label = ' .1', color = orange, style = filled, fontcolor = white]
+        V1_EA2 -> I1_EA2
+        V2_EA2 -> I2_EA2
+       }
+    }
+
[…]

### 2024-04-17 presentations/understanding the problem.qmd
add presentation

---
title: "Understanding the problem"
format: 
  revealjs:
    smaller: true
    scrollable: true
editor: visual
---

## Blåbærkrisa

:::: {.columns}

::: {.column width="30%"}
Same data gave different indicator values
:::

::: {.column width="70%"}
```{r}
#| fig-align: "left"
#| fig-height: 6
#| fig-width: 7
source("../R/example1.R")
example1
```

::: aside
Indicators with grey background
:::
:::

::::

## This project

-   250 000 kr

-   166 hours

-   1 month

-   Manuscript on pre-print server

-   Produce guidelines

---

Example with two indicators.
Scaling indicators at the sub-regions also makes it possible to
calculate an index for the sub-regions. 
But to get an index for the whole region, should we aggregate the

```{r}
#| fig-align: "left"
#| fig-height: 6
#| fig-width: 7
source("../R/example2.R")
example2
```

---

Example of indeces at EA scale aggregated to EAA scale

```{r}
#| fig-align: "left"
#| fig-height: 6
#| fig-width: 7
source("../R/example3.R")
example3
```

---

Example of indicators aggregated at EEA scale, then turned into an index

```{r}
#| fig-align: "left"
#| fig-height: 6
#| fig-width: 7
source("../R/example4.R")
example4
```
---

However, this result is area sensitive

```{r}
#| fig-align: "left"
#| fig-height: 6
#| fig-width: 7
source("../R/example5.R")
example5
```

---
Next, variables aggregated at EAA scale

### 2024-04-17 presentations/understanding the problem.qmd
more examples

@@
 
-Example with two indicators.
-Scaling indicators at the sub-regions also makes it possible to
-calculate an index for the sub-regions. 
-But to get an index for the whole region, should we aggregate the
-
-```{r}
-#| fig-align: "left"
-#| fig-height: 6
-#| fig-width: 7
-source("../R/example2.R")
-example2
-```
-
----
-
 Example of indeces at EA scale aggregated to EAA scale
@@
 ```
+
 ---
 
-However, this result is area sensitive
+Adding area weighting to the first example
+```{r}
+#| fig-align: "left"
+#| fig-height: 6
+#| fig-width: 7
+source("../R/example3b.R")
+example3b
+```
 
+---
+
+and to the second example
 ```{r}
@@
 ---
-Next, variables aggregated at EAA scale
+
+Next, variables aggregated at EAA scale by taking the area weighted mean. When would this make sense to do? Obviously not for count variables.
+
+```{r}
+#| fig-align: "left"
+#| fig-height: 6
+#| fig-width: 7
+source("../R/example6.R")
+example6
+```
+
+---
+
+Here, variable aggregated at EAA scale with no area weighting
+
+```{r}
+#| fig-align: "left"
+#| fig-height: 6
+#| fig-width: 7
+source("../R/example7.R")
+example7
+```
+
+---
+
+Example with two indicators.
+Scaling indicators at the sub-regions also makes it possible to
+calculate an index for the sub-regions. 
+But to get an index for the whole region, should we aggregate the
+
+```{r}
+#| fig-align: "left"
+#| fig-height: 6
+#| fig-width: 7
+source("../R/example2.R")
+example2
+```

### 2024-04-17 R/example3b.R

@@ -1,0 +1,89 @@
+
+
+
+library(DiagrammeR)
+
+# Example with two indicator and two EAs
+# where we scale variables at the EA level
+# and calculate an index at the EA level
+# and then aggregate the indeces for the EAA level
+# Version b. Adding area weighting
+
+example3b <- grViz("digraph{
+  
+   
+   graph [
+            compound = true, 
+            nodesep = .8, 
+            ranksep = .8,
+            color = crimson, 
+            rankdir = LR
+        ]
+
+  
+    node [
+            fontname = Helvetica, 
+            fontcolor = darkslategray,
+            shape = rectangle, 
+            fixedsize = true, 
+            width = 1,
+            color = darkslategray
+        ]
+        
+        
+    edge [
+            color = black
+        ]
+        
+# Orange part --------------------------
+
+   subgraph clusterTop {
+      graph[color = orange]
+      
+    subgraph cluster0 {
+      graph[color = green]
+      peripheries=0;
+       
+       subgraph cluster0_1 {
+        label='EA-1, 1 hectare';
+        labeljust=l;
+        peripheries=1;
+        V2_EA1[label = 'Value = 5\nRef. = 25', color = orange, style = filled, fontcolor = white]
+        V1_EA1[label = 'Value = 5\nRef. = 10', color = blue, style = filled, fontcolor = white]
+        I1_EA1[label = ' .5', color = blue, style = filled, fontcolor = white]
+        I2_EA1[label = ' .2', color = orange, style = filled, fontcolor = white]
+        X_EA1[label = ' .35', color = grey, style = filled]
+        V1_EA1 -> I1_EA1 -> X_EA1
+        V2_EA1 -> I2_EA1 -> X_EA1
+        
+       }
+       
+       subgraph cluster0_2 {
+        label='EA-2, 2 hectare';
+        labeljust=l;
+        peripheries=1;
+        V2_EA2[label = 'Value = 1\nRef. = 10', color = orange, style = filled, fontcolor = white]
+        V1_EA2[label = 'Value = 1\nRef. = 4', color = blue, style = filled, fontcolor = white]
+        I1_EA2[label = ' .25', color = blue, style = filled, fontcolor = white]
+        I2_EA2[label = ' .1', color = orange, style = filled, fontcolor = white]
[…]

### 2024-04-17 R/example6.R

@@ -1,0 +1,93 @@
+
+
+
+library(DiagrammeR)
+
+# Example with two indicator and two EAs
+# where we scale variables at the EAA level
+# and then aggregate the index also at the EAA level
+# giving UNequal weight to EAs
+# Does it ever make sense to get area weighted means
+# for variables across EAs?
+# Obviously not for count variables.
+
+example6 <-  grViz("digraph{
+  
+   
+   graph [
+            compound = true, 
+            nodesep = .8, 
+            ranksep = .8,
+            color = crimson, 
+            rankdir = LR
+        ]
+
+  
+    node [
+            fontname = Helvetica, 
+            fontcolor = darkslategray,
+            shape = rectangle, 
+            fixedsize = true, 
+            width = 1,
+            color = darkslategray
+        ]
+        
+        
+    edge [
+            color = black
+        ]
+        
+# Orange part --------------------------
+
+   subgraph clusterTop {
+      
+      graph[color = orange]
+       
+    subgraph cluster0 {
+      graph[color = green]
+      peripheries=0;
+       
+       subgraph cluster0_1 {
+        label='EA-1, 1 hectar';
+        labeljust=l;
+        peripheries=1;
+        V1_EA1[label = 'Value = 5\nRef. = 10', color = blue, style = filled, fontcolor = white]
+        V2_EA1[label = 'Value = 5\nRef. = 25', color = orange, style = filled, fontcolor = white]
+        #I1_EA1[label = ' .5', color = blue, style = filled, fontcolor = white]
+        #I2_EA1[label = ' .2', color = orange, style = filled, fontcolor = white]
+        #V1_EA1 -> I1_EA1
+        #V2_EA1 -> I2_EA1
+        
+       }
+       
+       subgraph cluster0_2 {
+        label='EA-2, 2 hectar';
+        labeljust=l;
+        peripheries=1;
+        V1_EA2[label = 'Value = 1\nRef. = 4', color = blue, style = filled, fontcolor = white]
+        V2_EA2[label = 'Value = 1\nRef. = 10', color = orange, style = filled, fontcolor = white]
+        #I1_EA2[label = ' .25', color = blue, style = filled, fontcolor = white]
[…]

### 2024-04-17 R/example7.R

@@ -1,0 +1,93 @@
+
+
+
+library(DiagrammeR)
+
+# Example with two indicator and two EAs
+# where we scale variables at the EAA level
+# and then aggregate the index also at the EAA level
+# giving UNequal weight to EAs
+# Does it ever make sense to get area weighted means
+# for variables across EAs?
+# Obviously not for count variables.
+
+example7 <-  grViz("digraph{
+  
+   
+   graph [
+            compound = true, 
+            nodesep = .8, 
+            ranksep = .8,
+            color = crimson, 
+            rankdir = LR
+        ]
+
+  
+    node [
+            fontname = Helvetica, 
+            fontcolor = darkslategray,
+            shape = rectangle, 
+            fixedsize = true, 
+            width = 1,
+            color = darkslategray
+        ]
+        
+        
+    edge [
+            color = black
+        ]
+        
+# Orange part --------------------------
+
+   subgraph clusterTop {
+      
+      graph[color = orange]
+       
+    subgraph cluster0 {
+      graph[color = green]
+      peripheries=0;
+       
+       subgraph cluster0_1 {
+        label='EA-1';
+        labeljust=l;
+        peripheries=1;
+        V1_EA1[label = 'Value = 5\nRef. = 10', color = blue, style = filled, fontcolor = white]
+        V2_EA1[label = 'Value = 5\nRef. = 25', color = orange, style = filled, fontcolor = white]
+        #I1_EA1[label = ' .5', color = blue, style = filled, fontcolor = white]
+        #I2_EA1[label = ' .2', color = orange, style = filled, fontcolor = white]
+        #V1_EA1 -> I1_EA1
+        #V2_EA1 -> I2_EA1
+        
+       }
+       
+       subgraph cluster0_2 {
+        label='EA-2';
+        labeljust=l;
+        peripheries=1;
+        V1_EA2[label = 'Value = 1\nRef. = 4', color = blue, style = filled, fontcolor = white]
+        V2_EA2[label = 'Value = 1\nRef. = 10', color = orange, style = filled, fontcolor = white]
+        #I1_EA2[label = ' .25', color = blue, style = filled, fontcolor = white]
+        #I2_EA2[label = ' .1', color = orange, style = filled, fontcolor = white]
[…]

### 2024-04-17 presentations/understanding the problem.qmd
commutativity

@@
 
+```{r setup}
+library(knitr)
+```
+
 ## Blåbærkrisa
 
-:::: {.columns}
-
+::: columns
 ::: {.column width="30%"}
@@
 :::
-
-::::
+:::
 
@@
 
----
+------------------------------------------------------------------------
 
@@
 
----
+------------------------------------------------------------------------
 
 Adding area weighting to the first example
+
 ```{r}
@@
 
----
+------------------------------------------------------------------------
 
 and to the second example
+
 ```{r}
@@
 
----
+## Commutativty
+
+These are examples of cummutative aggregation workflows
+
+![](img/commutativity.PNG)
+
+------------------------------------------------------------------------
 
@@
 
----
+------------------------------------------------------------------------
 
-Example with two indicators.
-Scaling indicators at the sub-regions also makes it possible to
-calculate an index for the sub-regions. 
-But to get an index for the whole region, should we aggregate the
+Example with two indicators. Scaling indicators at the sub-regions also makes it possible to calculate an index for the sub-regions. But to get an index for the whole region, should we aggregate the
 

### 2024-04-18 presentations/understanding the problem.qmd
presentation

@@
 ---
-title: "Understanding the problem"
+title: "Understanding the problem of when to scale"
 format: 
@@
     smaller: true
-    scrollable: true
+    scrollable: false
+    preview-links: true
 editor: visual
@@
 ```
-
-## Blåbærkrisa
-
-::: columns
-::: {.column width="30%"}
-Same data gave different indicator values
-:::
-
-::: {.column width="70%"}
-```{r}
-#| fig-align: "left"
-#| fig-height: 6
-#| fig-width: 7
-source("../R/example1.R")
-example1
-```
-
-::: aside
-Indicators with grey background
-:::
-:::
-:::
 
@@
 
+## Blåbærkrisa
+
+::: columns
+::: {.column width="50%"}
+
+- Same data gave different indicator values.
+- Reference values are at county scales (and probably only make sense at that scale)
+- IBECA normalised the variables (with truncation) at plot scale
+- NI normalised the variable (probably) at a regional scale after first taking a mean
+:::
+
+::: {.column width="50%"}
+
+![](img/billberry.PNG){width=100%}
+
+:::
+
+:::
+
+---
+
+![](img/billberry_graph.PNG)
+
+---
+
+![](img/billberry_ibeca.PNG)
+
+## It matters 
+if we  aggregate (takes means of) of the scaled or unscaled variabel:
+
+![](img/ASPT.png)
+
+## Arguments for scaling and truncating immediately
+
+- Positive (and negative) outliers in on area can not compensate for poor (or good) condition elsewhere
+- Gives higher-resolution indicator values
+
+## Arguments for scaling and truncating later at a level where the reference value is sensible
+
+- The reference value was maybe intended for that scale
+- The indicator uncertainty becomes (apparently) smaller because the mean or sum for a region is less variable.
+- Maybe, to many truncation events leds to accumulated dissplacement errors?
+
+## What if
+
+- The reference value is at the same scale as the variable
+- The reference value is the same all over
+- We scale locally before aggregating, and then truncate later
+- We also want to aggregate the original variable further, e.g. to national level - will the indicator and variable remain correlated
+
+## Resources
+
+[Sanvik 2019](https://brage.nina.no/nina-xmlui/handle/11250/2631056)
+
+## Why do we re-scale
+*notes from Bård*
+
+Reference values serve two purposes
+
+- to enable rescaling to a common measurement scale to facilitate the calculation of the mean.
+-	to set a limit for how much one indicator can compensate for other indicators being in a bad state.
+
+As a consequence, an index of rescaled indicators summarizes **negative deviations** from the reference state over a large set of indicators.
+
+Unscaled states above the reference value are not recognized as being better than the reference state.
+
+Rescaling is useful for making indeces, but why use it for individual indicators? [answer: it depends if we want to aggregate 'condition estimates' of actuall measurements]
+
+## Spatially aggregating 
+condition estimates (bottom) or actuall measurements (top)
+
+```{r}
+source("../R/example1.R")
+example1
+```
+
+::: aside
+Indicators with grey background
+:::
+
 ------------------------------------------------------------------------
 
-Example of indeces at EA scale aggregated to EAA scale
+## Early scaling 
+leads to cummutativity
 
@@
 
-Example of indicators aggregated at EEA scale, then turned into an index
+## Early scaling 
+leads to cummutativity
 
@@
 
-Adding area weighting to the first example
+## Also with area weighting
 
@@
 
-and to the second example
+## Also with area weighting
 
@@
 
-These are examples of cummutative aggregation workflows
-
 ![](img/commutativity.PNG)
 
-------------------------------------------------------------------------
-
-Next, variables aggregated at EAA scale by taking the area weighted mean. When would this make sense to do? Obviously not for count variables.
+## Alternativey,
+we can aggregate the actual measurements. When would this make sense to do?
 
@@
 
-Here, variable aggregated at EAA scale with no area weighting
+## Same, but with no area weighting and taking the sum (e.g. population sizes)
 
@@
 
-------------------------------------------------------------------------
+## What about
 
-Example with two indicators. Scaling indicators at the sub-regions also makes it possible to calculate an index for the sub-regions. But to get an index for the whole region, should we aggregate the
-
-```{r}
-#| fig-align: "left"
-#| fig-height: 6
-#| fig-width: 7
-source("../R/example2.R")
-example2
-```
+- Scaling, aggregating, and then truncating
+- Non-linear re-scaling functions with asymptotes

### 2024-04-18 R/pathways.R

@@ -1,0 +1,79 @@
+
+
+
+library(DiagrammeR)
+
+
+pathways <- grViz("digraph{
+  
+   
+   graph [
+            compound = true, 
+            nodesep = .8, 
+            ranksep = .8,
+            color = crimson, 
+            rankdir = LR
+        ]
+
+  
+    node [
+            fontname = Helvetica, 
+            fontcolor = darkslategray,
+            shape = rectangle, 
+            fixedsize = true, 
+            width = 1,
+            color = darkslategray
+        ]
+        
+        
+    edge [
+            color = black
+        ]
+
+subgraph clusterEAA {
+      graph[colour = black]
+      label='EAA scale';   
+   
+subgraph clusterRegion {
+      graph[style=dashed]
+      label='Regional scale';
+
+   
+subgraph clusterPlot {
+      graph[style=dashed]
+      style=filled; 
+      color=lightblue;
+      label='Plot scale';
+
+  Variable[label='Original\nvariable']
+  Variable2[label='Original\nvariable']
+  Scale2[label='Scale']
+  
+  
+}
+
+subgraph clusterHEA {
+      graph[style=dashed]
+      style=filled; 
+      color=grey90;
+      label='HEA';
+
+  Scale[label='Scale']
+
+} # end HEA scale
+} # end regional scale
+
+  Agg[label='Aggregate']
+  Agg2[label='Aggregate']
+
+} # end EAA scale
+  
+  Variable -> Scale -> Agg
+  Variable2 -> Scale2 -> Agg2
+
+   
+
+}")
+
+pathways
+

### 2024-04-19 presentations/understanding the problem.qmd
update presentation

@@
 ::: {.column width="50%"}
-
 -   Same data gave different indicator values.
@@
 ::: {.column width="50%"}
-
-![](img/billberry.PNG){width=100%}
-
+![](img/billberry.PNG){width="80%"}
+:::
 :::
 
-:::
-
----
+------------------------------------------------------------------------
 
@@
 ## It matters
+
 if we aggregate (takes means of) of the scaled or unscaled variabel:
@@
 
-## Arguments for scaling and truncating immediately
+## Pathways
 
-- Positive (and negative) outliers in on area can not compensate for poor (or good) condition elsewhere
-- Gives higher-resolution indicator values
+```{r}
+source("../R/pathways.R")
+pathways
+```
 
-## Arguments for scaling and truncating later at a level where the reference value is sensible
+## Glossery
 
-- The reference value was maybe intended for that scale
-- The indicator uncertainty becomes (apparently) smaller because the mean or sum for a region is less variable.
-- Maybe, to many truncation events leds to accumulated dissplacement errors?
+**Scaling** = Linear range scaling, usually reducing the range of the variable by using one or two reference value
 
-## What if
+**Truncating** = Cutting of value above or below a given threshold. In our case, making the metrices bound between 0 and 1. Truncation is also a type of scaling.
 
-- The reference value is at the same scale as the variable
-- The reference value is the same all over
-- We scale locally before aggregating, and then truncate later
-- We also want to aggregate the original variable further, e.g. to national level - will the indicator and variable remain correlated
+**Re-scaling** = not used, but often used as a synonym to normalisation.
 
-## Resources
+**Normalising** = a combination of scaling, followed by truncating (if needed). This lead to a non-linear transformation of the original variable and returns indicators that share the same scale.
 
-[Sanvik 2019](https://brage.nina.no/nina-xmlui/handle/11250/2631056)
-
-## Why do we re-scale
-*notes from Bård*
-
-Reference values serve two purposes
-
-- to enable rescaling to a common measurement scale to facilitate the calculation of the mean.
--	to set a limit for how much one indicator can compensate for other indicators being in a bad state.
-
-As a consequence, an index of rescaled indicators summarizes **negative deviations** from the reference state over a large set of indicators.
-
-Unscaled states above the reference value are not recognized as being better than the reference state.
-
-Rescaling is useful for making indeces, but why use it for individual indicators? [answer: it depends if we want to aggregate 'condition estimates' of actuall measurements]
+**Non-linear scaling**: Scaling functions like truncation, sigmoid, exponential or break-point types.
 
 ## Spatially aggregating
-condition estimates (bottom) or actuall measurements (top)
+
+condition estimates (bottom) or actual measurements (top)
 
@@
 ------------------------------------------------------------------------
+## Commutativty
+
+![](img/commutativity.PNG)
 
 ## Early scaling
+
 leads to cummutativity
@@
 ## Early scaling
+
 leads to cummutativity
@@
 
-## Commutativty
+## Alternativey,
 
-![](img/commutativity.PNG)
-
-## Alternativey,
 we can aggregate the actual measurements. When would this make sense to do?
@@
 
-## What about
+## Why do we normalise
 
-- Scaling, aggregating, and then truncating
-- Non-linear re-scaling functions with asymptotes
+*notes from Bård*
+
+Reference values serve two purposes
+
+-   to enable rescaling to a common measurement scale to facilitate the calculation of the mean.
+-   to set a limit for how much one indicator can compensate for other indicators being in a bad state.
+
+As a consequence, an index of rescaled indicators summarizes **negative deviations** from the reference state over a large set of indicators.
+
+Unscaled states above the reference value are not recognized as being better than the reference state.
+
+Rescaling is useful for making indeces, but why use it for individual indicators? \[answer: it depends if we want to aggregate 'condition estimates' of actuall measurements\]
+
+## Arguments for scaling and truncating immediately
+
+-   High or low *outliers* can not compensate for low or high values (i.e. condition) elsewhere, respectively.
+-   Gives higher-resolution indicator values and preserves the original spatial resolution of the variable.
+-   Means that subsequent aggregation is carrying information about the actual condition, and not the raw variable.
+
+## Arguments for scaling and truncating later at a level where the reference value is sensible
+
+-   The reference value was maybe intended for that scale
+-   The indicator uncertainty becomes (apparently) smaller because the mean or sum for a region is less variable.
+-   Maybe, too many truncation events leads to accumulated displacement errors?
+
+## What if
+
+-   The reference value is at the same scale as the variable
+-   The reference value is the same all over
+-   We also want to aggregate the original variable further, e.g. to national level - will the indicator and variable remain correlated
+-   We want to use non-linear re-scaling functions with asymptotes
+
+## Resources
+
+[Sanvik 2019](https://brage.nina.no/nina-xmlui/handle/11250/2631056)

### 2024-04-19 R/pathways.R

@@ -5,8 +5,7 @@ library(DiagrammeR)
 
 
 pathways <- grViz("digraph{
-  
-   
+
    graph [
             compound = true, 
             nodesep = .8, 
@@ -30,48 +29,49 @@ pathways <- grViz("digraph{
             color = black
         ]
 
-subgraph clusterEAA {
-      graph[colour = black]
-      label='EAA scale';   
-   
-subgraph clusterRegion {
-      graph[style=dashed]
-      label='Regional scale';
+   three[label='3) ', shape = plaintext]
+   two  [label='2) ', shape = plaintext]
+   one  [label='1) ', shape = plaintext]
 
-   
-subgraph clusterPlot {
-      graph[style=dashed]
-      style=filled; 
-      color=lightblue;
-      label='Plot scale';
+subgraph cluster0_1 {
+        label='Measurement scale';
+        labeljust=l;
+        peripheries=1;
 
-  Variable[label='Original\nvariable']
-  Variable2[label='Original\nvariable']
-  Scale2[label='Scale']
-  
-  
+   m3[label = 'measure']
+   m2[label = 'measure']
+   m1[label = 'measure']
+   n1[label = 'normalise']
+   scale
+
 }
 
-subgraph clusterHEA {
-      graph[style=dashed]
-      style=filled; 
-      color=grey90;
-      label='HEA';
+subgraph cluster0_2 {
+        label='Larger scale';
+        labeljust=l;
+        peripheries=1;
 
-  Scale[label='Scale']
+   a1[label = 'aggregate']
+   a2[label = 'aggregate']
+   a3[label = 'aggregate']
+   n2[label = 'normalise']
+   truncate
 
-} # end HEA scale
-} # end regional scale
+}
 
-  Agg[label='Aggregate']
-  Agg2[label='Aggregate']
 
-} # end EAA scale
-  
-  Variable -> Scale -> Agg
-  Variable2 -> Scale2 -> Agg2
+one -> m3 [style = invis]
+m3 ->a3 [minlen=2, weight=0]
+a3 -> n2
 
+two -> m2 [style = invis]
+m2 -> n1 -> a2
    
+three -> m1 [style = invis]
+m1 -> scale -> a1 -> truncate
+
+{rank=same; one; two; three;}
+
 
 }")
 

### 2024-04-23 R/scratch/pathways.R
try {here}

@@ -1,0 +1,79 @@
+
+
+
+library(DiagrammeR)
+
+
+pathways <- grViz("digraph{
+  
+   
+   graph [
+            compound = true, 
+            nodesep = .8, 
+            ranksep = .8,
+            color = crimson, 
+            rankdir = LR
+        ]
+
+  
+    node [
+            fontname = Helvetica, 
+            fontcolor = darkslategray,
+            shape = rectangle, 
+            fixedsize = true, 
+            width = 1,
+            color = darkslategray
+        ]
+        
+        
+    edge [
+            color = black
+        ]
+
+subgraph clusterEAA {
+      graph[colour = black]
+      label='EAA scale';   
+   
+subgraph clusterRegion {
+      graph[style=dashed]
+      label='Regional scale';
+
+   
+subgraph clusterPlot {
+      graph[style=dashed]
+      style=filled; 
+      color=lightblue;
+      label='Measurement scale';
+
+  Variable[label='Original\nvariable']
+  Variable2[label='Original\nvariable']
+  Scale2[label='Scale']
+  
+  
+}
+
+subgraph clusterHEA {
+      graph[style=dashed]
+      style=filled; 
+      color=grey90;
+      label='HEA';
+
+  Scale[label='Scale']
+
+} # end HEA scale
+} # end regional scale
+
+  Agg[label='Aggregate']
+  Agg2[label='Aggregate']
+
+} # end EAA scale
+  
+  Variable -> Scale -> Agg
+  Variable2 -> Scale2 -> Agg2
+{rank=same; Agg; Agg2;}
+
+
+}")
+
+pathways
+

### 2024-04-23 presentations/understanding the problem.qmd
try {here}
embed resources

@@
     preview-links: true
+    embed-resources: true
 editor: visual

### 2024-04-23 presentations/understanding the problem.qmd
print wd
selv contained

--- 2024-04-19 presentations/understanding the problem.qmd
@@
     preview-links: true
+    self-contained: true
 editor: visual
@@
 ------------------------------------------------------------------------
+
 ## Commutativty
@@
 
-## Same, but with no area weighting and taking the sum (e.g. population sizes)
+## Same, but with no area weighting 
+and taking the sum (e.g. population sizes)
 

### 2024-11-19 R/aggregationBiasFigure.R
fix figs

@@ -3,7 +3,7 @@ library(eaTools)
 library(sf)
 library(ggtext)
 
-var <- seq(50, 200, .1)
+var <- seq(50, 200, 0.1)
 dat <- data.frame(var = var)
 
 X0 <- 60
@@ -21,7 +21,7 @@ dat <- var |>
     var >X0 ~ (var -X0)/(X100-X0),
     var <= X0 ~ 0,
   ),
-  ind = round(ind, 2))
+  ind = round(ind, 3))
 
 low <- "red"
 high <- "green"
@@ -43,10 +43,16 @@ fVhatt <- dat |>
   pull(ind)
 fVbar <- ((fV2-fV1)/2)+fV1
 
+dat_x <- dat |>
+  distinct(ind,
+           .keep_all = T) |>
+  add_row(ind = 0, var = X0) |>
+  add_row(ind = 1, var = 200)
+
 
 # ggtext and element markdown doesnt work for some mathematical signs, like bar.
 ## PLOT 1 -------------
-dat |>
+dat_x |>
   ggplot(aes(
     x = var,
     y = ind,
@@ -130,7 +136,8 @@ dat |>
 ## PLOT 2 --------------
 # add X60
 
-X60 <- 155
+
+X60 <- 150
 X100 <- 160
 V2 <- 170
 Vbar <- ((V2-V1)/2)+V1
@@ -143,7 +150,7 @@ dat2 <- var |>
     var >X0 ~ (var-X0)/(X60-X0) *0.6,
     var <= X0 ~ 0,
   ),
-  ind = round(ind, 2))
+  ind = round(ind, 3))
 
 fV1x <- dat2 |>
   filter(var == V1) |>
@@ -156,31 +163,39 @@ fVhattx <- dat2 |>
   pull(ind)
 fVbarx <- ((fV2x-fV1x)/2)+fV1x
 
+dat3 <- dat2 |>
+  distinct(ind,
+           .keep_all = T) |>
+  add_row(ind = 0, var = X0) |>
+  add_row(ind = 1, var = 200)
 
-dat2 |>
+dat3 |>
   ggplot(aes(
     x = var,
     y = ind,
     colour = ind
   )) +
+  geom_point()+
   geom_line(
     lineend = "round",
+    linejoin = "bevel",
     size = mySize,
     alpha = myAlpha
   ) +
-  ylab("") +
-  xlab("") +
   scale_color_gradient("Indicator values",
                        low = low,
                        high = high
   ) +
+  ylab("") +
+  xlab("") +
[…]

### 2024-11-19 manuscript/manuscript.qmd
transfer from google docs

@@
 ---
-title: Short Paper
-subtitle: A Short Subtitle
+title: On the spatial aggregation of ecosystem condition indicators
 author:
@@
         postal-code: 7485
-  - name: Chloé R. Nater
-    email: chloe.nater@nina.no
-    orchid: https://orcid.org/0000-0002-7975-0108
+  - name: Bálint Czúcz
+    email: balint.czucz@nina.no
     affiliations:
@@
 abstract: |
-  This is the abstract.
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-Vestibulum augue turpis, dictum non malesuada a, volutpat eget velit.
-Nam placerat turpis purus, eu tristique ex tincidunt et.
-Mauris sed augue eget turpis ultrices tincidunt.
-Sed et mi in leo porta egestas.
-Aliquam non laoreet velit.
-Nunc quis ex vitae eros aliquet auctor nec ac libero.
-Duis laoreet sapien eu mi luctus, in bibendum leo molestie.
-Sed hendrerit diam diam, ac dapibus nisl volutpat vitae.
-Aliquam bibendum varius libero, eu efficitur justo rutrum at.
-Sed at tempus elit.
+  Ecosystem condition assessments (ECA) and accounts use variables and indicators to describe key ecosystem characteristics reflecting the ecosystem condition and any potential deviations from a reference condition.
+These metrices are rutinely aggregation spatially to produce values representing the condition of a larger area, such as a country.
+However, we experience little awareness about the potential pitfals arising from aggregation bias of highly modified indicators.
+Here we outline some consequences regarding the order of the steps involed in normalising variables aggregating them in space.
+We show that aggregation bias in non-trivial and has the potential to undermine the credibility and precition in ECAs, and to confuse the communication of their results.
+We introduce a common terminology for aggregatio pathways, spesific to ecosystem condition indicators follwing SEEA EA standards, and make some recomendations about which aggregation pathway to use in different settings, and how to report these choices.
+
 keywords: 
-  - keyword1
-  - keyword2
+  - SEEA EA
+  - ecosystem condition
+  - ecosystem accounting
+  - indicators
+  - aggregation bias
 date: last-modified
@@
     journal:
-      name: Ecological Indicators
+      name: ---
       formatting: preprint
@@
   message: false
+editor: 
+  markdown: 
+    wrap: sentence
 ---
@@
 
-# Methods
+In ecosystem accounting, fine-scale ecological information is routinely aggregated in space to produce single parameter estimates for entire regions.
+This simplification is necessary to show major changes in the extent and condition of ecosystems at a scale that is relevant for decision makers.
+The spatial aggregation of data can lead to biases arising from a number of sources, such as uneven sampling efforts, unequal sampling designs, or from the mathematical process of aggregating a variable that has a nonlinear relationship to a secondarily-derived metric (ref).
+The first two sources of error are usually explicitly addressed and managed, but the last case of aggregation bias is often overlooked.
+This error is most commonly encountered in ecosystem condition accounting when variables hold nonlinear relationships to the normalised indicators.
+In the statistical standard for EAs and the recommended guidelines for ecosystem accounts (EAs) (ref), there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
+Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (a thematic aggregation), but in this paper we are discussing spatial aggregation only.
 
-# Results
+Variables in ecosystem condition terminology are metrics describing ecosystem characteristics.
+They have a spatial component, tying them to unique Ecosystem Assets.
+Variables can be spatially aggregated across ecosystem assets (or across basal units), for example by taking the sum or an area-weighted mean, to produce variable estimates for larger regions, such as the entire Ecosystem Accounting Area (EAA).
+As a next step, variables can be normalised and turned into normative indicators for ecosystem condition.
+The mathematical part of this normalisation contains two, sometimes three, steps [@fig-term].
+These steps can be performed simultaneously, but for clarity we discuss them as separate steps.
+First the variable is scaled, using at least two reference levels, defining variable values to be coded as zero (X0) or one (X100) on the indicator scale.
+Secondly, the scaled variables are truncated to produce a bound indicator scale between 0 and 1.
+Sometimes a third step of transformation is required to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
+This is commonly done by anchoring specific variable values to predefined class boundaries (ref SELINA paper), but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
+When a variable is normalised in this way, its interpretation changes from descriptive to normative.
+The ecological significance of the indicator might also differ from that of the variable since additional ecological knowledge can be introduced via the reference values.
+The indicator is a function of the variable and the reference values, and this relationship is almost always nonlinear (e.g. due to truncation which is ubiquitous practice).
 
-# Discussion
+Similar to variables, the reference values used for normalising the variables also have a spatial assignment, but it can be different from that of the variable.
+For example, variable values may exist for unique 10x10m grid cells, but the reference values may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire EAA (e.g.
+a natural zero).
+A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition variables or indicators, is the choice about the order of the different steps in the normalisation and aggregation process, i.e.
+the aggregation pathway (Figure 2).
 
-# References {-}
+```{r fig-term}
+#| fig-cap: |
+#|   Examples of how important terms are used in this manuscript.
+The variable (1), in raw biophysical units, has a minimum and a maximum value (V), as well as points anchoring it to the indicator scale (X).
+Scaled variables (2) are scaled based on X0 and X100 only.
+Note that this is a different use of the term then the more common use that implies dividing a vector by its standard deviation.
+Truncation (3) implies assigning the value of 0 to values below X0, and the value 1 to value above X100.
+Transformation refers to the non-linear transformation of values within the 0-1 range, as in this piecewise-linear example by mapping X60 to 0.6 on the indicator scale.
+Transformations without anchoring points are also included in this term, such as exponential or sigmoid transformation.
+Although both truncation and scaling can be seen as types of transformations, we exclude these methods from the definition here.
+Examples 2, 3, and 4 are normative (hence the gradient colour scale from red to green) because they have declared that one end of the variable scale represents a good state and the other a poor state.
+All variables that are treated so that they conform to the definition of indicators in the SEEA EA are said to be normalised.
+For example, if a truncated variable is assumed to have a linear relationship with the indicator scale, then this variable has been normalised even though it has not been transformed.
+The term rescale is a synonym to normalise, but we will only use the latter from here on."
+#| echo: false
+knitr::include_graphics("../img/terminologi.png")
+```
+
+```{r fig-pathways}
+#| fig-cap: |
+#|   "Flowchart showing pathways for the spatial aggregation of variables in ecosystem condition accounting.
+Normalisation refers to the steps that make a variable conform to the definition of an indicator in SEEA EA, which may or may not include the transformation step.
+A truncated and/or transformed variable will have a different interpretation than the original variable.
+The aggregation bias depends on whether variables are truncated and/or transformed variables before or after spatial aggregation.
+Variables are never normalised (pathway 4), and hence the aggregation bias for variables will be different than for indicators.
+An important factor influencing the the choice whether to normalise early (as in pathway 1), or later, is the spatial validity of the reference levels, and sometimes raw variable values (e.g.
+billberry coverage in vegetation quadrats) are spatially aggregated to the scale that was considered when setting the reference levels (e.g.
+average regional billberry coverage)."
+#| out-width: 400
+#| echo: false
+knitr::include_graphics("../img/aggregationPathways.jpg")
+```
+
+```{r fig-agg1}
+#| echo: false
+#| fig-cap: |
+#|   "Example showing the difference in indicator values (y-axis) obtained from taking the mean of two variable values (V1 and V1) that are on the original scale (i hat) or on the normalised indicator scale (i bar).
+Due to truncation at x100, the latter results in a comparatively lower indicator value.
+The solid coloured line represents the normalisation function f(V).
+V = variable value, i = indicator value.
+Modified from Rastsetter (1991)."
+#| out-width: 400
+
+knitr::include_graphics("../img/aggregation_bias_truncations.PNG")
+
+```
+
+```{r fig-agg2}
+#| echo: false
+#| fig-cap: |
+#|   "Same figure as above, but showing an example where the normalisation function includes a transformation step (see Figure 1 for definition) that anchors X60 to 0.6 on the indicator scale. "
+#| out-width: 400
+
+knitr::include_graphics("../img/aggregation_bias_x60.PNG")
+```
+
+## Why do we normalise? {#sec-norm}
+
+Normalisation of ecosystem condition indicators, as defined in Figure 1, serves multiple purposes.
+Firstly, it gives a normative interpretation of a variable, defining a good and a bad state and a directionality to say when something is getting better or worse over time.
+This normative characteristic is an essential part of assessing ecosystem condition.
+Secondly, normalisation sets a limit to how much a high indicator value in one place can compensate for a low value somewhere else.
+This is because the truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviation from the reference levels, and positive deviations (which could compensate for the negative ones) are ignored.
+This is sometimes what we want, but not always something we are aware that we are doing.
+One reason to want to aggregate the negative deviations only, is because the reference levels are set (or should be set) so that values above these limits do not represent any further increase (or decrease) in ecosystem condition.
+Therefore, this way of aggregation summarises the presumed ecosystem condition (which is generally what we want), and not the variable itself.
+In other words, we normalise in order to facilitate the spatial (i.e. horizontal) aggregation of our ecosystem condition estimates (i.e. the indicator values), and not merely the variable values.
+Thirdly, we normalise in order to standardise the indicator on the same scale so that we can perform thematic (i.e. vertical) aggregation.
+This is commonly referred to as the reason for normalising variables, but as we have shown, it is but one of three main reasons, and also perhaps the least confusing part for many.
+
+## Main points to consider when deciding on aggregation pathway
+
+### Ease of interpretation {.unnumbered}
+
+The indicator should make intuitive sense.
+This is helped by using the same aggregation pathway for all indicators.
+Indicators using spatially aggregated population estimates are easily misinterpreted as representing population trends.
+
+### The ‘real’ spatial resolution of reference values {.unnumbered}
+
+To achieve an early normalisation of variables, we need reference values that make ecological sense at the same spatial resolution as the variable.
+Spatial disaggregation or interpolation of reference values needs to be well justified.
+
+### Scalability and comparability between assessment {.unnumbered}
+
+Separate assessments should give identical indicator values for the same areas, given the same underlying data.
+When this is not true it raises concerns about reproducibility and sacrifices the credibility of both assessments.
+
+### Making clear the differences in the interpretation of variables and indicators {.unnumbered}
+
+Often indicators and variables are both included side-by-side in ecosystem condition assessments/accounts.
+This is also recommended practice following the SEEA EA.
+However, if the indicators that are presented are normalised at an earlier stage, and subsequently spatially aggregated, same as the variable, then the interpretation of the indicators is not simply a normalised version of the variable, but rather a whole new beast.
+For example, these indicators typically now reflect the average ecosystem condition, where areas in very good or very poor condition have limited ability to compensate for opposite extreme values elsewhere.
+The spatially aggregated variable on the other hand, may reflect something like a sum of individuals, or some other aspect where values above and below the average value are able to compensate for each other.
+
+## Examples
+
+In this section we use some examples from our own experience to illustrate some of the issues arising from chosing different aggregation pathways,
+
+### Billberry example
+
+This example illustrates a variable normalised at plot level or aggregated to regions first.
+At what scale is the ref value intended?
+
+<!-- Indicator: Blåbæredekning (billberry coverage) -->
+
+<!-- Dataset: Norwegian national forest inventory. -->
+
+<!-- Reference: https://www.naturindeks.no/Indicators/blaabaer -->
+
+*Variable*: Horizontal coverage (vertical projection) of billberry (Vaccinum myrtillus) recorded in permanent vegetation plots.
+
+*Reference levels*: X~100~ was defined for each of 5 regions in Norway based on an expert elicitation.
+Experts were informed by the distribution of the variable values both within and outside protected areas, and the regional distribution of major forest types.
+They also used their general knowledge about the effect of forestry and on the general vegetation structure of old-growth forests.
+
+<!-- *Pathway*: 3 and 1 (on sperate occations) -->
+
+The special thing about the billberry indicator is the different spatial resolution for the variable and for the reference level.
+The variable is recorded at the scale of vegetation plots.
+The reference level however, is designed with a regional spatial scale in mind, especially because of how it encompasses the known variation in forest types in a region to estimate the mean billberry coverage under the reference condition.
+Because the normalisation includes a truncation step, scaling the variable at the plot scale (pathway 1; @fig-agg1) would not allow overshooting values to compensate for lower values when aggregating regional indicator values, which would then be negatively displaced ($\bar i$) relative to when aggregating variable values ($\hat i$; @fig-agg1).
+Scaling at the plot scale would require require unique reference levels for each forest type.
+However, both pathway 1 and 3 has been used for this variable in two different forest ecosystem condition assessments.
+Due to non-negligeable truncation of the variable during the normalisation, the two indicators were subject to different levels of aggregation bias, and consequently the spatially aggregated indicator values for the two assessments became different.
+This cause a general confusion about how how the same variable (and the same date) can produce different indicator values.
+
+*Conclusion*: Given that the reference value is the way it is, this indicator should have followed pathway 3, and spatially aggregated the variable, from plot scale to regional scale, before normalising the variable and potentially aggregating it further.
+It would also be possible to use pathway 6.
+
+### Alien species example
+
+*Variable*: The local ecological effect from alien species (mostly vascular plants) recorded on a 7-step ordinal scale (1 = no alien species, 7 = only alien species).
+The data comes from nature type monitoring with reference to individual occurences (polygons).
+
+*Reference levels*: X~0~ = total dominance from alien species (variable value 7).
+X~100~ = zero influence from alien species (variable value 1).
+X~60~ = \~2-5% alien plant cover (variable value 3).
+
+<!-- *Pathway*: 1 -->
+
+In this example, the reference levels are uniform throughout the EAA, and the normalisation includes a transformation in the form of a piecewise-linear transformation by anchoring X~60~ to 0.6 on the indicator scale.
+When performing an ECA, one has the option to either aggregate variable value for each region (each ecosystem asset), and then normalise that value (pathway 2 and 3), or to normalise the variable at the scale of the original nature type polygons, and then aggregate the indicator values to the regional level (pathway 1).
+
+*Conclusion*: One benefit of early normalisation is that two nested assessments, such as one national, and one assessment of region *Y* inside the same country, will get the same spatially aggregated value for region *Y*.
+The other thing to ask oneself here is: *Why do we normalise* in the first place (@sec-norm)?
+One reason is that we want to summarise information in space about the ecosystem condition.
+It is the indicator values that hold this information, and not the variable.
+We would therefore favor pathway 1 for this example.
+
+### Phytoplankton trophic index (PTI)
+
+<!-- Dataset: “Vannmiljø”, the Norwegian Environment Agency’s database for Water Framework Directive relevant measurements -->
+
+<!-- Reference: https://github.com/NINAnor/NI_vannf -->
+
+*Variable*: Mean score of algal species present, based on a set of indicator species scored for phosphorus requirements/tolerance.
+The variable is recorded in water bodies (lakes of 0,5 km\^2 or more).
+
+*Reference levels*: X~100~ = median variable value for water bodies in reference condition.
+X~0~, X~20~, X~40~, X~60~ and X~80~ = intercalibrated threshold values, based on dose–response curves.
+
+<!-- Pathway: 1 -->
+
+For use in water management, the Water Framework Directive (WFD) indicators uses aggregation pathway 7 [@fig-pathways], with truncation, scaling, and transformation.
+The value obtained after the first two steps is called an EQR (ecological quality ratio), and the value obtained after the third step nEQR (normalised EQR).
+Spatial aggregation is not done for the WFD, and therefore aggregation bias is not an issue.
+When WFD data are put into use in other contexts, however, aggregation becomes important.
+<!-- Relevant contexts in Norway are the Nature Index (ref) and ecological condition accounting. In both cases,  -->
+
+*Conclusion*: Because of how X~100~ is defined based on the median value across reference lakes, overshooting values (\> X~100~) should be preserved in the spatial aggregation.
+<!-- Otherwise, the condition of water bodies with X > X80 will be underreported.  --> Otherwise we get a negative displacement, as in the bilberry example.
+This means that neither EQR values (which are untransformed), nor nEQR values (which are truncated), can be used uncritically in ECAs.
+One solution, to enable the use of WFD indicators in ECA, would be to change to pathway 6, with (1) scaling, (2) transforming, (3) aggregating and (4) truncating the values [@fig.pathways].
+Furthermore, Eurostat @eurostat2022\
+<!-- requires  --> encourages reporting of variable values.
+These, however, cannot be aggregated spatially in a meaningful way.
+<!-- because the use of threshold values amounts to a nonlinear (specifically, a piecewise-linear) transformation.  -->
+
+# References {.unnumbered}

### 2024-11-20 manuscript/manuscript.qmd
First draft

@@
       - id: nina
-        name: Norwegian Institute for Nature Research
-        department: Department of Aquatic Biodiversity
-        address: Pb 5685 Torgarden
-        city: Trondheim
-        postal-code: 7485
   - name: Bálint Czúcz
@@
 However, we experience little awareness about the potential pitfals arising from aggregation bias of highly modified indicators.
-Here we outline some consequences regarding the order of the steps involed in normalising variables aggregating them in space.
-We show that aggregation bias in non-trivial and has the potential to undermine the credibility and precition in ECAs, and to confuse the communication of their results.
-We introduce a common terminology for aggregatio pathways, spesific to ecosystem condition indicators follwing SEEA EA standards, and make some recomendations about which aggregation pathway to use in different settings, and how to report these choices.
+Here we outline some consequences regarding the order of the steps involed in normalising variables and aggregating them in space.
+We show that the choice of  aggregation pathways in non-trivial and has the potential to undermine the credibility and precition in ECAs, and to confuse the communication of their results.
+We introduce a common terminology for aggregation pathways, spesific to ecosystem condition indicators following SEEA EA standards, and make some recomendations about which pathway to use in different settings, and how to report these choices.
 
@@
 execute:
+  cache: true
   warning: false
@@
 
-In ecosystem accounting, fine-scale ecological information is routinely aggregated in space to produce single parameter estimates for entire regions.
+In ecosystem accounting (EA), fine-scale ecological information is routinely aggregated in space to produce single parameter estimates for entire regions.
 This simplification is necessary to show major changes in the extent and condition of ecosystems at a scale that is relevant for decision makers.
-The spatial aggregation of data can lead to biases arising from a number of sources, such as uneven sampling efforts, unequal sampling designs, or from the mathematical process of aggregating a variable that has a nonlinear relationship to a secondarily-derived metric (ref).
+The spatial aggregation of data can lead to biases arising from a number of sources, such as uneven sampling efforts, unequal sampling designs.
+In ecosystem condition accounting (ECA) specifically, the aggregation of both variables and their normative indicators is common practice and raises the issue of aggregation bias when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
 The first two sources of error are usually explicitly addressed and managed, but the last case of aggregation bias is often overlooked.
-This error is most commonly encountered in ecosystem condition accounting when variables hold nonlinear relationships to the normalised indicators.
-In the statistical standard for EAs and the recommended guidelines for ecosystem accounts (EAs) (ref), there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
+In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2021; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
 Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (a thematic aggregation), but in this paper we are discussing spatial aggregation only.
 
-Variables in ecosystem condition terminology are metrics describing ecosystem characteristics.
-They have a spatial component, tying them to unique Ecosystem Assets.
-Variables can be spatially aggregated across ecosystem assets (or across basal units), for example by taking the sum or an area-weighted mean, to produce variable estimates for larger regions, such as the entire Ecosystem Accounting Area (EAA).
-As a next step, variables can be normalised and turned into normative indicators for ecosystem condition.
+Variables in ecosystem condition terminology are metrics describing ecosystem characteristics [@unitednations2021].
+They have a spatial component, tying them to concrete areas, which can be of varying size, from single pixels to entire regions.
+Variables can be spatially aggregated, for example by taking the sum or an area-weighted mean, to produce variable estimates for larger regions, such as the entire Ecosystem Accounting Area (EAA).
+Variables can also be normalised and turned into normative indicators for ecosystem condition.
 The mathematical part of this normalisation contains two, sometimes three, steps [@fig-term].
 These steps can be performed simultaneously, but for clarity we discuss them as separate steps.
-First the variable is scaled, using at least two reference levels, defining variable values to be coded as zero (X0) or one (X100) on the indicator scale.
-Secondly, the scaled variables are truncated to produce a bound indicator scale between 0 and 1.
-Sometimes a third step of transformation is required to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
-This is commonly done by anchoring specific variable values to predefined class boundaries (ref SELINA paper), but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
+The order of these steps can also change.
+Variable can be scaled using at least two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
+Variables can be truncated to produce a bound indicator scale between 0 and 1.
+Sometimes variables are transformed to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
+This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
 When a variable is normalised in this way, its interpretation changes from descriptive to normative.
-The ecological significance of the indicator might also differ from that of the variable since additional ecological knowledge can be introduced via the reference values.
+The ecological significance of the indicator might also differ from that of the variable since additional ecological knowledge can be introduced via the reference leveles.
 The indicator is a function of the variable and the reference values, and this relationship is almost always nonlinear (e.g. due to truncation which is ubiquitous practice).
@@
 Similar to variables, the reference values used for normalising the variables also have a spatial assignment, but it can be different from that of the variable.
-For example, variable values may exist for unique 10x10m grid cells, but the reference values may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire EAA (e.g.
+For example, variable values may exist for unique 10×10m grid cells, but the reference values may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire EAA (e.g.
 a natural zero).
-A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition variables or indicators, is the choice about the order of the different steps in the normalisation and aggregation process, i.e.
-the aggregation pathway (Figure 2).
+A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition indicators (@fig-agg1; @fig-agg2), is the choice about the order of the different steps in the normalisation and aggregation process, i.e.
+the aggregation pathway (@fig-pathways).
 
@@
 #|   Examples of how important terms are used in this manuscript.
+Note that the meaning of the terms may differ from how they are used in other fields, such as mathematics.
 The variable (1), in raw biophysical units, has a minimum and a maximum value (V), as well as points anchoring it to the indicator scale (X).
-Scaled variables (2) are scaled based on X0 and X100 only.
-Note that this is a different use of the term then the more common use that implies dividing a vector by its standard deviation.
-Truncation (3) implies assigning the value of 0 to values below X0, and the value 1 to value above X100.
-Transformation refers to the non-linear transformation of values within the 0-1 range, as in this piecewise-linear example by mapping X60 to 0.6 on the indicator scale.
+Scaled variables (2) are scaled based on X~0~ and X~100~ only.
+Truncation (3) implies assigning the value of 0 to values below X~0~, and the value 1 to value above X~100~.
+Transformation refers to the non-linear transformation of values within the 0-1 range, as in this piecewise-linear example by mapping X~60~ to 0.6 on the indicator scale.
 Transformations without anchoring points are also included in this term, such as exponential or sigmoid transformation.
 Although both truncation and scaling can be seen as types of transformations, we exclude these methods from the definition here.
-Examples 2, 3, and 4 are normative (hence the gradient colour scale from red to green) because they have declared that one end of the variable scale represents a good state and the other a poor state.
+Examples 2, 3, and 4 are normative (hence the gradient colour scale from red to green) because we assume they have declared that one end of the variable scale represents a good state and the other a poor state.
 All variables that are treated so that they conform to the definition of indicators in the SEEA EA are said to be normalised.
 For example, if a truncated variable is assumed to have a linear relationship with the indicator scale, then this variable has been normalised even though it has not been transformed.
-The term rescale is a synonym to normalise, but we will only use the latter from here on."
+The term rescale is a synonym to normalise, but we will only use the latter from here on.
 #| echo: false
@@
 
-```{r fig-pathways}
-#| fig-cap: |
-#|   "Flowchart showing pathways for the spatial aggregation of variables in ecosystem condition accounting.
-Normalisation refers to the steps that make a variable conform to the definition of an indicator in SEEA EA, which may or may not include the transformation step.
-A truncated and/or transformed variable will have a different interpretation than the original variable.
-The aggregation bias depends on whether variables are truncated and/or transformed variables before or after spatial aggregation.
-Variables are never normalised (pathway 4), and hence the aggregation bias for variables will be different than for indicators.
-An important factor influencing the the choice whether to normalise early (as in pathway 1), or later, is the spatial validity of the reference levels, and sometimes raw variable values (e.g.
-billberry coverage in vegetation quadrats) are spatially aggregated to the scale that was considered when setting the reference levels (e.g.
-average regional billberry coverage)."
-#| out-width: 400
-#| echo: false
-knitr::include_graphics("../img/aggregationPathways.jpg")
-```
+# Why do we normalise? {#sec-norm}
+
+Normalisation of ecosystem condition indicators, as defined in @fig-term, serves multiple purposes.
+Firstly, it gives a normative interpretation of a variable, defining a good and a bad state and a directionality to say when something is getting better or worse over time [@czucz_selection_2021].
+<!--# This normative characteristic is an essential part of assessing ecosystem condition.
+--> Secondly, normalisation sets a limit to how much a high indicator value in one place can compensate for a low value somewhere else, and vice versa.
+This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviation from the reference levels, and positive deviations (which could compensate for the negative ones) are ignored.
+Transformation can sometimes also have this effect, e.g. sigmoid transformations.
+<!--# This is often what we want (but not always), but not always something we are aware that we are doing.
+-->One reason to want to aggregate the negative deviations only, is because the reference levels are set (or should be set) so that values above these limits do not represent any further increase (or decrease) in ecosystem condition.
+(An exception if the WFD; see @sec-pti).
+Therefore, this way of aggregation summarises the estimated ecosystem condition (which is generally what we want), and not the variable itself.
+In other words, we normalise in order to facilitate the spatial (i.e. horizontal) aggregation of our ecosystem condition estimates (i.e. the indicator values), and not merely the variable values.
+Thirdly, we normalise in order to standardise the indicator on the same scale so that we can perform thematic (i.e. vertical) aggregation.
+This is commonly referred to as the reason for normalising variables, but as we have shown, it is but one of three main reasons, and also perhaps the least confusing part for many.
 
@@
 #| fig-cap: |
-#|   "Example showing the difference in indicator values (y-axis) obtained from taking the mean of two variable values (V1 and V1) that are on the original scale (i hat) or on the normalised indicator scale (i bar).
+#|   Example showing the difference in indicator values (y-axis) obtained from taking the mean of two variable values (V1 and V1) that are on the original scale (i hat) or on the normalised indicator scale (i bar).
 Due to truncation at x100, the latter results in a comparatively lower indicator value.
@@
 V = variable value, i = indicator value.
-Modified from Rastsetter (1991)."
+Modified from Rastsetter (1991).
 #| out-width: 400
@@
 #| fig-cap: |
-#|   "Same figure as above, but showing an example where the normalisation function includes a transformation step (see Figure 1 for definition) that anchors X60 to 0.6 on the indicator scale. "
+#|   Same figure as above, but showing an example where the normalisation function includes a transformation step (see Figure 1 for definition) that anchors X60 to 0.6 on the indicator scale.
 #| out-width: 400
@@
 
-## Why do we normalise? {#sec-norm}
+# Aggregation pathways
 
-Normalisation of ecosystem condition indicators, as defined in Figure 1, serves multiple purposes.
-Firstly, it gives a normative interpretation of a variable, defining a good and a bad state and a directionality to say when something is getting better or worse over time.
-This normative characteristic is an essential part of assessing ecosystem condition.
-Secondly, normalisation sets a limit to how much a high indicator value in one place can compensate for a low value somewhere else.
-This is because the truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviation from the reference levels, and positive deviations (which could compensate for the negative ones) are ignored.
-This is sometimes what we want, but not always something we are aware that we are doing.
-One reason to want to aggregate the negative deviations only, is because the reference levels are set (or should be set) so that values above these limits do not represent any further increase (or decrease) in ecosystem condition.
-Therefore, this way of aggregation summarises the presumed ecosystem condition (which is generally what we want), and not the variable itself.
-In other words, we normalise in order to facilitate the spatial (i.e. horizontal) aggregation of our ecosystem condition estimates (i.e. the indicator values), and not merely the variable values.
-Thirdly, we normalise in order to standardise the indicator on the same scale so that we can perform thematic (i.e. vertical) aggregation.
-This is commonly referred to as the reason for normalising variables, but as we have shown, it is but one of three main reasons, and also perhaps the least confusing part for many.
+We introduce the term aggregation pathway to describe the order of the steps used to go from a spatially explicit variable to a spatially aggregated metric, usually an indicator.
+In @fig-pathways we show some aggregation pathways that we have come across, and some that we we see as potential new pathways.
 
-## Main points to consider when deciding on aggregation pathway
+Pathway 1 involves early normalisation suing the perhaps most common order of the three steps scaling, truncating and transforming.
+The three steps may be done simultaneously as well.
+Then the indicator is spatially aggregated.
+Pathway 2 involves aggregating the variable before normalising.
+Pathway 3 is similar, but here there are two aggregation steps: one before and one after normalisation.
+This can for example be the case when variables are aggregated to the scale corresponding to the reference levels before they can be normalised.
+Pathway 4 illustrated the aggregation of a variable, with no normalisation.
+This is the pathway commonly used for variable accounts in the SEE EA. Pathway 5 is a common pathway in the WFD, which does not include spatial aggregation of indicator values.
+Instead one report the number of water bodies in each condition class (spatial statistics).
+Pathway 6 is a suggested pathway for handling WFD indicator in ECAs (see @sec-pti).
 
-### Ease of interpretation {.unnumbered}
+Pathways 7-8 describe the spatial aggregation of multiple indicators, as they all start with the normalisation step.
+These are not the main topic of this paper, but are included here for completeness.
+In pathway 7, indicators are thematically aggregated to an index, which is then again aggregated spatially.
+In pathway 8, spatial aggregation is only done on individual indicators.
+The two pathways differ in how they handle more complicated schemes for weighted averages, with pathway 8 being more flexible, but it may interfere with the principle of commutativity [@unitednations2021].
 
-The indicator should make intuitive sense.
-This is helped by using the same aggregation pathway for all indicators.
-Indicators using spatially aggregated population estimates are easily misinterpreted as representing population trends.
+Pathway 9 describe how a single indicators is aggregated hierarchically from region to region, whereas pathway 10 shows a case were the aggregation process always starts from scratch for each hierarchical spatial level.
+As for pathways 7 and 8, pathway 9 and 10 differ in how they propagate the indicator weight (typically area weight) between regions.
 
-### The ‘real’ spatial resolution of reference values {.unnumbered}
+The choice about which pathway to use is not trivial, and in this paper we want to highlight some of the issues that could arise from having an *ad hoc* approach to these considerations.
+Here we go through selected examples to illustrate some common pitfalls.
 
-To achieve an early normalisation of variables, we need reference values that make ecological sense at the same spatial resolution as the variable.
-Spatial disaggregation or interpolation of reference values needs to be well justified.
+<!--# what we see as four general points that developers of ecosystem condition indicators and assessments should reflect upon carefully when deciding on an aggregation pathway for their indicator or their assessment.
+-->
 
-### Scalability and comparability between assessment {.unnumbered}
+## Wolves example {#sec-wolves}
 
-Separate assessments should give identical indicator values for the same areas, given the same underlying data.
-When this is not true it raises concerns about reproducibility and sacrifices the credibility of both assessments.
+*Variable*: Number of wolves en each carnivore management area (CMA).
 
-### Making clear the differences in the interpretation of variables and indicators {.unnumbered}
+*Reference:* X~0~ = is zero wolves present.
+X~100~= Number of wolves equal to what experts think the ecosystem would support under the reference condition.
+The reference levels are unique to each (CMA).
 
-Often indicators and variables are both included side-by-side in ecosystem condition assessments/accounts.
-This is also recommended practice following the SEEA EA.
-However, if the indicators that are presented are normalised at an earlier stage, and subsequently spatially aggregated, same as the variable, then the interpretation of the indicators is not simply a normalised version of the variable, but rather a whole new beast.
-For example, these indicators typically now reflect the average ecosystem condition, where areas in very good or very poor condition have limited ability to compensate for opposite extreme values elsewhere.
-The spatially aggregated variable on the other hand, may reflect something like a sum of individuals, or some other aspect where values above and below the average value are able to compensate for each other.
+This example is from Norway, where wolves are extinct in most of the country.
+Therefore, most carnivore management areas will get an indicator value of 0.
+Wolves exists i designated wolf zones in south-east Norway, although in small numbers.
+In the Norwegian Natur Index [@jakobsson2020], the variable follows pathway 1 and is normalised at the CMA level.
+When aggregating the indicator to a national value, it is common for the general public to interpret this value as the status for wolves in Norway, and to connect this value to Norway's international obligations to maintain wolf numbers.
+But note that even an unlimited number of wolves in the wolf zone would not to bring the national indicator value up to a level indicator good condition.
+This is because the normalisation process truncates the variable, and does not allow overshooting values to compensate for lower values in the rest of the country.
 
-## Examples
+\*Conclusion\*: Indicators should make intuitive sense to the readers.
+However, there are sometimes good reasons for making things complicated.
+In this example, The Norwegian Nature Index uses the same pathway for all indicators, so this common missunderstanding about what indicator values represenets can be exaplin just one time, and not for each indicator spesifically.
+It can also be a good idea to sometime go back to the variable, and see if it can be defined differently to avoid these kinds of confusions.
+Wolf numbers could in this example be converted to wolf density.
 
-In this section we use some examples from our own experience to illustrate some of the issues arising from chosing different aggregation pathways,
-
-### Billberry example
+## Billberry example {#sec-bill}
 
@@
 
+The wolves example also highlight the importance of making clear the differences in the interpretation of variables and indicators.
+Often indicators and variables are both included side-by-side in ecosystem condition assessments/accounts.
+This is also recommended practice following the SEEA EA.
+However, if the indicators that are presented are normalised before any spatial aggregation, then the interpretation of the indicators is not simply as a normalised version of the variable.
+Indicators typically now reflect the average ecosystem condition, where areas in very good or very poor condition have limited ability to compensate for opposite extreme values elsewhere.
+The spatially aggregated variable on the other hand, may reflect something like a sum of individuals, or some other aspect where values above and below the average value are able to compensate for each other.
+
 ### Alien species example
@@
 
-### Phytoplankton trophic index (PTI)
+Separate assessments should give identical indicator values for the same areas, given the same underlying data.
+When this is not true it raises concerns about reproducibility and sacrifices the credibility of both assessments.
+This was also the case in the billberry example, but here we used another example to illustrate the point further.
+
+### Phytoplankton trophic index (PTI) {#sec-pti}
 
@@
 
+```{r fig-pathways}
+#| fig-cap: |
+#|   Flowchart showing pathways for the spatial aggregation of variables in ecosystem condition accounting.
+Normalisation refers to the steps that make a variable conform to the definition of an indicator in SEEA EA, which may or may not include the transformation step.
+A truncated and/or transformed variable will have a different interpretation than the original variable.
+The aggregation bias depends on whether variables are truncated and/or transformed variables before or after spatial aggregation.
+Variables are never normalised (pathway 4), and hence the aggregation bias for variables will be different than for indicators.
+An important factor influencing the the choice whether to normalise early (as in pathway 1), or later, is the spatial validity of the reference levels, and sometimes raw variable values (e.g.
+billberry coverage in vegetation quadrats) are spatially aggregated to the scale that was considered when setting the reference levels (e.g.
+average regional billberry coverage).
+#| out-width: 400
+#| echo: false
+knitr::include_graphics("../img/aggregationPathways.jpg")
+```
+
+# Recommendation
+
+Based on the example above, we suggest some recommendations for developers of ecosystem condition indicators and assessment for trying to avoid some of the pitfalls from having a too casual approach to the choice of aggregation pathways.
+
+### 1. Always report the aggregation pathway, using standardised terminology {.unnumbered}
+
+In indicator documentation or ECA repports, authors should make an effort to make a precise description of the steps in the aggregation process.
+To do this we suggest using the terms as described in @fig-term.
+The description can be made even easier by referencing specific aggregation pathways in @fig-pathways by number.
+Besides making the indicator workflow itself more transparrent and reproduciable, ee believe this recommendation would help raise the general awareness about aggregation bias in EAs, both among readers and researchers.
+It will make the interpretation of assessment less prone to misunderstanding.
+It will also make troubleshooting easier, as i the billberry example where researchers had to spend several days looking though old code to identify the reason why two assessment produces dissimilar indicator values from the same underlying data.
+
+### 2. Always normalise variables at the scale where the reference levels are relevant {.unnumbered}
+
+Normalising variables early in the aggregation pathway has several benefits, as we have seen.
+But same as variables and indicators, reference levels also have a spatial resolution, and they should only be used to normalise variables when the variable is at a scale which is relevant to the way the reference levels are defined.
+This was exeplified with the billberry example (@sec-bill).
+
+### 3. Use the same aggregation pathway for all indicators in the same assessment {.unnumbered}
+
+As we saw in the wolves example (@sec-wolves) this eases interpretation and reduces the change that the meaning of indicators are misunderstood.
+It may be premature, or not even possible, to prescribe an aggregation pathway to be used for all ECA, but internal consistency within assessment should be possible.
+
+### 4. Use unique indicator IDs on indicators that are similar, but use different pathways {.unnumbered}
+
+Indicators (and variables) often exist in multiple version, varying slightly in the raw data or in the methods used to produce the data or metrics.
+Yet different versions are often refered to by the same common name.
+This causes confusion about which indicator version is being used, and thus making it difficult to make out which aggregation pathway that has been used.
+We recommend making use of stable indicator IDs, unique for each version of the indicators sharing the same common name.
+This has for example been implemented in [ecRxiv](https://github.com/NINAnor/ecRxiv), a publishing platform for ecosystem condition indicators.
+
 # References {.unnumbered}

### 2024-11-20 manuscript/manuscript.qmd
add normalise early

@@
 
-### Alien species example
+## Alien species example {#sec-ali}
 
@@
 
-### Phytoplankton trophic index (PTI) {#sec-pti}
+## Phytoplankton trophic index (PTI) {#sec-pti}
 
@@
 
-### 1. Always report the aggregation pathway, using standardised terminology {.unnumbered}
+### 1. Report the aggregation pathway, using standardised terminology {.unnumbered}
 
@@
 
-### 2. Always normalise variables at the scale where the reference levels are relevant {.unnumbered}
+### 2. Normalise variables early, but at the scale where the reference levels are relevant {.unnumbered}
 
-Normalising variables early in the aggregation pathway has several benefits, as we have seen.
+Normalising variables early in the aggregation pathway has several benefits, such as comparability between nested assessments (@sec-ali).
+It also means that you are aggregating normative measures of condition, which is generally what you want in ECAs.
 But same as variables and indicators, reference levels also have a spatial resolution, and they should only be used to normalise variables when the variable is at a scale which is relevant to the way the reference levels are defined.
-This was exeplified with the billberry example (@sec-bill).
+This was exemplified with the billberry example (@sec-bill).
 
@@
 
-As we saw in the wolves example (@sec-wolves) this eases interpretation and reduces the change that the meaning of indicators are misunderstood.
-It may be premature, or not even possible, to prescribe an aggregation pathway to be used for all ECA, but internal consistency within assessment should be possible.
+It may be premature, or not even possible, to prescribe an aggregation pathway to be used for all ECAs, but internal consistency within assessments should be possible.
+As we saw in the wolves example (@sec-wolves), this eases interpretation and reduces the change that the meaning of indicators are misunderstood.
 

### 2024-11-20 manuscript/manuscript.qmd
fix typos

@@
 When a variable is normalised in this way, its interpretation changes from descriptive to normative.
-The ecological significance of the indicator might also differ from that of the variable since additional ecological knowledge can be introduced via the reference leveles.
+The ecological significance of the indicator might also differ from that of the variable since additional ecological knowledge can be introduced via the reference levels.
 The indicator is a function of the variable and the reference values, and this relationship is almost always nonlinear (e.g. due to truncation which is ubiquitous practice).
@@
 Wolves exists i designated wolf zones in south-east Norway, although in small numbers.
-In the Norwegian Natur Index [@jakobsson2020], the variable follows pathway 1 and is normalised at the CMA level.
+In the Norwegian Nature Index [@jakobsson2020], the variable follows pathway 1 and is normalised at the CMA level.
 When aggregating the indicator to a national value, it is common for the general public to interpret this value as the status for wolves in Norway, and to connect this value to Norway's international obligations to maintain wolf numbers.
@@
 However, there are sometimes good reasons for making things complicated.
-In this example, The Norwegian Nature Index uses the same pathway for all indicators, so this common missunderstanding about what indicator values represenets can be exaplin just one time, and not for each indicator spesifically.
+In this example, The Norwegian Nature Index uses the same pathway for all indicators, so this common misunderstandings about what indicator values represents can be explain just one time, and not for each indicator specifically.
 It can also be a good idea to sometime go back to the variable, and see if it can be defined differently to avoid these kinds of confusions.
@@
 
-*Variable*: Horizontal coverage (vertical projection) of billberry (Vaccinum myrtillus) recorded in permanent vegetation plots.
+*Variable*: Horizontal coverage (vertical projection) of billberry (*Vaccinum myrtillus*) recorded in permanent vegetation plots.
 
@@
 *Variable*: The local ecological effect from alien species (mostly vascular plants) recorded on a 7-step ordinal scale (1 = no alien species, 7 = only alien species).
-The data comes from nature type monitoring with reference to individual occurences (polygons).
+The data comes from nature type monitoring with reference to individual occurrences (polygons).
 
@@
 
-In indicator documentation or ECA repports, authors should make an effort to make a precise description of the steps in the aggregation process.
+In indicator documentation or ECA reports, authors should make an effort to make a precise description of the steps in the aggregation process.
 To do this we suggest using the terms as described in @fig-term.
 The description can be made even easier by referencing specific aggregation pathways in @fig-pathways by number.
-Besides making the indicator workflow itself more transparrent and reproduciable, ee believe this recommendation would help raise the general awareness about aggregation bias in EAs, both among readers and researchers.
+Besides making the indicator workflow itself more transparent and reproducible, ee believe this recommendation would help raise the general awareness about aggregation bias in EAs, both among readers and researchers.
 It will make the interpretation of assessment less prone to misunderstanding.
@@
 It may be premature, or not even possible, to prescribe an aggregation pathway to be used for all ECAs, but internal consistency within assessments should be possible.
-As we saw in the wolves example (@sec-wolves), this eases interpretation and reduces the change that the meaning of indicators are misunderstood.
+As we saw in the wolves example (@sec-wolves), this eases interpretation and reduces the chance that the meaning of indicators are misunderstood.
 
@@
 Indicators (and variables) often exist in multiple version, varying slightly in the raw data or in the methods used to produce the data or metrics.
-Yet different versions are often refered to by the same common name.
+Yet different versions are often refereed to by the same common name.
 This causes confusion about which indicator version is being used, and thus making it difficult to make out which aggregation pathway that has been used.

### 2024-11-25 manuscript/manuscript.qmd
inc base size

@@
     affiliations:
-      - id: nina
+      - ref: nina
   - name: Bálint Czúcz

### 2024-11-25 R/aggregationBiasFigure.R

@@ -90,7 +90,7 @@ dat_x |>
     )
   ) +
   guides(color = "none") +
-  theme_bw(base_size = 16) +
+  theme_bw(base_size = 30) +
   theme(panel.grid.minor = element_blank()) +
   geom_segment(aes(x = 50, y = fV1, xend = V1, yend = fV1),
     color = lc, linewidth = 2, linetype = lt
@@ -210,7 +210,7 @@ dat3 |>
     )
   ) +
   guides(color = "none") +
-  theme_bw(base_size = 16) +
+  theme_bw(base_size = 30) +
   theme(panel.grid.minor = element_blank()) +
   geom_segment(aes(x = 50, y = fV1x, xend = V1, yend = fV1x),
                color = lc, linewidth = 2, linetype = lt

### 2025-01-08 manuscript/manuscript.qmd
Change to authoryear

@@
       model: 3p
-      cite-style: super
+      cite-style: authoryear
     include-in-header:
@@
 This means that neither EQR values (which are untransformed), nor nEQR values (which are truncated), can be used uncritically in ECAs.
-One solution, to enable the use of WFD indicators in ECA, would be to change to pathway 6, with (1) scaling, (2) transforming, (3) aggregating and (4) truncating the values [@fig.pathways].
+One solution, to enable the use of WFD indicators in ECA, would be to change to pathway 6, with (1) scaling, (2) transforming, (3) aggregating and (4) truncating the values [@fig-pathways].
 Furthermore, Eurostat @eurostat2022\

### 2025-01-08 manuscript/manuscript.qmd
Why didnt this get merged

@@
 
-In ecosystem accounting (EA), fine-scale ecological information is routinely aggregated in space to produce single parameter estimates for entire regions.
-This simplification is necessary to show major changes in the extent and condition of ecosystems at a scale that is relevant for decision makers.
-The spatial aggregation of data can lead to biases arising from a number of sources, such as uneven sampling efforts, unequal sampling designs.
-In ecosystem condition accounting (ECA) specifically, the aggregation of both variables and their normative indicators is common practice and raises the issue of aggregation bias when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
-The first two sources of error are usually explicitly addressed and managed, but the last case of aggregation bias is often overlooked.
+In ecosystem accounting (EA), <!--# Maybe start more general than the EA case -->fine-scale ecological information is routinely aggregated in space to produce single parameter estimates for entire regions.
+This simplification is necessary <!--# at the expence of aiding debate, use in spatial planning, and increased risk of oversimlipfiction.
+So perhaps not 'necessary'.
+-->to show major changes in the extent and condition of ecosystems at a scale that is relevant for decision makers <!--# and for reporting -->.
+The spatial aggregation of data can lead to biases arising from a number of sources, such as uneven sampling efforts or sampling designs between regions.
+<!--# Spatial aggregation can also change the interprettation, descriptive properties and normative claimes one can make based on the indicator.
+Developer subjectivity is a big factor, and the process is not very democratic (Alain et al 2018 --> In ecosystem condition accounting (ECA) specifically, the aggregation of both variables and their normative indicators is common practice and raises the issue of aggregation bias when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
+The first two sources of error are usually explicitly addressed and managed <!--# what is ment here? -->, but the last case of aggregation bias is often overlooked.
 In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2021; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
-Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (a thematic aggregation), but in this paper we are discussing spatial aggregation only.
+Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (i.e.
+thematic or vertical aggregation; @langhans2014), but in this paper we are discussing spatial (i.e.
+horizontal) aggregation only.
 
-Variables in ecosystem condition terminology are metrics describing ecosystem characteristics [@unitednations2021].
+Variables in ecosystem condition terminology <!--# need to introduce SEEA EA first --> are metrics describing ecosystem characteristics [@unitednations2021].
 They have a spatial component, tying them to concrete areas, which can be of varying size, from single pixels to entire regions.
@@
 The order of these steps can also change.
-Variable can be scaled using at least two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
-Variables can be truncated to produce a bound indicator scale between 0 and 1.
-Sometimes variables are transformed to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
+Variable can be *scaled* using a minimum of two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
+Variables can be *truncated* to produce a bound indicator scale between 0 and 1.
+Sometimes variables are *transformed* to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
 This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
-When a variable is normalised in this way, its interpretation changes from descriptive to normative.
+When a variable is normalised in this way, its interpretation also changes and it becomes predominantly normative, designed to make claims about what is good and what is considered bad <!--# variables can also be normative in some sense, due to value judgement entering the variable selection or customisation process; Allain et al 2018 -->.
 The ecological significance of the indicator might also differ from that of the variable since additional ecological knowledge can be introduced via the reference levels.
-The indicator is a function of the variable and the reference values, and this relationship is almost always nonlinear (e.g. due to truncation which is ubiquitous practice).
+The indicator-variable relationship is almost always nonlinear (e.g. due to truncation which is ubiquitous practice).
 
@@
 the aggregation pathway (@fig-pathways).
+
+<!--# the intro could say something about the state of the art.
+Many have studied aggregation, mosty vertical aggregation, but also horisontal aggregation.
+E.g.
+Allain et al .
+But we know of none that explicily have looked at SEEA EA indicators and ecosystem accounting purposes.
+-->
 
@@
 <!--# This normative characteristic is an essential part of assessing ecosystem condition.
---> Secondly, normalisation sets a limit to how much a high indicator value in one place can compensate for a low value somewhere else, and vice versa.
+- But other framework dont use normalisation, and leave all value judgements to the users, so then it is not an essentiasl part then realy?
+--> Secondly, normalisation sets a limit to how much a high indicator value in one place can compensate for a low value somewhere else, and vice versa <!--# can we cite the Nature index?
+-->.
 This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviation from the reference levels, and positive deviations (which could compensate for the negative ones) are ignored.
@@
 Therefore, this way of aggregation summarises the estimated ecosystem condition (which is generally what we want), and not the variable itself.
-In other words, we normalise in order to facilitate the spatial (i.e. horizontal) aggregation of our ecosystem condition estimates (i.e. the indicator values), and not merely the variable values.
+In other words, normalising varables facilitate the spatial aggregation of our ecosystem condition estimates (i.e. the indicator values), and not merely the variable values.
 Thirdly, we normalise in order to standardise the indicator on the same scale so that we can perform thematic (i.e. vertical) aggregation.
@@
 ```
+
+# Why to we spatially aggregate data in ecosystem condition accounting
+
+<!--# We discussed why we normalise, so why not discuss the reason for aggregating -->
+
+Spatiall aggregation of data refers to the conversion of fine-resolution data into coarser-resolution data, and depending on both the purpose and the nature of the metric, there are several reasons for performing spatial aggregation, as well as for refraining [@allain2018].
+Generally, spatial aggregation increases the tangibility of a metric, because it simplifies or compresses the information, sometimes into just a single number (total aggregation).
+This is necessary in ecosystem accounting when information is required at a relatively high administrative level (e.g.
+nations) and when variables and indicators need to be presented in standardised tables.
+Sometimes total aggregation of a variable or indicator is required for the metric to actually fit its purpose and become relevant to what it is designed to describe.
+For example, a hypothetical metric "*proportion of total land area above critical nitrogen load",* informed from a spatial map of nitrogen loads, would only make sense it data was aggregated across some defined area.
+The reason and need for aggregation may therefore originate in the original idea for the metric.
+
+Data may also be only party aggregated, or the same metric may be presented at different levels of aggregation.
+The minimum level of aggregation may also be decided based on the uncertainties in the data.
+High spatially resolute data may be imprecise at fine scales, leading to false conclusion and potential miss-use, but these values seen over a larger area may be assumed to converge on the true value, making spatially aggregated data better suited for communication to users.
+Other common arguments for performing spatial aggregation are to ease visualisation and to allow relevant comparisons.
+For example, ecosystem condition metrics can be visualised on a map, but when both the the spatial resolution and spatial variation is high, maps become too complex, even unintelligible.
+It is therefore common to partially aggregate the data and provide a coarser map where different colours, representing values on the map, become distinguishable from each other.
+Similarly, one may chose to aggregate such that the spatial grit matches some meaningful delineation, notably ecosystem assets in SEEA EA, which makes it possible to easily compare the aggregated values between these areas.
+Sometimes variables are aggregated in order to become relevant for the
+
+However, spatial aggregation can have several consequences that may or may not be desirable, such as homogenisation of spatial variability which, depending on the aggregation function (e.g.
+weighted arithmetic mean, worst-rule, or others), influences the degree of compensation, i.e.
+the degree to which good condition in one place can compensate for bad condition somewhere else.
+Therefore, spatial aggregation can introduce substantial subjectivity from the side of the indicator developers, which also limits the different ways that a metric can be used and interpreted by indicator users (the target audience).
+In other words, spatial aggregation tasks indicator developers with defining narratives for underlying variables, and it is therefor a normative and value-laden operation.
+
+In conclusion, there are several requirements in the SEEA EA the makes spatial aggregation necessary, but spatial aggregation also comes with a whole suite of (sometimes undesirable) side-effects.
+Readers are referred to @allain2018 for a review of different reasons for spatially aggregating indicators.
 

### 2025-01-15 manuscript/manuscript.qmd
Writing

--- 2025-01-08 manuscript/manuscript.qmd
@@
 
-In ecosystem accounting (EA), fine-scale ecological information is routinely aggregated in space to produce single parameter estimates for entire regions.
-This simplification is necessary to show major changes in the extent and condition of ecosystems at a scale that is relevant for decision makers.
-The spatial aggregation of data can lead to biases arising from a number of sources, such as uneven sampling efforts, unequal sampling designs.
-In ecosystem condition accounting (ECA) specifically, the aggregation of both variables and their normative indicators is common practice and raises the issue of aggregation bias when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
-The first two sources of error are usually explicitly addressed and managed, but the last case of aggregation bias is often overlooked.
-In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2021; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
-Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (a thematic aggregation), but in this paper we are discussing spatial aggregation only.
+Ecosystem accounting (EA), and other types of environmental assessments, are generally aimed at communicating major changes and trends in the extent and condition of ecosystems at a scale that is relevant for decision makers.
+For ecosystem condition accounting (ECA) and assessments, this often requires that **metrics**[^1] used to describe the condition of ecosystems are aggregated spatially to produce a single value for a larger area.
+This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something actionable.
+However, spatial aggregation can and will often change the interpretation of a metric and the descriptive properties and normative claims one can make based on it.
+Without a good enough understand of the pitfalls in spatial aggregation of information, is can also lead to unintended and undetected systematic bias, as well as downright logical errors.
+Although much attention has been devoted to the vertical aggregation of environmental indicators into indices (ref), as well as to the spatial aggregation of relatively unmodified or raw variables, there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting.
+This paper therefore focus on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) - the most used ecosystem accounting (natural resource accounting) framework in the world (ref).
+We pay specific attention to the phenomenon of aggregation bias introduce by @rastetter1992.
 
+[^1]: Words in bold are explained in Table 1
+
+### SEEA EA
+
+SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services.
+We will focus on the SEEA EA condition accounts in this paper.
+In these accounts, information about ecosystem condition is made spatially explicit by assigning **variable** or **indicator** values to individual ecosystem assets.
+Unless the underlying data are wall-to-wall maps, for example from satellite images, then this process of assigning values to ecosystem assets already here requires some spatial manipulations in the form of aggregation, dissaggregation, extrapolation or interpolation (ref Ewert 2011?), of which averaging or summing are the most common methods.
+Subsequently, information from ecosystem assets may be spatially aggregated to larger areas, such as the entire ecosystem accounting area.
+When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels.
+The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.
+
+### Indicators vs. variables
+
+SEEA EA distinguishes between indicators and variables.
 Variables in ecosystem condition terminology are metrics describing ecosystem characteristics [@unitednations2021].
-They have a spatial component, tying them to concrete areas, which can be of varying size, from single pixels to entire regions.
-Variables can be spatially aggregated, for example by taking the sum or an area-weighted mean, to produce variable estimates for larger regions, such as the entire Ecosystem Accounting Area (EAA).
-Variables can also be normalised and turned into normative indicators for ecosystem condition.
+They are typically either in units or percentages or fractions, or as raw biophysical units, such as biomass, density, or cover.
+Indicators on the other hand, are always on a scale between 0 and 1, where 0 represent the worst condition, and 1 represents the best condition.
+Sometimes indicators are calculated directly, but usually indicators are made by **normalising** variables based on two or more **reference levels**.
+Although also variables may have normative properties (due to choices about how they are chosen (ref), defined, aggregated or presented) (Alain et al), the normalisation of variables into indicators makes indicators explicitly normative by introducing a clear directional interpretation, where increasing values always refer to an increase in condition.
+The information held by the indicator is also different from that of the variable since additional ecological knowledge can be introduced via the reference levels.
 The mathematical part of this normalisation contains two, sometimes three, steps [@fig-term].
@@
 The order of these steps can also change.
-Variable can be scaled using at least two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
-Variables can be truncated to produce a bound indicator scale between 0 and 1.
-Sometimes variables are transformed to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
+
+Variables can be **scaled** using a minimum of two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
+Variables can be **truncated** to produce a bound indicator scale between 0 and 1.
+Sometimes variables are **transformed** to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
 This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
-When a variable is normalised in this way, its interpretation changes from descriptive to normative.
-The ecological significance of the indicator might also differ from that of the variable since additional ecological knowledge can be introduced via the reference levels.
-The indicator is a function of the variable and the reference values, and this relationship is almost always nonlinear (e.g. due to truncation which is ubiquitous practice).
 
-Similar to variables, the reference values used for normalising the variables also have a spatial assignment, but it can be different from that of the variable.
-For example, variable values may exist for unique 10×10m grid cells, but the reference values may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire EAA (e.g.
+| Term             | Definition |
+|------------------|------------|
+| Metric           |            |
+| Variable         |            |
+| Indicator        |            |
+| Reference levels |            |
+| Scaling          |            |
+| Transformation   |            |
+| Truncation       |            |
+| Normalisation    |            |
+|                  |            |
+|                  |            |
+
+: Definition of terms
+
+# Why to we spatially aggregate data in ecosystem condition accounting
+
+Spatiall aggregation of data refers to the conversion of fine-resolution data into coarser-resolution data, and depending on both the purpose and the nature of the metric, there can be several reasons both for performing and for refraining from spatial aggregation[@allain2018].
+Generally, spatial aggregation increases the tangibility of a metric, because it simplifies or compresses the information, sometimes into just a single number (total aggregation).
+This is necessary in ecosystem accounting when information is required at a relatively high administrative level (e.g.
+nations) and when variables and indicators need to be presented in standardised tables.
+Sometimes total aggregation of a variable or indicator is required for the metric to actually fit its purpose and become relevant to what it is designed to describe.
+For example, a hypothetical metric "*proportion of total land area above critical nitrogen load",* informed from a spatial map of nitrogen loads, would only make sense if data was aggregated across some defined area.
+The reason and need for aggregation may therefore originate in the original idea for the metric.
+Data may also be only party aggregated, or the same metric may be presented at different levels of aggregation [@allain2018].
+
+A minimum level of aggregation may be decided on based on the uncertainties in the data.
+High spatially resolute data may be imprecise at fine scales, leading to false conclusion and potential miss-use [@lyytimäki2013], but these values seen over a larger area may be assumed to converge on some true value, making spatially aggregated data better suited for communication to users.
+Other common arguments for performing spatial aggregation are to ease visualisation and to allow relevant comparisons [@allain2018].
+For example, ecosystem condition metrics can be visualised on a map, but when both the the spatial resolution and spatial variation is high, maps become too complex, even unintelligible.
+It is therefore common to partially aggregate the data and provide a coarser map where different colours, representing values on the map, become distinguishable from each other.
+Similarly, one may chose to aggregate such that the spatial resolution matches some meaningful delineation, notably ecosystem assets in SEEA EA, which makes it possible to easily compare the aggregated values between these areas.
+
+However, spatial aggregation can have several consequences that may or may not be desirable, such as smoothing out spatial variability, which can mask local patterns and trends.
+This smoothing out makes the data and its inherent information less suitable for other uses, such as depicting spatial variation that can be used to support local land use planning, or for aiding a debate about spatial conflicts.
+
+### Matching the scale of variables and reference levels
+
+Similar to variables, the reference levels used for normalising variables also have a spatial assignment, but it can be different from that of the variable.
+For example, variable values may exist for unique 10 × 10 m grid cells, but the reference levels may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire accounting area (e.g.
 a natural zero).
-A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition indicators (@fig-agg1; @fig-agg2), is the choice about the order of the different steps in the normalisation and aggregation process, i.e.
-the aggregation pathway (@fig-pathways).
+Variables should not be normalised based on reference levels unless they are at a comparable scale.
+Let's illustrate this with an example using a hypothetical variable depicting wild animal densities.
+If the upper reference level (X~100~) for this variable is given by the assumed (or measured) animal densities under the **reference condition** (the optimal condition), then that reference level needs to be assigned a relevant spatial scale, for example a region where the densities are calculated over.
+One will probably acknowledge the fact that animals move and distribute themselves non-randomly within these regions, and it therefore makes little sense to normalise the variable at a finer scale than region.
+Doing so would mean that regions perfectly aligned with the reference condition (in terms of animal densities with the region) will show as having sub-regions with deteriorated condition simply due to natural variation in where the animals are at any give time.
+In addition, averaging the indicator values across sub-regions would not produce a regional indicator value that represents its true situation as being perfectly aligned with the reference condition, since truncation ensures that variable values above X~100~ (assuming X~100~ \> X~0~) means we are only averaging the deviations from the reference level, and that very high animal densities in some sub-regions are not allowed to compensate for low values elsewhere.
+This example with wild animal densities shows that sometimes variables need to be spatially aggregated before they can be normalised.
+
+In conclusion, there are several requirements in the SEEA EA that make spatial aggregation necessary, but spatial aggregation also comes with a whole suite of (sometimes undesirable) side-effects.
+Readers are referred to @allain2018 for a review of different reasons for spatially aggregating indicators.
 
@@
 Firstly, it gives a normative interpretation of a variable, defining a good and a bad state and a directionality to say when something is getting better or worse over time [@czucz_selection_2021].
-<!--# This normative characteristic is an essential part of assessing ecosystem condition.
---> Secondly, normalisation sets a limit to how much a high indicator value in one place can compensate for a low value somewhere else, and vice versa.
+SEEA EA encourages normalisation, but other environmental assessment framework leaves more of these kinds of value judgments to the end users (ref?).
+
+Secondly, normalisation sets a limit to how much a high indicator value in one place can compensate for a low value somewhere else, and vice versa [@pedersen2016].
 This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviation from the reference levels, and positive deviations (which could compensate for the negative ones) are ignored.
 Transformation can sometimes also have this effect, e.g. sigmoid transformations.
-<!--# This is often what we want (but not always), but not always something we are aware that we are doing.
--->One reason to want to aggregate the negative deviations only, is because the reference levels are set (or should be set) so that values above these limits do not represent any further increase (or decrease) in ecosystem condition.
-(An exception if the WFD; see @sec-pti).
-Therefore, this way of aggregation summarises the estimated ecosystem condition (which is generally what we want), and not the variable itself.
-In other words, we normalise in order to facilitate the spatial (i.e. horizontal) aggregation of our ecosystem condition estimates (i.e. the indicator values), and not merely the variable values.
-Thirdly, we normalise in order to standardise the indicator on the same scale so that we can perform thematic (i.e. vertical) aggregation.
-This is commonly referred to as the reason for normalising variables, but as we have shown, it is but one of three main reasons, and also perhaps the least confusing part for many.
+One reason to want to aggregate the negative deviations only, is because the reference levels are set (or should be set) so that values above these limits do not represent any further increase (or decrease) in ecosystem condition (an exception if the WFD; see @sec-pti).
+Therefore, this way of aggregation summarises the estimated ecosystem condition over an area (which is generally what we want), and not the variable itself.
+In other words, normalising variables facilitate the spatial aggregation of our ecosystem condition estimates (i.e. the indicator values).
+
+Thirdly, we normalise in order to standardise the indicator on the same scale so that we can perform thematic (i.e. vertical) aggregation (ref).
+This is commonly referred to as *the* reason for normalising variables, but as we have shown, it is but one of three main reasons, and also perhaps the least confusing part for many.
 
@@
 ```
+
+# Aggregation bias
+
+The spatial aggregation of data can lead to biases arising from a number of sources.
+For example, direct extrapolation based on uneven or spatially biased sampling leads to biased estimates [@kolstad_making_2024].
+However, we are here mostly interested in what we consider an extended case of aggregation error, sensu @rastetter1992.
+They showed that when we take the average of two variables, each of which have associated attribute values (e.g.
+indicator values), then the attribute of the average variable value differs from the average of the attributes of the original variables.
+This displacement is refereed to as aggregation error, and its effect become more pronounced when the non-linearity is strong between variables and indicators.
+
+Also, in ecosystem condition accounting (ECA) specifically, the aggregation of both variables and indicators is common practice and raises the issue of **aggregation bias** (@fig-agg1; @fig-agg2) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
+
+A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition indicators (@fig-agg1; @fig-agg2), is the choice about the order of the different steps in the normalisation and aggregation process, i.e.
+the aggregation pathway (@fig-pathways).
+
+# Indicator developer subjectivity
+
+Indicator developer subjectivity is a big factor as the process of designing indicators and estimating indicator values often left entirely to the them, with little real involvement from indicator users.
+It is therefore important to be aware of the role that indicator developers have in shaping the type of information, and the narratives, that is presented to decision makers.
+This paper is aimed at providing some recommendations for how to choose the right spatial aggregation method for variables and indicators, and how to report these choices.
+This will make it easier for indicator developers to avoid some of the pitfalls from having a too casual approach to the choice of aggregation pathways.
+
+The choice of aggregation function (e.g.
+weighted arithmetic mean, worst-rule, or others) also influences the degree of compensation, i.e.
+the degree to which good condition in one place can compensate for bad condition somewhere else (ref WFD?).
+Therefore, spatial aggregation can introduce substantial subjectivity from the side of the indicator developers, which also limits the different ways that a metric can be used and interpreted by indicator users (the target audience).
+In other words, spatial aggregation tasks indicator developers with defining narratives for underlying variables, and it is therefor a normative and value-laden operation.
+
+Something about sometimes spatial aggregation is an essential part of the indicator design ---
+
+In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2021; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
+Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (i.e.
+thematic or vertical aggregation; @langhans2014), but in this paper we are discussing spatial (i.e.
+horizontal) aggregation only.
 

### 2025-01-21 manuscript/manuscript.qmd
writing

@@
 | Normalisation    |            |
-|                  |            |
+| Compensation     |            |
 |                  |            |
@@
 Generally, spatial aggregation increases the tangibility of a metric, because it simplifies or compresses the information, sometimes into just a single number (total aggregation).
-This is necessary in ecosystem accounting when information is required at a relatively high administrative level (e.g.
-nations) and when variables and indicators need to be presented in standardised tables.
+This is sometimes necessary in ecosystem accounting when information is required at a relatively high administrative level (e.g.
+nations) and when variables and indicators need to be presented in standardised tables which dont allow for describing a lot of spatial variation.
 Sometimes total aggregation of a variable or indicator is required for the metric to actually fit its purpose and become relevant to what it is designed to describe.
@@
 However, spatial aggregation can have several consequences that may or may not be desirable, such as smoothing out spatial variability, which can mask local patterns and trends.
-This smoothing out makes the data and its inherent information less suitable for other uses, such as depicting spatial variation that can be used to support local land use planning, or for aiding a debate about spatial conflicts.
+This smoothing out makes the data and its inherent information less suitable for some uses, such as depicting spatial variation that can be used to support local land use planning, or for aiding a debate about spatial conflicts.
 
-### Matching the scale of variables and reference levels
+### Matching the scale of variables and reference levels {#sec-match}
 
@@
 Doing so would mean that regions perfectly aligned with the reference condition (in terms of animal densities with the region) will show as having sub-regions with deteriorated condition simply due to natural variation in where the animals are at any give time.
-In addition, averaging the indicator values across sub-regions would not produce a regional indicator value that represents its true situation as being perfectly aligned with the reference condition, since truncation ensures that variable values above X~100~ (assuming X~100~ \> X~0~) means we are only averaging the deviations from the reference level, and that very high animal densities in some sub-regions are not allowed to compensate for low values elsewhere.
+In addition, averaging the indicator values across sub-regions would not produce a regional indicator value that represents its true situation as being perfectly aligned with the reference condition, since truncation ensures that variable values above X~100~ (assuming X~100~ \> X~0~) means we are only averaging the deviations from the reference level, and that very high animal densities in some sub-regions are not allowed to **compensate** for low values elsewhere.
 This example with wild animal densities shows that sometimes variables need to be spatially aggregated before they can be normalised.
 
-In conclusion, there are several requirements in the SEEA EA that make spatial aggregation necessary, but spatial aggregation also comes with a whole suite of (sometimes undesirable) side-effects.
+There are several requirements in the SEEA EA that make spatial aggregation necessary, but spatial aggregation also comes with a whole suite of (sometimes undesirable) side-effects.
 Readers are referred to @allain2018 for a review of different reasons for spatially aggregating indicators.
@@
 For example, direct extrapolation based on uneven or spatially biased sampling leads to biased estimates [@kolstad_making_2024].
-However, we are here mostly interested in what we consider an extended case of aggregation error, sensu @rastetter1992.
+In this paper, however, we are mostly interested in what we consider an extended case of aggregation error, sensu @rastetter1992.
 They showed that when we take the average of two variables, each of which have associated attribute values (e.g.
@@
 
-Also, in ecosystem condition accounting (ECA) specifically, the aggregation of both variables and indicators is common practice and raises the issue of **aggregation bias** (@fig-agg1; @fig-agg2) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
+## Aggregation error and discrepancies between variables and indicators
 
+In ecosystem condition accounting the aggregation of both variables and indicators is common practice and raises the issue of **aggregation bias** (@fig-agg1; @fig-agg2) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
+This means that the spatial aggregation of metrics leads to a divergence in the type of information held in variables and in their associated indicators, to a point where they become qualitatively very different.
+As an example, consider an indicator based on spatial information on the abundance of a large herbivore.
+The ecosystem characteristic that this metric is meant to describe is "*Appropriate herbivore pressure, balancing overgrazing and shrubification"*.
+The variable shows relatively high densities in half the area and low densities in the other half.
+When spatially aggregating the variable, either by taking the mean density or the sum of individuals, will produce an estimate that is somewhere in the middle of the two regions.
+The user will have no information about the carrying capacity for each of the regions, and will only see the spatially aggregated value.
+From this, many different conclusion can be made, depending on what the user consider high or low densities.
+In any case, the judgment will not reflect a density that actually exists anywhere.
+The indicator on the other hand, can have different reference levels for each region, introducing additional ecological information about the carrying capacity of the ecosystems.
+The levels may be such that the high density region is considered as having an overabundance of herbivores, and the low density region as having underabundance.
+When normalised, the indicator value will be low for both regions, and the spatially aggregated value will also be low.
+The indicator will describe the average condition for the intended ecosystem characteristic, whereas the variable will in essence not show anything that can be interpreted as such.
+
+The indicator arguably has much less room for miss-interpretation (relative to its intended use).
+Although one common misunderstanding is that indicators named by an animal (e.g. reindeer) is taken by untrained users to represent a population status (typically the number of individuals).
+This would, however, imply that the spatial aggregation function used for the indicator is summation, but the recommended, on assumed most used function, is an area weighted mean.
+With this aggregation function, indicators will always refer to the mean indicator value for all the areas considered, and if properly defined, refer to the degree of fulfillment for the ecosystem characteristic.
+Sometimes changing the indicator name, or redefining the indicator, goes a long way towards easing the interpretation and avoiding miss-use (e.g. *mean reindeer densities*).
+
+Although the variable and the indicator may be reference by the same common name, and may be presented side-by-side, it is not so that they can replace each other or act as alternative evidence.
+Presenting the same metric both in its indicator form and its variable form opens up for alternative interpretations.
+While this may initially seem like a democratic and transparent approach to reporting, this practice actually places a significant burden on both the end user and the indicator developer.
+The developer must clarify to the user how the two metrics have diverged significantly during spatial aggregation and normalisation, complicating the comparison of their respective information.
+This can lead to miss-use, for example by taking the variable value and assigning it the same interpretation as the indicator value.
+
+Generally, because of the normalisation process, indicators are more aligned with the purpose of ECA with describing ecosystem condition in normative terms.
+Although variable accounts are allowed under the SEEA EA framework, they may be seen as a type of interim account for metrics that cannot or have not yet been normalised.
+We would therefore argue that when an indicator has been developed, the original variable is not presented.
+And when only variable estimates are given, extra care is give to describing the spatial aggregation pathway, and how this will have evened out any spatial variation that is relevant for interpreting the variable.
+
+## Aggregation bias in indicator development
+
+As we have seen, aggregation bias causes a divergence in the type of information made available through the spatial aggregation of variables versus indicators.
+In addition, unavoidable aggregation bias (for good and bad) will also affect the indicator values and the inferences that can be drawn from these.
+If not careful, aggregation bias may also cause errors to sneak into the indicator logic.
 A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition indicators (@fig-agg1; @fig-agg2), is the choice about the order of the different steps in the normalisation and aggregation process, i.e.
 the aggregation pathway (@fig-pathways).
+If normalisation occurs early in the aggregation process, the amount of compensation is limited.
+However, sometimes complete or partial compensation is justified, as in the example in @sec-pti, or as we saw in @sec-match where indicator development involved an initial step of spatially aggregating variables to align with the spatial configuration of reference levels (pathways 2 and 3; @fig-pathways).
+Failure to harmonise the spatial scale between variables and reference levels before normalising the variable can lead to aggregation bias, as described in @sec-bill.
+
+Another cause of potential errors or confusion arising during the spatial aggregation of indicators, that may not relate directly to the aggregation bias problem, has to do with deciding on the appropriate area weights to use.
+The SEEA EA uses a principle of commensurability, meaning that the order of certain operations should not affect the resulting estimate or value.
+Spesifically, it should not matter whether you aggregated vertically first (to create an index) and then horizontally (i.e. spatial), or the other way around.
+Note that this prisiple may not always hold for other cases, such as aggregating across nested areas, e.g. from ecosystem assets to EAAs, although this is commonly agreed upon as desirable.
+This has to do with the for appropriate data to inform the area weighting in the spatial aggregation.
+Consider an indicator originating from point-based field surveys.
+The sampling location are selected using a stratified protocol that ensures a representative sample of similar size (same *n*) is made for each ecosystem asset.
+If the variable is normalised early in the aggregation pathway to create indicator values for each locality, then the mean of indicator values for each region will be representative for those regions, and no area weighting is required.
+For the next spatial level, when information from all ecosystem assets should be used to create an estimate for the EAA, ecosystem assets with large extents should weight more in the calculations (following the SEEA EA prinsiples of area weighting).
+However, if in a later ECA, ecosystem assets are defined as subsets of the initial assets (a finer scale account), then the same indicator, when aggregated to the EAA level, will have a different value.
+Cases when two accounts or assessments use what is considered the same indicator but find different values for the same areas, is cause for both confusion, and potentially mistrust.
+(something here about indicator versions and good documentation)
+
+## Aggregating to account for spatial autocorrelation
 

### 2025-01-22 manuscript/manuscript.qmd
more writing

@@
   - aggregation bias
+  - upscaling
 date: last-modified
@@
 For ecosystem condition accounting (ECA) and assessments, this often requires that **metrics**[^1] used to describe the condition of ecosystems are aggregated spatially to produce a single value for a larger area.
-This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something actionable.
+This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something actionable at a political level.
 However, spatial aggregation can and will often change the interpretation of a metric and the descriptive properties and normative claims one can make based on it.
 Without a good enough understand of the pitfalls in spatial aggregation of information, is can also lead to unintended and undetected systematic bias, as well as downright logical errors.
-Although much attention has been devoted to the vertical aggregation of environmental indicators into indices (ref), as well as to the spatial aggregation of relatively unmodified or raw variables, there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting.
+Although much attention has been devoted to the vertical aggregation of environmental indicators into indices (e.g.
+@langhans2014), as well as to the spatial aggregation of relatively unmodified or raw variables (e.g @allain2018) , there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting.
 This paper therefore focus on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) - the most used ecosystem accounting (natural resource accounting) framework in the world (ref).
-We pay specific attention to the phenomenon of aggregation bias introduce by @rastetter1992.
+We pay specific attention to the phenomenon of aggregation bias [@rastetter1992] and on the different ways to interpret the information in ecosystem condition variables versus indicators.
 
 [^1]: Words in bold are explained in Table 1
+
+@stein2001
+
+@finke2002
+
+<https://www.sciencedirect.com/science/article/pii/S0016706198000779?casa_token=UTwC6C3DthYAAAAA:ZZ6eao2vlMClkMtTrdA043xgkyjfjwjpLsitDzfohdqiB3GvVPyQoDFzY1jgu4o176jpDjbWCZE>
+
+read bib in knotters...
+
+### Other types of aggregation not considered in this paper
+
+Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (i.e.
+thematic or vertical aggregation; @langhans2014), but in this paper we are discussing spatial (i.e.
+horizontal) aggregation only.
+The term spatial aggregation is also sometimes used to refer to **upscaling** in the sense of taking local information and making it applicable of translate it to larger, e.g.
+landscape or regional, scales [@knotters2010].
+This process often involves predicting values across larger areas or surfaces, generally through models, including krieging-based models and distribution models.
+When the data (e.g. indicator values) are made wall-to-wall, one can then more easily get point estimates for larger areas via simple aggregation functions such as summation or by taking the mean.
+Using models to predict variable or indicator values is allowed under the SEEA EA.
+However, in this paper we are addressing the second part of this process only: of simplifying the spatial variation, from many to few data points.
 
@@
 The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.
+
+In ECAs, data are often not wall-to-wall, and when upscaling is not done, as defined above, the spatial aggregation needs to account for the spatial structure of the data in a different way, such as design-based approached (relying on random or balanced samples across the entire EEA), or a step-wise aggregation via ecosystem assets, homogenous ecosystem areas [@vallecillo2022] or homogenous impact areas [@kolstad_making_2024].
+Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means.
+When using means, area weighting is required when following SEEA EA.
 
@@
 | Term                | Definition |
-|------------------|------------|
+|---------------------|------------|
 | Metric              |            |
@@
 | Compensation        |            |
-|                  |            |
+| Aggregation pathway |            |
 
@@
 This smoothing out makes the data and its inherent information less suitable for some uses, such as depicting spatial variation that can be used to support local land use planning, or for aiding a debate about spatial conflicts.
+
+### Accounting for spatial autocorrelation {#-sec-auto}
+
+One reason why we might want to spatially aggregate ecosystem condition metrics is to fix issues with spatial auto-correlation, that is to remove problematic spatial structure in the data.
+For example if data is collected opportunistically with sampling size varying between areas, then performing partial spatial aggregation to some intermediate size areaa will give equal weight or importance to to all areas regardless of how many data points are included in each.
 
@@
 
-# Why do we normalise? {#sec-norm}
+# Why do we normalise variables? {#sec-norm}
 
@@
 This displacement is refereed to as aggregation error, and its effect become more pronounced when the non-linearity is strong between variables and indicators.
+In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2021; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
+We now describe two ways that aggregation bias affects ECAs: by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators, and secondly, how it can cause confusion about the effects of chosing different **aggregation pathways** during indicator development.
 
-## Aggregation error and discrepancies between variables and indicators
+## Discrepancies between variables and indicators
 
@@
 
-Another cause of potential errors or confusion arising during the spatial aggregation of indicators, that may not relate directly to the aggregation bias problem, has to do with deciding on the appropriate area weights to use.
+# Compliance across nested assessments
+
+Another cause of potential errors or confusion arising during the spatial aggregation of indicators has to do with the number of aggregation levels.
 The SEEA EA uses a principle of commensurability, meaning that the order of certain operations should not affect the resulting estimate or value.
 Spesifically, it should not matter whether you aggregated vertically first (to create an index) and then horizontally (i.e. spatial), or the other way around.
-Note that this prisiple may not always hold for other cases, such as aggregating across nested areas, e.g. from ecosystem assets to EAAs, although this is commonly agreed upon as desirable.
-This has to do with the for appropriate data to inform the area weighting in the spatial aggregation.
+Note that this prisiple may not always hold for other cases, such as aggregating across nested areas, e.g. from ecosystem assets to EAAs, although this is perhaps considered desirable.
+This has to do with the use of area weighting in the spatial aggregation.
 Consider an indicator originating from point-based field surveys.
@@
 For the next spatial level, when information from all ecosystem assets should be used to create an estimate for the EAA, ecosystem assets with large extents should weight more in the calculations (following the SEEA EA prinsiples of area weighting).
+Note that one cannot skip this first level of spatial aggregation via ecosystem assets, since this is an integral part of the sampling design, leading to spatial autocorrelation (se @sec-auto).
 However, if in a later ECA, ecosystem assets are defined as subsets of the initial assets (a finer scale account), then the same indicator, when aggregated to the EAA level, will have a different value.
+A similar case would be is sampling was random and dispersed across the EAA, lets say a country, and indicator values were first calculated as some average across all sampling points, and then in a later account or assessment, the same indicator was estimated for separate regions.
+When aggregating the indicator values from regions to the whole country, using area weighting, the national estimate will differ, albeit slightly, from the initial national estimate that did not include this step-wise aggregation.
 Cases when two accounts or assessments use what is considered the same indicator but find different values for the same areas, is cause for both confusion, and potentially mistrust.
-(something here about indicator versions and good documentation)
-
-## Aggregating to account for spatial autocorrelation
+In our experience this can also lead to time consuming detective work, trying to find the causes for such discrepancies.
+We suggest two immediate solution to this issue.
+First, to find better ways to document the aggregation pathways that are being used for each indicator, as part of the indicator metadata.
+Second, to find ways to distinguish between apparently identical indicators, for example using different version numbers or IDs.
+We provide practical solution to both these suggestions in @sec-rec.
 
@@
 
-Indicator developer subjectivity is a big factor as the process of designing indicators and estimating indicator values often left entirely to the them, with little real involvement from indicator users.
+Indicator developer subjectivity affects ECAs, in part because the process of designing indicators and estimating indicator values often left entirely to the them, with little real involvement from indicator users.
 It is therefore important to be aware of the role that indicator developers have in shaping the type of information, and the narratives, that is presented to decision makers.
@@
 
-Something about sometimes spatial aggregation is an essential part of the indicator design ---
-
-In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2021; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
-Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (i.e.
-thematic or vertical aggregation; @langhans2014), but in this paper we are discussing spatial (i.e.
-horizontal) aggregation only.
-
 # Aggregation pathways
@@
 
-Pathway 1 involves early normalisation suing the perhaps most common order of the three steps scaling, truncating and transforming.
+Pathway 1 involves early normalisation using the perhaps most common order of the three steps scaling, truncating and transforming.
 The three steps may be done simultaneously as well.
@@
 
+# Examples
+
 ## Wolves example {#sec-wolves}
@@
 Therefore, most carnivore management areas will get an indicator value of 0.
-Wolves exists i designated wolf zones in south-east Norway, although in small numbers.
+Wolves exist in designated wolf zones in south-east Norway, although in small numbers.
 In the Norwegian Nature Index [@jakobsson2020], the variable follows pathway 1 and is normalised at the CMA level.
@@
 However, there are sometimes good reasons for making things complicated.
-In this example, The Norwegian Nature Index uses the same pathway for all indicators, so this common misunderstandings about what indicator values represents can be explain just one time, and not for each indicator specifically.
+In this example, The Norwegian Nature Index uses the same pathway for all indicators, so this common misunderstandings about what indicator values represents can be explaind just one time, and not for each indicator specifically.
 It can also be a good idea to sometime go back to the variable, and see if it can be defined differently to avoid these kinds of confusions.
@@
 ## Billberry example {#sec-bill}
-
-This example illustrates a variable normalised at plot level or aggregated to regions first.
-At what scale is the ref value intended?
 
@@
 
-The special thing about the billberry indicator is the different spatial resolution for the variable and for the reference level.
+The billberry indicator exemplifies different spatial resolutions for the variable and for the reference levels.
 The variable is recorded at the scale of vegetation plots.
@@
 Due to non-negligeable truncation of the variable during the normalisation, the two indicators were subject to different levels of aggregation bias, and consequently the spatially aggregated indicator values for the two assessments became different.
-This cause a general confusion about how how the same variable (and the same date) can produce different indicator values.
+This caused a general confusion about how how the same variable (and the same data) can produce different indicator values.
 
-*Conclusion*: Given that the reference value is the way it is, this indicator should have followed pathway 3, and spatially aggregated the variable, from plot scale to regional scale, before normalising the variable and potentially aggregating it further.
+*Conclusion*: Given that the reference value is the way it is, this indicator should have followed pathway 3, and spatially aggregated the variable from plot scale to regional scale before normalising the variable and potentially aggregating it further.
 It would also be possible to use pathway 6.
@@
 The wolves example also highlight the importance of making clear the differences in the interpretation of variables and indicators.
-Often indicators and variables are both included side-by-side in ecosystem condition assessments/accounts.
-This is also recommended practice following the SEEA EA.
-However, if the indicators that are presented are normalised before any spatial aggregation, then the interpretation of the indicators is not simply as a normalised version of the variable.
+Indicators and variables are sometimes presented side-by-side in ecosystem condition assessments or accounts.
+This is also recommended <!--# is this true?
+--> practice following the SEEA EA.
+However, due to aggregation bias, spatially aggregated indicators should not be interpreted as normalised versions of aggregated variables.
 Indicators typically now reflect the average ecosystem condition, where areas in very good or very poor condition have limited ability to compensate for opposite extreme values elsewhere.
-The spatially aggregated variable on the other hand, may reflect something like a sum of individuals, or some other aspect where values above and below the average value are able to compensate for each other.
+The spatially aggregated variable on the other hand, may reflect something like a sum of individuals, an average of a population, or some other aspect where values above and below the reference levels are able to compensate for each other.
 
@@
 In this example, the reference levels are uniform throughout the EAA, and the normalisation includes a transformation in the form of a piecewise-linear transformation by anchoring X~60~ to 0.6 on the indicator scale.
-When performing an ECA, one has the option to either aggregate variable value for each region (each ecosystem asset), and then normalise that value (pathway 2 and 3), or to normalise the variable at the scale of the original nature type polygons, and then aggregate the indicator values to the regional level (pathway 1).
+When performing an ECA, one has the option to either aggregate variable values for each region (each ecosystem asset), and then normalise that value (pathway 2 and 3), or to normalise the variable at the scale of the original nature type polygons, and then aggregate the indicator values to the regional level (pathway 1).
 
@@
 
-Separate assessments should give identical indicator values for the same areas, given the same underlying data.
+<!--# Separate assessments should give identical indicator values for the same areas, given the same underlying data.
 When this is not true it raises concerns about reproducibility and sacrifices the credibility of both assessments.
 This was also the case in the billberry example, but here we used another example to illustrate the point further.
+-->
 
@@
 
-# Recommendation
+# Recommendation {#sec-rec}
 

### 2025-01-28 manuscript/manuscript.qmd
Note 1

@@
 For ecosystem condition accounting (ECA) and assessments, this often requires that **metrics**[^1] used to describe the condition of ecosystems are aggregated spatially to produce a single value for a larger area.
-This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something actionable at a political level.
+This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something that is relevant to decision makers and perhaps actionable at a political level.
 However, spatial aggregation can and will often change the interpretation of a metric and the descriptive properties and normative claims one can make based on it.
@@
 Although much attention has been devoted to the vertical aggregation of environmental indicators into indices (e.g.
-@langhans2014), as well as to the spatial aggregation of relatively unmodified or raw variables (e.g @allain2018) , there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting.
-This paper therefore focus on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) - the most used ecosystem accounting (natural resource accounting) framework in the world (ref).
+@langhans2014; @maes2020), as well as to the spatial aggregation of relatively unmodified or raw variables (e.g @allain2018) , there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting (but see @vanbeurden1999).
+This paper therefore focus on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) [@unitednations2021] - the most used ecosystem accounting (natural resource accounting) framework in the world (ref).
 We pay specific attention to the phenomenon of aggregation bias [@rastetter1992] and on the different ways to interpret the information in ecosystem condition variables versus indicators.
 
-[^1]: Words in bold are explained in Table 1
+[^1]: Words in bold are explained in @tbl-def
 
-@stein2001
+This paper has a rather narrow and defined focus.
+Related concepts are briefly mentioned in @nte-rel.
 
-@finke2002
+::: {#nte-rel .callout-note appearance="minimal" icon="false"}
+### Related concepts
 
-<https://www.sciencedirect.com/science/article/pii/S0016706198000779?casa_token=UTwC6C3DthYAAAAA:ZZ6eao2vlMClkMtTrdA043xgkyjfjwjpLsitDzfohdqiB3GvVPyQoDFzY1jgu4o176jpDjbWCZE>
+##### Vertical aggregation {.unnumbered}
 
-read bib in knotters...
-
-### Other types of aggregation not considered in this paper
-
-Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (i.e.
+It is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (i.e.
 thematic or vertical aggregation; @langhans2014), but in this paper we are discussing spatial (i.e.
 horizontal) aggregation only.
-The term spatial aggregation is also sometimes used to refer to **upscaling** in the sense of taking local information and making it applicable of translate it to larger, e.g.
-landscape or regional, scales [@knotters2010].
+
+##### Upscaling {.unnumbered}
+
+The term spatial aggregation is also sometimes used to refer to **upscaling** in the sense of taking local information and making it applicable or translate it to larger, e.g.
+landscape or regional, scales [@knotters2010; @finke2002].
 This process often involves predicting values across larger areas or surfaces, generally through models, including krieging-based models and distribution models.
@@
 Using models to predict variable or indicator values is allowed under the SEEA EA.
-However, in this paper we are addressing the second part of this process only: of simplifying the spatial variation, from many to few data points.
+However, in this paper we are addressing the second part of this process only, of simplifying the spatial variation, from many to few data points.
+
+##### Modifiable area unit problem {.unnumbered}
+
+The modifiable area unit problem (MAUP; [@openshaw1984, @wong2009]) describes the sensitivity that spatially aggregated information has on the choice about the spatial units used for either collecting the data or for aggregating the data to.
+These units can be arbitrary (*modifiable*), yet different sizes and configuration of units can have large effects on the outcome of aggregation.
+Administrative units are often used as the ecosystem assets in SEEA EAs, yet they are in an ecological sense almost always arbitrary, and can also be subject to change over time.
+The original spatial units that data are collected at may be a pragmatic choice in much the same way.
+Further still, data may not be available at the original spatial scale, but may only be downloadable in preaggregated form [@vanbeurden1999].
+All these cases lead to some form of aggregation bias [@rastetter1992] where the aggregated information has a non-linear relationship with the true variable value.
+However, the issue of MAUP in SEEA EA deserves an in-depth analyses that we cannot provide here.
+
+##### Ecological fallacy {.unnumbered}
+
+The ecological fallacy [@robinson2009] explains the common misinterpretation of correlations between group (*i.e.* aggregated information) and some phenomena.
+Simply said, the mean of a group, or an area, cannot accurately describe the qualities of individuals in that group, or pin-points inside that area, yet it is common for people to interpret it that way.
+:::
 
@@
 
-SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services.
+SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2021 @unitednations2022].
 We will focus on the SEEA EA condition accounts in this paper.
@@
 |---------------------|------------|
-| Metric              |            |
+| Metric              | blablabla  |
 | Variable            |            |
@@
 
-: Definition of terms
+: Definition of terms {#tbl-def}
 
@@
 
-### Accounting for spatial autocorrelation {#-sec-auto}
+### Accounting for spatial autocorrelation {#sec-auto}
 

### 2025-01-29 manuscript/manuscript.qmd
sea ch

@@
 Using models to predict variable or indicator values is allowed under the SEEA EA.
-However, in this paper we are addressing the second part of this process only, of simplifying the spatial variation, from many to few data points.
+However, in this paper we are addressing the second part of this process only, of simplifying the spatial variation, from many to few data points (sometimes also included in the terms upscaling).
 
@@
 SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2021 @unitednations2022].
-We will focus on the SEEA EA condition accounts in this paper.
-In these accounts, information about ecosystem condition is made spatially explicit by assigning **variable** or **indicator** values to individual ecosystem assets.
-Unless the underlying data are wall-to-wall maps, for example from satellite images, then this process of assigning values to ecosystem assets already here requires some spatial manipulations in the form of aggregation, dissaggregation, extrapolation or interpolation (ref Ewert 2011?), of which averaging or summing are the most common methods.
-Subsequently, information from ecosystem assets may be spatially aggregated to larger areas, such as the entire ecosystem accounting area.
-When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels.
+We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is a relevant topic also for the other parts of the account.
+
+In ECAs, information about ecosystem condition is made spatially explicit by assigning **variable** or **indicator** values to individual ecosystem assets.
+The process of assigning values to ecosystem assets usually requires some spatial manipulations of data, or scale changes [@ewert2011].
+The type of manipulation depends on the spatial resolution of the underlying data, as well as the size and configuration of the ecosystem assets.
+To main types of data can be distinguished: sparse data and complete data.
+
+Complete, or wall-to-wall data, typically include remote sensing products.
+It can also include other types of pre-aggregated data, where sparse data, such as field survey summaries, are assigned to areal units.
+This is very common.
+A lot of demographic information for example is made available at the scale of administrative units, rather than at the individual level (which would be very ineffective and problematic also in terms of protecting sensitive information).
+Similar pre-aggregation is, however, ambiguous and true for most complete data.
+Consider as an example data from LiDAR (Light Detection And Ranging).
+The raw data is a point cloud, yet the most commonly available datasets are surface or canopy models; wall-to-wall generalisations of the point cloud.
+Also data such as rasters derived from satellite images, such as NDVI (extensively used in ECAs), have some inherent pre-aggregation where values are averaged for each raster cell.
+Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (@nte-rel).
+This issue will affect our ability to detect true maximum or minimum values (as used for *worst case* aggregation, for example in the WFD) for the phenomena our variable is describing, since these extremes are already smoothed out in the pre-aggregation step.
+This will clearly affect the suitability of such data in looking for areas of high risk (very degraded areas, for example).
+The problem is the same when we deliberately aggregate data later in the indicator development process, and we always need to ensure that the spatial aggregation function we use does not have any unintentional effect on the type of inference that we can draw from the metric.
+
+For sparse data we are considering type point data, with variable values assigned to singular points or small areas.
+In these cases, there will be a lot of area inside each ecosystem asset for which we have no direct information.
+To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some generalisations.
+One can assign values to areas via an aggregation functions used on data values that fall within a given ecosystem asset.
+Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means.
+When using means, area weighting is required when following SEEA EA. This is what we refer to as deliberate spatial aggregation, and what is the main focus of this paper.
+When doing this, we need to account for the spatial structure (spatial correlations) of the data (@sec-auto).
+One way is to rely on design-based approaches (relying on random or balanced samples across the entire EEA), or using a step-wise aggregation via ecosystem assets.
+In our experience, the choice of which areal units to use when doing this step-wise aggregating is often not reflected much upon.
+It may be a limitation with the data related to the type of pre-aggregation performed (see above).
+In some cases the areal units are decided politically, wit little scientific reasoning behind it.
+However, the arbitrariness of this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; @nte-rel).
+It arguable important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods (REF).
+The MAUP can, however, lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
+One option to reduce this discrepancy and arbitrariness is to use ecologically motivated areal units in the step-wise aggregation such as homogeneous ecosystem areas [@vallecillo2022] or homogeneous impact areas [@kolstad_making_2024].
+
+Interpolation or other modeling can also be used to impute or predict data values between the points where we have information.
+However, this is not the focus of this paper, and in our experience this has not been common practice in SEEA EA accounts produced to date.
+
+When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels, thus invoking the issue of the ecological fallacy, as well as many more concrete pitfalls in spatial aggregation of information.
 The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.
-
-In ECAs, data are often not wall-to-wall, and when upscaling is not done, as defined above, the spatial aggregation needs to account for the spatial structure of the data in a different way, such as design-based approached (relying on random or balanced samples across the entire EEA), or a step-wise aggregation via ecosystem assets, homogenous ecosystem areas [@vallecillo2022] or homogenous impact areas [@kolstad_making_2024].
-Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means.
-When using means, area weighting is required when following SEEA EA.
 
@@
 
-# Compliance across nested assessments
+# Compliance across nested assessments {#sec-comp}
 

### 2025-01-29 manuscript/manuscript.qmd
update un24

@@
 Ecosystem accounting (EA), and other types of environmental assessments, are generally aimed at communicating major changes and trends in the extent and condition of ecosystems at a scale that is relevant for decision makers.
-For ecosystem condition accounting (ECA) and assessments, this often requires that **metrics**[^1] used to describe the condition of ecosystems are aggregated spatially to produce a single value for a larger area.
+For ecosystem condition accounting (ECA) and assessments, this often requires that **metrics**[^1] used to describe the condition of ecosystems are aggregated spatially, sometimes to produce a single value for a larger area.
 This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something that is relevant to decision makers and perhaps actionable at a political level.
-However, spatial aggregation can and will often change the interpretation of a metric and the descriptive properties and normative claims one can make based on it.
-Without a good enough understand of the pitfalls in spatial aggregation of information, is can also lead to unintended and undetected systematic bias, as well as downright logical errors.
+However, spatial aggregation affect how a metric should be interpreted, and the descriptive properties and normative claims one can make based on it.
+Without an understanding of the pitfalls during spatial aggregation of information, one risks introducing unintended and undetected systematic bias, as well as downright logical errors.
 Although much attention has been devoted to the vertical aggregation of environmental indicators into indices (e.g.
-@langhans2014; @maes2020), as well as to the spatial aggregation of relatively unmodified or raw variables (e.g @allain2018) , there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting (but see @vanbeurden1999).
-This paper therefore focus on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) [@unitednations2021] - the most used ecosystem accounting (natural resource accounting) framework in the world (ref).
-We pay specific attention to the phenomenon of aggregation bias [@rastetter1992] and on the different ways to interpret the information in ecosystem condition variables versus indicators.
+@langhans2014; @maes2020; @oecd_2008\], as well as to the spatial aggregation of relatively unmodified or raw variables (e.g @allain2018) , there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting (but see @vanbeurden1999).
+This paper therefore focus on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) [@unitednations2024] - the most used ecosystem accounting (natural resource accounting) framework in the world [@comte2022].
+SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2024 @unitednations2022].
+We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is a relevant topic also for the other parts of the account.
 
@@
 
+We pay specific attention to the phenomenon of aggregation bias [@rastetter1992] and on the different ways to interpret the information in ecosystem condition variables versus indicators.
+
 This paper has a rather narrow and defined focus.
 Related concepts are briefly mentioned in @nte-rel.
+
+## Indicators vs. variables
+
+SEEA EA distinguishes between indicators and variables, and to be able to describe how these are affected by spatial aggregation, we need to spend some time to explaining the difference between them.
+Variables are metrics describing ecosystem characteristics [@unitednations2024 §5.41].
+They are typically either in percentages or fractions, or as raw biophysical units, such as biomass, density, or cover.
+Indicators on the other hand, are always on a unitless scale between 0 and 1, where 0 represent the worst condition, and 1 represents the best condition.
+Sometimes indicators are calculated directly, but usually indicators are made by **normalising** variables based on two or more **reference levels** [@unitednations2024 §5.60].
+Although also variables may have normative properties (due to choices about how they are chosen, defined, aggregated or presented [@allain2018]) , the normalisation of variables into indicators makes indicators explicitly normative by introducing a clear directional interpretation, where increasing values always refer to an increase in condition.
+The information held by the indicator is also different from that of the variable since additional ecological knowledge can be introduced via the reference levels.
+The mathematical part of this normalisation contains two, sometimes three, steps [@fig-term].
+These steps can be performed simultaneously, but for clarity we discuss them as separate steps.
+The order of these steps can also change.
+
+Variables can be **scaled** using a minimum of two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
+Variables can be **truncated** to produce a bound indicator scale between 0 and 1.
+Sometimes variables are **transformed** to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
+This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz; @jakobsson_setting_2020], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
+
+## Spatial aggregation
+
+In ECAs, information about ecosystem condition is made spatially explicit by assigning variable or indicator values to individual ecosystem assets.
+The process of assigning values to ecosystem assets requires some spatial manipulations of data, or scale changes [@ewert2011].
+The type of manipulation depends on the spatial resolution of the underlying data, as well as the size and configuration of the ecosystem assets.
+Two main types of data can be distinguished: sparse (i.e. spatially incomplete) data and complete data.
+
+<!--# Subsequently, spatial aggregation is required when going from ecosystem assets to large spaatial scale, ultimately the scale of the entire ecosystem accounting area (EAA).  -->
+
+Complete, or wall-to-wall data, include remote sensing products, as well as other types of pre-aggregated data, where sparse data are assigned to areal units.
+This last type of data is very common.
+A lot of demographic information for example is made available at the scale of administrative units, rather than at the individual level (which would be very ineffective and problematic also in terms of protecting sensitive information).
+Similar pre-aggregation is, however, ambiguous and true for most complete datasets.
+Consider as an example data from LiDAR (Light Detection And Ranging).
+The raw data is a point cloud, yet the most commonly available datasets are surface or canopy models; wall-to-wall generalisations of the point cloud.
+Also data such as rasters from satellite images ,or maps derived from such, have some inherent pre-aggregation where values are averaged for each raster cell.
+Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (@nte-rel).
+This issue will for example affect our ability to detect true maximum or minimum values (used in *worst case* aggregation, for example in the WFD) for the phenomena our variable is describing, since these extremes are already smoothed out in the pre-aggregation step.
+This will clearly affect the suitability of such data for identifying areas of high risk (very degraded areas, for example).
+The problem is the same when we deliberately aggregate data later in the indicator development process, and we always need to ensure that the spatial aggregation function we use does not have any unintentional effect on the type of inference that we can draw from the metric.
+
+For sparse data we are considering data where variable values are assigned to singular points or small areas.
+In these cases, there will be a lot of area inside each ecosystem asset for which we have no direct information.
+To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some generalisations.
+One can assign values to areas via an aggregation functions used on data values that fall within a given ecosystem asset.
+Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54] .
+This deliberate spatial aggregation is the main focus of this paper.
+When doing this aggregation, we need to account for the spatial structure (spatial correlations) of the data (@sec-auto).
+One way is to rely on design-based approaches (relying on random or balanced samples across the entire EEA), or using a step-wise aggregation via ecosystem assets.
+In our experience, the choice of which area units to use when doing this step-wise aggregation is often not reflected much upon.
+There may be limitations with the data, related to the type of pre-aggregation performed (see above).
+In other cases the areal units are decided politically, with little scientific reasoning behind it.
+However, the arbitrariness of this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; @nte-rel).
+It is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
+The MAUP can, however, lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
+One option to reduce this discrepancy and arbitrariness is to use ecologically motivated areal units in the step-wise aggregation such as homogeneous ecosystem areas [@vallecillo2022] or homogeneous impact areas [@kolstad_making_2024].
+
+[^2]: When using means across areas, area weighting is required when following SEEA EA.
+
+Interpolation or other modeling can also be used to impute or predict data values between the points where we have information [@finke2002; @stein2001].
+However, this is not the focus of this paper, and in our experience this has not been common practice in SEEA EA accounts produced to date.
+
+When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels, thus invoking the issue of the ecological fallacy, and many other potential pitfalls in the spatial aggregation of information.
+The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.
 
@@
 
-### SEEA EA
-
-SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2021 @unitednations2022].
-We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is a relevant topic also for the other parts of the account.
-
-In ECAs, information about ecosystem condition is made spatially explicit by assigning **variable** or **indicator** values to individual ecosystem assets.
-The process of assigning values to ecosystem assets usually requires some spatial manipulations of data, or scale changes [@ewert2011].
-The type of manipulation depends on the spatial resolution of the underlying data, as well as the size and configuration of the ecosystem assets.
-To main types of data can be distinguished: sparse data and complete data.
-
-Complete, or wall-to-wall data, typically include remote sensing products.
-It can also include other types of pre-aggregated data, where sparse data, such as field survey summaries, are assigned to areal units.
-This is very common.
-A lot of demographic information for example is made available at the scale of administrative units, rather than at the individual level (which would be very ineffective and problematic also in terms of protecting sensitive information).
-Similar pre-aggregation is, however, ambiguous and true for most complete data.
-Consider as an example data from LiDAR (Light Detection And Ranging).
-The raw data is a point cloud, yet the most commonly available datasets are surface or canopy models; wall-to-wall generalisations of the point cloud.
-Also data such as rasters derived from satellite images, such as NDVI (extensively used in ECAs), have some inherent pre-aggregation where values are averaged for each raster cell.
-Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (@nte-rel).
-This issue will affect our ability to detect true maximum or minimum values (as used for *worst case* aggregation, for example in the WFD) for the phenomena our variable is describing, since these extremes are already smoothed out in the pre-aggregation step.
-This will clearly affect the suitability of such data in looking for areas of high risk (very degraded areas, for example).
-The problem is the same when we deliberately aggregate data later in the indicator development process, and we always need to ensure that the spatial aggregation function we use does not have any unintentional effect on the type of inference that we can draw from the metric.
-
-For sparse data we are considering type point data, with variable values assigned to singular points or small areas.
-In these cases, there will be a lot of area inside each ecosystem asset for which we have no direct information.
-To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some generalisations.
-One can assign values to areas via an aggregation functions used on data values that fall within a given ecosystem asset.
-Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means.
-When using means, area weighting is required when following SEEA EA. This is what we refer to as deliberate spatial aggregation, and what is the main focus of this paper.
-When doing this, we need to account for the spatial structure (spatial correlations) of the data (@sec-auto).
-One way is to rely on design-based approaches (relying on random or balanced samples across the entire EEA), or using a step-wise aggregation via ecosystem assets.
-In our experience, the choice of which areal units to use when doing this step-wise aggregating is often not reflected much upon.
-It may be a limitation with the data related to the type of pre-aggregation performed (see above).
-In some cases the areal units are decided politically, wit little scientific reasoning behind it.
-However, the arbitrariness of this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; @nte-rel).
-It arguable important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods (REF).
-The MAUP can, however, lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
-One option to reduce this discrepancy and arbitrariness is to use ecologically motivated areal units in the step-wise aggregation such as homogeneous ecosystem areas [@vallecillo2022] or homogeneous impact areas [@kolstad_making_2024].
-
-Interpolation or other modeling can also be used to impute or predict data values between the points where we have information.
-However, this is not the focus of this paper, and in our experience this has not been common practice in SEEA EA accounts produced to date.
-
-When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels, thus invoking the issue of the ecological fallacy, as well as many more concrete pitfalls in spatial aggregation of information.
-The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.
-
-### Indicators vs. variables
-
-SEEA EA distinguishes between indicators and variables.
-Variables in ecosystem condition terminology are metrics describing ecosystem characteristics [@unitednations2021].
-They are typically either in units or percentages or fractions, or as raw biophysical units, such as biomass, density, or cover.
-Indicators on the other hand, are always on a scale between 0 and 1, where 0 represent the worst condition, and 1 represents the best condition.
-Sometimes indicators are calculated directly, but usually indicators are made by **normalising** variables based on two or more **reference levels**.
-Although also variables may have normative properties (due to choices about how they are chosen (ref), defined, aggregated or presented) (Alain et al), the normalisation of variables into indicators makes indicators explicitly normative by introducing a clear directional interpretation, where increasing values always refer to an increase in condition.
-The information held by the indicator is also different from that of the variable since additional ecological knowledge can be introduced via the reference levels.
-The mathematical part of this normalisation contains two, sometimes three, steps [@fig-term].
-These steps can be performed simultaneously, but for clarity we discuss them as separate steps.
-The order of these steps can also change.
-
-Variables can be **scaled** using a minimum of two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
-Variables can be **truncated** to produce a bound indicator scale between 0 and 1.
-Sometimes variables are **transformed** to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
-This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
-
 | Term | Definition |
-|---------------------|------------|
-| Metric              | blablabla  |
-| Variable            |            |
-| Indicator           |            |
-| Reference levels    |            |
-| Scaling             |            |
+|----|----|
+| Metric | A collective term for both variables and indicators |
+| Reference levels | Variable values corresponding and anchoring the variable to specific indicator values.
+A special terminology is used, based on @czúcz, which differs somewhat from that in @unitednations2024.
+See @fig-term.
+|
+| Scaling | A linear shrinking or enlarging the variable data range by anchoring X~0~ and X~100~ to the values 0 and 1, respectively. See @fig-term. |
 | Transformation |  |
 | Truncation |  |
-| Normalisation       |            |
+| Normalisation | Referred to as rescaling in @unitednations2024. This is the process of converting a variable to an indicator. |
 | Compensation |  |

### 2025-01-30 manuscript/manuscript.qmd
remove table with definitions

@@
 Ecosystem accounting (EA), and other types of environmental assessments, are generally aimed at communicating major changes and trends in the extent and condition of ecosystems at a scale that is relevant for decision makers.
-For ecosystem condition accounting (ECA) and assessments, this often requires that **metrics**[^1] used to describe the condition of ecosystems are aggregated spatially, sometimes to produce a single value for a larger area.
+For ecosystem condition accounting (ECA) and assessments, this often requires that metrics[^1] used to describe the condition of ecosystems are aggregated spatially, sometimes to produce a single value for a larger area.
 This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something that is relevant to decision makers and perhaps actionable at a political level.
@@
 
-[^1]: Words in bold are explained in @tbl-def
+[^1]: We use the SEEA EA definitions of the terms variables, indicators (normalised variables) and metrics (refering to either a variable or a metric) [@unitednations2024]
 
-We pay specific attention to the phenomenon of aggregation bias [@rastetter1992] and on the different ways to interpret the information in ecosystem condition variables versus indicators.
+<!--# We pay specific attention to the phenomenon of aggregation bias [@rastetter1992] and on the different ways to interpret the information in ecosystem condition variables versus indicators. -->
 
@@
 Indicators on the other hand, are always on a unitless scale between 0 and 1, where 0 represent the worst condition, and 1 represents the best condition.
-Sometimes indicators are calculated directly, but usually indicators are made by **normalising** variables based on two or more **reference levels** [@unitednations2024 §5.60].
+Sometimes indicators are calculated directly, but usually indicators are made by normalising variables based on two or more reference levels [@unitednations2024 §5.60].
 Although also variables may have normative properties (due to choices about how they are chosen, defined, aggregated or presented [@allain2018]) , the normalisation of variables into indicators makes indicators explicitly normative by introducing a clear directional interpretation, where increasing values always refer to an increase in condition.
@@
 
-Variables can be **scaled** using a minimum of two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
-Variables can be **truncated** to produce a bound indicator scale between 0 and 1.
-Sometimes variables are **transformed** to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
+Variables can be scaled using a minimum of two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
+Variables can be truncated to produce a bound indicator scale between 0 and 1.
+Sometimes variables are transformed to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
 This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz; @jakobsson_setting_2020], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
 
-## Spatial aggregation
+# Why do we spatially aggregate data in ecosystem condition accounting
 
-In ECAs, information about ecosystem condition is made spatially explicit by assigning variable or indicator values to individual ecosystem assets.
+Spatiall aggregation of data refers to the conversion of fine-resolution data into coarser-resolution data, and depending on both the purpose and the nature of the metric, there can be several reasons both for performing and for refraining from spatial aggregation.
+Readers are referred to @allain2018 for a more comprehensive review of different reasons for spatially aggregating indicators.
+There are also several requirements in the SEEA EA that make spatial aggregation necessary.
+However, spatial aggregation also comes with a whole suite of (sometimes undesirable) side-effects, which we shall focus in now.
+
+Generally, spatial aggregation increases the tangibility of a metric, because it simplifies or compresses the information, sometimes into just a single number (total aggregation).
+This is sometimes necessary in ecosystem accounting when information is required at a relatively high administrative level (e.g.
+nations) and when variables and indicators need to be presented in standardised tables which don't allow for describing a lot of spatial variation.
+
+Sometimes total aggregation of a variable or indicator is required for the metric to actually fit its purpose and become relevant to what it is designed to describe.
+For example, a hypothetical metric "*proportion of total land area above critical nitrogen load",* informed from a spatial map of nitrogen loads, would only make sense if data was aggregated across some defined area.
+The reason and need for aggregation may therefore originate in the original idea for the metric.
+
+Data may also be only party aggregated, or the same metric may be presented at different levels of aggregation [@allain2018].
+A minimum level of aggregation may be decided on based on the uncertainties in the data.
+High spatially resolute data may be imprecise at fine scales, leading to false conclusion and potential miss-use [@lyytimäki2013], but these values seen over a larger area may be assumed to converge on some true value, making spatially aggregated data better suited for communication to users.
+Other common arguments for performing spatial aggregation are to ease visualisation and to allow relevant comparisons [@allain2018].
+For example, ecosystem condition metrics can be visualised on a map, but when both the the spatial resolution and spatial variation is high, maps become too complex, even unintelligible.
+It is therefore common to partially aggregate the data and provide a coarser map where different colours, representing values on the map, become distinguishable from each other.
+Similarly, one may chose to aggregate such that the spatial resolution matches some meaningful delineation, notably ecosystem assets in SEEA EA, which makes it possible to easily compare the aggregated values between these areas.
+
+However, spatial aggregation can have several consequences that may or may not be desirable, such as smoothing out spatial variability, which can mask local patterns and trends.
+This smoothing out makes the data and its inherent information less suitable for some use cases, such as depicting spatial variation that can be used to support local land use planning, or for aiding a debate about spatial conflicts or identification of high risk areas.
+
+Spatially aggregating ecosystem condition metrics can be used to alleviate issues of spatial auto-correlation.
+For example if data is collected opportunistically with sampling sizes varying between areas, then performing partial spatial aggregation to some intermediate sized areas will give equal weight or importance to all areas regardless of how many data points are included in each.
+This approach represents a form of post-stratification.
+
+Interpolation or other types of modeling can also be used to impute or predict data values between the points where we have information [@finke2002; @stein2001].
+This approach can be seen as a type of post-stratifiction to get better variable or indicator estimates for the ecosystem assets as a whole.
+It can also be motivated by a decire to know and perhaps visualise the fine-scale variation in variable value on a map.
+However, this type of spatial aggregation, or perhaps better named upscaling, is not the focus of this paper (but see @nte-rel), and in our experience this has not been common practice in SEEA EA accounts produced to date.
+
+## Matching the scale of variables and reference levels {#sec-match}
+
+Similar to variables, the reference levels used for normalising variables also have a spatial assignment, but it can be different from that of the variable.
+For example, variable values may exist for unique 10 × 10 m grid cells, but the reference levels may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire accounting area (e.g.
+a natural zero).
+Variables should not be normalised based on reference levels unless they are at a commensurable spatial scale.
+Let's illustrate this with an example using a hypothetical variable depicting wild animal densities.
+If the upper reference level (X~100~) for this variable is given by the assumed (or measured) animal densities under the reference condition (the optimal condition), then that reference level needs to be assigned a relevant spatial scale, for example a region where the densities are calculated over.
+One will probably acknowledge the fact that animals move and distribute themselves non-randomly within these regions, and it therefore makes little sense to normalise the variable at a finer scale than region.
+Doing so would mean that regions perfectly aligned with the reference condition (in terms of animal densities with the region) will show as having sub-regions with sub-optimal condition simply due to natural variation in where the animals are at any give time.
+In addition, averaging the indicator values across sub-regions would not produce a regional indicator value that represents its true situation as being perfectly aligned with the reference condition, since truncation of variable values above X~100~ (assuming X~100~ \> X~0~) means we are only aggregating the negative deviations from the reference level, and that very high animal densities in some sub-regions are not allowed to compensate for low values elsewhere.
+This example with wild animal densities shows that sometimes variables need to be spatially aggregated before they can be normalised.
+
+## Some pitfalls with spatial aggregation
+
+When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels, thus invoking the issue of the ecological fallacy, and many other potential pitfalls in the spatial aggregation of information.
+The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.
+
+In ECAs that follow SEEA EA, information about ecosystem condition is made spatially explicit by assigning variable or indicator values to individual ecosystem assets.
 The process of assigning values to ecosystem assets requires some spatial manipulations of data, or scale changes [@ewert2011].
@@
 
-<!--# Subsequently, spatial aggregation is required when going from ecosystem assets to large spaatial scale, ultimately the scale of the entire ecosystem accounting area (EAA).  -->
+<!--# Subsequently, spatial aggregation is required when going from ecosystem assets to large spatial scale, ultimately the scale of the entire ecosystem accounting area (EAA).  -->
 
-Complete, or wall-to-wall data, include remote sensing products, as well as other types of pre-aggregated data, where sparse data are assigned to areal units.
+Complete, or wall-to-wall data, include remote sensing products, but also other types of pre-aggregated data, where sparse data are assigned to areal units.
 This last type of data is very common.
@@
 Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54] .
-This deliberate spatial aggregation is the main focus of this paper.
 When doing this aggregation, we need to account for the spatial structure (spatial correlations) of the data (@sec-auto).
@@
 In our experience, the choice of which area units to use when doing this step-wise aggregation is often not reflected much upon.
-There may be limitations with the data, related to the type of pre-aggregation performed (see above).
-In other cases the areal units are decided politically, with little scientific reasoning behind it.
+In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
 However, the arbitrariness of this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; @nte-rel).
@@
 [^2]: When using means across areas, area weighting is required when following SEEA EA.
-
-Interpolation or other modeling can also be used to impute or predict data values between the points where we have information [@finke2002; @stein2001].
-However, this is not the focus of this paper, and in our experience this has not been common practice in SEEA EA accounts produced to date.
-
-When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels, thus invoking the issue of the ecological fallacy, and many other potential pitfalls in the spatial aggregation of information.
-The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.
 
@@
 
-| Term | Definition |
-|----|----|
-| Metric | A collective term for both variables and indicators |
-| Reference levels | Variable values corresponding and anchoring the variable to specific indicator values.
-A special terminology is used, based on @czúcz, which differs somewhat from that in @unitednations2024.
-See @fig-term.
-|
-| Scaling | A linear shrinking or enlarging the variable data range by anchoring X~0~ and X~100~ to the values 0 and 1, respectively. See @fig-term. |
-| Transformation |  |
-| Truncation |  |
-| Normalisation | Referred to as rescaling in @unitednations2024. This is the process of converting a variable to an indicator. |
-| Compensation |  |
-| Aggregation pathway |  |
-
-: Definition of terms {#tbl-def}
-
-# Why to we spatially aggregate data in ecosystem condition accounting
-
-Spatiall aggregation of data refers to the conversion of fine-resolution data into coarser-resolution data, and depending on both the purpose and the nature of the metric, there can be several reasons both for performing and for refraining from spatial aggregation[@allain2018].
-Generally, spatial aggregation increases the tangibility of a metric, because it simplifies or compresses the information, sometimes into just a single number (total aggregation).
-This is sometimes necessary in ecosystem accounting when information is required at a relatively high administrative level (e.g.
-nations) and when variables and indicators need to be presented in standardised tables which dont allow for describing a lot of spatial variation.
-Sometimes total aggregation of a variable or indicator is required for the metric to actually fit its purpose and become relevant to what it is designed to describe.
-For example, a hypothetical metric "*proportion of total land area above critical nitrogen load",* informed from a spatial map of nitrogen loads, would only make sense if data was aggregated across some defined area.
-The reason and need for aggregation may therefore originate in the original idea for the metric.
-Data may also be only party aggregated, or the same metric may be presented at different levels of aggregation [@allain2018].
-
-A minimum level of aggregation may be decided on based on the uncertainties in the data.
-High spatially resolute data may be imprecise at fine scales, leading to false conclusion and potential miss-use [@lyytimäki2013], but these values seen over a larger area may be assumed to converge on some true value, making spatially aggregated data better suited for communication to users.
-Other common arguments for performing spatial aggregation are to ease visualisation and to allow relevant comparisons [@allain2018].
-For example, ecosystem condition metrics can be visualised on a map, but when both the the spatial resolution and spatial variation is high, maps become too complex, even unintelligible.
-It is therefore common to partially aggregate the data and provide a coarser map where different colours, representing values on the map, become distinguishable from each other.
-Similarly, one may chose to aggregate such that the spatial resolution matches some meaningful delineation, notably ecosystem assets in SEEA EA, which makes it possible to easily compare the aggregated values between these areas.
-
-However, spatial aggregation can have several consequences that may or may not be desirable, such as smoothing out spatial variability, which can mask local patterns and trends.
-This smoothing out makes the data and its inherent information less suitable for some uses, such as depicting spatial variation that can be used to support local land use planning, or for aiding a debate about spatial conflicts.
-
-### Accounting for spatial autocorrelation {#sec-auto}
-
-One reason why we might want to spatially aggregate ecosystem condition metrics is to fix issues with spatial auto-correlation, that is to remove problematic spatial structure in the data.
-For example if data is collected opportunistically with sampling size varying between areas, then performing partial spatial aggregation to some intermediate size areaa will give equal weight or importance to to all areas regardless of how many data points are included in each.
-
-### Matching the scale of variables and reference levels {#sec-match}
-
-Similar to variables, the reference levels used for normalising variables also have a spatial assignment, but it can be different from that of the variable.
-For example, variable values may exist for unique 10 × 10 m grid cells, but the reference levels may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire accounting area (e.g.
-a natural zero).
-Variables should not be normalised based on reference levels unless they are at a comparable scale.
-Let's illustrate this with an example using a hypothetical variable depicting wild animal densities.
-If the upper reference level (X~100~) for this variable is given by the assumed (or measured) animal densities under the **reference condition** (the optimal condition), then that reference level needs to be assigned a relevant spatial scale, for example a region where the densities are calculated over.
-One will probably acknowledge the fact that animals move and distribute themselves non-randomly within these regions, and it therefore makes little sense to normalise the variable at a finer scale than region.
-Doing so would mean that regions perfectly aligned with the reference condition (in terms of animal densities with the region) will show as having sub-regions with deteriorated condition simply due to natural variation in where the animals are at any give time.
-In addition, averaging the indicator values across sub-regions would not produce a regional indicator value that represents its true situation as being perfectly aligned with the reference condition, since truncation ensures that variable values above X~100~ (assuming X~100~ \> X~0~) means we are only averaging the deviations from the reference level, and that very high animal densities in some sub-regions are not allowed to **compensate** for low values elsewhere.
-This example with wild animal densities shows that sometimes variables need to be spatially aggregated before they can be normalised.
-
-There are several requirements in the SEEA EA that make spatial aggregation necessary, but spatial aggregation also comes with a whole suite of (sometimes undesirable) side-effects.
-Readers are referred to @allain2018 for a review of different reasons for spatially aggregating indicators.
-
 ```{r fig-term}
@@
 Firstly, it gives a normative interpretation of a variable, defining a good and a bad state and a directionality to say when something is getting better or worse over time [@czucz_selection_2021].
-SEEA EA encourages normalisation, but other environmental assessment framework leaves more of these kinds of value judgments to the end users (ref?).
+SEEA EA encourages normalisation (which they refer to as rescaling), but other environmental assessment framework leaves more of these kinds of value judgments to the end users (ref?).
 
@@
 In other words, normalising variables facilitate the spatial aggregation of our ecosystem condition estimates (i.e. the indicator values).
+The flip-side of this is that spatially aggregating variable values does not directly summarise information about ecosystem condition for that area.
 
-Thirdly, we normalise in order to standardise the indicator on the same scale so that we can perform thematic (i.e. vertical) aggregation (ref).
+Thirdly, we normalise in order to standardise the indicator on the same scale so that we can perform thematic (i.e. vertical) aggregation [@jakobsson2021; @unitednations2024 §5.81].
 This is commonly referred to as *the* reason for normalising variables, but as we have shown, it is but one of three main reasons, and also perhaps the least confusing part for many.

### 2025-02-04 manuscript/manuscript.qmd
update terminology figure

@@
 Note that the meaning of the terms may differ from how they are used in other fields, such as mathematics.
-The variable (1), in raw biophysical units, has a minimum and a maximum value (V), as well as points anchoring it to the indicator scale (X).
+The variable (1), in raw biophysical units, has a minimum and a maximum value (V), as well as points anchoring it to the indicator scale (reference levels; X).
 Scaled variables (2) are scaled based on X~0~ and X~100~ only.
 Truncation (3) implies assigning the value of 0 to values below X~0~, and the value 1 to value above X~100~.
-Transformation refers to the non-linear transformation of values within the 0-1 range, as in this piecewise-linear example by mapping X~60~ to 0.6 on the indicator scale.
+Transformation (4) refers to the non-linear transformation of values within the 0-1 range, as in this piecewise-linear example by mapping X~60~ to 0.6 on the indicator scale.
 Transformations without anchoring points are also included in this term, such as exponential or sigmoid transformation.
-Although both truncation and scaling can be seen as types of transformations, we exclude these methods from the definition here.
-Examples 2, 3, and 4 are normative (hence the gradient colour scale from red to green) because we assume they have declared that one end of the variable scale represents a good state and the other a poor state.
+Categorisation (5) refers to the creation of discrete bins (here five) where incremental change in uncategorised indicator values within a set of class boundaries is ignored.
+Although truncation, scaling and categorisation can be seen as types of transformations, we exclude these methods from the definition here.
+Examples 2-5 are normative (hence the gradient colour scale from red to green) because we assume it has been  declared that one end of the scale represents a good state and the other a poor state.
 All variables that are treated so that they conform to the definition of indicators in the SEEA EA are said to be normalised.

### 2025-02-04 manuscript/manuscript.qmd
update agg bias fig

@@
 Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54] .
-When doing this aggregation, we need to account for the spatial structure (spatial correlations) of the data (@sec-auto).
+When doing this aggregation, we need to account for the spatial structure (spatial correlations) of the data.
 One way is to rely on design-based approaches (relying on random or balanced samples across the entire EEA), or using a step-wise aggregation via ecosystem assets.
@@
 #| fig-cap: |
-#|   Example showing the difference in indicator values (y-axis) obtained from taking the mean of two variable values (V1 and V1) that are on the original scale (i hat) or on the normalised indicator scale (i bar).
-Due to truncation at x100, the latter results in a comparatively lower indicator value.
+#|   Examples showing the difference in indicator values (y-axis) obtained from taking the mean of two variable values (V1 and V1) that are on the original scale (i hat) or on the normalised indicator scale (i bar).
+Pane A: Due to truncation at X~100~, the latter results in a comparatively lower indicator value.
+Pane B: Example where the normalisation function includes a transformation step (see Figure 1 for definition) that anchors X~60~ to 0.6 on the indicator scale.
+This reverses the order of i hat and i bar compared to example in pane A.
 The solid coloured line represents the normalisation function f(V).
@@
 Modified from Rastsetter (1991).
-#| out-width: 400
+#| out-width: 600
 
-knitr::include_graphics("../img/aggregation_bias_truncations.PNG")
+knitr::include_graphics("../img/aggregation_bias.PNG")
 
-```
-
-```{r fig-agg2}
-#| echo: false
-#| fig-cap: |
-#|   Same figure as above, but showing an example where the normalisation function includes a transformation step (see Figure 1 for definition) that anchors X60 to 0.6 on the indicator scale.
-#| out-width: 400
-
-knitr::include_graphics("../img/aggregation_bias_x60.PNG")
 ```
@@
 
-In ecosystem condition accounting the aggregation of both variables and indicators is common practice and raises the issue of **aggregation bias** (@fig-agg1; @fig-agg2) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
+In ecosystem condition accounting the aggregation of both variables and indicators is common practice and raises the issue of **aggregation bias** (@fig-agg1) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
 This means that the spatial aggregation of metrics leads to a divergence in the type of information held in variables and in their associated indicators, to a point where they become qualitatively very different.
@@
 If not careful, aggregation bias may also cause errors to sneak into the indicator logic.
-A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition indicators (@fig-agg1; @fig-agg2), is the choice about the order of the different steps in the normalisation and aggregation process, i.e.
+A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition indicators (@fig-agg1), is the choice about the order of the different steps in the normalisation and aggregation process, i.e.
 the aggregation pathway (@fig-pathways).
@@
 For the next spatial level, when information from all ecosystem assets should be used to create an estimate for the EAA, ecosystem assets with large extents should weight more in the calculations (following the SEEA EA prinsiples of area weighting).
-Note that one cannot skip this first level of spatial aggregation via ecosystem assets, since this is an integral part of the sampling design, leading to spatial autocorrelation (se @sec-auto).
+Note that one cannot skip this first level of spatial aggregation via ecosystem assets, since this is an integral part of the sampling design, leading to spatial autocorrelation.
 However, if in a later ECA, ecosystem assets are defined as subsets of the initial assets (a finer scale account), then the same indicator, when aggregated to the EAA level, will have a different value.

### 2025-02-04 manuscript/manuscript.qmd
down to ch 3

@@
 
-Ecosystem accounting (EA), and other types of environmental assessments, are generally aimed at communicating major changes and trends in the extent and condition of ecosystems at a scale that is relevant for decision makers.
+The world is facing a nature crisis of a magnitude that requires urgent political decision making and international coordination.\
+Ecosystem accounting (EA) is a type of environmental assessments aimed at communicating on the status and trends in the extent and condition of ecosystems at a scale that is relevant for decision makers.
 For ecosystem condition accounting (ECA) and assessments, this often requires that metrics[^1] used to describe the condition of ecosystems are aggregated spatially, sometimes to produce a single value for a larger area.
@@
 However, spatial aggregation affect how a metric should be interpreted, and the descriptive properties and normative claims one can make based on it.
-Without an understanding of the pitfalls during spatial aggregation of information, one risks introducing unintended and undetected systematic bias, as well as downright logical errors.
+Without an understanding of the pitfalls during spatial aggregation of information, one risks introducing unintended and undetected systematic bias, as well as downright logical errors, into the decition making process.
 Although much attention has been devoted to the vertical aggregation of environmental indicators into indices (e.g.
 @langhans2014; @maes2020; @oecd_2008\], as well as to the spatial aggregation of relatively unmodified or raw variables (e.g @allain2018) , there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting (but see @vanbeurden1999).
-This paper therefore focus on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) [@unitednations2024] - the most used ecosystem accounting (natural resource accounting) framework in the world [@comte2022].
-SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2024 @unitednations2022].
-We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is a relevant topic also for the other parts of the account.
+This paper therefore focus on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) [@unitednations2024] - the most used ecosystem accounting framework in the world [@comte2022].
 
@@
 
+SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2024 @unitednations2022].
+We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is a relevant topic also for the other parts of the account.
+When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels, thus invoking the issue of the ecological fallacy (@nte-rel), and many other potential pitfalls in the spatial aggregation of information.
+The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.
+
 <!--# We pay specific attention to the phenomenon of aggregation bias [@rastetter1992] and on the different ways to interpret the information in ecosystem condition variables versus indicators. -->
 
-This paper has a rather narrow and defined focus.
-Related concepts are briefly mentioned in @nte-rel.
+```{=html}
+<!-- This paper has a rather narrow and defined focus.
+Related concepts are briefly mentioned in @nte-rel. -->
+```
 
-## Indicators vs. variables
+## Terminology
 
-SEEA EA distinguishes between indicators and variables, and to be able to describe how these are affected by spatial aggregation, we need to spend some time to explaining the difference between them.
+There is a jungle of terms used for similar metrics or operations across relevant theoretical frameworks for ecosystem assessments, such as the SEEA EA, WFD, MSFD, IBECA, OSPAR, PAEC, HD, CFP, and others.
+We see a need to a common, more standardised terminology across these fields of practice in order to ease communication and avoid misunderstandings.
+For now we will now define the terms and how we use them in this paper.
+
+SEEA EA distinguishes between indicators and variables, as do we.
 Variables are metrics describing ecosystem characteristics [@unitednations2024 §5.41].
@@
 Sometimes indicators are calculated directly, but usually indicators are made by normalising variables based on two or more reference levels [@unitednations2024 §5.60].
-Although also variables may have normative properties (due to choices about how they are chosen, defined, aggregated or presented [@allain2018]) , the normalisation of variables into indicators makes indicators explicitly normative by introducing a clear directional interpretation, where increasing values always refer to an increase in condition.
+Although also variables may have normative properties (due to choices about how they are chosen, defined, aggregated or presented [@allain2018]), the normalisation of variables into indicators makes them explicitly normative by introducing a clear directional interpretation, where increasing values always refer to an increase in condition in the direction of a reference condition.
 The information held by the indicator is also different from that of the variable since additional ecological knowledge can be introduced via the reference levels.
-The mathematical part of this normalisation contains two, sometimes three, steps [@fig-term].
-These steps can be performed simultaneously, but for clarity we discuss them as separate steps.
-The order of these steps can also change.
+The mathematical part of this normalisation contains several steps [@fig-term].
+These can be performed simultaneously, but for clarity we discuss them as separate steps.
+The order of the steps can also change.
 
-Variables can be scaled using a minimum of two reference levels, defining variable values to be coded as zero (X~0~) or one (X~100~) on the indicator scale.
+Variables can be scaled using a minimum of two reference levels that defines which variable values to be coded as zero (X~0~) and as one (X~100~) on the indicator scale.
 Variables can be truncated to produce a bound indicator scale between 0 and 1.
@@
 This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz; @jakobsson_setting_2020], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
+Finally, the indicator scale can also be divided into discrete fractions, either based on additional reference levels, or some other rule (e.g. equal range for each class).
+We refer to this step as categorisation.
+
+```{r fig-term}
+#| fig-cap: |
+#|   Examples of how important terms are used in this manuscript.
+Note that the meaning of the terms may differ from how they are used in other fields, such as mathematics.
+The variable (1), in raw biophysical units, has a minimum and a maximum value (V), as well as points anchoring it to the indicator scale (reference levels; X).
+Scaled variables (2) are scaled based on X~0~ and X~100~ only.
+Truncation (3) implies assigning the value of 0 to values below X~0~, and the value 1 to value above X~100~.
+Transformation (4) refers to the non-linear transformation of values within the 0-1 range, as in this piecewise-linear example by mapping X~60~ to 0.6 on the indicator scale.
+Transformations without anchoring points are also included in this term, such as exponential or sigmoid transformation.
+Categorisation (5) refers to the creation of discrete bins (here five) where incremental change in uncategorised indicator values within a set of class boundaries is ignored.
+Although truncation, scaling and categorisation can be seen as types of transformations, we exclude these methods from the definition here.
+Examples 2-5 are normative (hence the gradient colour scale from red to green) because we assume it has been  declared that one end of the scale represents a good state and the other a poor state.
+All variables that are treated so that they conform to the definition of indicators in the SEEA EA are said to be normalised.
+For example, if a truncated variable is assumed to have a linear relationship with the indicator scale, then this variable has been normalised even though it has not been transformed.
+The term rescale is a synonym to normalise, but we will only use the latter from here on.
+#| echo: false
+knitr::include_graphics("../img/terminologi.png")
+```
 
@@
 Spatiall aggregation of data refers to the conversion of fine-resolution data into coarser-resolution data, and depending on both the purpose and the nature of the metric, there can be several reasons both for performing and for refraining from spatial aggregation.
-Readers are referred to @allain2018 for a more comprehensive review of different reasons for spatially aggregating indicators.
-There are also several requirements in the SEEA EA that make spatial aggregation necessary.
-However, spatial aggregation also comes with a whole suite of (sometimes undesirable) side-effects, which we shall focus in now.
+Readers are referred to @allain2018 for a more comprehensive review of different reasons for spatially aggregating variables.
+As we shall see, there are several requirements in the SEEA EA that make spatial aggregation necessary.
+However, spatial aggregation also comes with a whole suite of (sometimes undesirable) side-effects.
 
@@
 
-Sometimes total aggregation of a variable or indicator is required for the metric to actually fit its purpose and become relevant to what it is designed to describe.
-For example, a hypothetical metric "*proportion of total land area above critical nitrogen load",* informed from a spatial map of nitrogen loads, would only make sense if data was aggregated across some defined area.
+Sometimes total aggregation of a metric is required for the it to actually fit its purpose and become relevant to what it is designed to describe.
+For example, a hypothetical variable "*proportion of total land area above critical nitrogen load",* informed from a spatial map of nitrogen loads, would only make sense if data was aggregated across some defined area.
 The reason and need for aggregation may therefore originate in the original idea for the metric.
@@
 Data may also be only party aggregated, or the same metric may be presented at different levels of aggregation [@allain2018].
-A minimum level of aggregation may be decided on based on the uncertainties in the data.
-High spatially resolute data may be imprecise at fine scales, leading to false conclusion and potential miss-use [@lyytimäki2013], but these values seen over a larger area may be assumed to converge on some true value, making spatially aggregated data better suited for communication to users.
+A minimum level of aggregation may be decided based on uncertainties in the data: High spatially resolute data may be imprecise at fine scales, leading to false conclusion and potential miss-use [@lyytimäki2013], but these values seen over a larger area may be assumed to converge on some true value, making spatially aggregated data better suited for communication to users.
 Other common arguments for performing spatial aggregation are to ease visualisation and to allow relevant comparisons [@allain2018].
@@
 
-However, spatial aggregation can have several consequences that may or may not be desirable, such as smoothing out spatial variability, which can mask local patterns and trends.
+Spatial aggregation can have several consequences that may or may not be desirable, such as smoothing out spatial variability, which can mask local patterns and trends.
 This smoothing out makes the data and its inherent information less suitable for some use cases, such as depicting spatial variation that can be used to support local land use planning, or for aiding a debate about spatial conflicts or identification of high risk areas.
 
-Spatially aggregating ecosystem condition metrics can be used to alleviate issues of spatial auto-correlation.
+Spatial aggregating of ecosystem condition metrics is applied to alleviate issues of spatial auto-correlation.
 For example if data is collected opportunistically with sampling sizes varying between areas, then performing partial spatial aggregation to some intermediate sized areas will give equal weight or importance to all areas regardless of how many data points are included in each.
@@
 
-Interpolation or other types of modeling can also be used to impute or predict data values between the points where we have information [@finke2002; @stein2001].
-This approach can be seen as a type of post-stratifiction to get better variable or indicator estimates for the ecosystem assets as a whole.
-It can also be motivated by a decire to know and perhaps visualise the fine-scale variation in variable value on a map.
-However, this type of spatial aggregation, or perhaps better named upscaling, is not the focus of this paper (but see @nte-rel), and in our experience this has not been common practice in SEEA EA accounts produced to date.
+Interpolation, or other types of modeling, can also be used to impute or predict data values between the points where we have information [@finke2002; @stein2001].
+This approach can also be seen as a type of post-stratifiction, and it can result in better variable or indicator estimates for the ecosystem assets as a whole.
+It can also be motivated by a desire to know more about the fine-scale variation in variable value, and perhaps visualise this on a map.
+This type of spatial aggregation, or perhaps better named upscaling, is not the focus of this paper (but see @nte-rel), and in our experience this has not been common practice in SEEA EA accounts produced to date.
 
@@
 
-Similar to variables, the reference levels used for normalising variables also have a spatial assignment, but it can be different from that of the variable.
+Similar to variables, the reference levels used for normalising variables also have a spatial assignment, which can be different from that of the variable.
 For example, variable values may exist for unique 10 × 10 m grid cells, but the reference levels may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire accounting area (e.g.
@@
 Let's illustrate this with an example using a hypothetical variable depicting wild animal densities.
-If the upper reference level (X~100~) for this variable is given by the assumed (or measured) animal densities under the reference condition (the optimal condition), then that reference level needs to be assigned a relevant spatial scale, for example a region where the densities are calculated over.
+If the upper reference level (X~100~) for this variable is given by the assumed (or measured) animal densities under the reference condition, then that reference level needs to be assigned a relevant spatial scale, for example a region where the densities are calculated over.
 One will probably acknowledge the fact that animals move and distribute themselves non-randomly within these regions, and it therefore makes little sense to normalise the variable at a finer scale than region.
-Doing so would mean that regions perfectly aligned with the reference condition (in terms of animal densities with the region) will show as having sub-regions with sub-optimal condition simply due to natural variation in where the animals are at any give time.
-In addition, averaging the indicator values across sub-regions would not produce a regional indicator value that represents its true situation as being perfectly aligned with the reference condition, since truncation of variable values above X~100~ (assuming X~100~ \> X~0~) means we are only aggregating the negative deviations from the reference level, and that very high animal densities in some sub-regions are not allowed to compensate for low values elsewhere.
+Doing so would mean that regions perfectly aligned with the reference condition (in terms of animal densities with the region) will show as having sub-regions with sub-optimal condition, simply due to natural variation in where the animals are at any give time.
+Averaging (arguably the most common spatial aggregation function) the indicator values across sub-regions would not produce a regional indicator value that shows the region as being perfectly aligned with the reference condition, since truncation of variable values above X~100~ (assuming X~100~ \> X~0~) means we are only aggregating the negative deviations from the reference level, and that very high animal densities in some sub-regions are not allowed to compensate for low values elsewhere.
 This example with wild animal densities shows that sometimes variables need to be spatially aggregated before they can be normalised.
+See @nte-billberry for another example.
 
-## Some pitfalls with spatial aggregation
+::: {#nte-billberry .callout-note appearance="minimal" icon="false"}
+### Billberry example
 
-When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels, thus invoking the issue of the ecological fallacy, and many other potential pitfalls in the spatial aggregation of information.
-The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.
+*Variable*: Horizontal coverage (vertical projection) of billberry (*Vaccinum myrtillus*) recorded in permanent vegetation plots.\
 
-In ECAs that follow SEEA EA, information about ecosystem condition is made spatially explicit by assigning variable or indicator values to individual ecosystem assets.
-The process of assigning values to ecosystem assets requires some spatial manipulations of data, or scale changes [@ewert2011].
+*Reference levels*: X~100~ was defined for each of 5 regions in Norway based on an expert elicitation.
+Experts were informed by the distribution of the variable values both within and outside protected areas, and the regional distribution of major forest types.
+They also used their general knowledge about the effect of forestry and on the general vegetation structure of old-growth forests.\
+
+The billberry indicator exemplifies different spatial resolutions for the variable and for the reference levels.
+The variable is recorded at the scale of vegetation plots.
+The reference level however, is designed with a regional spatial scale in mind, especially because of how it encompasses the known variation in forest types in a region to estimate the mean billberry coverage under the reference condition.
+Because the normalisation includes a truncation step, normalisation the variable at the plot scale (pathway 1; @fig-agg1) would not allow overshooting values to compensate for lower values when aggregating regional indicator values, which would then be negatively displaced ($\bar i$) relative to when aggregating variable values ($\hat i$; @fig-agg1).\
+<!-- Scaling at the plot scale would require unique reference levels for each forest type.
+However, both pathway 1 and 3 has been used for this variable in two different forest ecosystem condition assessments.
+Due to non-negligeable truncation of the variable during the normalisation, the two indicators were subject to different levels of aggregation bias, and consequently the spatially aggregated indicator values for the two assessments became different.
+This caused a general confusion about how how the same variable (and the same data) can produce different indicator values. -->
+
+*Conclusion*: Given that the reference value is the way it is, this indicator should have followed pathway 3 (@fig-agg1), and spatially aggregated the variable from plot scale to regional scale before normalising the variable and potentially aggregating it further.
+:::
+
+## Fitting data to ecosystem assets
+
+In SEEA EA, information about ecosystem condition is made spatially explicit by assigning variable or indicator values to ecosystem assets.
+This process requires some spatial manipulations of data, or scale changes [@ewert2011].
 The type of manipulation depends on the spatial resolution of the underlying data, as well as the size and configuration of the ecosystem assets.
@@
 Consider as an example data from LiDAR (Light Detection And Ranging).
-The raw data is a point cloud, yet the most commonly available datasets are surface or canopy models; wall-to-wall generalisations of the point cloud.
-Also data such as rasters from satellite images ,or maps derived from such, have some inherent pre-aggregation where values are averaged for each raster cell.
+The raw data is a point cloud, yet the most commonly available datasets are rasters, such as surface or canopy models; wall-to-wall generalisations of the point cloud.
+Also data such as satellite images ,or maps derived from such, have some inherent pre-aggregation where values are averaged for each raster cell.
 Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (@nte-rel).
@@
 In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
-However, the arbitrariness of this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; @nte-rel).
+However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; @nte-rel).
 It is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
@@
 
-```{r fig-term}
-#| fig-cap: |
-#|   Examples of how important terms are used in this manuscript.
-Note that the meaning of the terms may differ from how they are used in other fields, such as mathematics.
-The variable (1), in raw biophysical units, has a minimum and a maximum value (V), as well as points anchoring it to the indicator scale (reference levels; X).
-Scaled variables (2) are scaled based on X~0~ and X~100~ only.
-Truncation (3) implies assigning the value of 0 to values below X~0~, and the value 1 to value above X~100~.
-Transformation (4) refers to the non-linear transformation of values within the 0-1 range, as in this piecewise-linear example by mapping X~60~ to 0.6 on the indicator scale.
-Transformations without anchoring points are also included in this term, such as exponential or sigmoid transformation.
-Categorisation (5) refers to the creation of discrete bins (here five) where incremental change in uncategorised indicator values within a set of class boundaries is ignored.
-Although truncation, scaling and categorisation can be seen as types of transformations, we exclude these methods from the definition here.
-Examples 2-5 are normative (hence the gradient colour scale from red to green) because we assume it has been  declared that one end of the scale represents a good state and the other a poor state.
-All variables that are treated so that they conform to the definition of indicators in the SEEA EA are said to be normalised.
-For example, if a truncated variable is assumed to have a linear relationship with the indicator scale, then this variable has been normalised even though it has not been transformed.
-The term rescale is a synonym to normalise, but we will only use the latter from here on.
-#| echo: false
-knitr::include_graphics("../img/terminologi.png")
-```
-
 # Why do we normalise variables? {#sec-norm}
@@
 Firstly, it gives a normative interpretation of a variable, defining a good and a bad state and a directionality to say when something is getting better or worse over time [@czucz_selection_2021].
-SEEA EA encourages normalisation (which they refer to as rescaling), but other environmental assessment framework leaves more of these kinds of value judgments to the end users (ref?).
+SEEA EA encourages normalisation (which they refer to as rescaling), but other environmental assessment or natural resource accounting frameworks, refrains from these kinds of value judgments, or leaves more of this task up to the end users (ref?).
 
 Secondly, normalisation sets a limit to how much a high indicator value in one place can compensate for a low value somewhere else, and vice versa [@pedersen2016].
-This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviation from the reference levels, and positive deviations (which could compensate for the negative ones) are ignored.
+This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviation from the reference levels, and positive deviations (which could compensate for the negative ones) are ignored (see example in @sec-match).
 Transformation can sometimes also have this effect, e.g. sigmoid transformations.
@@
 
-In ecosystem condition accounting the aggregation of both variables and indicators is common practice and raises the issue of **aggregation bias** (@fig-agg1) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
+In ecosystem condition accounting the aggregation of both variables and indicators is common practice and raises the issue of aggregation bias (@fig-agg1) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
 This means that the spatial aggregation of metrics leads to a divergence in the type of information held in variables and in their associated indicators, to a point where they become qualitatively very different.
@@
 
-## Billberry example {#sec-bill}
-
-<!-- Indicator: Blåbæredekning (billberry coverage) -->
-
-<!-- Dataset: Norwegian national forest inventory. -->
-
-<!-- Reference: https://www.naturindeks.no/Indicators/blaabaer -->
-
-*Variable*: Horizontal coverage (vertical projection) of billberry (*Vaccinum myrtillus*) recorded in permanent vegetation plots.
-
-*Reference levels*: X~100~ was defined for each of 5 regions in Norway based on an expert elicitation.
-Experts were informed by the distribution of the variable values both within and outside protected areas, and the regional distribution of major forest types.
-They also used their general knowledge about the effect of forestry and on the general vegetation structure of old-growth forests.
-
-<!-- *Pathway*: 3 and 1 (on sperate occations) -->
-
-The billberry indicator exemplifies different spatial resolutions for the variable and for the reference levels.
-The variable is recorded at the scale of vegetation plots.
-The reference level however, is designed with a regional spatial scale in mind, especially because of how it encompasses the known variation in forest types in a region to estimate the mean billberry coverage under the reference condition.
-Because the normalisation includes a truncation step, scaling the variable at the plot scale (pathway 1; @fig-agg1) would not allow overshooting values to compensate for lower values when aggregating regional indicator values, which would then be negatively displaced ($\bar i$) relative to when aggregating variable values ($\hat i$; @fig-agg1).
-Scaling at the plot scale would require require unique reference levels for each forest type.
-However, both pathway 1 and 3 has been used for this variable in two different forest ecosystem condition assessments.
-Due to non-negligeable truncation of the variable during the normalisation, the two indicators were subject to different levels of aggregation bias, and consequently the spatially aggregated indicator values for the two assessments became different.
-This caused a general confusion about how how the same variable (and the same data) can produce different indicator values.
-
-*Conclusion*: Given that the reference value is the way it is, this indicator should have followed pathway 3, and spatially aggregated the variable from plot scale to regional scale before normalising the variable and potentially aggregating it further.
-It would also be possible to use pathway 6.
-
 The wolves example also highlight the importance of making clear the differences in the interpretation of variables and indicators.
@@
 
-## Alien species example {#sec-ali}
+## Alien species example
 

### 2025-02-05 manuscript/manuscript.qmd
Callouts as boxes

@@
     wrap: sentence
+callout-appearance: simple
+callout-icon: false
 ---
@@
 This example with wild animal densities shows that sometimes variables need to be spatially aggregated before they can be normalised.
-See @nte-billberry for another example.
+See Box 1 for another example.
 
-::: {#nte-billberry .callout-note appearance="minimal" icon="false"}
-### Billberry example
+::: {.callout-note}
+
+**Box 1 -- Billberry example** \
 
@@
 Also data such as satellite images ,or maps derived from such, have some inherent pre-aggregation where values are averaged for each raster cell.
-Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (@nte-rel).
+Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (Box 2).
 This issue will for example affect our ability to detect true maximum or minimum values (used in *worst case* aggregation, for example in the WFD) for the phenomena our variable is describing, since these extremes are already smoothed out in the pre-aggregation step.
@@
 In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
-However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; @nte-rel).
+However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; Box 2).
 It is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
@@
 
-::: {#nte-rel .callout-note appearance="minimal" icon="false"}
-### Related concepts
+::: {.callout-note}
 
-##### Vertical aggregation {.unnumbered}
+**Box 2 -- Related concepts** \
+
+**Vertical aggregation**
 
@@
 
-##### Upscaling {.unnumbered}
+**Upscaling**
 
@@
 
-##### Modifiable area unit problem {.unnumbered}
+**Modifiable area unit problem**
 
@@
 
-##### Ecological fallacy {.unnumbered}
+**Ecological fallacy**
 
@@
 This displacement is refereed to as aggregation error, and its effect become more pronounced when the non-linearity is strong between variables and indicators.
-In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2021; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
+In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2024; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
 We now describe two ways that aggregation bias affects ECAs: by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators, and secondly, how it can cause confusion about the effects of chosing different **aggregation pathways** during indicator development.
@@
 In pathway 8, spatial aggregation is only done on individual indicators.
-The two pathways differ in how they handle more complicated schemes for weighted averages, with pathway 8 being more flexible, but it may interfere with the principle of commutativity [@unitednations2021].
+The two pathways differ in how they handle more complicated schemes for weighted averages, with pathway 8 being more flexible, but it may interfere with the principle of commutativity [@unitednations2024].
 

### 2025-02-06 manuscript/manuscript.qmd
down to recommendations

@@
 
-The world is facing a nature crisis of a magnitude that requires urgent political decision making and international coordination.\
+The world is facing a nature crisis of a magnitude that requires urgent political decision making and international coordination.
 Ecosystem accounting (EA) is a type of environmental assessments aimed at communicating on the status and trends in the extent and condition of ecosystems at a scale that is relevant for decision makers.
@@
 Without an understanding of the pitfalls during spatial aggregation of information, one risks introducing unintended and undetected systematic bias, as well as downright logical errors, into the decition making process.
+
+[^1]: We use the SEEA EA definitions of the terms variables, indicators (normalised variables) and metrics (refering to either a variable or a metric) [@unitednations2024]
+
 Although much attention has been devoted to the vertical aggregation of environmental indicators into indices (e.g.
@@
 
-[^1]: We use the SEEA EA definitions of the terms variables, indicators (normalised variables) and metrics (refering to either a variable or a metric) [@unitednations2024]
-
 SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2024 @unitednations2022].
 We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is a relevant topic also for the other parts of the account.
-When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels, thus invoking the issue of the ecological fallacy (@nte-rel), and many other potential pitfalls in the spatial aggregation of information.
+When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels, thus invoking the issue of the ecological fallacy (Box 1), and many other potential pitfalls in the spatial aggregation of information.
 The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.
 
-<!--# We pay specific attention to the phenomenon of aggregation bias [@rastetter1992] and on the different ways to interpret the information in ecosystem condition variables versus indicators. -->
-
-```{=html}
-<!-- This paper has a rather narrow and defined focus.
-Related concepts are briefly mentioned in @nte-rel. -->
-```
+This paper is aimed primarily at indicator developers, providing them with some recommendations for how to choose the right spatial aggregation method for variables and indicators, and how to report these choices.
+Indicator developer subjectivity affects ECAs, in part because the process of designing indicators and estimating indicator values often left entirely to the them, with little real involvement from indicator users.
+It is therefore important to be aware of the role that indicator developers have in shaping the type of information, and the narratives, that is presented to decision makers.
+This paper will make it easier for indicator developers to avoid some of the pitfalls from having a too casual approach to the choice of aggregation pathways.
 
@@
 This example with wild animal densities shows that sometimes variables need to be spatially aggregated before they can be normalised.
-See Box 1 for another example.
+See Example 1 for another example.
 
-::: {.callout-note}
-
-**Box 1 -- Billberry example** \
+::: callout-note
+**Example 1 -- Billberry**\
 
@@
 Also data such as satellite images ,or maps derived from such, have some inherent pre-aggregation where values are averaged for each raster cell.
-Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (Box 2).
+Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (Box 1).
 This issue will for example affect our ability to detect true maximum or minimum values (used in *worst case* aggregation, for example in the WFD) for the phenomena our variable is describing, since these extremes are already smoothed out in the pre-aggregation step.
@@
 In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
-However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; Box 2).
+However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; Box 1).
 It is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
@@
 
-::: {.callout-note}
-
-**Box 2 -- Related concepts** \
+::: callout-note
+**Box 1 -- Related concepts**\
 
@@
 
-Secondly, normalisation sets a limit to how much a high indicator value in one place can compensate for a low value somewhere else, and vice versa [@pedersen2016].
-This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviation from the reference levels, and positive deviations (which could compensate for the negative ones) are ignored (see example in @sec-match).
+Secondly, normalisation sets a limit to how much a high variable value in one place can compensate for a low value somewhere else, and vice versa [@pedersen2016].
+This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviations from the upper reference level X~100~, and ignoring any positive deviations which would otherwise compensate for the negative ones (see example in @sec-match).
 Transformation can sometimes also have this effect, e.g. sigmoid transformations.
@@
 For example, direct extrapolation based on uneven or spatially biased sampling leads to biased estimates [@kolstad_making_2024].
-In this paper, however, we are mostly interested in what we consider an extended case of aggregation error, sensu @rastetter1992.
+We will, however, focus now on what we consider an extended case of aggregation error, sensu @rastetter1992.
 They showed that when we take the average of two variables, each of which have associated attribute values (e.g.
@@
 In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2024; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
-We now describe two ways that aggregation bias affects ECAs: by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators, and secondly, how it can cause confusion about the effects of chosing different **aggregation pathways** during indicator development.
+We now describe two ways that aggregation bias affects ECAs: by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators, and secondly, how it can cause confusion about the effects of chosing different aggregation pathways (see @sec-pathways) during indicator development.
 
@@
 The ecosystem characteristic that this metric is meant to describe is "*Appropriate herbivore pressure, balancing overgrazing and shrubification"*.
-The variable shows relatively high densities in half the area and low densities in the other half.
+Imagine the variable shows relatively high densities in half the area and low densities in the other half.
 When spatially aggregating the variable, either by taking the mean density or the sum of individuals, will produce an estimate that is somewhere in the middle of the two regions.
@@
 
-The indicator arguably has much less room for miss-interpretation (relative to its intended use).
-Although one common misunderstanding is that indicators named by an animal (e.g. reindeer) is taken by untrained users to represent a population status (typically the number of individuals).
-This would, however, imply that the spatial aggregation function used for the indicator is summation, but the recommended, on assumed most used function, is an area weighted mean.
+The wolves example (Example 2) also highlight the importance of making clear the differences in the interpretation of variables and indicators.
+Indicators and variables are sometimes presented side-by-side in ecosystem condition assessments or accounts.
+This is also recommended practice following the SEEA EA [@unitednations2024; ch. 5.3.4].
+However, due to aggregation bias, spatially aggregated indicators should not be interpreted as normalised versions of aggregated variables.
+Aggregated indicators reflect the average ecosystem condition, where areas in very good or very poor condition have limited ability to compensate for opposite extreme values elsewhere.
+The spatially aggregated variable on the other hand, may reflect something like a sum of individuals, an average of a population, or some other aspect where values above and below the reference levels are able to compensate for each other.
+
+Aggregated variables and the associated aggregated indicators cannot act as alternative evidence or interpretations for the state of nature.
+While this may initially seem like a democratic and transparent approach to reporting, this practice places a significant burden on both the end user and the indicator developer.
+Miss-use can ensue, for example by taking the variable value and assigning it the same interpretation as the indicator value.
+t the very least, the developer must clarify to the user how the two metrics have diverged significantly during spatial aggregation and normalisation, complicating the comparison of their respective information.
+
+::: callout-note
+**Example 2 -- Wolves**\
+
+*Variable*: Number of wolves en each carnivore management area (CMA).\
+
+*Reference levels:*   X~0~ = zero wolves present.\
+X~100~= Number of wolves equal to what experts think the ecosystem would support under the reference condition.
+The reference levels are unique to each (CMA).\
+
+This example is from Norway, where wolves are extinct in most of the country.
+Therefore, most carnivore management areas will get an indicator value of 0.
+Wolves exist in designated wolf zones in south-east Norway, although in small numbers.
+In the Norwegian Nature Index [@jakobsson2020], the variable follows pathway 1 and is normalised at the CMA level.
+When aggregating the indicator to a national value, it is common for the general public to interpret this value as the status for wolves in Norway, and to connect this value to Norway's international obligations to maintain wolf numbers.
+But note that even an unlimited number of wolves inside the wolf zones would not to bring the national indicator value up to a very hight level.
+This is because the normalisation process truncates the variable, and does not allow overshooting values to compensate for lower values in the rest of the country.
+The Norwegian Nature Index uses (approximately) the same pathway for all indicators, so this common misunderstandings about what indicator values represents can be explaind just one time, and not for each indicator specifically.\
+
+```{=html}
+<!-- \*Conclusion\*: Indicators should make intuitive sense to the readers.
+However, there are sometimes good reasons for making things complicated. -->
+```
+
+It can, however, be a good idea to sometimes go back to the variable, and see if it can be defined differently to avoid these kinds of confusions.
+Wolf numbers could in this example be converted to wolf density.
+:::
+
+```{=html}
+<!-- The indicator arguably has much less room for miss-interpretation (relative to its intended use).
+Although one common misunderstanding is that indicators named by an animal (e.g. wolves; Example 2) is taken by untrained users to represent a population status (typically the number of individuals).
+This would, however, imply that the spatial aggregation function used for the indicator is summation, 
+but the recommended, on arguably the most used function, is an area weighted mean.
 With this aggregation function, indicators will always refer to the mean indicator value for all the areas considered, and if properly defined, refer to the degree of fulfillment for the ecosystem characteristic.
-Sometimes changing the indicator name, or redefining the indicator, goes a long way towards easing the interpretation and avoiding miss-use (e.g. *mean reindeer densities*).
-
-Although the variable and the indicator may be reference by the same common name, and may be presented side-by-side, it is not so that they can replace each other or act as alternative evidence.
-Presenting the same metric both in its indicator form and its variable form opens up for alternative interpretations.
-While this may initially seem like a democratic and transparent approach to reporting, this practice actually places a significant burden on both the end user and the indicator developer.
-The developer must clarify to the user how the two metrics have diverged significantly during spatial aggregation and normalisation, complicating the comparison of their respective information.
-This can lead to miss-use, for example by taking the variable value and assigning it the same interpretation as the indicator value.
+Sometimes changing the indicator name, or redefining the indicator, goes a long way towards easing the interpretation and avoiding miss-use (e.g. *mean reindeer densities*). -->
+```
 
 Generally, because of the normalisation process, indicators are more aligned with the purpose of ECA with describing ecosystem condition in normative terms.
-Although variable accounts are allowed under the SEEA EA framework, they may be seen as a type of interim account for metrics that cannot or have not yet been normalised.
+Although variable accounts are allowed under the SEEA EA framework, ans is indeed a central feature, we belive they should be seen as interim accounts for metrics that cannot or have not yet been normalised.
 We would therefore argue that when an indicator has been developed, the original variable is not presented.
-And when only variable estimates are given, extra care is give to describing the spatial aggregation pathway, and how this will have evened out any spatial variation that is relevant for interpreting the variable.
+And when only variable estimates are given, extra care is given to describing the spatial aggregation pathway.
 
@@
 the aggregation pathway (@fig-pathways).
-If normalisation occurs early in the aggregation process, the amount of compensation is limited.
-However, sometimes complete or partial compensation is justified, as in the example in @sec-pti, or as we saw in @sec-match where indicator development involved an initial step of spatially aggregating variables to align with the spatial configuration of reference levels (pathways 2 and 3; @fig-pathways).
-Failure to harmonise the spatial scale between variables and reference levels before normalising the variable can lead to aggregation bias, as described in @sec-bill.
+If normalisation occurs early in the aggregation process, the amount of compensation, or off-setting, is limited.
+However, sometimes complete or partial compensation is justified.
+In @sec-match we saw how indicator development involved an initial step of spatially aggregating variables to align with the spatial configuration of reference levels (pathways 2 and 3; @fig-pathways).
+In Example 3 we see that compensation can be warranted, depending on how high the reference levels are set.
 
-# Compliance across nested assessments {#sec-comp}
-
-Another cause of potential errors or confusion arising during the spatial aggregation of indicators has to do with the number of aggregation levels.
-The SEEA EA uses a principle of commensurability, meaning that the order of certain operations should not affect the resulting estimate or value.
-Spesifically, it should not matter whether you aggregated vertically first (to create an index) and then horizontally (i.e. spatial), or the other way around.
-Note that this prisiple may not always hold for other cases, such as aggregating across nested areas, e.g. from ecosystem assets to EAAs, although this is perhaps considered desirable.
-This has to do with the use of area weighting in the spatial aggregation.
-Consider an indicator originating from point-based field surveys.
-The sampling location are selected using a stratified protocol that ensures a representative sample of similar size (same *n*) is made for each ecosystem asset.
-If the variable is normalised early in the aggregation pathway to create indicator values for each locality, then the mean of indicator values for each region will be representative for those regions, and no area weighting is required.
-For the next spatial level, when information from all ecosystem assets should be used to create an estimate for the EAA, ecosystem assets with large extents should weight more in the calculations (following the SEEA EA prinsiples of area weighting).
-Note that one cannot skip this first level of spatial aggregation via ecosystem assets, since this is an integral part of the sampling design, leading to spatial autocorrelation.
-However, if in a later ECA, ecosystem assets are defined as subsets of the initial assets (a finer scale account), then the same indicator, when aggregated to the EAA level, will have a different value.
-A similar case would be is sampling was random and dispersed across the EAA, lets say a country, and indicator values were first calculated as some average across all sampling points, and then in a later account or assessment, the same indicator was estimated for separate regions.
-When aggregating the indicator values from regions to the whole country, using area weighting, the national estimate will differ, albeit slightly, from the initial national estimate that did not include this step-wise aggregation.
-Cases when two accounts or assessments use what is considered the same indicator but find different values for the same areas, is cause for both confusion, and potentially mistrust.
-In our experience this can also lead to time consuming detective work, trying to find the causes for such discrepancies.
-We suggest two immediate solution to this issue.
-First, to find better ways to document the aggregation pathways that are being used for each indicator, as part of the indicator metadata.
-Second, to find ways to distinguish between apparently identical indicators, for example using different version numbers or IDs.
-We provide practical solution to both these suggestions in @sec-rec.
-
-# Indicator developer subjectivity
-
-Indicator developer subjectivity affects ECAs, in part because the process of designing indicators and estimating indicator values often left entirely to the them, with little real involvement from indicator users.
-It is therefore important to be aware of the role that indicator developers have in shaping the type of information, and the narratives, that is presented to decision makers.
-This paper is aimed at providing some recommendations for how to choose the right spatial aggregation method for variables and indicators, and how to report these choices.
-This will make it easier for indicator developers to avoid some of the pitfalls from having a too casual approach to the choice of aggregation pathways.
-
-The choice of aggregation function (e.g.
-weighted arithmetic mean, worst-rule, or others) also influences the degree of compensation, i.e.
-the degree to which good condition in one place can compensate for bad condition somewhere else (ref WFD?).
-Therefore, spatial aggregation can introduce substantial subjectivity from the side of the indicator developers, which also limits the different ways that a metric can be used and interpreted by indicator users (the target audience).
-In other words, spatial aggregation tasks indicator developers with defining narratives for underlying variables, and it is therefor a normative and value-laden operation.
-
-# Aggregation pathways
-
-We introduce the term aggregation pathway to describe the order of the steps used to go from a spatially explicit variable to a spatially aggregated metric, usually an indicator.
-In @fig-pathways we show some aggregation pathways that we have come across, and some that we we see as potential new pathways.
-
-Pathway 1 involves early normalisation using the perhaps most common order of the three steps scaling, truncating and transforming.
-The three steps may be done simultaneously as well.
-Then the indicator is spatially aggregated.
-Pathway 2 involves aggregating the variable before normalising.
-Pathway 3 is similar, but here there are two aggregation steps: one before and one after normalisation.
-This can for example be the case when variables are aggregated to the scale corresponding to the reference levels before they can be normalised.
-Pathway 4 illustrated the aggregation of a variable, with no normalisation.
-This is the pathway commonly used for variable accounts in the SEE EA. Pathway 5 is a common pathway in the WFD, which does not include spatial aggregation of indicator values.
-Instead one report the number of water bodies in each condition class (spatial statistics).
-Pathway 6 is a suggested pathway for handling WFD indicator in ECAs (see @sec-pti).
-
-Pathways 7-8 describe the spatial aggregation of multiple indicators, as they all start with the normalisation step.
-These are not the main topic of this paper, but are included here for completeness.
-In pathway 7, indicators are thematically aggregated to an index, which is then again aggregated spatially.
-In pathway 8, spatial aggregation is only done on individual indicators.
-The two pathways differ in how they handle more complicated schemes for weighted averages, with pathway 8 being more flexible, but it may interfere with the principle of commutativity [@unitednations2024].
-
-Pathway 9 describe how a single indicators is aggregated hierarchically from region to region, whereas pathway 10 shows a case were the aggregation process always starts from scratch for each hierarchical spatial level.
-As for pathways 7 and 8, pathway 9 and 10 differ in how they propagate the indicator weight (typically area weight) between regions.
-
-The choice about which pathway to use is not trivial, and in this paper we want to highlight some of the issues that could arise from having an *ad hoc* approach to these considerations.
-Here we go through selected examples to illustrate some common pitfalls.
-
-<!--# what we see as four general points that developers of ecosystem condition indicators and assessments should reflect upon carefully when deciding on an aggregation pathway for their indicator or their assessment.
--->
-
-# Examples
-
-## Wolves example {#sec-wolves}
-
-*Variable*: Number of wolves en each carnivore management area (CMA).
-
-*Reference:* X~0~ = is zero wolves present.
-X~100~= Number of wolves equal to what experts think the ecosystem would support under the reference condition.
-The reference levels are unique to each (CMA).
-
-This example is from Norway, where wolves are extinct in most of the country.
-Therefore, most carnivore management areas will get an indicator value of 0.
-Wolves exist in designated wolf zones in south-east Norway, although in small numbers.
-In the Norwegian Nature Index [@jakobsson2020], the variable follows pathway 1 and is normalised at the CMA level.
-When aggregating the indicator to a national value, it is common for the general public to interpret this value as the status for wolves in Norway, and to connect this value to Norway's international obligations to maintain wolf numbers.
-But note that even an unlimited number of wolves in the wolf zone would not to bring the national indicator value up to a level indicator good condition.
-This is because the normalisation process truncates the variable, and does not allow overshooting values to compensate for lower values in the rest of the country.
-
-\*Conclusion\*: Indicators should make intuitive sense to the readers.
-However, there are sometimes good reasons for making things complicated.
-In this example, The Norwegian Nature Index uses the same pathway for all indicators, so this common misunderstandings about what indicator values represents can be explaind just one time, and not for each indicator specifically.
-It can also be a good idea to sometime go back to the variable, and see if it can be defined differently to avoid these kinds of confusions.
-Wolf numbers could in this example be converted to wolf density.
-
-The wolves example also highlight the importance of making clear the differences in the interpretation of variables and indicators.
-Indicators and variables are sometimes presented side-by-side in ecosystem condition assessments or accounts.
-This is also recommended <!--# is this true?
---> practice following the SEEA EA.
-However, due to aggregation bias, spatially aggregated indicators should not be interpreted as normalised versions of aggregated variables.
-Indicators typically now reflect the average ecosystem condition, where areas in very good or very poor condition have limited ability to compensate for opposite extreme values elsewhere.
-The spatially aggregated variable on the other hand, may reflect something like a sum of individuals, an average of a population, or some other aspect where values above and below the reference levels are able to compensate for each other.
-
-## Alien species example
-
-*Variable*: The local ecological effect from alien species (mostly vascular plants) recorded on a 7-step ordinal scale (1 = no alien species, 7 = only alien species).
-The data comes from nature type monitoring with reference to individual occurrences (polygons).
-
-*Reference levels*: X~0~ = total dominance from alien species (variable value 7).
-X~100~ = zero influence from alien species (variable value 1).
-X~60~ = \~2-5% alien plant cover (variable value 3).
-
-<!-- *Pathway*: 1 -->
-
-In this example, the reference levels are uniform throughout the EAA, and the normalisation includes a transformation in the form of a piecewise-linear transformation by anchoring X~60~ to 0.6 on the indicator scale.
-When performing an ECA, one has the option to either aggregate variable values for each region (each ecosystem asset), and then normalise that value (pathway 2 and 3), or to normalise the variable at the scale of the original nature type polygons, and then aggregate the indicator values to the regional level (pathway 1).
-
-*Conclusion*: One benefit of early normalisation is that two nested assessments, such as one national, and one assessment of region *Y* inside the same country, will get the same spatially aggregated value for region *Y*.
-The other thing to ask oneself here is: *Why do we normalise* in the first place (@sec-norm)?
-One reason is that we want to summarise information in space about the ecosystem condition.
-It is the indicator values that hold this information, and not the variable.
-We would therefore favor pathway 1 for this example.
-
-<!--# Separate assessments should give identical indicator values for the same areas, given the same underlying data.
-When this is not true it raises concerns about reproducibility and sacrifices the credibility of both assessments.
-This was also the case in the billberry example, but here we used another example to illustrate the point further.
--->
-
-## Phytoplankton trophic index (PTI) {#sec-pti}
+::: callout-note
+**Example 3 -- Phytoplankton trophic index (PTI)**\
 
@@
 *Reference levels*: X~100~ = median variable value for water bodies in reference condition.
-X~0~, X~20~, X~40~, X~60~ and X~80~ = intercalibrated threshold values, based on dose–response curves.
+X~0~, X~20~, X~40~, X~60~ and X~80~ = intercalibrated threshold values, based on dose–response curves.\
 
@@
 
-For use in water management, the Water Framework Directive (WFD) indicators uses aggregation pathway 7 [@fig-pathways], with truncation, scaling, and transformation.
+The Water Framework Directive (WFD) indicators uses aggregation pathway 7 [@fig-pathways], with truncation, scaling, and transformation, in that order.
 The value obtained after the first two steps is called an EQR (ecological quality ratio), and the value obtained after the third step nEQR (normalised EQR).
-Spatial aggregation is not done for the WFD, and therefore aggregation bias is not an issue.
-When WFD data are put into use in other contexts, however, aggregation becomes important.
+Spatial aggregation, the way that we use the term here, is not done for the WFD.
+Instead, the proportion of water bodies in each condition class is the aggregated reporting metric.
+Therefore aggregation bias is not an issue.
+When WFD data are put into use in other contexts, however, aggregation becomes important.\
 <!-- Relevant contexts in Norway are the Nature Index (ref) and ecological condition accounting. In both cases,  -->
@@
 *Conclusion*: Because of how X~100~ is defined based on the median value across reference lakes, overshooting values (\> X~100~) should be preserved in the spatial aggregation.
-<!-- Otherwise, the condition of water bodies with X > X80 will be underreported.  --> Otherwise we get a negative displacement, as in the bilberry example.
-This means that neither EQR values (which are untransformed), nor nEQR values (which are truncated), can be used uncritically in ECAs.
-One solution, to enable the use of WFD indicators in ECA, would be to change to pathway 6, with (1) scaling, (2) transforming, (3) aggregating and (4) truncating the values [@fig-pathways].
-Furthermore, Eurostat @eurostat2022\
-<!-- requires  --> encourages reporting of variable values.
-These, however, cannot be aggregated spatially in a meaningful way.
-<!-- because the use of threshold values amounts to a nonlinear (specifically, a piecewise-linear) transformation.  -->
+<!-- Otherwise, the condition of water bodies with X > X80 will be underreported.  --> Otherwise we get a negative displacement, making it practically impossible to reach an indicator value of 1.
+
+This means that neither EQR values, nor nEQR values, can be uncritically aggregated spatially to be sued in a SEEA EA comlient in ECA.
+One solution, to enable the use of WFD indicators in ECAs, would be to change to pathway 6, with (1) scaling, (2) transforming, (3) aggregating and (4) truncating the values [@fig-pathways].\
+
+Furthermore, Eurostat [@eurostat2022] <!-- requires  --> encourages reporting of variable values.
+These, however, cannot be aggregated spatially in any meaningful way for WFD indicators.
+The issue is the same as for all other variables for which normalised indicators exist, but is perhaps more extreme for WFD indicators due to different reference conditions and levels for different water body types, and strong non-linear transformation functions that inflate the aggreation bias issuse.
+<!-- Can we have a figure of calculation exmaple here? -->
+:::
+
+# Aggreements across nested ECAs {#sec-comp}
+
+We do not presume that ECA end users will have the capacity, nor interest, to refect very much upon the effect that spatial aggregation has had on the properties of the information they are presneted with.
+It should simply work, without the need for elaborate explenations.
+Cases when two accounts or assessments use what is considered the same indicator but find different values for the same areas, is cause for both confusion, and potentially mistrust.
+This can, however, happen, and has happened, when nested ECA use different aggregation pathways for the same indicator, for example when nested assessments use a different number of aggregation levels.
+
+Consider a case ECA was to be made for a country.
+For one of the varaibles, data sampling was random and dispersed across the EAA, and a national indicator values was calculated as some average across all sampling points.
+Then in a later regional ECA, the same indicator was estimated for separate regions, as well as for the whole country.
+When aggregating the indicator values from regions to the whole country, using area weighting, the national estimate will differ, albeit slightly, from the initial national estimate that did not include this step-wise aggregation, causing potentially the two nested assessment to reach different conclusions.
+A third account, focusing on municipalities (using municipalities as ecosystem assets, rather than the larger regions), will create a third answer to the same question.
+We believe it a very desirable property of ECA to be compliant and harmonised assessment scales.
+In this spesific case, one could in the first national account have performed a step-wise aggregtion via municipalities, regions, and finally to the wntire EAA, but it would not always be possibel to know which ecosystem assest are to be used in future accounts.
+We belive one solution to this is to aqknowledge that the aggregation pathway, including the number of aggregation levels used, can affect the indicators so much, sometimes as much as variation in the variable itself, that the use of different pathways imply that metrics are named and treated uniquely, for example using different common names, but perhaps better still, different version numbers or indicator IDs.
+Fr each unique indicator the precise aggregation pathway should be specified as part of the indicator metadata.
+This would show ECA users if two accounts are using different metrics that cannot be assumed to behave similarly or report the same thing, even if they have the same common name and data source.
+
+Besides changing the number of aggregation levels, the choice about when in the aggregation pathway to normalise variables is a possible factor leading to discrepancies across nested accounts (Example 4).
+
+```{=html}
+<!-- The SEEA EA uses a principle of commensurability, meaning that the order of certain operations should not affect the resulting estimate or value.
+Spesifically, it should not matter whether you aggregated vertically first (to create an index) and then horizontally (i.e. spatial), or the other way around.
+Note that this prisiple may not always hold for other cases, such as aggregating across nested areas, e.g. from ecosystem assets to EAAs, although this is perhaps considered desirable.
+This has to do with the use of area weighting in the spatial aggregation. -->
+```
+
+```{=html}
+<!-- 
+Consider an indicator originating from point-based field surveys.
+The sampling location are selected using a stratified protocol that ensures a representative sample of similar size (same *n*) is made for each ecosystem asset.
+If the variable is normalised early in the aggregation pathway to create indicator values for each locality, then the mean of indicator values for each region will be representative for those regions, and no area weighting is required.
+For the next spatial level, when information from all ecosystem assets should be used to create an estimate for the EAA, ecosystem assets with large extents should weight more in the calculations (following the SEEA EA prinsiples of area weighting).
+Note that one cannot skip this first level of spatial aggregation via ecosystem assets, since this is an integral part of the sampling design, leading to spatial autocorrelation.
+However, if in a later ECA, ecosystem assets are defined differently, e.g. as subsets of the initial assets (a finer scale account), 
+then the same indicator, when aggregated to the EAA level, will have a different value. -->
+```
+
+```{=html}
+<!-- In our experience this can also lead to time consuming detective work, trying to find the causes for such discrepancies.
+We suggest two immediate solution to this issue.
+First, to find better ways to document the aggregation pathways that are being used for each indicator, as part of the indicator metadata.
+Second, to find ways to distinguish between apparently identical indicators, for example using different version numbers or IDs.
+We provide practical solution to both these suggestions in @sec-rec. -->
+```
+
+::: callout-note
+**Example 3 -- Alien species**\
+
+*Variable*: The local ecological effect from alien species (mostly vascular plants) recorded on a 7-step ordinal scale (1 = no alien species, 7 = only alien species).
+The data comes from nature type monitoring with reference to individual occurrences (polygons).\
+
+*Reference levels*: X~0~ = total dominance from alien species (variable value 7).
+X~100~ = zero influence from alien species (variable value 1).
+X~60~ = \~2-5% alien plant cover (variable value 3).\
+
+<!-- *Pathway*: 1 -->
+
+In this example, the reference levels are uniform throughout the EAA, and the normalisation includes a transformation in the form of a piecewise-linear transformation by anchoring X~60~ to 0.6 on the indicator scale (similar to @fig-agg1 pane B, but without truncation, as that is not needed in the alien species example).
+When performing an ECA, one has the option to either aggregate variable values for each ecosystem asset, and then normalise that value (pathway 2 and 3), or to normalise the variable at the scale of the original nature type polygons (Basal spatial unit (BSU) scale), and then aggregate the indicator values to the regional level (pathway 1).
+Arguably the variable and the reference levels are at commensurable scales in both cases (cf. Example 1), and other arguments for choosing one over the other must be found.\
+
+One benefit of early normalisation is that two nested assessments, such as one national, and one assessment of region *Y* inside the same country, will get the same spatially aggregated value for region *Y*.
+Separate assessments should give identical indicator values for the same areas, given the same underlying data.
+When this is not true it raises concerns about reproducibility and sacrifices the credibility of both assessments.
+The other thing to ask oneself here is: *Why do we normalise* in the first place (@sec-norm)?
+One reason is that we want to summarise information in space about the ecosystem condition.
+It is the indicator values that hold this information, and not the variable.
+We would therefore favor pathway 1 for this example.
+:::
+
+# Aggregation pathways {#sec-pathways}
+
+We introduce the term aggregation pathway to describe the order of the steps used to go from a spatially explicit variable to a spatially aggregated metric, usually an indicator.
+In @fig-pathways we show some aggregation pathways that we have come across, and some that we we see as potential new pathways.
+The choice about which pathway to use is not trivial, and we want to highlight some of the issues that could arise from having an *ad hoc* approach to these considerations.
+
+Pathway 1 involves early normalisation using the perhaps most common order of the three steps scaling, truncating and transforming.
+The three steps may be done simultaneously or in sequence.
+Finanlly, the indicator is spatially aggregated.
+
+Pathway 2 involves aggregating the variable before normalising.
+This preserves the ability for very high or very low variable values to compensate for very low or very high values elsewhere, respectively.
+
+Pathway 3 is similar, but here there are two aggregation steps: one before and one after normalisation.
+This can for example be the case when variables are aggregated to the scale corresponding to the reference levels before they can be normalised (Example 1).
+
+Pathway 4 illustrates the aggregation of a variable, with no normalisation.
+This is the pathway commonly used for variable accounts in the SEE EA.
+For this, and all other aggregation pathways, the picture can be made more complicated by adding additional aggregation levels (i.e.
+a step-wise aggregation, @sec-comp)
+
+Pathway 5 is a common pathway in the WFD, which does not include spatial aggregation of indicator values.
+Instead one report the number of water bodies in each condition class (spatial statistics).
+Pathway 6 is a suggested pathway for handling WFD indicator in ECAs (see Example 3), where truncation is postponed until after the spatial aggregation for in that way to allow overshooting values to compensate for low indicator values esewhere (what we refer to as compensation).
+
+Pathways 7-8 describe the spatial aggregation of composite indicators (indices).
+These are not the main topic of this paper, but are included here for completeness.
+In pathway 7, indicators are thematically aggregated to an index, which is then again aggregated spatially.
+In pathway 8, spatial aggregation is only done on individual indicators.
+The two pathways differ in how they handle more complicated schemes for weighted averages.
+The spatial aggregation of indices (pathway 7), using area weighted means, does not allow different area weights to be applied to each underlying indicator.
+Yet this would in many cases be needed if the individual indicators do not all represents the entire ecosystem inside each ecosystem asset.
+This is often the case, since ECA are frequently geared towards high level ecosystem types (e.g.
+level 2 in the IUCN GET [@keith_iucn_2020]) that includes several lower level ecosystems that may need unique indicators to describe them.
+Pathway 8 is therefore more flexible in this regard, but it may interfere with the principle of commutativity [@unitednations2024].
+We feel this is a very much unresolved issue in the SEEA EA.
+
+Pathway 9 describe how a single indicators can be aggregated hierarchically from region to region, whereas pathway 10 shows a case were the aggregation process always starts from scratch for each hierarchical spatial level.
+As for pathways 7 and 8, pathway 9 and 10 differ in how they propagate the indicator weight (typically area weight) between regions.
+SEEA EA describes and prescribes a hierarchical spatial aggregation (pathway 9).
+Note that can cause issues with comparability across nested assessment, if one alters the number or size of sub-regions (@sec-comp).
 
@@
 #|   Flowchart showing pathways for the spatial aggregation of variables in ecosystem condition accounting.
-Normalisation refers to the steps that make a variable conform to the definition of an indicator in SEEA EA, which may or may not include the transformation step.
-A truncated and/or transformed variable will have a different interpretation than the original variable.
-The aggregation bias depends on whether variables are truncated and/or transformed variables before or after spatial aggregation.
-Variables are never normalised (pathway 4), and hence the aggregation bias for variables will be different than for indicators.
-An important factor influencing the the choice whether to normalise early (as in pathway 1), or later, is the spatial validity of the reference levels, and sometimes raw variable values (e.g.
-billberry coverage in vegetation quadrats) are spatially aggregated to the scale that was considered when setting the reference levels (e.g.
-average regional billberry coverage).
+Normalisation refers to the steps that make a variable conform to the definition of an indicator in SEEA EA, which may include some, but not necesarily all the steps scale, truncate and transform.
 #| out-width: 400

### 2025-02-07 manuscript/manuscript.qmd
abstract

@@
 abstract: |
-  Ecosystem condition assessments (ECA) and accounts use variables and indicators to describe key ecosystem characteristics reflecting the ecosystem condition and any potential deviations from a reference condition.
-These metrices are rutinely aggregation spatially to produce values representing the condition of a larger area, such as a country.
-However, we experience little awareness about the potential pitfals arising from aggregation bias of highly modified indicators.
-Here we outline some consequences regarding the order of the steps involed in normalising variables and aggregating them in space.
-We show that the choice of  aggregation pathways in non-trivial and has the potential to undermine the credibility and precition in ECAs, and to confuse the communication of their results.
-We introduce a common terminology for aggregation pathways, spesific to ecosystem condition indicators following SEEA EA standards, and make some recomendations about which pathway to use in different settings, and how to report these choices.
+  Ecosystem condition accounts (ECA) make use of variables and indicators to describe key ecosystem characteristics, reflecting their condition and deviations from a reference condition.
+These metrics are often spatially aggregated to represent larger areas, such as countries.
+However, spatial aggregation of information has the potential to alter the descriptive and normative interpretations one can make from these metrics.
+For example, aggregation bias causes the information held in variables and indicators to diverge when these are aggregated spatially.
+This process is influenced also by the order of steps involved in normalising and aggregating variables.
+This paper outlines the consequences of different aggregation pathways, emphasising their impact on the credibility of ECAs, and how these are interpreted by users.
+We introduce a standardized terminology for aggregation pathways specific to ecosystem condition indicators following the SEEA EA standard and provide recommendations for selecting appropriate pathways in various contexts.
+Our discussion of this topic is aimed at raising the general awareness of spatial aggregation issues and to guide indicator developers in choosing and reporting spatial aggregation methods.
 
@@
 Instead one report the number of water bodies in each condition class (spatial statistics).
-Pathway 6 is a suggested pathway for handling WFD indicator in ECAs (see Example 3), where truncation is postponed until after the spatial aggregation for in that way to allow overshooting values to compensate for low indicator values esewhere (what we refer to as compensation).
+Pathway 6 is a suggested pathway for handling WFD indicator in ECAs (see Example 3), where truncation is postponed until after the spatial aggregation for in that way to allow overshooting values to compensate for low indicator values elsewhere (what we refer to as compensation).
 
@@
 
-Based on the example above, we suggest some recommendations for developers of ecosystem condition indicators and assessment for trying to avoid some of the pitfalls from having a too casual approach to the choice of aggregation pathways.
+Based on the above discussion and examples, we here summarise four recommendations for developers of ecosystem condition indicators and assessments for avoiding some of the pitfalls from having a too casual approach to the choice of aggregation pathways.
 
@@
 
-In indicator documentation or ECA reports, authors should make an effort to make a precise description of the steps in the aggregation process.
-To do this we suggest using the terms as described in @fig-term.
-The description can be made even easier by referencing specific aggregation pathways in @fig-pathways by number.
-Besides making the indicator workflow itself more transparent and reproducible, ee believe this recommendation would help raise the general awareness about aggregation bias in EAs, both among readers and researchers.
-It will make the interpretation of assessment less prone to misunderstanding.
-It will also make troubleshooting easier, as i the billberry example where researchers had to spend several days looking though old code to identify the reason why two assessment produces dissimilar indicator values from the same underlying data.
+ECAs should be accompanied by detailed indicator documentation where a precise description of the steps in the aggregation process is presented.
+Verbose descriptions can be supplemented or replaced by a standardised short-hand annotation that we shall present now.
+We suggest using the terms as described in @fig-term and elsewhere in the paper: scale, truncate (abbreviated to *trunc.*), transform (*trans.*), spatial aggregation (*sp.
+agg.*), thematic aggregation (*th.
+agg.*) and spatial statistics (*sp.
+stat.*).
+These terms can be abbreviated as shown in the parentheses, and placed togeth into a single string, where operations performed simultaneously are enclosed in square brackets.
+For example, pathway 1 in @fig-pathways can be annotated as *scale - trunc.
+- trans.
+- sp.
+agg..* If the three normalisation steps are performed simultaneously this can be written as \[*scale - trunc.
+- trans.\] - sp.
+agg.* If there are more aggregation steps, then the spatial level at each step should be included in the description, e.g.
+for pathway 3: *sp.
+agg.
+(municipalities) - scale - sp.
+agg.
+(country).
+"*Municipality" and "country" could here be included as footnotes.
+Note also that in this example, normalisation only involved a scaling step.
+The description of indicator pathways can perhaps made even easier by referencing specific aggregation pathways in @fig-pathways by number, as we have done in this paper.
+
+Besides making the indicator workflow itself more transparent and reproducible, we believe this recommendation would help raise the general awareness about aggregation bias in EAs, both among users and developers.
+It will make the interpretation of ECAs less prone to misunderstanding and miss-use.
+It will also make troubleshooting easier, for example when trying to identify the reason why two ECAs produces dissimilar indicator values from the same underlying data.
 
@@
 
-Normalising variables early in the aggregation pathway has several benefits, such as comparability between nested assessments (@sec-ali).
-It also means that you are aggregating normative measures of condition, which is generally what you want in ECAs.
-But same as variables and indicators, reference levels also have a spatial resolution, and they should only be used to normalise variables when the variable is at a scale which is relevant to the way the reference levels are defined.
-This was exemplified with the billberry example (@sec-bill).
+Normalising variables early in the aggregation pathway has several benefits, such as comparability between nested assessments (Example 4, @sec-comp).
+It also means that you are aggregating normative measures of condition, which is generally what you want in ECAs (@sec-norm).
+But same as for variables and indicators, reference levels also have a spatial resolution, and they should only be used to normalise variables when the variable is at a scale which is relevant to the way the reference levels are defined, i.e.
+they are at a commensurable scale (Example 1).
 
-### 3. Use the same aggregation pathway for all indicators in the same assessment {.unnumbered}
+### 3. Use a similar aggregation approach for all indicators in the same assessment {.unnumbered}
 
-It may be premature, or not even possible, to prescribe an aggregation pathway to be used for all ECAs, but internal consistency within assessments should be possible.
-As we saw in the wolves example (@sec-wolves), this eases interpretation and reduces the chance that the meaning of indicators are misunderstood.
+It may be premature, or not even possible, to prescribe an aggregation pathway to be used for all ECAs, or a type of aggregation pathway to be used given the data at hand.
+However, internal consistency within assessments should be possible, and this can make it easier to communicate and interpret the information conveyed though the indicator values.
+This can be achieved by for example always scaling early (see recommendation number 2 above) and using the same aggregation levels and same approach to step-wise aggregation across levels (@sec-comp), for all indicators.
 
-### 4. Use unique indicator IDs on indicators that are similar, but use different pathways {.unnumbered}
+### 4. Use unique indicator IDs , also for indicators that are similar, but use different aggregation pathways {.unnumbered}
 
-Indicators (and variables) often exist in multiple version, varying slightly in the raw data or in the methods used to produce the data or metrics.
+Indicators (and variables) often exist in multiple versions, varying slightly in the raw data or in the methods used to produce the data or metrics.
 Yet different versions are often refereed to by the same common name.
 This causes confusion about which indicator version is being used, and thus making it difficult to make out which aggregation pathway that has been used.
-We recommend making use of stable indicator IDs, unique for each version of the indicators sharing the same common name.
-This has for example been implemented in [ecRxiv](https://github.com/NINAnor/ecRxiv), a publishing platform for ecosystem condition indicators.
+We recommend making use of stable and unique indicator IDs, also and especially for indicators that use the same common name.
+This has already been implemented in ecRxiv, a GitHub-based publishing platform for ecosystem condition indicators [@Kolstad_ecRxiv].
+On ecRxiv, indicator documentation follow a structured approach to both reporting and peer review (including code review).
+Each indicator version is given its own ID, and its own fact sheet with a persistent URL, so that ECA that report these indicators unequivocally can cite specific indicator versions, and how they looked at a given point in time.
+ecRxiv also includes standardised metadata reporting with fields, including ecosystem type (IUCN GET) and aggregation pathways (see recommendation number 1 above).
 

### 2025-02-07 manuscript/manuscript.qmd
start revise

@@
 This paper outlines the consequences of different aggregation pathways, emphasising their impact on the credibility of ECAs, and how these are interpreted by users.
-We introduce a standardized terminology for aggregation pathways specific to ecosystem condition indicators following the SEEA EA standard and provide recommendations for selecting appropriate pathways in various contexts.
+We introduce a standardised terminology for aggregation pathways specific to ecosystem condition indicators following the SEEA EA standard and provide recommendations for selecting appropriate pathways in various contexts.
 Our discussion of this topic is aimed at raising the general awareness of spatial aggregation issues and to guide indicator developers in choosing and reporting spatial aggregation methods.
@@
 
-The world is facing a nature crisis of a magnitude that requires urgent political decision making and international coordination.
-Ecosystem accounting (EA) is a type of environmental assessments aimed at communicating on the status and trends in the extent and condition of ecosystems at a scale that is relevant for decision makers.
-For ecosystem condition accounting (ECA) and assessments, this often requires that metrics[^1] used to describe the condition of ecosystems are aggregated spatially, sometimes to produce a single value for a larger area.
-This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something that is relevant to decision makers and perhaps actionable at a political level.
-However, spatial aggregation affect how a metric should be interpreted, and the descriptive properties and normative claims one can make based on it.
-Without an understanding of the pitfalls during spatial aggregation of information, one risks introducing unintended and undetected systematic bias, as well as downright logical errors, into the decition making process.
+The world is facing a nature crisis of a magnitude that requires urgent political decision making and international coordination [@ipbes_global_2019].
+Ecosystem accounting (EA) is a type of environmental assessments aimed at communicating on the status and trends in the extent and condition of ecosystems at a scale and in a format that is relevant for decision makers [@edens2022, @comte2022].
+For ecosystem condition accounting (ECA) and assessments, this often requires that metrics[^1] used to describe the condition of ecosystems are aggregated spatially to produce a single value for a larger area.
+This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something that is relevant to decision makers and perhaps actionable at a political level [@king2024] <!--# read more -->.
+However, spatial aggregation affect how a metric should be interpreted, and the descriptive properties and normative claims one can make based on it [@allain2018], and without an understanding of the pitfalls during spatial aggregation of information, one risks introducing unintended and undetected systematic bias, as well as downright logical errors, into the decition making process.
 
@@
 
-Although much attention has been devoted to the vertical aggregation of environmental indicators into indices (e.g.
-@langhans2014; @maes2020; @oecd_2008\], as well as to the spatial aggregation of relatively unmodified or raw variables (e.g @allain2018) , there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting (but see @vanbeurden1999).
+Although much attention has been devoted to the vertical aggregation of environmental indicators into indices [e.g.
+@langhans2014; @maes2020; @oecd_2008], as well as to the spatial aggregation of relatively unmodified or raw variables [e.g.
+@allain2018], there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting [but see @vanbeurden1999].
 This paper therefore focus on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) [@unitednations2024] - the most used ecosystem accounting framework in the world [@comte2022].
 
-SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2024 @unitednations2022].
-We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is a relevant topic also for the other parts of the account.
-When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels, thus invoking the issue of the ecological fallacy (Box 1), and many other potential pitfalls in the spatial aggregation of information.
+SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2024; @unitednations2022].
+We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is a relevant topic also for the other parts of the account, and to other types of indicator-based ecosystem assessments.
+When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels.
 The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.
+<!--# thus invoking the issue of the ecological fallacy (Box 1), and many other potential pitfalls in the spatial aggregation of information. -->
 
-This paper is aimed primarily at indicator developers, providing them with some recommendations for how to choose the right spatial aggregation method for variables and indicators, and how to report these choices.
+This paper is aimed primarily at indicator developers, providing them with some recommendations for how to choose the right spatial aggregation approach for variables and indicators, and how to report these choices.
 Indicator developer subjectivity affects ECAs, in part because the process of designing indicators and estimating indicator values often left entirely to the them, with little real involvement from indicator users.
@@
 Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (Box 1).
-This issue will for example affect our ability to detect true maximum or minimum values (used in *worst case* aggregation, for example in the WFD) for the phenomena our variable is describing, since these extremes are already smoothed out in the pre-aggregation step.
+This issue will for example affect our ability to detect true maximum or minimum values (used in *worst case* aggregation, for example in the WFD) for the phenomena our variable is describing, since these extremes are already smoothed out in the pre-aggregation step [@vanbeurden1999].
 This will clearly affect the suitability of such data for identifying areas of high risk (very degraded areas, for example).
@@
 
-Pathway 2 involves aggregating the variable before normalising.
+Pathway 2 involves aggregating the variable before normalising[^3].
 This preserves the ability for very high or very low variable values to compensate for very low or very high values elsewhere, respectively.
+
+[^3]: In a parallel terminology, normalisation can be considered modeling, and spatial aggregation before and after normalisation can be referred to as pre or post processing, respectively [@vanbeurden1999].
 

### 2025-02-17 manuscript/manuscript.qmd
working on text

@@
 This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something that is relevant to decision makers and perhaps actionable at a political level [@king2024] <!--# read more -->.
-However, spatial aggregation affect how a metric should be interpreted, and the descriptive properties and normative claims one can make based on it [@allain2018], and without an understanding of the pitfalls during spatial aggregation of information, one risks introducing unintended and undetected systematic bias, as well as downright logical errors, into the decition making process.
+However, spatial aggregation affect how a metric should be interpreted, and the descriptive properties and normative claims one can make based on it [@allain2018], and without an understanding of the pitfalls during spatial aggregation of information, one risks introducing unintended and undetected systematic bias, as well as downright logical errors, into the decision making process.
 
@@
 
-SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2024; @unitednations2022].
+SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2024; -@unitednations2022].
 We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is a relevant topic also for the other parts of the account, and to other types of indicator-based ecosystem assessments.
@@
 This paper is aimed primarily at indicator developers, providing them with some recommendations for how to choose the right spatial aggregation approach for variables and indicators, and how to report these choices.
-Indicator developer subjectivity affects ECAs, in part because the process of designing indicators and estimating indicator values often left entirely to the them, with little real involvement from indicator users.
+Indicator developer subjectivity affects indicator interpretation [@allain2018] and finally the conclusion from ECAs.
+Still, the process of designing indicators and estimating indicator values often left entirely to the them, with little real involvement from indicator users.
 It is therefore important to be aware of the role that indicator developers have in shaping the type of information, and the narratives, that is presented to decision makers.
 This paper will make it easier for indicator developers to avoid some of the pitfalls from having a too casual approach to the choice of aggregation pathways.
+We use four examples to illustrate some of our main points (@tbl-examples).
+In @sec-pathways we describe a new reporting terminology for spatial aggregation pathways.
+And finally, in @sec-rec we give four concrete recommendations aimed at indicator developers that will help ensure a sensible and informed approach to spatial scaling of ecosystem condition metrics.
 
@@
 Variables are metrics describing ecosystem characteristics [@unitednations2024 §5.41].
-They are typically either in percentages or fractions, or as raw biophysical units, such as biomass, density, or cover.
-Indicators on the other hand, are always on a unitless scale between 0 and 1, where 0 represent the worst condition, and 1 represents the best condition.
-Sometimes indicators are calculated directly, but usually indicators are made by normalising variables based on two or more reference levels [@unitednations2024 §5.60].
+They are typically either in percentages or fractions, or as raw biophysical units, such as biomass or density.
+Indicators on the other hand, are always on a unitless scale between 0 and 1, where 0 represent very poor condition, and 1 represents the very good, or the best possible condition.
+Sometimes indicators are calculated directly [e.g. @åström], but usually indicators are made by normalising variables based on two or more reference levels [@unitednations2024 §5.60].
 Although also variables may have normative properties (due to choices about how they are chosen, defined, aggregated or presented [@allain2018]), the normalisation of variables into indicators makes them explicitly normative by introducing a clear directional interpretation, where increasing values always refer to an increase in condition in the direction of a reference condition.
 The information held by the indicator is also different from that of the variable since additional ecological knowledge can be introduced via the reference levels.
-The mathematical part of this normalisation contains several steps [@fig-term].
+The mathematical part of this normalisation contains several steps (@fig-term).
 These can be performed simultaneously, but for clarity we discuss them as separate steps.
@@
 
-Variables can be scaled using a minimum of two reference levels that defines which variable values to be coded as zero (X~0~) and as one (X~100~) on the indicator scale.
-Variables can be truncated to produce a bound indicator scale between 0 and 1.
+Variables can be scaled using a minimum of two reference levels, a lower and an upper, that defines which variable values to be coded as zero (X~0~) and as one (X~100~) on the indicator scale.
+Variables can, if needed, truncated to produce a bound indicator scale between 0 and 1.
 Sometimes variables are transformed to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
-This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz; @jakobsson_setting_2020], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation.
+This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz; @jakobsson_setting_2020], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation [@mienna2024].
 Finally, the indicator scale can also be divided into discrete fractions, either based on additional reference levels, or some other rule (e.g. equal range for each class).
@@
 Note that the meaning of the terms may differ from how they are used in other fields, such as mathematics.
-The variable (1), in raw biophysical units, has a minimum and a maximum value (V), as well as points anchoring it to the indicator scale (reference levels; X).
-Scaled variables (2) are scaled based on X~0~ and X~100~ only.
+The variable (1), in original units, has a minimum (V~min~) and a maximum value (V~max~), as well as points anchoring it to the indicator scale (reference levels; X~n~).
+Scaled variables (2) are scaled based on X~0~ and X~100~.
 Truncation (3) implies assigning the value of 0 to values below X~0~, and the value 1 to value above X~100~.
@@
 It is therefore common to partially aggregate the data and provide a coarser map where different colours, representing values on the map, become distinguishable from each other.
-Similarly, one may chose to aggregate such that the spatial resolution matches some meaningful delineation, notably ecosystem assets in SEEA EA, which makes it possible to easily compare the aggregated values between these areas.
+Similarly, one may chose to aggregate such that the spatial resolution matches some meaningful delineation, notably ecosystem assets in SEEA EA, which makes it possible to easily compare the ecosystem condition between these areas.
 
 Spatial aggregation can have several consequences that may or may not be desirable, such as smoothing out spatial variability, which can mask local patterns and trends.
-This smoothing out makes the data and its inherent information less suitable for some use cases, such as depicting spatial variation that can be used to support local land use planning, or for aiding a debate about spatial conflicts or identification of high risk areas.
+This smoothing out makes the data and its inherent information less suitable for some use cases, such as depicting spatial variation that can be used to support local land-use planning, for aiding a debate about spatial conflicts or identification of high risk areas [@vanbeurden1999].
 
@@
 For example if data is collected opportunistically with sampling sizes varying between areas, then performing partial spatial aggregation to some intermediate sized areas will give equal weight or importance to all areas regardless of how many data points are included in each.
-This approach represents a form of post-stratification.
+This approach represents a form of post-stratification that is also relevant to process data from hierarchical (e.g. blocked) sampling designs.
 
-Interpolation, or other types of modeling, can also be used to impute or predict data values between the points where we have information [@finke2002; @stein2001].
-This approach can also be seen as a type of post-stratifiction, and it can result in better variable or indicator estimates for the ecosystem assets as a whole.
-It can also be motivated by a desire to know more about the fine-scale variation in variable value, and perhaps visualise this on a map.
-This type of spatial aggregation, or perhaps better named upscaling, is not the focus of this paper (but see @nte-rel), and in our experience this has not been common practice in SEEA EA accounts produced to date.
+| Example | Main illustrating point |
+|-------------------------|-----------------------------------------------|
+| Billbery | Commensurable spatial scales for variable and reference levels (@sec-match) |
+| Deer | Compensation, and discrepancy between variable and indicator (@sec-discrep) |
+| PTI | Compensation and the order of operations (@sec-pathways) |
+| Alien species | Agreements across scales (@sec-comp) |
+
+: Examples used in this paper, and how they illustrate specific points raised {#tbl-examples}
 
@@
 Variables should not be normalised based on reference levels unless they are at a commensurable spatial scale.
-Let's illustrate this with an example using a hypothetical variable depicting wild animal densities.
-If the upper reference level (X~100~) for this variable is given by the assumed (or measured) animal densities under the reference condition, then that reference level needs to be assigned a relevant spatial scale, for example a region where the densities are calculated over.
+The bilberry example (**Example 1**) shows that sometimes variables need to be spatially aggregated before they can be normalised.
+
+<!--# Let's illustrate this with an example using a hypothetical variable depicting wild animal densities.
+If the upper reference level (X100) for this variable is given by the assumed (or measured) animal densities under the reference condition, then that reference level needs to be assigned a relevant spatial scale, for example a region where the densities are calculated over.
 One will probably acknowledge the fact that animals move and distribute themselves non-randomly within these regions, and it therefore makes little sense to normalise the variable at a finer scale than region.
 Doing so would mean that regions perfectly aligned with the reference condition (in terms of animal densities with the region) will show as having sub-regions with sub-optimal condition, simply due to natural variation in where the animals are at any give time.
-Averaging (arguably the most common spatial aggregation function) the indicator values across sub-regions would not produce a regional indicator value that shows the region as being perfectly aligned with the reference condition, since truncation of variable values above X~100~ (assuming X~100~ \> X~0~) means we are only aggregating the negative deviations from the reference level, and that very high animal densities in some sub-regions are not allowed to compensate for low values elsewhere.
-This example with wild animal densities shows that sometimes variables need to be spatially aggregated before they can be normalised.
-See Example 1 for another example.
+Averaging (arguably the most common spatial aggregation function) the indicator values across sub-regions would not produce a regional indicator value that shows the region as being perfectly aligned with the reference condition, since truncation of variable values above X100 (assuming X100 > X0) means we are only aggregating the negative deviations from the reference level, and that very high animal densities in some sub-regions are not allowed to compensate for low values elsewhere.
+-->
 
@@
 
-*Variable*: Horizontal coverage (vertical projection) of billberry (*Vaccinum myrtillus*) recorded in permanent vegetation plots.\
+*Variable*: Horizontal coverage (vertical projection) of bilberry (*Vaccinum myrtillus*) recorded in permanent vegetation plots.\
 
@@
 Experts were informed by the distribution of the variable values both within and outside protected areas, and the regional distribution of major forest types.
-They also used their general knowledge about the effect of forestry and on the general vegetation structure of old-growth forests.\
+They also used their general knowledge about the effect of forestry and on the general vegetation structure of old-growth forests.
+X~0~ is a natural zero (bilberry completely absent).\
 
-The billberry indicator exemplifies different spatial resolutions for the variable and for the reference levels.
+The bilberry indicator exemplifies different spatial resolutions for the variable and for the reference levels.
 The variable is recorded at the scale of vegetation plots.
 The reference level however, is designed with a regional spatial scale in mind, especially because of how it encompasses the known variation in forest types in a region to estimate the mean billberry coverage under the reference condition.
-Because the normalisation includes a truncation step, normalisation the variable at the plot scale (pathway 1; @fig-agg1) would not allow overshooting values to compensate for lower values when aggregating regional indicator values, which would then be negatively displaced ($\bar i$) relative to when aggregating variable values ($\hat i$; @fig-agg1).\
+Because the normalisation includes a truncation step, normalisation the variable at the plot scale (pathway 1; @fig-agg1) would not allow overshooting values to compensate for lower values when aggregating regional indicator values, which would then be negatively displaced ($\bar i$) relative to when aggregating variable values ($\hat i$; @fig-agg1).
+This means one is essentially aggregating the negative deviations from the upper reference level X~100~.\
 <!-- Scaling at the plot scale would require unique reference levels for each forest type.
@@
 
-*Conclusion*: Given that the reference value is the way it is, this indicator should have followed pathway 3 (@fig-agg1), and spatially aggregated the variable from plot scale to regional scale before normalising the variable and potentially aggregating it further.
+*Conclusion*: Given that the reference levels are the way they are, this indicator should follow pathway 3 (@fig-agg1), and the variable should be spatially aggregated from plot scale to regional scale before normalising the variable and potentially aggregating it further.
 :::
@@
 The type of manipulation depends on the spatial resolution of the underlying data, as well as the size and configuration of the ecosystem assets.
-Two main types of data can be distinguished: sparse (i.e. spatially incomplete) data and complete data.
+To elaborate on this issue we first make a distinction between two main types of data: sparse (i.e. spatially incomplete) data and complete data.
 
@@
 Also data such as satellite images ,or maps derived from such, have some inherent pre-aggregation where values are averaged for each raster cell.
-Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (Box 1).
+Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (**Box 1**).
 This issue will for example affect our ability to detect true maximum or minimum values (used in *worst case* aggregation, for example in the WFD) for the phenomena our variable is describing, since these extremes are already smoothed out in the pre-aggregation step [@vanbeurden1999].
@@
 When doing this aggregation, we need to account for the spatial structure (spatial correlations) of the data.
-One way is to rely on design-based approaches (relying on random or balanced samples across the entire EEA), or using a step-wise aggregation via ecosystem assets.
+One way is to rely on design-based approaches (e.g.
+relying on random or balanced samples across the entire EEA), or using a step-wise aggregation via ecosystem assets or some smaller basal spatial units.
 In our experience, the choice of which area units to use when doing this step-wise aggregation is often not reflected much upon.
@@
 
+Interpolation, or other types of modeling, can also be used to impute or predict data values between the points where we have information [@finke2002; @stein2001].
+This approach can also be seen as a type of post-stratifiction, and it can result in better variable or indicator estimates for the ecosystem assets as a whole.
+It can also be motivated by a desire to know more about the fine-scale variation in variable values, and perhaps visualise this on a map.
+This type of spatial aggregation, or perhaps better named upscaling, is not the focus of this paper (but see **Box 1**), and in our experience this has not been common practice in SEEA EA accounts produced to date.
+
 ::: callout-note
@@
 
-It is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (i.e.
-thematic or vertical aggregation; @langhans2014), but in this paper we are discussing spatial (i.e.
-horizontal) aggregation only.
+It is common in the ECA community of practice to use the term aggregation about the process of aggregating normalised indicators into indices (i.e.
+thematic or vertical aggregation; @langhans2014), but in this paper we are primarily discussing spatial (i.e.
+horizontal) aggregation.
+\\
 
@@
 Using models to predict variable or indicator values is allowed under the SEEA EA.
-However, in this paper we are addressing the second part of this process only, of simplifying the spatial variation, from many to few data points (sometimes also included in the terms upscaling).
+However, in this paper we are addressing the second part of this process only, of simplifying the spatial variation, from many to few data points (sometimes also included in the terms upscaling).\
 
@@
 
-The modifiable area unit problem (MAUP; [@openshaw1984, @wong2009]) describes the sensitivity that spatially aggregated information has on the choice about the spatial units used for either collecting the data or for aggregating the data to.
+The modifiable area unit problem (MAUP) describes the sensitivity of spatially aggregated information to the choice of spatial units used for data collection or aggregation [@openshaw1984, @wong2009].
 These units can be arbitrary (*modifiable*), yet different sizes and configuration of units can have large effects on the outcome of aggregation.
@@
 However, the issue of MAUP in SEEA EA deserves an in-depth analyses that we cannot provide here.
+\\
 
@@
 
-The ecological fallacy [@robinson2009] explains the common misinterpretation of correlations between group (*i.e.* aggregated information) and some phenomena.
+The ecological fallacy [@robinson2009] explains the common misinterpretation of correlations between groups (*i.e.* aggregated information) and some phenomena.
 Simply said, the mean of a group, or an area, cannot accurately describe the qualities of individuals in that group, or pin-points inside that area, yet it is common for people to interpret it that way.
@@
 
-Normalisation of ecosystem condition indicators, as defined in @fig-term, serves multiple purposes.
-Firstly, it gives a normative interpretation of a variable, defining a good and a bad state and a directionality to say when something is getting better or worse over time [@czucz_selection_2021].
+Normalisation of ecosystem condition indicators, as defined in @fig-term, serves at least three purposes.
+Firstly, it gives a normative interpretation of a variable, defining a good and a bad state, and simultaneously a directionality to say when something is getting better or worse over time [@czucz_selection_2021].
 SEEA EA encourages normalisation (which they refer to as rescaling), but other environmental assessment or natural resource accounting frameworks, refrains from these kinds of value judgments, or leaves more of this task up to the end users (ref?).
@@
 Secondly, normalisation sets a limit to how much a high variable value in one place can compensate for a low value somewhere else, and vice versa [@pedersen2016].
-This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviations from the upper reference level X~100~, and ignoring any positive deviations which would otherwise compensate for the negative ones (see example in @sec-match).
+This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviations from the upper reference level X~100~, and ignoring any positive deviations which would otherwise compensate for the negative ones (see **Example 2**).
 Transformation can sometimes also have this effect, e.g. sigmoid transformations.
-One reason to want to aggregate the negative deviations only, is because the reference levels are set (or should be set) so that values above these limits do not represent any further increase (or decrease) in ecosystem condition (an exception if the WFD; see @sec-pti).
-Therefore, this way of aggregation summarises the estimated ecosystem condition over an area (which is generally what we want), and not the variable itself.
+One reason to want to aggregate the negative deviations only, is because the X~100~ is set (or should be set) so that values above this limit (assuming X~100~ \> X~0~) do not represent any further increase in ecosystem condition (for an exception, see **Example 3**).
+Therefore, this way of aggregation summarises the estimated ecosystem condition over an area, and not the variable itself.
 In other words, normalising variables facilitate the spatial aggregation of our ecosystem condition estimates (i.e. the indicator values).
@@
 This is commonly referred to as *the* reason for normalising variables, but as we have shown, it is but one of three main reasons, and also perhaps the least confusing part for many.
+
+<!--# Try add using !expr to get hat and bar sympols into caption:  -->
 
@@
 
-## Discrepancies between variables and indicators
+## Discrepancies between variables and indicators {#sec-discrep}
 
@@
 Generally, because of the normalisation process, indicators are more aligned with the purpose of ECA with describing ecosystem condition in normative terms.
-Although variable accounts are allowed under the SEEA EA framework, ans is indeed a central feature, we belive they should be seen as interim accounts for metrics that cannot or have not yet been normalised.
+Although variable accounts are allowed under the SEEA EA framework, ans is indeed a central feature, we believe they should be seen as interim accounts for metrics that cannot or have not yet been normalised.
 We would therefore argue that when an indicator has been developed, the original variable is not presented.
@@
 In @sec-match we saw how indicator development involved an initial step of spatially aggregating variables to align with the spatial configuration of reference levels (pathways 2 and 3; @fig-pathways).
-In Example 3 we see that compensation can be warranted, depending on how high the reference levels are set.
+In **Example 3** we see that compensation can be warranted, depending on how high the reference levels are set.
 
@@
 
-We do not presume that ECA end users will have the capacity, nor interest, to refect very much upon the effect that spatial aggregation has had on the properties of the information they are presneted with.
-It should simply work, without the need for elaborate explenations.
+We do not presume that ECA end users will have the capacity, nor interest, to reflect very much upon the effect that spatial aggregation has had on the properties of the information they are presented with.
+It should simply work, without the need for elaborate explanations.
 Cases when two accounts or assessments use what is considered the same indicator but find different values for the same areas, is cause for both confusion, and potentially mistrust.
@@
 Consider a case ECA was to be made for a country.
-For one of the varaibles, data sampling was random and dispersed across the EAA, and a national indicator values was calculated as some average across all sampling points.
+For one of the variables, data sampling was random and dispersed across the EAA, and a national indicator values was calculated as some average across all sampling points.
 Then in a later regional ECA, the same indicator was estimated for separate regions, as well as for the whole country.
@@
 We believe it a very desirable property of ECA to be compliant and harmonised assessment scales.
-In this spesific case, one could in the first national account have performed a step-wise aggregtion via municipalities, regions, and finally to the wntire EAA, but it would not always be possibel to know which ecosystem assest are to be used in future accounts.
-We belive one solution to this is to aqknowledge that the aggregation pathway, including the number of aggregation levels used, can affect the indicators so much, sometimes as much as variation in the variable itself, that the use of different pathways imply that metrics are named and treated uniquely, for example using different common names, but perhaps better still, different version numbers or indicator IDs.
+In this specific case, one could in the first national account have performed a step-wise aggregation via municipalities, regions, and finally to the entire EAA, but it would not always be possible to know which ecosystem asset are to be used in future accounts.
+We believe one solution to this is to acknowledge that the aggregation pathway, including the number of aggregation levels used, can affect the indicators so much, sometimes as much as variation in the variable itself, that the use of different pathways imply that metrics are named and treated uniquely, for example using different common names, but perhaps better still, different version numbers or indicator IDs.
 Fr each unique indicator the precise aggregation pathway should be specified as part of the indicator metadata.
@@
 ::: callout-note
-**Example 3 -- Alien species**\
+**Example 4 -- Alien species**\
 

### 2025-02-18 manuscript/manuscript.qmd
aggregation displacement

@@
 However, spatial aggregation of information has the potential to alter the descriptive and normative interpretations one can make from these metrics.
-For example, aggregation bias causes the information held in variables and indicators to diverge when these are aggregated spatially.
+For example, aggregation displacement causes the information held in variables and indicators to diverge when these are aggregated spatially.
 This process is influenced also by the order of steps involved in normalising and aggregating variables.
@@
   - aggregation bias
+  - aggregation error
+  - aggregation displacement
   - upscaling
@@
 | Example | Main illustrating point |
-|-------------------------|-----------------------------------------------|
+|-------------------------|----------------------------------------------|
 | Billbery | Commensurable spatial scales for variable and reference levels (@sec-match) |
@@
 Further still, data may not be available at the original spatial scale, but may only be downloadable in preaggregated form [@vanbeurden1999].
-All these cases lead to some form of aggregation bias [@rastetter1992] where the aggregated information has a non-linear relationship with the true variable value.
+All these cases lead to some form of aggregation error[@rastetter1992] where the aggregated information has a non-linear relationship with the true variable value.
 However, the issue of MAUP in SEEA EA deserves an in-depth analyses that we cannot provide here.
@@
 Firstly, it gives a normative interpretation of a variable, defining a good and a bad state, and simultaneously a directionality to say when something is getting better or worse over time [@czucz_selection_2021].
-SEEA EA encourages normalisation (which they refer to as rescaling), but other environmental assessment or natural resource accounting frameworks, refrains from these kinds of value judgments, or leaves more of this task up to the end users (ref?).
-
-Secondly, normalisation sets a limit to how much a high variable value in one place can compensate for a low value somewhere else, and vice versa [@pedersen2016].
+SEEA EA encourages normalisation (which they refer to as rescaling), but other environmental assessment or natural resource accounting frameworks, refrains from these kinds of value judgments, or leaves more of this task up to the end users.
+<!-- ref? --> Secondly, normalisation sets a limit to how much a high variable value in one place can compensate for a low value somewhere else, and vice versa [@pedersen2016].
 This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviations from the upper reference level X~100~, and ignoring any positive deviations which would otherwise compensate for the negative ones (see **Example 2**).
@@
 #| fig-cap: |
-#|   Examples showing the difference in indicator values (y-axis) obtained from taking the mean  of two variable values (V1 and V1) that are on the original scale (i hat) or on the normalised indicator scale (i bar).
+#|   Examples showing the difference in indicator values (y-axis) obtained from taking the mean (V bar) of two variable values (V1 and V1) that are on the original scale (i hat) or on the normalised indicator scale (i bar).
 Pane A: Due to truncation at X~100~, the latter results in a comparatively lower indicator value.
@@
 
-# Aggregation bias
+# Aggregation displacement and spatial compensation {#sec-agg}
 
@@
 We will, however, focus now on what we consider an extended case of aggregation error, sensu @rastetter1992.
-They showed that when we take the average of two variables, each of which have associated attribute values (e.g.
-indicator values), then the attribute of the average variable value differs from the average of the attributes of the original variables.
-This displacement is refereed to as aggregation error, and its effect become more pronounced when the non-linearity is strong between variables and indicators.
-In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2024; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
-We now describe two ways that aggregation bias affects ECAs: by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators, and secondly, how it can cause confusion about the effects of chosing different aggregation pathways (see @sec-pathways) during indicator development.
+They showed that when we take the average of two variables, each of which have attribute values (e.g.
+indicator values) associated with them via some non-linear function, then the attribute of the average variable value differs from the average of the attributes of the original variables.
+@rastetter1992 referred to this as aggregation error and used it to describe some of the dangers associated with upscaling in the sense of using precise fine-scale information to explain or predict large scale phenomena.
+However, in our case the difference between variables and indicators is an intentional one, and we therefore to refer to what is essentially the same mathematical effect, with the less derogatory term aggregation displacement.
+
+In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2024; -@unitednations2022], there is no mention of aggregation displacement, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
+We now describe two ways that aggregation displacement affects ECAs: by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators (@sec-discrep), and secondly, how it can cause confusion about the effects of choosing different aggregation pathways during indicator development (@sec-develop).
 
@@
 
-In ecosystem condition accounting the aggregation of both variables and indicators is common practice and raises the issue of aggregation bias (@fig-agg1) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
+In ecosystem condition accounting the aggregation of both variables and indicators is common practice and raises the issue of aggregation displacement (@fig-agg1) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
 This means that the spatial aggregation of metrics leads to a divergence in the type of information held in variables and in their associated indicators, to a point where they become qualitatively very different.
@@
 
-The wolves example (Example 2) also highlight the importance of making clear the differences in the interpretation of variables and indicators.
+The wolves example (**Example 2**) also highlight the importance of making clear the differences in the interpretation of variables and indicators.
 Indicators and variables are sometimes presented side-by-side in ecosystem condition assessments or accounts.
 This is also recommended practice following the SEEA EA [@unitednations2024; ch. 5.3.4].
-However, due to aggregation bias, spatially aggregated indicators should not be interpreted as normalised versions of aggregated variables.
+However, due to aggregation displacement, spatially aggregated indicators should not be interpreted as normalised versions of aggregated variables.
 Aggregated indicators reflect the average ecosystem condition, where areas in very good or very poor condition have limited ability to compensate for opposite extreme values elsewhere.
@@
 Miss-use can ensue, for example by taking the variable value and assigning it the same interpretation as the indicator value.
-t the very least, the developer must clarify to the user how the two metrics have diverged significantly during spatial aggregation and normalisation, complicating the comparison of their respective information.
+At the very least, the developer must clarify to the user how the two metrics have diverged significantly during spatial aggregation and normalisation, complicating the comparison of their respective information.
 
@@
 
-## Aggregation bias in indicator development
+## Aggregation displacement in indicator development {#sec-develop}
 
-As we have seen, aggregation bias causes a divergence in the type of information made available through the spatial aggregation of variables versus indicators.
-In addition, unavoidable aggregation bias (for good and bad) will also affect the indicator values and the inferences that can be drawn from these.
-If not careful, aggregation bias may also cause errors to sneak into the indicator logic.
-A major determinant of the resulting aggregation bias from the spatial aggregation of ecosystem condition indicators (@fig-agg1), is the choice about the order of the different steps in the normalisation and aggregation process, i.e.
+As we have seen, aggregation displacement causes a divergence in the type of information made available through the spatial aggregation of variables versus indicators.
+In addition, unavoidable aggregation displacement (for good and bad) will also affect the indicator values and the inferences that can be drawn from these.
+If not careful, aggregation displacement may also cause errors to sneak into the indicator logic.
+A major determinant of the resulting aggregation displacement from the spatial aggregation of ecosystem condition indicators (@fig-agg1), is the choice about the order of the different steps in the normalisation and aggregation process, i.e.
 the aggregation pathway (@fig-pathways).
@@
 Instead, the proportion of water bodies in each condition class is the aggregated reporting metric.
-Therefore aggregation bias is not an issue.
+Therefore aggregation displacement is not an issue.
 When WFD data are put into use in other contexts, however, aggregation becomes important.\
@@
 These, however, cannot be aggregated spatially in any meaningful way for WFD indicators.
-The issue is the same as for all other variables for which normalised indicators exist, but is perhaps more extreme for WFD indicators due to different reference conditions and levels for different water body types, and strong non-linear transformation functions that inflate the aggreation bias issuse.
+The issue is the same as for all other variables for which normalised indicators exist, but is perhaps more extreme for WFD indicators due to different reference conditions and levels for different water body types, and strong non-linear transformation functions that inflate the aggreation displacement issuse.
 <!-- Can we have a figure of calculation exmaple here? -->
@@
 
-Besides making the indicator workflow itself more transparent and reproducible, we believe this recommendation would help raise the general awareness about aggregation bias in EAs, both among users and developers.
+Besides making the indicator workflow itself more transparent and reproducible, we believe this recommendation would help raise the general awareness about aggregation displacement in EAs, both among users and developers.
 It will make the interpretation of ECAs less prone to misunderstanding and miss-use.

### 2025-02-18 manuscript/manuscript.qmd
wolves

@@
 | Example | Main illustrating point |
-|-------------------------|----------------------------------------------|
+|--------------------------|----------------------------------------------|
 | Billbery | Commensurable spatial scales for variable and reference levels (@sec-match) |
-| Deer | Compensation, and discrepancy between variable and indicator (@sec-discrep) |
+| Wolves | Spatial compensation and discrepancy between variable and indicator (@sec-discrep) |
 | PTI | Compensation and the order of operations (@sec-pathways) |
@@
 
-*Variable*: Number of wolves en each carnivore management area (CMA).\
+*Variable*: Number of wolves.\
 
-*Reference levels:*   X~0~ = zero wolves present.\
-X~100~= Number of wolves equal to what experts think the ecosystem would support under the reference condition.
-The reference levels are unique to each (CMA).\
+*Reference levels:* X~100~= Number of wolves equal to what experts think the ecosystem would support under the reference condition.
+This is number is informed by data on habitat quality, food supply, and more.
+X~0~ = no wolves.
 
-This example is from Norway, where wolves are extinct in most of the country.
-Therefore, most carnivore management areas will get an indicator value of 0.
-Wolves exist in designated wolf zones in south-east Norway, although in small numbers.
-In the Norwegian Nature Index [@jakobsson2020], the variable follows pathway 1 and is normalised at the CMA level.
-When aggregating the indicator to a national value, it is common for the general public to interpret this value as the status for wolves in Norway, and to connect this value to Norway's international obligations to maintain wolf numbers.
-But note that even an unlimited number of wolves inside the wolf zones would not to bring the national indicator value up to a very hight level.
-This is because the normalisation process truncates the variable, and does not allow overshooting values to compensate for lower values in the rest of the country.
-The Norwegian Nature Index uses (approximately) the same pathway for all indicators, so this common misunderstandings about what indicator values represents can be explaind just one time, and not for each indicator specifically.\
+|                      | EA 1 | EA 2 | EA 3 | EAA aggregate |
+|----------------------|------|------|------|---------------|
+| **Variable values**  | 20   | 5    | 0    | Sum = 25      |
+| **X~100~**           | 10   | 10   | 5    | Sum = 25      |
+| **Indicator values** | 1.0  | 0.5  | 0.0  | Mean = 0.5    |
+
+: Dummy data fro the wolves example. EA = ecosystem asset (e.g. a municipality). EAA = ecosystem accounting area (e.g. country). {#tbl-wolf}
+
+This hypothetical example (@tbl-wolf) shows how different the interpretation of a variable and an indicator can become when the information is spatially aggregated.
+When species-based indicators are aggregated to the EAA level, it is common for the general public to interpret this value (0.5) as the status for this species, for example the conservation status[^3].
+However, for the EAA seen overall, the population size is actually equal to the expected population size in the reference condition.
+Instead, the spatially aggregated indicator value describes the ecosystem condition for the different *areas* in terms of wolf under-abundance.
+Then we see that the first ecosystem asset have more wolves than than we would expect under the reference condition, whereas the other two have less.
+Because variable values above X~100~ are truncated, the high wolf numbers in ecosystem asset one does not compensate for the low indicator value in assets two and three, and hence the mean indicator value is less then the sum of the variable values divided by the sum of the X~100~values.
+But the high wolf numbers in asset 1 *does compensate* when looking at the aggregated variable value.
 
@@
 ```
+:::
 
-It can, however, be a good idea to sometimes go back to the variable, and see if it can be defined differently to avoid these kinds of confusions.
+[^3]:
+    ::: callout-note
+    It is sometimes worth going back to the variable, and see if it can be defined differently to avoid this kind of confusions.
     Wolf numbers could in this example be converted to wolf density.
@@
 
-Pathway 2 involves aggregating the variable before normalising[^3].
+Pathway 2 involves aggregating the variable before normalising[^4].
 This preserves the ability for very high or very low variable values to compensate for very low or very high values elsewhere, respectively.
 
-[^3]: In a parallel terminology, normalisation can be considered modeling, and spatial aggregation before and after normalisation can be referred to as pre or post processing, respectively [@vanbeurden1999].
+[^4]: In a parallel terminology, normalisation can be considered modeling, and spatial aggregation before and after normalisation can be referred to as pre or post processing, respectively [@vanbeurden1999].
 

### 2025-02-18 manuscript/manuscript.qmd
PTI start

@@
 
-*Reference levels*: X~100~ was defined for each of 5 regions in Norway based on an expert elicitation.
+*Reference levels*: X~100~ is defined for each ecosystem asset based on an expert elicitation.
 Experts were informed by the distribution of the variable values both within and outside protected areas, and the regional distribution of major forest types.
@@
 
-The bilberry indicator exemplifies different spatial resolutions for the variable and for the reference levels.
+| \-                   | Plot 1 | Plot 2 | EA 1         |
+|----------------------|--------|--------|--------------|
+| **Variable values**  | 2      | 36     | 18           |
+| **X~100~**           | 20\*   | 20\*   | 20           |
+| **Indicator values** | 0.1\*  | 1.0\*  | 0.9 (0.55\*) |
+
+: Dummy data for the bilberry example.
+Variable values (% cover of bilberry) is recorded for each vegetation plot, and aggregated (arithmetic mean) to the ecosystem asset (EA) level.
+The X~100~ value is defined for the EA level, but in this hypothetical case it is also dis-aggregated to each vegetation plot and used for normalising variable values also at the plot level.
+The dis-aggregated X~100~ values, and subsequent values that rely on these, are marked with an asterisk (\*).
+{#tbl-billbery}
+
+The bilberry indicator exemplifies different spatial resolutions for the variable and for the reference levels (@tbl-billbery).
 The variable is recorded at the scale of vegetation plots.
-The reference level however, is designed with a regional spatial scale in mind, especially because of how it encompasses the known variation in forest types in a region to estimate the mean billberry coverage under the reference condition.
-Because the normalisation includes a truncation step, normalisation the variable at the plot scale (pathway 1; @fig-agg1) would not allow overshooting values to compensate for lower values when aggregating regional indicator values, which would then be negatively displaced ($\bar i$) relative to when aggregating variable values ($\hat i$; @fig-agg1).
-This means one is essentially aggregating the negative deviations from the upper reference level X~100~.\
-<!-- Scaling at the plot scale would require unique reference levels for each forest type.
-However, both pathway 1 and 3 has been used for this variable in two different forest ecosystem condition assessments.
-Due to non-negligeable truncation of the variable during the normalisation, the two indicators were subject to different levels of aggregation bias, and consequently the spatially aggregated indicator values for the two assessments became different.
-This caused a general confusion about how how the same variable (and the same data) can produce different indicator values. -->
-
-*Conclusion*: Given that the reference levels are the way they are, this indicator should follow pathway 3 (@fig-agg1), and the variable should be spatially aggregated from plot scale to regional scale before normalising the variable and potentially aggregating it further.
+The reference level however, is designed with a regional spatial scale in mind, especially because of how it encompasses the known variation in forest types in each ecosystem asset to estimate the mean bilberry coverage under the reference condition.
+Because the normalisation includes a truncation step, normalisation the variable at the plot scale (pathway 1; @fig-agg1) does not allow overshooting values in plot 2 to compensate for lower values in plot 1 when spatially aggregating the indicator values.
+When normalising at the plot scale in this case, one is essentially aggregating only the negative deviations from the upper reference level X~100~, and the aggregated indicator value becomes negatively displaced (0.55) relative to when aggregating variable values (0.9).
+Given that X~100~ is defined the way it is, this indicator should follow pathway 3 (@fig-agg1) by spatially aggregating the variable from plot scale to ecosystem asset to allow spatial compensation between plots.
+At the scale of ecosystem assets, the variable and the reference levels are at a commensurable scale, and the variable can be normalised and potentially aggregating it further.
 :::
@@
 It can also be motivated by a desire to know more about the fine-scale variation in variable values, and perhaps visualise this on a map.
-This type of spatial aggregation, or perhaps better named upscaling, is not the focus of this paper (but see **Box 1**), and in our experience this has not been common practice in SEEA EA accounts produced to date.
+This type of spatial aggregation, or perhaps better named upscaling, is not the focus of this paper (but see **Box 1**), and in our experience this has not been common practice in SEEA EA accounts produced to date [but see @kolstad_making_2024].
 
@@
 
-: Dummy data fro the wolves example. EA = ecosystem asset (e.g. a municipality). EAA = ecosystem accounting area (e.g. country). {#tbl-wolf}
+: Dummy data for the wolves example. EA = ecosystem asset (e.g. a municipality). EAA = ecosystem accounting area (e.g. country). {#tbl-wolf}
 
@@
 the aggregation pathway (@fig-pathways).
-If normalisation occurs early in the aggregation process, the amount of compensation, or off-setting, is limited.
-However, sometimes complete or partial compensation is justified.
-In @sec-match we saw how indicator development involved an initial step of spatially aggregating variables to align with the spatial configuration of reference levels (pathways 2 and 3; @fig-pathways).
-In **Example 3** we see that compensation can be warranted, depending on how high the reference levels are set.
+If normalisation occurs early in the aggregation process, the amount of compensation, or off-setting, is limited (**Example 2**).
+However, sometimes complete or partial compensation is justified (**Example 1 and 3**).
 
@@
 *Variable*: Mean score of algal species present, based on a set of indicator species scored for phosphorus requirements/tolerance.
-The variable is recorded in water bodies (lakes of 0,5 km\^2 or more).
+The variable is recorded in water bodies (lakes of 0,5 km^2^ or more).
 
@@
 
-<!-- Pathway: 1 -->
-
-The Water Framework Directive (WFD) indicators uses aggregation pathway 7 [@fig-pathways], with truncation, scaling, and transformation, in that order.
+The Water Framework Directive (WFD) indicators uses aggregation pathway 7 (@fig-pathways), with truncation, scaling, and transformation, in that order.
 The value obtained after the first two steps is called an EQR (ecological quality ratio), and the value obtained after the third step nEQR (normalised EQR).
@@
 When WFD data are put into use in other contexts, however, aggregation becomes important.\
-<!-- Relevant contexts in Norway are the Nature Index (ref) and ecological condition accounting. In both cases,  -->
 
-*Conclusion*: Because of how X~100~ is defined based on the median value across reference lakes, overshooting values (\> X~100~) should be preserved in the spatial aggregation.
-<!-- Otherwise, the condition of water bodies with X > X80 will be underreported.  --> Otherwise we get a negative displacement, making it practically impossible to reach an indicator value of 1.
+Because of how X~100~ is defined based on the median value across reference lakes, overshooting values (\> X~100~) should be preserved in the spatial aggregation.
+Otherwise we get a negative displacement, making it practically impossible to reach an indicator value of 1.
 
-This means that neither EQR values, nor nEQR values, can be uncritically aggregated spatially to be sued in a SEEA EA comlient in ECA.
+This means that neither EQR values, nor nEQR values, can be uncritically aggregated spatially to be used in a SEEA EA compliant in ECA.
 One solution, to enable the use of WFD indicators in ECAs, would be to change to pathway 6, with (1) scaling, (2) transforming, (3) aggregating and (4) truncating the values [@fig-pathways].\
 
-Furthermore, Eurostat [@eurostat2022] <!-- requires  --> encourages reporting of variable values.
+Furthermore, Eurostat [@eurostat2022] encourages reporting of variable values.
 These, however, cannot be aggregated spatially in any meaningful way for WFD indicators.
-The issue is the same as for all other variables for which normalised indicators exist, but is perhaps more extreme for WFD indicators due to different reference conditions and levels for different water body types, and strong non-linear transformation functions that inflate the aggreation displacement issuse.
+The issue is the same as for all other variables for which normalised indicators exist, but is perhaps more extreme for WFD indicators due to different reference conditions and levels for different water body types, and strong non-linear transformation functions that inflate the aggregation displacement issue.
 <!-- Can we have a figure of calculation exmaple here? -->

### 2025-02-19 manuscript/manuscript.qmd
latex table in pti example

@@
 ---
+
+```{r setup, include=FALSE}
+library(tidyverse)
+library(knitr)
+library(kableExtra)
+```
 
@@
 
-::: callout-note
+::: {.callout-note}
 **Box 1 -- Related concepts**\
@@
 horizontal) aggregation.
-\\
+\
 
@@
 All these cases lead to some form of aggregation error[@rastetter1992] where the aggregated information has a non-linear relationship with the true variable value.
-However, the issue of MAUP in SEEA EA deserves an in-depth analyses that we cannot provide here.
-\\
+However, the issue of MAUP in SEEA EA deserves an in-depth analyses that we cannot provide here. \
 
@@
 Generally, because of the normalisation process, indicators are more aligned with the purpose of ECA with describing ecosystem condition in normative terms.
-Although variable accounts are allowed under the SEEA EA framework, ans is indeed a central feature, we believe they should be seen as interim accounts for metrics that cannot or have not yet been normalised.
+Although variable accounts are allowed under the SEEA EA framework, and is indeed a central feature, we believe they should be seen as interim accounts for metrics that cannot or have not yet been normalised.
 We would therefore argue that when an indicator has been developed, the original variable is not presented.
 And when only variable estimates are given, extra care is given to describing the spatial aggregation pathway.
+Furthermore, Eurostat [@eurostat2022] encourages reporting of variable values.
+Sometimes, however, variable values cannot be aggregated spatially in any meaningful way (see **Example 3**).
 
@@
 If normalisation occurs early in the aggregation process, the amount of compensation, or off-setting, is limited (**Example 2**).
-However, sometimes complete or partial compensation is justified (**Example 1 and 3**).
+However, sometimes complete or partial compensation is justified (**Example 1** and **3**).
 
@@
 
-<!-- Dataset: “Vannmiljø”, the Norwegian Environment Agency’s database for Water Framework Directive relevant measurements -->
-
-<!-- Reference: https://github.com/NINAnor/NI_vannf -->
-
 *Variable*: Mean score of algal species present, based on a set of indicator species scored for phosphorus requirements/tolerance.
-The variable is recorded in water bodies (lakes of 0,5 km^2^ or more).
+The variable is recorded in water bodies (lakes of 0.5 km^2^ or more).
 
@@
 
-The Water Framework Directive (WFD) indicators uses aggregation pathway 7 (@fig-pathways), with truncation, scaling, and transformation, in that order.
+```{r read}
+#| include: false
+dat <- readxl::read_xlsx(here::here("data/PTI_data.xlsx")) |>
+  mutate(X = X |> str_replace_all("%", "\\\\%"))
+cn <- colnames(dat) |>
+    str_replace_all("2", "") |>
+    str_replace_all("X", "") |>
+    str_replace_all("RL", "X\\\\textsubscript{100}")
+```
+
+```{r tbl-pti}
+#| echo: false
+#| tbl-cap: "Dummy data for the PTI example.
+value = value for a hypothetical variable.
+agg.
+= spatial aggregation (arithemtic mean); trunc.
+truncation; stat.
+= spatial statistics.
+The transformation step is left out for simplicity.
+Pathways refer to figure 3."
+
+kbl(dat, format = "latex",
+    booktabs = TRUE,
+    col.names = cn,
+    escape = F
+    ) |>
+  kable_styling() %>%
+  add_header_above(c(
+    " " = 3, 
+    "Pathway 5" = 4,
+     "Pathway 6" = 3)) |>
+  collapse_rows(
+    columns = c(9,10),
+    latex_hline = "major"
+  )
+
+```
+
+The Water Framework Directive (WFD) indicators uses aggregation pathway 5 (@fig-pathways), with truncation, scaling, and transformation, in that order (see worked example in @tbl-pti).
 The value obtained after the first two steps is called an EQR (ecological quality ratio), and the value obtained after the third step nEQR (normalised EQR).
@@
 Therefore aggregation displacement is not an issue.
-When WFD data are put into use in other contexts, however, aggregation becomes important.\
+When WFD data are put into use in other contexts, however, aggregation becomes important.
 
+\
 Because of how X~100~ is defined based on the median value across reference lakes, overshooting values (\> X~100~) should be preserved in the spatial aggregation.
 Otherwise we get a negative displacement, making it practically impossible to reach an indicator value of 1.
-
 This means that neither EQR values, nor nEQR values, can be uncritically aggregated spatially to be used in a SEEA EA compliant in ECA.
-One solution, to enable the use of WFD indicators in ECAs, would be to change to pathway 6, with (1) scaling, (2) transforming, (3) aggregating and (4) truncating the values [@fig-pathways].\
-
-Furthermore, Eurostat [@eurostat2022] encourages reporting of variable values.
-These, however, cannot be aggregated spatially in any meaningful way for WFD indicators.
-The issue is the same as for all other variables for which normalised indicators exist, but is perhaps more extreme for WFD indicators due to different reference conditions and levels for different water body types, and strong non-linear transformation functions that inflate the aggregation displacement issue.
-<!-- Can we have a figure of calculation exmaple here? -->
+One solution, to enable the use of WFD indicators in ECAs, would be to change to pathway 6, with (1) scaling, (2) transforming, (3) aggregating and (4) truncating the values.
+In our hypothetical example (@tbl-pti) this gives and aggregated value of 0.85, whereas the mean value for the nEQRs would be 0.75 (i.e. negatively displaced).
 :::

### 2025-02-20 manuscript/manuscript.qmd
clearpage

@@
 execute:
-  cache: true
+  cache: false
   warning: false
@@
 
+\clearpage
+
 ::: callout-note
@@
 
-::: {.callout-note}
+\clearpage
+
+::: callout-note
 **Box 1 -- Related concepts**\
@@
 thematic or vertical aggregation; @langhans2014), but in this paper we are primarily discussing spatial (i.e.
-horizontal) aggregation.
-\
+horizontal) aggregation.\
 
@@
 All these cases lead to some form of aggregation error[@rastetter1992] where the aggregated information has a non-linear relationship with the true variable value.
-However, the issue of MAUP in SEEA EA deserves an in-depth analyses that we cannot provide here. \
+However, the issue of MAUP in SEEA EA deserves an in-depth analyses that we cannot provide here.\
 
@@
 
-```{=html}
-<!-- The indicator arguably has much less room for miss-interpretation (relative to its intended use).
-Although one common misunderstanding is that indicators named by an animal (e.g. wolves; Example 2) is taken by untrained users to represent a population status (typically the number of individuals).
-This would, however, imply that the spatial aggregation function used for the indicator is summation, 
-but the recommended, on arguably the most used function, is an area weighted mean.
-With this aggregation function, indicators will always refer to the mean indicator value for all the areas considered, and if properly defined, refer to the degree of fulfillment for the ecosystem characteristic.
-Sometimes changing the indicator name, or redefining the indicator, goes a long way towards easing the interpretation and avoiding miss-use (e.g. *mean reindeer densities*). -->
-```
-
 Generally, because of the normalisation process, indicators are more aligned with the purpose of ECA with describing ecosystem condition in normative terms.
@@
 #| include: false
+#| cache: false
 dat <- readxl::read_xlsx(here::here("data/PTI_data.xlsx")) |>
@@
 Pathways refer to figure 3."
+#| cache: false
 
@@
 
-We do not presume that ECA end users will have the capacity, nor interest, to reflect very much upon the effect that spatial aggregation has had on the properties of the information they are presented with.
-It should simply work, without the need for elaborate explanations.
+We believe it is a very desirable property of ECAs to be comparable and harmonised across assessment scales, and that accounts from sub-areas should be able to recreate the account of the entire area.
 Cases when two accounts or assessments use what is considered the same indicator but find different values for the same areas, is cause for both confusion, and potentially mistrust.
-This can, however, happen, and has happened, when nested ECA use different aggregation pathways for the same indicator, for example when nested assessments use a different number of aggregation levels.
+This can happen when nested ECA use different aggregation pathways for the same indicator, for example when nested assessments use a different number of aggregation levels.
 
-Consider a case ECA was to be made for a country.
+Consider a case where an ECA was to be made for a country.
 For one of the variables, data sampling was random and dispersed across the EAA, and a national indicator values was calculated as some average across all sampling points.
@@
 A third account, focusing on municipalities (using municipalities as ecosystem assets, rather than the larger regions), will create a third answer to the same question.
-We believe it a very desirable property of ECA to be compliant and harmonised assessment scales.
 In this specific case, one could in the first national account have performed a step-wise aggregation via municipalities, regions, and finally to the entire EAA, but it would not always be possible to know which ecosystem asset are to be used in future accounts.
+
+We do not presume that ECA end users will have the capacity, nor interest, to reflect very much upon the effect that spatial aggregation has had on the properties of the information they are presented with.
+It should simply work, without the need for elaborate explanations.
+
 We believe one solution to this is to acknowledge that the aggregation pathway, including the number of aggregation levels used, can affect the indicators so much, sometimes as much as variation in the variable itself, that the use of different pathways imply that metrics are named and treated uniquely, for example using different common names, but perhaps better still, different version numbers or indicator IDs.
-Fr each unique indicator the precise aggregation pathway should be specified as part of the indicator metadata.
+For each unique indicator the precise aggregation pathway should be specified as part of the indicator metadata.
 This would show ECA users if two accounts are using different metrics that cannot be assumed to behave similarly or report the same thing, even if they have the same common name and data source.
@@
 Besides changing the number of aggregation levels, the choice about when in the aggregation pathway to normalise variables is a possible factor leading to discrepancies across nested accounts (Example 4).
-
-```{=html}
-<!-- The SEEA EA uses a principle of commensurability, meaning that the order of certain operations should not affect the resulting estimate or value.
-Spesifically, it should not matter whether you aggregated vertically first (to create an index) and then horizontally (i.e. spatial), or the other way around.
-Note that this prisiple may not always hold for other cases, such as aggregating across nested areas, e.g. from ecosystem assets to EAAs, although this is perhaps considered desirable.
-This has to do with the use of area weighting in the spatial aggregation. -->
-```
-
-```{=html}
-<!-- 
-Consider an indicator originating from point-based field surveys.
-The sampling location are selected using a stratified protocol that ensures a representative sample of similar size (same *n*) is made for each ecosystem asset.
-If the variable is normalised early in the aggregation pathway to create indicator values for each locality, then the mean of indicator values for each region will be representative for those regions, and no area weighting is required.
-For the next spatial level, when information from all ecosystem assets should be used to create an estimate for the EAA, ecosystem assets with large extents should weight more in the calculations (following the SEEA EA prinsiples of area weighting).
-Note that one cannot skip this first level of spatial aggregation via ecosystem assets, since this is an integral part of the sampling design, leading to spatial autocorrelation.
-However, if in a later ECA, ecosystem assets are defined differently, e.g. as subsets of the initial assets (a finer scale account), 
-then the same indicator, when aggregated to the EAA level, will have a different value. -->
-```
 
@@
 
-<!-- *Pathway*: 1 -->
-
-In this example, the reference levels are uniform throughout the EAA, and the normalisation includes a transformation in the form of a piecewise-linear transformation by anchoring X~60~ to 0.6 on the indicator scale (similar to @fig-agg1 pane B, but without truncation, as that is not needed in the alien species example).
+In this example, the reference levels are uniform throughout the EAA, and the normalisation includes a transformation in the form of a piecewise-linear transformation by anchoring X~60~ to 0.6 on the indicator scale (similar to @fig-agg1 pane B, but without truncation, as that is not needed in this example).
 When performing an ECA, one has the option to either aggregate variable values for each ecosystem asset, and then normalise that value (pathway 2 and 3), or to normalise the variable at the scale of the original nature type polygons (Basal spatial unit (BSU) scale), and then aggregate the indicator values to the regional level (pathway 1).

### 2025-02-20 manuscript/manuscript.qmd
pathways

@@
 
-We believe it is a very desirable property of ECAs to be comparable and harmonised across assessment scales, and that accounts from sub-areas should be able to recreate the account of the entire area.
-Cases when two accounts or assessments use what is considered the same indicator but find different values for the same areas, is cause for both confusion, and potentially mistrust.
-This can happen when nested ECA use different aggregation pathways for the same indicator, for example when nested assessments use a different number of aggregation levels.
+We believe it is a very desirable property that variable and indicator values are comparable across repeated ECAs, and harmonised across assessment scales and so that accounts from sub-areas should be able to recreate the account of the larger area.
+This can happen when ECAs use different aggregation pathways for the same indicator.
+Cases when two accounts or assessments use what is considered the same metric, but find different values for the same areas, is cause for both confusion, and could result in reduced general credibility for ECAs.
 
-Consider a case where an ECA was to be made for a country.
-For one of the variables, data sampling was random and dispersed across the EAA, and a national indicator values was calculated as some average across all sampling points.
-Then in a later regional ECA, the same indicator was estimated for separate regions, as well as for the whole country.
-When aggregating the indicator values from regions to the whole country, using area weighting, the national estimate will differ, albeit slightly, from the initial national estimate that did not include this step-wise aggregation, causing potentially the two nested assessment to reach different conclusions.
-A third account, focusing on municipalities (using municipalities as ecosystem assets, rather than the larger regions), will create a third answer to the same question.
-In this specific case, one could in the first national account have performed a step-wise aggregation via municipalities, regions, and finally to the entire EAA, but it would not always be possible to know which ecosystem asset are to be used in future accounts.
-
-We do not presume that ECA end users will have the capacity, nor interest, to reflect very much upon the effect that spatial aggregation has had on the properties of the information they are presented with.
-It should simply work, without the need for elaborate explanations.
-
-We believe one solution to this is to acknowledge that the aggregation pathway, including the number of aggregation levels used, can affect the indicators so much, sometimes as much as variation in the variable itself, that the use of different pathways imply that metrics are named and treated uniquely, for example using different common names, but perhaps better still, different version numbers or indicator IDs.
+We believe one solution to this is to acknowledge that the aggregation pathway can affect the indicators so much, sometimes as much as variation in the variable itself, that the use of different pathways imply that metrics are named and treated uniquely, for example using different common names, but perhaps better still, different version numbers or indicator IDs.
 For each unique indicator the precise aggregation pathway should be specified as part of the indicator metadata.
@@
 
-Besides changing the number of aggregation levels, the choice about when in the aggregation pathway to normalise variables is a possible factor leading to discrepancies across nested accounts (Example 4).
+SEEA EA uses a the term aggregation commutativity for the property that the order of vertical and horizontal aggregation is irrelevant to the end result.
+This is indeed a desirable property, however, when indicators that make up an index have different extents (e.g.
+they relate to different ecosystem sub-types or have sparse data coverage), then the indicator-specific areal weights will not be transferable in the spatial aggregation of indices (see description of pathways 7 and 8 in @sec-pathways).
+A similar commutativity can be envisioned for step-wise aggregation of a single indicator, where it does not matter if the original data is aggregated directly from the original measurement scale to the EAA level, or via some intermittent scale, such as ecosystem assets or larger regions.
+This is achievable when aggregating indicators using area weighting, and when one is aware of the need for post-stratification when using stratified data sampling.
 
-```{=html}
-<!-- In our experience this can also lead to time consuming detective work, trying to find the causes for such discrepancies.
-We suggest two immediate solution to this issue.
-First, to find better ways to document the aggregation pathways that are being used for each indicator, as part of the indicator metadata.
-Second, to find ways to distinguish between apparently identical indicators, for example using different version numbers or IDs.
-We provide practical solution to both these suggestions in @sec-rec. -->
-```
+<!--# We do not presume that ECA end users will have the capacity, nor interest, to reflect very much upon the effect that spatial aggregation has had on the properties of the information they are presented with.
+It should simply work, without the need for elaborate explanations.
+-->
 
@@
 
-We introduce the term aggregation pathway to describe the order of the steps used to go from a spatially explicit variable to a spatially aggregated metric, usually an indicator.
+We introduce the term aggregation pathway to describe the order of the steps used to go from a variable at the measurement scale, to a spatially aggregated metric, usually an indicator, that describes the ecosystem condition for a larger area.
 In @fig-pathways we show some aggregation pathways that we have come across, and some that we we see as potential new pathways.
+The examples are non-exhaustive, and serves mainly to illustrate that many different pathways can be created using a limited number of unique operations (@sec-rec).
+We then describe how the aggregation pathways can formally, and succinctly, be described using the names of these operations.
 The choice about which pathway to use is not trivial, and we want to highlight some of the issues that could arise from having an *ad hoc* approach to these considerations.
@@
 
-Pathway 3 is similar, but here there are two aggregation steps: one before and one after normalisation.
-This can for example be the case when variables are aggregated to the scale corresponding to the reference levels before they can be normalised (Example 1).
+Pathway 3 is similar as 2, but here there are two aggregation steps: one before and one after normalisation.
+This can for example be the case when variables are aggregated to the scale corresponding to the reference levels before they can be normalised (**Example 1**).
 
@@
 Instead one report the number of water bodies in each condition class (spatial statistics).
-Pathway 6 is a suggested pathway for handling WFD indicator in ECAs (see Example 3), where truncation is postponed until after the spatial aggregation for in that way to allow overshooting values to compensate for low indicator values elsewhere (what we refer to as compensation).
+Pathway 6 is a suggested pathway for handling WFD indicator in ECAs (see **Example 3**), where truncation is postponed until after the spatial aggregation for in that way to allow overshooting values to compensate for low indicator values elsewhere (what we refer to as compensation).
 
@@
 In pathway 7, indicators are thematically aggregated to an index, which is then again aggregated spatially.
+The order of the two aggregation steps can be reversed if the process is truly commutative (@sec-comp).
 In pathway 8, spatial aggregation is only done on individual indicators.
@@
 
-Normalising variables early in the aggregation pathway has several benefits, such as comparability between nested assessments (Example 4, @sec-comp).
-It also means that you are aggregating normative measures of condition, which is generally what you want in ECAs (@sec-norm).
+Normalising variables early in the aggregation pathway means that you are aggregating normative measures of condition, which is generally what you want in ECAs (@sec-norm).
 But same as for variables and indicators, reference levels also have a spatial resolution, and they should only be used to normalise variables when the variable is at a scale which is relevant to the way the reference levels are defined, i.e.
-they are at a commensurable scale (Example 1).
+they are at a commensurable scale (**Example 1**).
 

### 2025-02-20 manuscript/manuscript.qmd
rm ex 4

@@
 | PTI | Compensation and the order of operations (@sec-pathways) |
-| Alien species | Agreements across scales (@sec-comp) |
 
@@
 
-<!--# We do not presume that ECA end users will have the capacity, nor interest, to reflect very much upon the effect that spatial aggregation has had on the properties of the information they are presented with.
-It should simply work, without the need for elaborate explanations.
--->
-
-::: callout-note
-**Example 4 -- Alien species**\
-
-*Variable*: The local ecological effect from alien species (mostly vascular plants) recorded on a 7-step ordinal scale (1 = no alien species, 7 = only alien species).
-The data comes from nature type monitoring with reference to individual occurrences (polygons).\
-
-*Reference levels*: X~0~ = total dominance from alien species (variable value 7).
-X~100~ = zero influence from alien species (variable value 1).
-X~60~ = \~2-5% alien plant cover (variable value 3).\
-
-In this example, the reference levels are uniform throughout the EAA, and the normalisation includes a transformation in the form of a piecewise-linear transformation by anchoring X~60~ to 0.6 on the indicator scale (similar to @fig-agg1 pane B, but without truncation, as that is not needed in this example).
-When performing an ECA, one has the option to either aggregate variable values for each ecosystem asset, and then normalise that value (pathway 2 and 3), or to normalise the variable at the scale of the original nature type polygons (Basal spatial unit (BSU) scale), and then aggregate the indicator values to the regional level (pathway 1).
-Arguably the variable and the reference levels are at commensurable scales in both cases (cf. Example 1), and other arguments for choosing one over the other must be found.\
-
-One benefit of early normalisation is that two nested assessments, such as one national, and one assessment of region *Y* inside the same country, will get the same spatially aggregated value for region *Y*.
-Separate assessments should give identical indicator values for the same areas, given the same underlying data.
-When this is not true it raises concerns about reproducibility and sacrifices the credibility of both assessments.
-The other thing to ask oneself here is: *Why do we normalise* in the first place (@sec-norm)?
-One reason is that we want to summarise information in space about the ecosystem condition.
-It is the indicator values that hold this information, and not the variable.
-We would therefore favor pathway 1 for this example.
-:::
-
 # Aggregation pathways {#sec-pathways}

### 2025-02-20 manuscript/manuscript.qmd
rm pathways 9 and 10

@@
 
-Pathway 9 describe how a single indicators can be aggregated hierarchically from region to region, whereas pathway 10 shows a case were the aggregation process always starts from scratch for each hierarchical spatial level.
-As for pathways 7 and 8, pathway 9 and 10 differ in how they propagate the indicator weight (typically area weight) between regions.
-SEEA EA describes and prescribes a hierarchical spatial aggregation (pathway 9).
-Note that can cause issues with comparability across nested assessment, if one alters the number or size of sub-regions (@sec-comp).
-
 ```{r fig-pathways}
@@
 #| echo: false
-knitr::include_graphics("../img/aggregationPathways.jpg")
+knitr::include_graphics("../img/aggregationPathways.png")
 ```

### 2025-02-20 manuscript/manuscript.qmd
first draft with pdf

@@
     email: balint.czucz@nina.no
+    affiliations:
+      - ref: nina
+  - name: Chloé R. Nater
+    email: chloe.nater@nina.no
     affiliations:
@@
 This paper will make it easier for indicator developers to avoid some of the pitfalls from having a too casual approach to the choice of aggregation pathways.
-We use four examples to illustrate some of our main points (@tbl-examples).
+We use three examples to illustrate some of our main points (@tbl-examples).
 In @sec-pathways we describe a new reporting terminology for spatial aggregation pathways.
@@
 |----------------------|--------|--------|--------------|
-| **Variable values**  | 2      | 36     | 18           |
-| **X~100~**           | 20\*   | 20\*   | 20           |
-| **Indicator values** | 0.1\*  | 1.0\*  | 0.9 (0.55\*) |
+| Variable values  | 2      | 36     | 18           |
+| X~100~           | 20\*   | 20\*   | 20           |
+| Indicator values | 0.1\*  | 1.0\*  | 0.9 (0.55\*) |
 
@@
 |----------------------|------|------|------|---------------|
-| **Variable values**  | 20   | 5    | 0    | Sum = 25      |
-| **X~100~**           | 10   | 10   | 5    | Sum = 25      |
-| **Indicator values** | 1.0  | 0.5  | 0.0  | Mean = 0.5    |
+| Variable values  | 20   | 5    | 0    | Sum = 25      |
+| X~100~           | 10   | 10   | 5    | Sum = 25      |
+| Indicator values | 1.0  | 0.5  | 0.0  | Mean = 0.5    |
 
@@
 stat.*).
-These terms can be abbreviated as shown in the parentheses, and placed togeth into a single string, where operations performed simultaneously are enclosed in square brackets.
-For example, pathway 1 in @fig-pathways can be annotated as *scale - trunc.
+These terms can be abbreviated as shown in the parentheses, and placed together into a single string, where operations performed simultaneously are enclosed in square brackets.
+For example, pathway 1 in @fig-pathways can be annotated as $scale - trunc.
 - trans.
 - sp.
-agg..* If the three normalisation steps are performed simultaneously this can be written as \[*scale - trunc.
-- trans.\] - sp.
-agg.* If there are more aggregation steps, then the spatial level at each step should be included in the description, e.g.
-for pathway 3: *sp.
+agg.$.
+If the three normalisation steps are performed simultaneously this can be written as $[scale - trunc.
+- trans.] - sp.
+agg.$.
+If there are more aggregation steps, then the spatial level at each step should be included in the description, e.g.
+for pathway 3: $sp.
 agg.
@@
 agg.
-(country).
+(country)$.
 "*Municipality" and "country" could here be included as footnotes.
-Note also that in this example, normalisation only involved a scaling step.
-The description of indicator pathways can perhaps made even easier by referencing specific aggregation pathways in @fig-pathways by number, as we have done in this paper.
+Note also that in the last example, normalisation only involved a scaling step.
+The description of indicator pathways can be made even easier by referencing specific aggregation pathways in by number, as we have done in this paper using @fig-pathways in this paper.
 
 Besides making the indicator workflow itself more transparent and reproducible, we believe this recommendation would help raise the general awareness about aggregation displacement in EAs, both among users and developers.
-It will make the interpretation of ECAs less prone to misunderstanding and miss-use.
-It will also make troubleshooting easier, for example when trying to identify the reason why two ECAs produces dissimilar indicator values from the same underlying data.
+It will make the interpretation of ECAs less prone to misunderstanding and miss-use, as well as making troubleshooting easier, for example when trying to identify the reason why two ECAs produces dissimilar indicator values from the same underlying data.
 
@@
 
-It may be premature, or not even possible, to prescribe an aggregation pathway to be used for all ECAs, or a type of aggregation pathway to be used given the data at hand.
+It may be premature, or not even possible, to prescribe an aggregation pathway to be used for all ECAs, or even a type of aggregation pathway to be used given the data at hand.
 However, internal consistency within assessments should be possible, and this can make it easier to communicate and interpret the information conveyed though the indicator values.
-This can be achieved by for example always scaling early (see recommendation number 2 above) and using the same aggregation levels and same approach to step-wise aggregation across levels (@sec-comp), for all indicators.
+This can be achieved by for example always scaling early (see recommendation number 2 above) and using the same aggregation levels (same spatial units) and same approach to step-wise aggregation across levels (@sec-comp), for all indicators.
 
@@
 
-Indicators (and variables) often exist in multiple versions, varying slightly in the raw data or in the methods used to produce the data or metrics.
+Indicators (and variables) often exist in multiple versions, varying slightly in the raw data or in the methods used to produce the data or the metrics.
 Yet different versions are often refereed to by the same common name.
 This causes confusion about which indicator version is being used, and thus making it difficult to make out which aggregation pathway that has been used.
-We recommend making use of stable and unique indicator IDs, also and especially for indicators that use the same common name.
-This has already been implemented in ecRxiv, a GitHub-based publishing platform for ecosystem condition indicators [@Kolstad_ecRxiv].
+We recommend making use of stable and unique indicator IDs.
+This has for example already been implemented in ecRxiv, a GitHub-based publishing platform for ecosystem condition indicators [@Kolstad_ecRxiv].
 On ecRxiv, indicator documentation follow a structured approach to both reporting and peer review (including code review).
-Each indicator version is given its own ID, and its own fact sheet with a persistent URL, so that ECA that report these indicators unequivocally can cite specific indicator versions, and how they looked at a given point in time.
-ecRxiv also includes standardised metadata reporting with fields, including ecosystem type (IUCN GET) and aggregation pathways (see recommendation number 1 above).
+Each indicator version is given its own ID, and its own fact sheet with a persistent URL, so that ECAs that use these indicators can unequivocally cite specific indicator versions, and how they looked at a given point in time.
+ecRxiv also includes standardised metadata reporting schemes with fields for  ecosystem type (IUCN GET) and aggregation pathways (see recommendation number 1 above).
+
+# Conclusion
+
+...
 

### 2025-02-21 manuscript/manuscript.qmd
trackdown edits

@@
 
-[^1]: We use the SEEA EA definitions of the terms variables, indicators (normalised variables) and metrics (refering to either a variable or a metric) [@unitednations2024]
+[^1]: We use the SEEA EA definitions of the terms variables, indicators (normalised variables) and metrics (referring to either a variable or a metric) [@unitednations2024]
 
@@
 @allain2018], there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting [but see @vanbeurden1999].
-This paper therefore focus on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) [@unitednations2024] - the most used ecosystem accounting framework in the world [@comte2022].
+This paper therefore focuses on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) [@unitednations2024] - the most used ecosystem accounting framework in the world [@comte2022].
 
@@
 Indicator developer subjectivity affects indicator interpretation [@allain2018] and finally the conclusion from ECAs.
-Still, the process of designing indicators and estimating indicator values often left entirely to the them, with little real involvement from indicator users.
-It is therefore important to be aware of the role that indicator developers have in shaping the type of information, and the narratives, that is presented to decision makers.
+Still, the process of designing indicators and estimating indicator values is often left entirely to them, with little real involvement from indicator users.
+It is therefore important to be aware of the role that indicator developers have in shaping the type of information, and the narratives, that are presented to decision makers.
 This paper will make it easier for indicator developers to avoid some of the pitfalls from having a too casual approach to the choice of aggregation pathways.
@@
 There is a jungle of terms used for similar metrics or operations across relevant theoretical frameworks for ecosystem assessments, such as the SEEA EA, WFD, MSFD, IBECA, OSPAR, PAEC, HD, CFP, and others.
-We see a need to a common, more standardised terminology across these fields of practice in order to ease communication and avoid misunderstandings.
+We see a need for a common, more standardised terminology across these fields of practice in order to ease communication and avoid misunderstandings.
 For now we will now define the terms and how we use them in this paper.
@@
 They are typically either in percentages or fractions, or as raw biophysical units, such as biomass or density.
-Indicators on the other hand, are always on a unitless scale between 0 and 1, where 0 represent very poor condition, and 1 represents the very good, or the best possible condition.
+Indicators on the other hand, are always on a unitless scale between 0 and 1, where 0 represents very poor condition, and 1 represents the very good, or the best possible condition.
 Sometimes indicators are calculated directly [e.g. @åström], but usually indicators are made by normalising variables based on two or more reference levels [@unitednations2024 §5.60].
@@
 Variables can be scaled using a minimum of two reference levels, a lower and an upper, that defines which variable values to be coded as zero (X~0~) and as one (X~100~) on the indicator scale.
-Variables can, if needed, truncated to produce a bound indicator scale between 0 and 1.
+Variables can, if needed, be truncated to produce a bound indicator scale between 0 and 1.
 Sometimes variables are transformed to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
@@
 
-Spatiall aggregation of data refers to the conversion of fine-resolution data into coarser-resolution data, and depending on both the purpose and the nature of the metric, there can be several reasons both for performing and for refraining from spatial aggregation.
+Spatial aggregation of data refers to the conversion of fine-resolution data into coarser-resolution data, and depending on both the purpose and the nature of the metric, there can be several reasons both for performing and for refraining from spatial aggregation.
 Readers are referred to @allain2018 for a more comprehensive review of different reasons for spatially aggregating variables.
@@
 
-Sometimes total aggregation of a metric is required for the it to actually fit its purpose and become relevant to what it is designed to describe.
+Sometimes total aggregation of a metric is required for it to actually fit its purpose and become relevant to what it is designed to describe.
 For example, a hypothetical variable "*proportion of total land area above critical nitrogen load",* informed from a spatial map of nitrogen loads, would only make sense if data was aggregated across some defined area.
@@
 Other common arguments for performing spatial aggregation are to ease visualisation and to allow relevant comparisons [@allain2018].
-For example, ecosystem condition metrics can be visualised on a map, but when both the the spatial resolution and spatial variation is high, maps become too complex, even unintelligible.
+For example, ecosystem condition metrics can be visualised on a map, but when both the spatial resolution and spatial variation is high, maps become too complex, even unintelligible.
 It is therefore common to partially aggregate the data and provide a coarser map where different colours, representing values on the map, become distinguishable from each other.
-Similarly, one may chose to aggregate such that the spatial resolution matches some meaningful delineation, notably ecosystem assets in SEEA EA, which makes it possible to easily compare the ecosystem condition between these areas.
+Similarly, one may choose to aggregate such that the spatial resolution matches some meaningful delineation, notably ecosystem assets in SEEA EA, which makes it possible to easily compare the ecosystem condition between these areas.
 
@@
 |--------------------------|----------------------------------------------|
-| Billbery | Commensurable spatial scales for variable and reference levels (@sec-match) |
+| Bilberry | Commensurable spatial scales for variable and reference levels (@sec-match) |
 | Wolves | Spatial compensation and discrepancy between variable and indicator (@sec-discrep) |
@@
 To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some generalisations.
-One can assign values to areas via an aggregation functions used on data values that fall within a given ecosystem asset.
+One can assign values to areas via an aggregation function used on data values that fall within a given ecosystem asset.
 Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54] .
@@
 Interpolation, or other types of modeling, can also be used to impute or predict data values between the points where we have information [@finke2002; @stein2001].
-This approach can also be seen as a type of post-stratifiction, and it can result in better variable or indicator estimates for the ecosystem assets as a whole.
+This approach can also be seen as a type of post-stratification, and it can result in better variable or indicator estimates for the ecosystem assets as a whole.
 It can also be motivated by a desire to know more about the fine-scale variation in variable values, and perhaps visualise this on a map.
@@
 
-The term spatial aggregation is also sometimes used to refer to **upscaling** in the sense of taking local information and making it applicable or translate it to larger, e.g.
+The term spatial aggregation is also sometimes used to refer to **upscaling** in the sense of taking local information and making it applicable or translating it to larger, e.g.
 landscape or regional, scales [@knotters2010; @finke2002].
@@
 The original spatial units that data are collected at may be a pragmatic choice in much the same way.
-Further still, data may not be available at the original spatial scale, but may only be downloadable in preaggregated form [@vanbeurden1999].
+Further still, data may not be available at the original spatial scale, but may only be downloadable in pre aggregated form [@vanbeurden1999].
 All these cases lead to some form of aggregation error[@rastetter1992] where the aggregated information has a non-linear relationship with the true variable value.
@@
 #| out-width: 600
-
 knitr::include_graphics("../img/aggregation_bias.PNG")
-
 ```
@@
 @rastetter1992 referred to this as aggregation error and used it to describe some of the dangers associated with upscaling in the sense of using precise fine-scale information to explain or predict large scale phenomena.
-However, in our case the difference between variables and indicators is an intentional one, and we therefore to refer to what is essentially the same mathematical effect, with the less derogatory term aggregation displacement.
+However, in our case the difference between variables and indicators is an intentional one, and we therefore refer to what is essentially the same mathematical effect, with the less derogatory term aggregation displacement.
 
@@
 The user will have no information about the carrying capacity for each of the regions, and will only see the spatially aggregated value.
-From this, many different conclusion can be made, depending on what the user consider high or low densities.
+From this, many different conclusions can be made, depending on what the user considers high or low densities.
 In any case, the judgment will not reflect a density that actually exists anywhere.
-The indicator on the other hand, can have different reference levels for each region, introducing additional ecological information about the carrying capacity of the ecosystems.
+The indicator, on the other hand, can have different reference levels for each region, introducing additional ecological information about the carrying capacity of the ecosystems.
 The levels may be such that the high density region is considered as having an overabundance of herbivores, and the low density region as having underabundance.
@@
 Instead, the spatially aggregated indicator value describes the ecosystem condition for the different *areas* in terms of wolf under-abundance.
-Then we see that the first ecosystem asset have more wolves than than we would expect under the reference condition, whereas the other two have less.
+Then we see that the first ecosystem asset has more wolves than we would expect under the reference condition, whereas the other two have less.
 Because variable values above X~100~ are truncated, the high wolf numbers in ecosystem asset one does not compensate for the low indicator value in assets two and three, and hence the mean indicator value is less then the sum of the variable values divided by the sum of the X~100~values.
@@
     ::: callout-note
-    It is sometimes worth going back to the variable, and see if it can be defined differently to avoid this kind of confusions.
+    It is sometimes worth going back to the variable, and see if it can be defined differently to avoid this kind of confusion.
     Wolf numbers could in this example be converted to wolf density.
@@
   )
-
 ```
@@
 Therefore aggregation displacement is not an issue.
-When WFD data are put into use in other contexts, however, aggregation becomes important.
+When WFD data is put into use in other contexts, however, aggregation becomes important.
 
@@
 One solution, to enable the use of WFD indicators in ECAs, would be to change to pathway 6, with (1) scaling, (2) transforming, (3) aggregating and (4) truncating the values.
-In our hypothetical example (@tbl-pti) this gives and aggregated value of 0.85, whereas the mean value for the nEQRs would be 0.75 (i.e. negatively displaced).
+In our hypothetical example (@tbl-pti) this gives an aggregated value of 0.85, whereas the mean value for the nEQRs would be 0.75 (i.e. negatively displaced).
 :::
 
-# Aggreements across nested ECAs {#sec-comp}
+# Aggregements across nested ECAs {#sec-comp}
 
@@
 
-SEEA EA uses a the term aggregation commutativity for the property that the order of vertical and horizontal aggregation is irrelevant to the end result.
+SEEA EA uses the term aggregation commutativity for the property that the order of vertical and horizontal aggregation is irrelevant to the end result.
 This is indeed a desirable property, however, when indicators that make up an index have different extents (e.g.
@@
 We introduce the term aggregation pathway to describe the order of the steps used to go from a variable at the measurement scale, to a spatially aggregated metric, usually an indicator, that describes the ecosystem condition for a larger area.
-In @fig-pathways we show some aggregation pathways that we have come across, and some that we we see as potential new pathways.
-The examples are non-exhaustive, and serves mainly to illustrate that many different pathways can be created using a limited number of unique operations (@sec-rec).
+In @fig-pathways we show some aggregation pathways that we have come across, and some that we see as potential new pathways.
+The examples are non-exhaustive, and serve mainly to illustrate that many different pathways can be created using a limited number of unique operations (@sec-rec).
 We then describe how the aggregation pathways can formally, and succinctly, be described using the names of these operations.
@@
 
-Pathway 1 involves early normalisation using the perhaps most common order of the three steps scaling, truncating and transforming.
+Pathway 1 involves early normalisation using the perhaps most common order of the three steps: scaling, truncating and transforming.
 The three steps may be done simultaneously or in sequence.
-Finanlly, the indicator is spatially aggregated.
+Finally, the indicator is spatially aggregated.
 
@@
 
-Pathway 3 is similar as 2, but here there are two aggregation steps: one before and one after normalisation.
+Pathway 3 is similar to 2, but here there are two aggregation steps: one before and one after normalisation.
 This can for example be the case when variables are aggregated to the scale corresponding to the reference levels before they can be normalised (**Example 1**).
@@
 Pathway 5 is a common pathway in the WFD, which does not include spatial aggregation of indicator values.
-Instead one report the number of water bodies in each condition class (spatial statistics).
+Instead one reports the number of water bodies in each condition class (spatial statistics).
 Pathway 6 is a suggested pathway for handling WFD indicator in ECAs (see **Example 3**), where truncation is postponed until after the spatial aggregation for in that way to allow overshooting values to compensate for low indicator values elsewhere (what we refer to as compensation).
@@
 The spatial aggregation of indices (pathway 7), using area weighted means, does not allow different area weights to be applied to each underlying indicator.
-Yet this would in many cases be needed if the individual indicators do not all represents the entire ecosystem inside each ecosystem asset.
+Yet this would in many cases be needed if the individual indicators do not all represent the entire ecosystem inside each ecosystem asset.
 This is often the case, since ECA are frequently geared towards high level ecosystem types (e.g.
@@
 Note also that in the last example, normalisation only involved a scaling step.
-The description of indicator pathways can be made even easier by referencing specific aggregation pathways in by number, as we have done in this paper using @fig-pathways in this paper.
+The description of indicator pathways can be made even easier by referencing specific aggregation pathways by number, as we have done in this paper using @fig-pathways in this paper.
 
 Besides making the indicator workflow itself more transparent and reproducible, we believe this recommendation would help raise the general awareness about aggregation displacement in EAs, both among users and developers.
-It will make the interpretation of ECAs less prone to misunderstanding and miss-use, as well as making troubleshooting easier, for example when trying to identify the reason why two ECAs produces dissimilar indicator values from the same underlying data.
+It will make the interpretation of ECAs less prone to misunderstanding and miss-use, as well as making troubleshooting easier, for example when trying to identify the reason why two ECAs produce dissimilar indicator values from the same underlying data.
 
@@
 It may be premature, or not even possible, to prescribe an aggregation pathway to be used for all ECAs, or even a type of aggregation pathway to be used given the data at hand.
-However, internal consistency within assessments should be possible, and this can make it easier to communicate and interpret the information conveyed though the indicator values.
+However, internal consistency within assessments should be possible, and this can make it easier to communicate and interpret the information conveyed through the indicator values.
 This can be achieved by for example always scaling early (see recommendation number 2 above) and using the same aggregation levels (same spatial units) and same approach to step-wise aggregation across levels (@sec-comp), for all indicators.
@@
 Indicators (and variables) often exist in multiple versions, varying slightly in the raw data or in the methods used to produce the data or the metrics.
-Yet different versions are often refereed to by the same common name.
+Yet different versions are often referred to by the same common name.
 This causes confusion about which indicator version is being used, and thus making it difficult to make out which aggregation pathway that has been used.
@@
 This has for example already been implemented in ecRxiv, a GitHub-based publishing platform for ecosystem condition indicators [@Kolstad_ecRxiv].
-On ecRxiv, indicator documentation follow a structured approach to both reporting and peer review (including code review).
+On ecRxiv, indicator documentation follows a structured approach to both reporting and peer review (including code review).
 Each indicator version is given its own ID, and its own fact sheet with a persistent URL, so that ECAs that use these indicators can unequivocally cite specific indicator versions, and how they looked at a given point in time.

### 2025-02-21 manuscript/manuscript.qmd
rm comments

@@
 The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.
-<!--# thus invoking the issue of the ecological fallacy (Box 1), and many other potential pitfalls in the spatial aggregation of information. -->
 
@@
 
-<!--# Let's illustrate this with an example using a hypothetical variable depicting wild animal densities.
-If the upper reference level (X100) for this variable is given by the assumed (or measured) animal densities under the reference condition, then that reference level needs to be assigned a relevant spatial scale, for example a region where the densities are calculated over.
-One will probably acknowledge the fact that animals move and distribute themselves non-randomly within these regions, and it therefore makes little sense to normalise the variable at a finer scale than region.
-Doing so would mean that regions perfectly aligned with the reference condition (in terms of animal densities with the region) will show as having sub-regions with sub-optimal condition, simply due to natural variation in where the animals are at any give time.
-Averaging (arguably the most common spatial aggregation function) the indicator values across sub-regions would not produce a regional indicator value that shows the region as being perfectly aligned with the reference condition, since truncation of variable values above X100 (assuming X100 > X0) means we are only aggregating the negative deviations from the reference level, and that very high animal densities in some sub-regions are not allowed to compensate for low values elsewhere.
--->
-
 \clearpage
@@
 
-<!--# Subsequently, spatial aggregation is required when going from ecosystem assets to large spatial scale, ultimately the scale of the entire ecosystem accounting area (EAA).  -->
-
 Complete, or wall-to-wall data, include remote sensing products, but also other types of pre-aggregated data, where sparse data are assigned to areal units.
@@
 In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
-However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; Box 1).
+However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; **Box 1**).
 It is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
@@
 
-```{=html}
-<!-- \*Conclusion\*: Indicators should make intuitive sense to the readers.
-However, there are sometimes good reasons for making things complicated. -->
-```
 :::

### 2025-03-05 manuscript/manuscript.qmd
download from docs

@@
 
+We follow the definition in Bierkens et al 2002 and treat spatial aggregation as the process of increasing the support (i.e.
+reducing the grain).
+We see this as a separate process to models that create spatially continuous data from point data by either increasing the extent (extrapolation) or cover (interpolation).
+However, these operations are in practice often intertwined with spatial aggregation.
+Either explicitly, for example by using model based spatial estimation methods (e.g.
+geostatistics) and subsequent spatial sampling of this field of random values (Brus et al 1997).
+Inter- or extrapolation can also be combined with spatial aggregation more inadvertently, for example by averaging point data and attributing this value to entire polygons, thus relying on a design-based estimation process (Brus 1997, Gruijter et al 1990).
+The term upscaling is sometimes used synonymously to our definition of spatial aggregation, but it has also a wider use, related especially to the ambiguous use of the terms scale to mean either grain or extent.
+
 SEEA EA distinguishes between indicators and variables, as do we.
@@
 Spatial aggregating of ecosystem condition metrics is applied to alleviate issues of spatial auto-correlation.
-For example if data is collected opportunistically with sampling sizes varying between areas, then performing partial spatial aggregation to some intermediate sized areas will give equal weight or importance to all areas regardless of how many data points are included in each.
+For example if data is collected opportunistically with sampling sizes varying between areas, then performing partial spatial aggregation to some intermediate sized areas will give equal weight or importance to all areas regardless of how many data points are included in each (Brus 1997 again here?).
 This approach represents a form of post-stratification that is also relevant to process data from hierarchical (e.g. blocked) sampling designs.
@@
 ::: callout-note
-**Example 1 -- Billberry**\
+**Example 1 -- Bilberry**\
 
@@
 The reference level however, is designed with a regional spatial scale in mind, especially because of how it encompasses the known variation in forest types in each ecosystem asset to estimate the mean bilberry coverage under the reference condition.
-Because the normalisation includes a truncation step, normalisation the variable at the plot scale (pathway 1; @fig-agg1) does not allow overshooting values in plot 2 to compensate for lower values in plot 1 when spatially aggregating the indicator values.
+Because the normalisation includes a truncation step, normalising the variable at the plot scale (pathway 1; @fig-agg1) does not allow overshooting values in plot 2 to compensate for lower values in plot 1 when spatially aggregating the indicator values.
 When normalising at the plot scale in this case, one is essentially aggregating only the negative deviations from the upper reference level X~100~, and the aggregated indicator value becomes negatively displaced (0.55) relative to when aggregating variable values (0.9).
@@
 The type of manipulation depends on the spatial resolution of the underlying data, as well as the size and configuration of the ecosystem assets.
-To elaborate on this issue we first make a distinction between two main types of data: sparse (i.e. spatially incomplete) data and complete data.
+To elaborate on this issue we first make a distinction between two main types of data: sparse (i.e. spatially discontinuous) data and spatially exhaustive data.
 
-Complete, or wall-to-wall data, include remote sensing products, but also other types of pre-aggregated data, where sparse data are assigned to areal units.
+Spatially exhaustive, or wall-to-wall data, include remote sensing products, but also other types of pre-aggregated data, where sparse data are assigned to areal units.
 This last type of data is very common.
-A lot of demographic information for example is made available at the scale of administrative units, rather than at the individual level (which would be very ineffective and problematic also in terms of protecting sensitive information).
-Similar pre-aggregation is, however, ambiguous and true for most complete datasets.
+A lot of demographic information for example is made available at the scale of administrative units, rather than at the level of individuals (which would be very ineffective and also problematic in terms of protecting sensitive information).
+Similar pre-aggregation is, however, ubiquitous and true for most complete datasets.
 Consider as an example data from LiDAR (Light Detection And Ranging).
@@
 For sparse data we are considering data where variable values are assigned to singular points or small areas.
-In these cases, there will be a lot of area inside each ecosystem asset for which we have no direct information.
-To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some generalisations.
-One can assign values to areas via an aggregation function used on data values that fall within a given ecosystem asset.
-Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54] .
-When doing this aggregation, we need to account for the spatial structure (spatial correlations) of the data.
-One way is to rely on design-based approaches (e.g.
-relying on random or balanced samples across the entire EEA), or using a step-wise aggregation via ecosystem assets or some smaller basal spatial units.
+In these cases, there will be a lot of area inside each ecosystem asset for which we have no direct information, and some ecosystem assets may have no data at all.
+To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some statistical inference.
+Both design-based and model-based approaches (Brus 2021) are valid under SEEA EA.
+However, transparency, validity and simplicity are all very desirable properties from ECAs, which favours a design-based approach that makes no modeling assumptions (Brus 2021).
+We will therefore consider design-based approaches here.
+Model-based approaches also generally means that the variables are simpler to aggregate spatially since the spatial structure of the variables will already be accounted for via the prediction process.
+For example, if using krieging to create local predictions for the variable across the ecosystem asset, then the variable is considered random, and the sampling locations may be considered as fixed.
+Thus a simple systematic sampling of the field of values across the ecosystem asset will be valid, and give representative grounds for statistical inference (Brus 2021).
+
+For design-based approaches, one can assign values to areas via simple aggregation functions used on data values that fall within a given ecosystem asset.
+Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54].
+If the inclusion probabilities are known, or they can be estimated, then the Horvitz-Thomson estimator for the sum or the mean (and variance) of a spatial sample is appropriate and useful, especially when the data originated from a stratified sample.
+For balanced samples, or samples from a simple random sampling process, we can apply these functions directly, but depending on the random process used to select the sampling points we may need to account for the spatial structure (spatial autocorrelation) of the data as well, when doing this aggregation.
+This can be addressed in several ways, such as model-assisted approaches (Brus 2021), e.g.
+using regression models (Brus 2008).
+One can also use a stepwise aggregation via homogeneous areas (domains), ecosystem assets or some smaller basal spatial units, to control for this spatial non-independence, and thus reduce the sampling bias and variance.
+
 In our experience, the choice of which area units to use when doing this step-wise aggregation is often not reflected much upon.
@@
 However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; **Box 1**).
-It is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
-The MAUP can, however, lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
+The MAUP can lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
+This is unfortunate, as it is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
 One option to reduce this discrepancy and arbitrariness is to use ecologically motivated areal units in the step-wise aggregation such as homogeneous ecosystem areas [@vallecillo2022] or homogeneous impact areas [@kolstad_making_2024].
@@
 [^2]: When using means across areas, area weighting is required when following SEEA EA.
-
-Interpolation, or other types of modeling, can also be used to impute or predict data values between the points where we have information [@finke2002; @stein2001].
-This approach can also be seen as a type of post-stratification, and it can result in better variable or indicator estimates for the ecosystem assets as a whole.
-It can also be motivated by a desire to know more about the fine-scale variation in variable values, and perhaps visualise this on a map.
-This type of spatial aggregation, or perhaps better named upscaling, is not the focus of this paper (but see **Box 1**), and in our experience this has not been common practice in SEEA EA accounts produced to date [but see @kolstad_making_2024].
 
@@
 
-**Upscaling**
-
-The term spatial aggregation is also sometimes used to refer to **upscaling** in the sense of taking local information and making it applicable or translating it to larger, e.g.
-landscape or regional, scales [@knotters2010; @finke2002].
-This process often involves predicting values across larger areas or surfaces, generally through models, including krieging-based models and distribution models.
-When the data (e.g. indicator values) are made wall-to-wall, one can then more easily get point estimates for larger areas via simple aggregation functions such as summation or by taking the mean.
-Using models to predict variable or indicator values is allowed under the SEEA EA.
-However, in this paper we are addressing the second part of this process only, of simplifying the spatial variation, from many to few data points (sometimes also included in the terms upscaling).\
-
 **Modifiable area unit problem**
@@
 The original spatial units that data are collected at may be a pragmatic choice in much the same way.
-Further still, data may not be available at the original spatial scale, but may only be downloadable in pre aggregated form [@vanbeurden1999].
+In addition, data may not be available in the original spatial grain, but may only be downloadable in pre aggregated form [@vanbeurden1999].
 All these cases lead to some form of aggregation error[@rastetter1992] where the aggregated information has a non-linear relationship with the true variable value.
@@
 The ecological fallacy [@robinson2009] explains the common misinterpretation of correlations between groups (*i.e.* aggregated information) and some phenomena.
-Simply said, the mean of a group, or an area, cannot accurately describe the qualities of individuals in that group, or pin-points inside that area, yet it is common for people to interpret it that way.
+Simply said, the mean of a group(or an area) cannot accurately describe the qualities of individuals in that group(or pin-points inside that area), yet it is common for people to interpret it that way.
 :::
@@
 SEEA EA encourages normalisation (which they refer to as rescaling), but other environmental assessment or natural resource accounting frameworks, refrains from these kinds of value judgments, or leaves more of this task up to the end users.
-<!-- ref? --> Secondly, normalisation sets a limit to how much a high variable value in one place can compensate for a low value somewhere else, and vice versa [@pedersen2016].
+<!-- ref?
+Not sure this is true, and whether to drop the last part of this sentence or not --> Secondly, normalisation sets a limit to how much a high variable value in one place can compensate for a low value somewhere else, and vice versa [@pedersen2016].
 This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviations from the upper reference level X~100~, and ignoring any positive deviations which would otherwise compensate for the negative ones (see **Example 2**).
@@
 The flip-side of this is that spatially aggregating variable values does not directly summarise information about ecosystem condition for that area.
-
-Thirdly, we normalise in order to standardise the indicator on the same scale so that we can perform thematic (i.e. vertical) aggregation [@jakobsson2021; @unitednations2024 §5.81].
+<!-- ref? Is this list of three exhaustive? -->
+Thirdly, we normalise in order to standardise the indicator on the same unitless scale so that we can perform thematic (i.e. vertical) aggregation [@jakobsson2021; @unitednations2024 §5.81].
 This is commonly referred to as *the* reason for normalising variables, but as we have shown, it is but one of three main reasons, and also perhaps the least confusing part for many.
@@
 @rastetter1992 referred to this as aggregation error and used it to describe some of the dangers associated with upscaling in the sense of using precise fine-scale information to explain or predict large scale phenomena.
-However, in our case the difference between variables and indicators is an intentional one, and we therefore refer to what is essentially the same mathematical effect, with the less derogatory term aggregation displacement.
+However, in our case, the difference between variables and indicators is an intentional one, and we therefore choose to refer to what is essentially the same mathematical effect with the less derogatory term aggregation displacement.
 
-In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2024; -@unitednations2022], there is no mention of aggregation displacement, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
+In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2024; -@unitednations2022], there is no mention of aggregation displacement, and there seems to be little awareness in general of the ramification of choosing different spatial aggregation pathways for variables.
 We now describe two ways that aggregation displacement affects ECAs: by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators (@sec-discrep), and secondly, how it can cause confusion about the effects of choosing different aggregation pathways during indicator development (@sec-develop).
@@
 In ecosystem condition accounting the aggregation of both variables and indicators is common practice and raises the issue of aggregation displacement (@fig-agg1) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
-This means that the spatial aggregation of metrics leads to a divergence in the type of information held in variables and in their associated indicators, to a point where they become qualitatively very different.
-As an example, consider an indicator based on spatial information on the abundance of a large herbivore.
-The ecosystem characteristic that this metric is meant to describe is "*Appropriate herbivore pressure, balancing overgrazing and shrubification"*.
-Imagine the variable shows relatively high densities in half the area and low densities in the other half.
-When spatially aggregating the variable, either by taking the mean density or the sum of individuals, will produce an estimate that is somewhere in the middle of the two regions.
-The user will have no information about the carrying capacity for each of the regions, and will only see the spatially aggregated value.
-From this, many different conclusions can be made, depending on what the user considers high or low densities.
-In any case, the judgment will not reflect a density that actually exists anywhere.
-The indicator, on the other hand, can have different reference levels for each region, introducing additional ecological information about the carrying capacity of the ecosystems.
-The levels may be such that the high density region is considered as having an overabundance of herbivores, and the low density region as having underabundance.
-When normalised, the indicator value will be low for both regions, and the spatially aggregated value will also be low.
-The indicator will describe the average condition for the intended ecosystem characteristic, whereas the variable will in essence not show anything that can be interpreted as such.
+This means that the spatial aggregation of metrics leads to a divergence in the type of information held in variables and in their associated indicators, to a point where they become qualitatively very different (Example 2).
 
-The wolves example (**Example 2**) also highlight the importance of making clear the differences in the interpretation of variables and indicators.
 Indicators and variables are sometimes presented side-by-side in ecosystem condition assessments or accounts.
@@
 Miss-use can ensue, for example by taking the variable value and assigning it the same interpretation as the indicator value.
-At the very least, the developer must clarify to the user how the two metrics have diverged significantly during spatial aggregation and normalisation, complicating the comparison of their respective information.
+At the very least, the developer must clarify to the user how the two metrics may have diverged during spatial aggregation and normalisation, complicating the comparison of their respective information.
 
@@
 We would therefore argue that when an indicator has been developed, the original variable is not presented.
-And when only variable estimates are given, extra care is given to describing the spatial aggregation pathway.
+And when only variable estimates are given, extra care is given to describing the spatial aggregation pathway, since unfavourable local variable values (in terms of ecosystem condition) could have become smoothed out during aggregation, hiding potential environmental problems .
 Furthermore, Eurostat [@eurostat2022] encourages reporting of variable values.
@@
 #| cache: false
-
 kbl(dat, format = "latex",
@@
 
-# Aggregements across nested ECAs {#sec-comp}
+# Agreement across nested ECAs {#sec-comp}
 
 We believe it is a very desirable property that variable and indicator values are comparable across repeated ECAs, and harmonised across assessment scales and so that accounts from sub-areas should be able to recreate the account of the larger area.
-This can happen when ECAs use different aggregation pathways for the same indicator.
+This can, however, happen when ECAs use different aggregation pathways for the same indicator.
 Cases when two accounts or assessments use what is considered the same metric, but find different values for the same areas, is cause for both confusion, and could result in reduced general credibility for ECAs.
@@
 SEEA EA uses the term aggregation commutativity for the property that the order of vertical and horizontal aggregation is irrelevant to the end result.
-This is indeed a desirable property, however, when indicators that make up an index have different extents (e.g.
-they relate to different ecosystem sub-types or have sparse data coverage), then the indicator-specific areal weights will not be transferable in the spatial aggregation of indices (see description of pathways 7 and 8 in @sec-pathways).
-A similar commutativity can be envisioned for step-wise aggregation of a single indicator, where it does not matter if the original data is aggregated directly from the original measurement scale to the EAA level, or via some intermittent scale, such as ecosystem assets or larger regions.
-This is achievable when aggregating indicators using area weighting, and when one is aware of the need for post-stratification when using stratified data sampling.
+This is indeed a desirable propertyA similar commutativity can be envisioned for step-wise aggregation of a single indicator, where it does not matter if the original data is aggregated directly from the original measurement scale to the EAA level, or via some intermittent scale, such as ecosystem assets or larger regions.
+This is achievable when aggregating metrics using area weighting, and when one is aware of the need for post-stratification when using stratified data sampling.
 
@@
 In @fig-pathways we show some aggregation pathways that we have come across, and some that we see as potential new pathways.
-The examples are non-exhaustive, and serve mainly to illustrate that many different pathways can be created using a limited number of unique operations (@sec-rec).
-We then describe how the aggregation pathways can formally, and succinctly, be described using the names of these operations.
+The examples are non-exhaustive, and serve mainly as illustration.
+We go on to describe how the aggregation pathways can formally, and succinctly, be described using the names of these operations.
 The choice about which pathway to use is not trivial, and we want to highlight some of the issues that could arise from having an *ad hoc* approach to these considerations.
@@
 
-Pathway 3 is similar to 2, but here there are two aggregation steps: one before and one after normalisation.
+Pathway 3 is similar to pathway 2, but here there are two aggregation steps: one before and one after normalisation.
 This can for example be the case when variables are aggregated to the scale corresponding to the reference levels before they can be normalised (**Example 1**).
@@
 
-Pathways 7-8 describe the spatial aggregation of composite indicators (indices).
-These are not the main topic of this paper, but are included here for completeness.
-In pathway 7, indicators are thematically aggregated to an index, which is then again aggregated spatially.
-The order of the two aggregation steps can be reversed if the process is truly commutative (@sec-comp).
-In pathway 8, spatial aggregation is only done on individual indicators.
-The two pathways differ in how they handle more complicated schemes for weighted averages.
-The spatial aggregation of indices (pathway 7), using area weighted means, does not allow different area weights to be applied to each underlying indicator.
-Yet this would in many cases be needed if the individual indicators do not all represent the entire ecosystem inside each ecosystem asset.
-This is often the case, since ECA are frequently geared towards high level ecosystem types (e.g.
-level 2 in the IUCN GET [@keith_iucn_2020]) that includes several lower level ecosystems that may need unique indicators to describe them.
-Pathway 8 is therefore more flexible in this regard, but it may interfere with the principle of commutativity [@unitednations2024].
-We feel this is a very much unresolved issue in the SEEA EA.
+Pathway 7 describe an alternative to pathway 1 for a model-assisted aggregation pathway, where the original sample is the result of a defined sampling strategy, but where a model is used to extrapolate or interpolate (depending on the case) values to other spatial points or locations that were not sampled.
+The resulting map (field of values) is then normalised, (re)sampled (which could imply taking all the discrete population units (e.g.
+all rivers), and finally aggregated.
 
@@
 
+ <!--# a possible fifth recommendation is to not present aggregated variables and indicators side-by-side as alternative to choose from -->
+
 Based on the above discussion and examples, we here summarise four recommendations for developers of ecosystem condition indicators and assessments for avoiding some of the pitfalls from having a too casual approach to the choice of aggregation pathways.
@@
 ECAs should be accompanied by detailed indicator documentation where a precise description of the steps in the aggregation process is presented.
-Verbose descriptions can be supplemented or replaced by a standardised short-hand annotation that we shall present now.
+Verbose descriptions can be supplemented or replaced by a standardised short-hand notation that we shall present now.
 We suggest using the terms as described in @fig-term and elsewhere in the paper: scale, truncate (abbreviated to *trunc.*), transform (*trans.*), spatial aggregation (*sp.
@@
 (country)$.
-"*Municipality" and "country" could here be included as footnotes.
+"Municipality" and "country" could here be included as footnotes.
 Note also that in the last example, normalisation only involved a scaling step.
-The description of indicator pathways can be made even easier by referencing specific aggregation pathways by number, as we have done in this paper using @fig-pathways in this paper.
+The description of indicator pathways can be made even easier by referencing specific aggregation pathways by number, as we have done in this paper using @fig-pathways in this paper.Methods used to change either the cover or the extent of the indicator data (see @sec-term) can also be included in the shorthand notation scheme.
+The three operations we conceive as possible options are extrapolation (extr.), interpolation (inter.), and (sub)sampling.
+We use the term spatial projection (projection for short) to encompass both extra- and interpolation.
+Thus, pathway 7 in Fig.
+3 can be written as $project - scale - sample - sp.agg.
+- trunc$.
+The type of model used for the spatial projection can be added as a footnote.
 

### 2025-03-05 manuscript/manuscript.qmd
handle first round of trackdown

@@
 
-## Terminology
+## Terminology {#sec-term}
 
@@
 
-We follow the definition in Bierkens et al 2002 and treat spatial aggregation as the process of increasing the support (i.e.
-reducing the grain).
+We follow the definition in @bierkens2002 and treat spatial aggregation as the process of increasing the support (i.e. reducing the grain).
 We see this as a separate process to models that create spatially continuous data from point data by either increasing the extent (extrapolation) or cover (interpolation).
 However, these operations are in practice often intertwined with spatial aggregation.
-Either explicitly, for example by using model based spatial estimation methods (e.g.
-geostatistics) and subsequent spatial sampling of this field of random values (Brus et al 1997).
-Inter- or extrapolation can also be combined with spatial aggregation more inadvertently, for example by averaging point data and attributing this value to entire polygons, thus relying on a design-based estimation process (Brus 1997, Gruijter et al 1990).
+Either explicitly, for example by using model-based spatial estimation methods (e.g. geostatistics) and subsequent spatial sampling of this field of random values [@brus1997].
+Inter- or extrapolation can also be combined with spatial aggregation more inadvertently, for example by averaging point data and attributing this value to entire polygons, thus relying on a design-based estimation process [@gruijter1990].
 The term upscaling is sometimes used synonymously to our definition of spatial aggregation, but it has also a wider use, related especially to the ambiguous use of the terms scale to mean either grain or extent.
@@
 Spatial aggregating of ecosystem condition metrics is applied to alleviate issues of spatial auto-correlation.
-For example if data is collected opportunistically with sampling sizes varying between areas, then performing partial spatial aggregation to some intermediate sized areas will give equal weight or importance to all areas regardless of how many data points are included in each (Brus 1997 again here?).
+For example if data is collected opportunistically with sampling sizes or sampling effort varying between areas, then performing partial spatial aggregation to some intermediate sized, homogeneous areas, will give equal weight or importance to all areas regardless of how many data points are included in each.
 This approach represents a form of post-stratification that is also relevant to process data from hierarchical (e.g. blocked) sampling designs.
@@
 
-\clearpage
-
 ::: callout-note
@@
 | \-               | Plot 1 | Plot 2 | EA 1         |
-|----------------------|--------|--------|--------------|
+|------------------|--------|--------|--------------|
 | Variable values  | 2      | 36     | 18           |
@@
 To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some statistical inference.
-Both design-based and model-based approaches (Brus 2021) are valid under SEEA EA.
-However, transparency, validity and simplicity are all very desirable properties from ECAs, which favours a design-based approach that makes no modeling assumptions (Brus 2021).
+Both design-based and model-based approaches are valid under SEEA EA.
+However, transparency, validity and simplicity are all very desirable properties from ECAs, which favours a design-based approach that makes no modeling assumptions [@brus2021].
 We will therefore consider design-based approaches here.
 Model-based approaches also generally means that the variables are simpler to aggregate spatially since the spatial structure of the variables will already be accounted for via the prediction process.
-For example, if using krieging to create local predictions for the variable across the ecosystem asset, then the variable is considered random, and the sampling locations may be considered as fixed.
-Thus a simple systematic sampling of the field of values across the ecosystem asset will be valid, and give representative grounds for statistical inference (Brus 2021).
+For example, if using kriging to create local predictions for a variable across an ecosystem asset, then the variable is considered random, and the sampling locations may be considered as fixed.
+Thus a simple systematic sampling of the field of values across the ecosystem asset will be valid, and give representative grounds for statistical inference [@brus2021].
 
@@
 For balanced samples, or samples from a simple random sampling process, we can apply these functions directly, but depending on the random process used to select the sampling points we may need to account for the spatial structure (spatial autocorrelation) of the data as well, when doing this aggregation.
-This can be addressed in several ways, such as model-assisted approaches (Brus 2021), e.g.
-using regression models (Brus 2008).
-One can also use a stepwise aggregation via homogeneous areas (domains), ecosystem assets or some smaller basal spatial units, to control for this spatial non-independence, and thus reduce the sampling bias and variance.
+This can be addressed in several ways, such as model-assisted approaches [@brus2021], e.g.
+using regression models @brus2000.
+One can also use add an intermediate aggregation step via homogeneous areas (domains), ecosystem assets or some smaller basal spatial units, to control for this spatial non-independence, and thus reduce the sampling bias and variance.
+
+[^2]: When using means across areas, area weighting is required when following SEEA EA.
 
@@
 One option to reduce this discrepancy and arbitrariness is to use ecologically motivated areal units in the step-wise aggregation such as homogeneous ecosystem areas [@vallecillo2022] or homogeneous impact areas [@kolstad_making_2024].
-
-[^2]: When using means across areas, area weighting is required when following SEEA EA.
-
-\clearpage
 
@@
 The ecological fallacy [@robinson2009] explains the common misinterpretation of correlations between groups (*i.e.* aggregated information) and some phenomena.
-Simply said, the mean of a group(or an area) cannot accurately describe the qualities of individuals in that group(or pin-points inside that area), yet it is common for people to interpret it that way.
+Simply said, the mean of a group (or an area) cannot accurately describe the qualities of individuals in that group(or pin-points inside that area), yet it is common for people to interpret it that way.
 :::
@@
 SEEA EA encourages normalisation (which they refer to as rescaling), but other environmental assessment or natural resource accounting frameworks, refrains from these kinds of value judgments, or leaves more of this task up to the end users.
-<!-- ref?
+<!--# ref?
 Not sure this is true, and whether to drop the last part of this sentence or not  --> Secondly, normalisation sets a limit to how much a high variable value in one place can compensate for a low value somewhere else, and vice versa [@pedersen2016].
@@
 The flip-side of this is that spatially aggregating variable values does not directly summarise information about ecosystem condition for that area.
-<!-- ref? Is this list of three exhaustive? -->
-Thirdly, we normalise in order to standardise the indicator on the same unitless scale so that we can perform thematic (i.e. vertical) aggregation [@jakobsson2021; @unitednations2024 §5.81].
+<!--# ref?
+Is this list of three exhaustive?
+--> Thirdly, we normalise in order to standardise the indicator on the same unitless scale so that we can perform thematic (i.e.
+vertical) aggregation [@jakobsson2021; @unitednations2024 §5.81].
 This is commonly referred to as *the* reason for normalising variables, but as we have shown, it is but one of three main reasons, and also perhaps the least confusing part for many.
@@
 For example, direct extrapolation based on uneven or spatially biased sampling leads to biased estimates [@kolstad_making_2024].
-We will, however, focus now on what we consider an extended case of aggregation error, sensu @rastetter1992.
+We will, however, focus now on what we consider an extended case of aggregation error, *sensu* @rastetter1992.
 They showed that when we take the average of two variables, each of which have attribute values (e.g.
@@
 |                  | EA 1 | EA 2 | EA 3 | EAA aggregate |
-|----------------------|------|------|------|---------------|
+|------------------|------|------|------|---------------|
 | Variable values  | 20   | 5    | 0    | Sum = 25      |
@@
 But the high wolf numbers in asset 1 *does compensate* when looking at the aggregated variable value.
-
 :::
@@
 SEEA EA uses the term aggregation commutativity for the property that the order of vertical and horizontal aggregation is irrelevant to the end result.
-This is indeed a desirable propertyA similar commutativity can be envisioned for step-wise aggregation of a single indicator, where it does not matter if the original data is aggregated directly from the original measurement scale to the EAA level, or via some intermittent scale, such as ecosystem assets or larger regions.
+This is indeed a desirable property.
+A similar commutativity can be envisioned for step-wise aggregation of a single indicator, where it does not matter if the original data is aggregated directly from the original measurement scale to the EAA level, or via some intermittent scale, such as ecosystem assets or larger regions.
 This is achievable when aggregating metrics using area weighting, and when one is aware of the need for post-stratification when using stratified data sampling.
@@
 Pathway 7 describe an alternative to pathway 1 for a model-assisted aggregation pathway, where the original sample is the result of a defined sampling strategy, but where a model is used to extrapolate or interpolate (depending on the case) values to other spatial points or locations that were not sampled.
-The resulting map (field of values) is then normalised, (re)sampled (which could imply taking all the discrete population units (e.g.
-all rivers), and finally aggregated.
+The resulting map (field of values) is then normalised, (re)sampled (which could imply taking all the discrete population units (e.g. all rivers), and finally aggregated.
 
@@
 These terms can be abbreviated as shown in the parentheses, and placed together into a single string, where operations performed simultaneously are enclosed in square brackets.
-For example, pathway 1 in @fig-pathways can be annotated as $scale - trunc.
-- trans.
-- sp.
-agg.$.
-If the three normalisation steps are performed simultaneously this can be written as $[scale - trunc.
-- trans.] - sp.
-agg.$.
-If there are more aggregation steps, then the spatial level at each step should be included in the description, e.g.
-for pathway 3: $sp.
-agg.
-(municipalities) - scale - sp.
-agg.
-(country)$.
+For example, pathway 1 in @fig-pathways can be annotated as $scale - trunc. - trans. - sp. agg.$.
+If the three normalisation steps are performed simultaneously this can be written as $[scale - trunc. - trans.] - sp. agg.$.
+If there are more aggregation steps, then the spatial level at each step should be included in the description, e.g. for pathway 3: $sp. agg. (municipalities) - scale - sp. agg. (country)$.
 "Municipality" and "country" could here be included as footnotes.
@@
 Thus, pathway 7 in Fig.
-3 can be written as $project - scale - sample - sp.agg.
-- trunc$.
+3 can be written as $project - scale - sample - sp.agg. - trunc$.
 The type of model used for the spatial projection can be added as a footnote.
@@
 Each indicator version is given its own ID, and its own fact sheet with a persistent URL, so that ECAs that use these indicators can unequivocally cite specific indicator versions, and how they looked at a given point in time.
-ecRxiv also includes standardised metadata reporting schemes with fields for  ecosystem type (IUCN GET) and aggregation pathways (see recommendation number 1 above).
+ecRxiv also includes standardised metadata reporting schemes with fields for ecosystem type (IUCN GET; @keith_iucn_2020) and aggregation pathways (see recommendation number 1 above).
 

### 2025-04-07 manuscript/manuscript.qmd
download

@@
 We will therefore consider design-based approaches here.
-Model-based approaches also generally means that the variables are simpler to aggregate spatially since the spatial structure of the variables will already be accounted for via the prediction process.
+Model-based approaches (e.g.
+interpolation) also generally means that the variables become simpler to aggregate spatially since the spatial structure of the variables will already be accounted for via the prediction process.
 For example, if using kriging to create local predictions for a variable across an ecosystem asset, then the variable is considered random, and the sampling locations may be considered as fixed.
@@
 Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54].
+In addition, non-paramteric bootstrapping (GLACIERS) and bayesian updating (KOLSTAD et al ) can be used to possibly obtain better descriptions of the variation or the uncertainty in the sample.
 If the inclusion probabilities are known, or they can be estimated, then the Horvitz-Thomson estimator for the sum or the mean (and variance) of a spatial sample is appropriate and useful, especially when the data originated from a stratified sample.
@@
 using regression models @brus2000.
-One can also use add an intermediate aggregation step via homogeneous areas (domains), ecosystem assets or some smaller basal spatial units, to control for this spatial non-independence, and thus reduce the sampling bias and variance.
+One can also add an intermediate aggregation step via homogeneous areas (domains), ecosystem assets or some smaller basal spatial units, to control for this spatial non-independence, and thus reduce the sampling bias and variance.
 

### 2025-04-07 manuscript/manuscript.qmd
ref agg methods

@@
 Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54].
-In addition, non-paramteric bootstrapping (GLACIERS) and bayesian updating (KOLSTAD et al ) can be used to possibly obtain better descriptions of the variation or the uncertainty in the sample.
+In addition, non-paramteric bootstrapping [@kolstad_area_2025] and bayesian updating [@kolstad_making_2024] can be used to possibly obtain better descriptions of the variation or the uncertainty in the sample.
 If the inclusion probabilities are known, or they can be estimated, then the Horvitz-Thomson estimator for the sum or the mean (and variance) of a spatial sample is appropriate and useful, especially when the data originated from a stratified sample.

### 2025-04-07 manuscript/manuscript.qmd
add more pathways

@@
 
-We believe it is a very desirable property that variable and indicator values are comparable across repeated ECAs, and harmonised across assessment scales and so that accounts from sub-areas should be able to recreate the account of the larger area.
-This can, however, happen when ECAs use different aggregation pathways for the same indicator.
+We believe it is a very desirable property that variable and indicator values are comparable across repeated ECAs and harmonised across assessment scales so that accounts from sub-areas should be able to recreate the account of the larger area.
+This can, however, fail to happen when ECAs use different aggregation pathways for the same indicator.
 Cases when two accounts or assessments use what is considered the same metric, but find different values for the same areas, is cause for both confusion, and could result in reduced general credibility for ECAs.
@@
 SEEA EA uses the term aggregation commutativity for the property that the order of vertical and horizontal aggregation is irrelevant to the end result.
-This is indeed a desirable property.
-A similar commutativity can be envisioned for step-wise aggregation of a single indicator, where it does not matter if the original data is aggregated directly from the original measurement scale to the EAA level, or via some intermittent scale, such as ecosystem assets or larger regions.
-This is achievable when aggregating metrics using area weighting, and when one is aware of the need for post-stratification when using stratified data sampling.
+This is indeed a desirable property, although not always possible.
+**Example 4** shows a case where the spatial aggregation of an indicator uses area weighs that are different from the total extent of the EA (which is what is recommended by SEEA EA).
+Such alternative weights can be appropriate when the indicator only covers part of the ecosystem covered in the account, for example if it covers only one of many habitat types.
+In these cases, each indicator in an index might require different weights when spatially aggregated, but this is of coarse not possible after the indicators are combined into a single index.
+
+A similar aggregation commutativity can be envisioned for step-wise aggregation of a single indicator, where it does not matter if the original data is aggregated directly from the original measurement scale to the EAA level, or via some intermittent scale, such as ecosystem assets or larger regions.
+This is possible when aggregating metrics using area weighting, and when one is aware of the need for post-stratification when using stratified data sampling.
+However, **Example 4** again shows how
+
+::: callout-note
+**Example 4 -- Glacier area**\
+*Variable*: Extent of glacier area *anno* 2018-2019 (@fig-mapsGlaciers; A).
+
+*Reference levels*: X~100~ is defined as the extent of glacier area *anno* 1947-1985. X~0~ is an absence of glaciers, and X~60~ is a 90% loss of glaciers.
+
+This indicator [@kolstad_area_2025] uses historical data as the reference levels against which to normalise the variable.
+Both the variable and the reference levels are maps, but the metric itself needs to be numbers assigned to distinct ecosystem assets (EA).
+Seventy EAs (@fig-mapsGlaciers; B) were defined as unique combinations of geographical regions (@fig-mapsGlaciers; C) and bioclimatic regions.
+There is then some spatial aggregation done to get the summed glacier area for each EA.
+Of the 70 EAs, 29 had glaciers in the historical glacier map.
+The remaining EAs did not get assigned indicator values.
+
+The point estimates for each EA was then turned into probability distributions by sampling from a normal distribution with a standard deviation equal to 3% of the mean (the uncertainty reported for the maps).
+To further aggregate the data to geographical regions (@fig-mapsGlaciers; C), these distributions were resampled with replacement and with an *n* corresponding to the extent of glaciers in the historical map, i.e.
+using historical glacier extent as the area weights.
+The same bootstrapping procedure was done for aggregation to the country (EAA) level.
+
+In SEEA EA it says that "\[...\] the values recorded in an ecosystem condition variable account should be calculated as the area-weighted arithmetic mean of ecosystem assets belonging to the particular ecosystem type within the EAA" [@unitednations2024, p.
+104].
+However, in this case the weighs were not the total area (i.e. the alpine area, since that is the ecosystem that this indicator is designed for).
+Instead the weights were defined by the extent of glaciers in the historic map.
+This implied that EAs with no glaciers in the historic map were coded as *NA* and did not contribute to the EAA estimate, with the argument that the indicator is not valid or relevant for those areas.
+It also ment that areas inside the EA with no initial glacier cover was seen as similarly non-relevant to the indicator.
+Alternatively, using alpine extent as the weight during the spatial aggregation caused the indicator values to become considerably lower (@fig-ridge).
+This was partly due to cases where EAs with large alpine extent, but marginal glacier extent in the historic map, had lost all its glaciers in recent time.
+These EA would then get a very low indicator value and also weight very heavy towards the EAA estimate, even though the actual change in absolute glacier extent for that EA was very little.
+It was therefore argued that the area weight should be defined based on the area within the alpine ecosystem for which the indicator was considered relevant, and not defined base on the total EA extent.
+
+The spatial delineation of EAs can affect the indicator estimate at the EAA level.
+In this case the EAA estimates were not identical, but qualitatively similar (@fig-ridge).
+However, using larger EAs generally led to more narrow probability distributions which we believe is an artifact, and not a real reduction in uncertainty.
+In any case, this exemplifies how the MAUP and the choice of spatial units can affect the aggregated values.
+
+```{r fig-mapsGlaciers}
+#| fig-cap: 'Extent of glaciers in Norway can be visualised either as actuall occurences (A) or as the summed area for fifferent spatial units: B) ecosystem asset; C) gegrapical regions; D) the whole country.
+The legend is omitted for simplicity.
+The different maps exemplifies the modifiable area unit problem.
+From Kolstad (2025)'
+knitr::include_graphics("../img/maps.png")
+```
+
+```{r fig-ridge}
+#| fig-cap: 'Indicator values (probability distributions; vertical lines are quartiles) for the indicator _Glacier extent_ aggregated to EAA (country) level.
+The *indicator* is first normalised at the scale of the original EAs; *indicator_level2* is pre-aggregated to geographical regions before it it normalised, and; *indicator_level3* is normalised at the EAA scale with no further spatial aggregation.
+The difference between these three distributions show that the definition of the spatial untis and and especially the sice of the ecosystem assets does matter to the EAA estimate.
+*indicator_mtArea* is an indicator normalised at the EA scale (similar to *indicator*), but which is aggregted using the total alpine extent for each EA rather than the intial glacer extent as the weights.
+This approach puts a lot more emphasis on areas especially in the south of the country where there is a lot of alpine area but little intial glacer extent.
+Any loss of glacers, even if quite small in absolute terms, had a great effect on lowering the indicator estimates which where then propagated to the EAA level.'
+knitr::include_graphics("../img/ridge.png")
+```
+:::
 
@@
 
-Pathway 7 describe an alternative to pathway 1 for a model-assisted aggregation pathway, where the original sample is the result of a defined sampling strategy, but where a model is used to extrapolate or interpolate (depending on the case) values to other spatial points or locations that were not sampled.
+Pathway 7 describe an alternative to pathway 1 for a model-assisted aggregation pathway, where the original sample is the result of a defined sampling strategy, but where a model is used to project (extrapolate or interpolate, depending on the case) values to other spatial points or locations that were not sampled.
 The resulting map (field of values) is then normalised, (re)sampled (which could imply taking all the discrete population units (e.g. all rivers), and finally aggregated.
+
+Pathways 8 and 9 describe alternative ways to create indecies at different spatial scales,
+either by spatially aggregating the indecies (pathway 8), or by aggregating the indicators seperately and
+recalcuating the indecies at differet spatial scales (pathway 9).
+If the
 

### 2025-04-08 manuscript/manuscript.qmd
add recommodations

@@
 For ecosystem condition accounting (ECA) and assessments, this often requires that metrics[^1] used to describe the condition of ecosystems are aggregated spatially to produce a single value for a larger area.
-This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something that is relevant to decision makers and perhaps actionable at a political level [@king2024] <!--# read more -->.
+This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something that is relevant to decision makers and perhaps actionable at a political level [@king2024].
 However, spatial aggregation affect how a metric should be interpreted, and the descriptive properties and normative claims one can make based on it [@allain2018], and without an understanding of the pitfalls during spatial aggregation of information, one risks introducing unintended and undetected systematic bias, as well as downright logical errors, into the decision making process.
@@
 In @sec-pathways we describe a new reporting terminology for spatial aggregation pathways.
-And finally, in @sec-rec we give four concrete recommendations aimed at indicator developers that will help ensure a sensible and informed approach to spatial scaling of ecosystem condition metrics.
+And finally, in @sec-rec we give five concrete recommendations aimed at indicator developers that will help ensure a sensible and informed approach to spatial scaling of ecosystem condition metrics.
 
@@
 
-| Example | Main illustrating point |
-|--------------------------|----------------------------------------------|
-| Bilberry | Commensurable spatial scales for variable and reference levels (@sec-match) |
-| Wolves | Spatial compensation and discrepancy between variable and indicator (@sec-discrep) |
-| PTI | Compensation and the order of operations (@sec-pathways) |
+|   | Example | Main illustrating point |
+|--|--------------------------|---------------------------------------------|
+| 1 | Bilberry | Commensurable spatial scales for variable and reference levels (@sec-match) |
+| 2 | Wolves | Spatial compensation and discrepancy between variable and indicator (@sec-discrep) |
+| 3 | PTI | Compensation and the order of operations (@sec-pathways) |
+| 4 | Glaciers | Aggregation commutativity and alternative indicator weights (@sec-comp) |
 
@@
 
-We believe one solution to this is to acknowledge that the aggregation pathway can affect the indicators so much, sometimes as much as variation in the variable itself, that the use of different pathways imply that metrics are named and treated uniquely, for example using different common names, but perhaps better still, different version numbers or indicator IDs.
+We believe one solution to this is to acknowledge that the aggregation pathway can affect the indicators (and indices) so much, sometimes as much as variation in the variable itself, that the use of different pathways imply that metrics are named and treated uniquely, for example using different common names, but perhaps better still, different version numbers or indicator IDs.
 For each unique indicator the precise aggregation pathway should be specified as part of the indicator metadata.
@@
 This is possible when aggregating metrics using area weighting, and when one is aware of the need for post-stratification when using stratified data sampling.
-However, **Example 4** again shows how
+However, **Example 4** demonstrates two factors that lead to this indicator-level commutativity is not being fulfilled in this case.
+
+First, some indicators are possible normalise at the measurement scale, i.e. the finest resolution of the raw data.
+But others may need to be (pre-) aggregated, for example if the variable is designed to depict a density or a proportion that retains to a spatial unit.
+In Example 4, there is no obvious definition of the ecosystem assets, and so one must first define those before calculating the variable values for each of them.
+It turn out that in this case the indicator value at the EAA level is not insensitive to the spatial delineation of the EAs.
+
+Secondly, SEEA EA mentions several arithmetic or parametric aggregation functions, such as different means, median, min or max values.
+It does not mention the possibility of bootstrap resampling [@jakobsson2021; @kolstad_making_2024].
+This method allows for using probability distributions to represent indicator values and uncertainties, and to use weighted resampling to create new joint probability distributions for the spatially aggregated indicator.
+In Example 4, point estimates for the variable and reference levels are replaced with probability distributions where a 3% uncertainty is introduced based on the reported mapping uncertainty of the raw data.
+The shape of the aggregated joint distributions depend on the spatial size and configuration of the spatial units where this uncertainty is introduced.
 
@@
 
-*Reference levels*: X~100~ is defined as the extent of glacier area *anno* 1947-1985. X~0~ is an absence of glaciers, and X~60~ is a 90% loss of glaciers.
+*Reference levels*: X~100~ is defined as the extent of glacier area *anno* 1947-1985.
+X~0~ is an absence of glaciers, and X~60~ is a 90% loss of glaciers.
 
 This indicator [@kolstad_area_2025] uses historical data as the reference levels against which to normalise the variable.
-Both the variable and the reference levels are maps, but the metric itself needs to be numbers assigned to distinct ecosystem assets (EA).
+Both the variable and the reference levels are vector maps, but the metric itself needs to be numbers assigned to distinct ecosystem assets (EA).
 Seventy EAs (@fig-mapsGlaciers; B) were defined as unique combinations of geographical regions (@fig-mapsGlaciers; C) and bioclimatic regions.
@@
 
-The point estimates for each EA was then turned into probability distributions by sampling from a normal distribution with a standard deviation equal to 3% of the mean (the uncertainty reported for the maps).
+The point estimates for the variable and the reference levels for each EA was then turned into probability distributions by sampling from a normal distribution with a standard deviation equal to 3% of the mean (the uncertainty reported for the maps).
 To further aggregate the data to geographical regions (@fig-mapsGlaciers; C), these distributions were resampled with replacement and with an *n* corresponding to the extent of glaciers in the historical map, i.e.
@@
 This implied that EAs with no glaciers in the historic map were coded as *NA* and did not contribute to the EAA estimate, with the argument that the indicator is not valid or relevant for those areas.
-It also ment that areas inside the EA with no initial glacier cover was seen as similarly non-relevant to the indicator.
+It also meant that areas inside the EA with no initial glacier cover was seen as similarly non-relevant to the indicator.
 Alternatively, using alpine extent as the weight during the spatial aggregation caused the indicator values to become considerably lower (@fig-ridge).
@@
 From Kolstad (2025)'
+#| echo: false
 knitr::include_graphics("../img/maps.png")
@@
 Any loss of glacers, even if quite small in absolute terms, had a great effect on lowering the indicator estimates which where then propagated to the EAA level.'
+#| echo: false
+#| out-width: 500
 knitr::include_graphics("../img/ridge.png")
@@
 
-Pathways 8 and 9 describe alternative ways to create indecies at different spatial scales,
-either by spatially aggregating the indecies (pathway 8), or by aggregating the indicators seperately and
-recalcuating the indecies at differet spatial scales (pathway 9).
-If the
+Pathways 8 and 9 describe alternative ways to create indices at different spatial scales, either by spatially aggregating the indices (pathway 8), or by aggregating the indicators separately and recalculating the indices at different spatial scales (pathway 9).
+If perfectly commutative then the two approaches will yield the same result.
+However, if the indicators that make up the indices use different area weights as part of their spatial aggregation (**Example 4**), then pathway 9 is the only one that gives the same index values across nested accounts (@sec-comp).
 
@@
 
-Based on the above discussion and examples, we here summarise four recommendations for developers of ecosystem condition indicators and assessments for avoiding some of the pitfalls from having a too casual approach to the choice of aggregation pathways.
+Based on the above discussion and examples, we here summarise five recommendations for developers of ecosystem condition indicators and assessments for avoiding some of the pitfalls from having a too casual approach to the choice of aggregation pathways.
 
@@
 
+### 5. Don't aggregate indices, unless you know all aggregation operations are commutative
+
+When spatially aggregating indices it is no longer possible to use different area weights to each single indicator.
+If all indicators making up n index are spatially aggregated using the same weights, typically the extent of the spatial units, then the aggregation operations will be commutative and it will not matter if you aggregate vertically then horisontally , or the other way around (\@fig-pathways; Pathway 8 and 9, respectively).
+However, indicators are not always valid for the entire ecosystem type that the account is made over.
+They can for example retain to only certain habitat or nature types within the ecosystem type, in which case one might want to use different weights when spatially aggregating the indicator that reflect the extent of only the relevant habitat or nature types, and not of the entire ecosystem (see **Example 4**).
+But when combined with other indicators into a common index, any subsequent aggregation of that index will give equal spatial weight to all indicators (see next paragraph for an exception to this).
+We therefore recommend caution, and to default to the solution of instead spatially aggregating indicators separately, and then re-calculating indices when needed (\@fig-pathways; Pathway 9).
+This will ensure that nested accounts will be comparable, i.e. that they will present the same index values independent of which spatial units that the indices first are calculated for.
+
+An alternative solution to ensure agreement across nested accounts is to use weights also when performing the vertical or thematic aggregation of indicators into indicators, where indicators are weighted respective of the area that they are valid over.
+With this approach one can down-weigh indicators that only cover only smaller areas (specific nature types etc.), or which have incomplete spatial coverage.
+When this weighting is already performed at the vertical aggregation stage, any subsequent horisontal aggregation does not need to account for the unique indicator weights again.
+However, weighting indicators in vertical aggregation opens a whole other can of worms, and therefore we do not offer this as a generic recommendation.
+
 # Conclusion

### 2025-09-11 manuscript/manuscript.qmd
pull changes from google docs

@@
       - ref: nina
-  - name: Bálint Czúcz
+  - name: BC!lint CzC:cz
     email: balint.czucz@nina.no
@@
       - ref: nina
-  - name: Chloé R. Nater
+  - name: ChloC) R. Nater
     email: chloe.nater@nina.no
@@
 The world is facing a nature crisis of a magnitude that requires urgent political decision making and international coordination [@ipbes_global_2019].
-Ecosystem accounting (EA) is a type of environmental assessments aimed at communicating on the status and trends in the extent and condition of ecosystems at a scale and in a format that is relevant for decision makers [@edens2022, @comte2022].
+Ecosystem accounting (EA) has been fronted as a framework for providing a knowledge base for this.
+EAs are environmental assessments aimed at communicating on the status and trends in the extent and condition of ecosystems at a scale and in a format that is relevant for decision makers [@edens2022, @comte2022].
 For ecosystem condition accounting (ECA) and assessments, this often requires that metrics[^1] used to describe the condition of ecosystems are aggregated spatially to produce a single value for a larger area.
-This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something that is relevant to decision makers and perhaps actionable at a political level [@king2024].
-However, spatial aggregation affect how a metric should be interpreted, and the descriptive properties and normative claims one can make based on it [@allain2018], and without an understanding of the pitfalls during spatial aggregation of information, one risks introducing unintended and undetected systematic bias, as well as downright logical errors, into the decision making process.
+This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something that is relevant to decision makers and potentially actionable at a political level [@king2024].
+However, the way spatial aggregation is done affects how a metric should be interpreted, and the descriptive properties and normative claims one can make based on it [@allain2018].
+Without an understanding of the pitfalls during spatial aggregation of information, one risks introducing unintended and undetected systematic bias into the decision making process.
 
-[^1]: We use the SEEA EA definitions of the terms variables, indicators (normalised variables) and metrics (referring to either a variable or a metric) [@unitednations2024]
+[^1]: We use the SEEA EA definitions of the terms variables, indicators (normalised variables) and metrics (referring to either a variable or an indicator) [@unitednations2024]
 
-Although much attention has been devoted to the vertical aggregation of environmental indicators into indices [e.g.
+Although much attention has been devoted to the vertical aggregation of environmental indicators into composite indices [e.g.
 @langhans2014; @maes2020; @oecd_2008], as well as to the spatial aggregation of relatively unmodified or raw variables [e.g.
-@allain2018], there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting [but see @vanbeurden1999].
-This paper therefore focuses on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) [@unitednations2024] - the most used ecosystem accounting framework in the world [@comte2022].
+@allain2018], there is a lack of explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting [but see @vanbeurden1999].
+This paper therefore focuses on explaining and providing guidelines for the spatial aggregation of ecosystem indicators in the context of the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) [@unitednations2024] - the most used ecosystem accounting framework in the world [@comte2022].
 
 SEEA EA includes principles for compiling accounts for ecosystem extent, ecosystem condition, and ecosystem services [@unitednations2024; -@unitednations2022].
-We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is a relevant topic also for the other parts of the account, and to other types of indicator-based ecosystem assessments.
-When EU and EEA countries are in 2026 required to report ecosystem condition accounts to Eurostat and the European Commision (?), these accounts will be spatially aggregated to national levels.
-The point of having a robust methodological framework for the spatial aggregation of data and information is therefore quite pertinent.
+We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is also a relevant topic for the other parts of the account, and for other types of indicator-based ecosystem assessments.
+In 2026, EU and EEA countries will be required to report ecosystem condition accounts to Eurostat and the European Commission (?), and these accounts will have to be spatially aggregated to national levels.
+This underlines the importance of having a robust methodological framework for the spatial aggregation of data and information.
 
-This paper is aimed primarily at indicator developers, providing them with some recommendations for how to choose the right spatial aggregation approach for variables and indicators, and how to report these choices.
-Indicator developer subjectivity affects indicator interpretation [@allain2018] and finally the conclusion from ECAs.
-Still, the process of designing indicators and estimating indicator values is often left entirely to them, with little real involvement from indicator users.
-It is therefore important to be aware of the role that indicator developers have in shaping the type of information, and the narratives, that are presented to decision makers.
-This paper will make it easier for indicator developers to avoid some of the pitfalls from having a too casual approach to the choice of aggregation pathways.
-We use three examples to illustrate some of our main points (@tbl-examples).
-In @sec-pathways we describe a new reporting terminology for spatial aggregation pathways.
-And finally, in @sec-rec we give five concrete recommendations aimed at indicator developers that will help ensure a sensible and informed approach to spatial scaling of ecosystem condition metrics.
+This paper is aimed primarily at indicator developers, providing them with recommendations for how to choose the right spatial aggregation approach for variables and indicators, and how to report these choices in an accessible and transparent way.
+Indicator developer subjectivity affects indicator interpretation [@allain2018] and thus ultimately  the conclusion from ECAs.
+Still, the process of designing indicators and estimating indicator values is often left entirely to them, with little real involvement from indicator users, such as regional and local land managers and policy makers.
+It is therefore important to be aware of the role that indicator developers play in shaping the type of information and the narratives that are presented to decision makers.
+This paper will make it easier for indicator developers to avoid some common pitfalls that are associated with choosing aggregation pathways.
+We use four examples to illustrate some of our main points (@tbl-examples).
+In @sec-pathways we describe a new reporting terminology for spatial aggregation pathways, and finally, in @sec-rec we give five concrete recommendations aimed at indicator developers that will help ensure a sensible and informed approach to spatial scaling of ecosystem condition metrics.
 
@@
 We see a need for a common, more standardised terminology across these fields of practice in order to ease communication and avoid misunderstandings.
-For now we will now define the terms and how we use them in this paper.
+In the following, we define the terms as we use them in this paper.
 
@@
 However, these operations are in practice often intertwined with spatial aggregation.
-Either explicitly, for example by using model-based spatial estimation methods (e.g. geostatistics) and subsequent spatial sampling of this field of random values [@brus1997].
-Inter- or extrapolation can also be combined with spatial aggregation more inadvertently, for example by averaging point data and attributing this value to entire polygons, thus relying on a design-based estimation process [@gruijter1990].
-The term upscaling is sometimes used synonymously to our definition of spatial aggregation, but it has also a wider use, related especially to the ambiguous use of the terms scale to mean either grain or extent.
+Intertwinement can be explicit, for example when using model-based spatial estimation methods (e.g.
+geostatistics) and subsequent spatial sampling of this field of random values [@brus1997].
+In other cases, inter- or extrapolation can also be combined with spatial aggregation more inadvertently, for example by averaging point data and attributing this value to entire polygons, thus relying on a design-based estimation process [@gruijter1990].
+The term “upscaling” is sometimes used synonymously to our definition of spatial aggregation, but it also has a wider use, related especially to the ambiguous use of the terms scale to mean either grain or extent.
 
@@
 They are typically either in percentages or fractions, or as raw biophysical units, such as biomass or density.
-Indicators on the other hand, are always on a unitless scale between 0 and 1, where 0 represents very poor condition, and 1 represents the very good, or the best possible condition.
+Indicators, on the other hand, are always on a unitless scale between 0 and 1, where 0 represents very poor (or the worst possible) condition, and 1 represents very good (or the best possible) condition.
 Sometimes indicators are calculated directly [e.g. @åström], but usually indicators are made by normalising variables based on two or more reference levels [@unitednations2024 §5.60].
-Although also variables may have normative properties (due to choices about how they are chosen, defined, aggregated or presented [@allain2018]), the normalisation of variables into indicators makes them explicitly normative by introducing a clear directional interpretation, where increasing values always refer to an increase in condition in the direction of a reference condition.
+Although some variables may have normative properties (due to decisions about how they are chosen, defined, aggregated or presented [@allain2018]), the normalisation of variables into indicators makes them explicitly normative by introducing a clear directional interpretation, where increasing values always refer to an increase in condition in the direction of a reference condition.
 The information held by the indicator is also different from that of the variable since additional ecological knowledge can be introduced via the reference levels.
 The mathematical part of this normalisation contains several steps (@fig-term).
-These can be performed simultaneously, but for clarity we discuss them as separate steps.
-The order of the steps can also change.
+These can be performed simultaneously, but for clarity we discuss them as separate steps that may be implemented in different order.
 
-Variables can be scaled using a minimum of two reference levels, a lower and an upper, that defines which variable values to be coded as zero (X~0~) and as one (X~100~) on the indicator scale.
+Variables can be scaled using a minimum of two reference levels, a lower and an upper, that define which variable values are to be coded as zero (X~0~) and as one (X~100~) on the indicator scale.
 Variables can, if needed, be truncated to produce a bound indicator scale between 0 and 1.
 Sometimes variables are transformed to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
-This is commonly done by anchoring specific variable values to predefined class boundaries [@czúcz; @jakobsson_setting_2020], but may also be done without any additional reference levels, for example by using a sigmoid or exponential transformation [@mienna2024].
+This is commonly done by anchoring specific variable values to predefined class boundaries (sensu WFD) which we call reference levels [@czúcz; @jakobsson_setting_2020], but may also be done without explicit reference levels, for example by using a sigmoid or exponential transformation [@mienna2024].
 Finally, the indicator scale can also be divided into discrete fractions, either based on additional reference levels, or some other rule (e.g. equal range for each class).
@@
 
-# Why do we spatially aggregate data in ecosystem condition accounting
+# Spatial aggregation of data in ecosystem condition accounting
 
-Spatial aggregation of data refers to the conversion of fine-resolution data into coarser-resolution data, and depending on both the purpose and the nature of the metric, there can be several reasons both for performing and for refraining from spatial aggregation.
+Spatial aggregation of data refers to the conversion of fine-resolution data into coarser-resolution data and, depending on both the purpose and the nature of the metric, there can be several reasons for both performing and refraining from spatial aggregation.
 Readers are referred to @allain2018 for a more comprehensive review of different reasons for spatially aggregating variables.
-As we shall see, there are several requirements in the SEEA EA that make spatial aggregation necessary.
-However, spatial aggregation also comes with a whole suite of (sometimes undesirable) side-effects.
+As we will illustrate below, there are several requirements in the SEEA EA that make spatial aggregation necessary.
 
-Generally, spatial aggregation increases the tangibility of a metric, because it simplifies or compresses the information, sometimes into just a single number (total aggregation).
-This is sometimes necessary in ecosystem accounting when information is required at a relatively high administrative level (e.g.
+Generally, spatial aggregation increases the clarity and accessibility of a metric because it simplifies or compresses the information, sometimes into just a single number (total aggregation).
+This may be necessary in some cases of ecosystem accounting when information is required at a relatively high administrative level (e.g.
 nations) and when variables and indicators need to be presented in standardised tables which don't allow for describing a lot of spatial variation.
@@
 Data may also be only party aggregated, or the same metric may be presented at different levels of aggregation [@allain2018].
-A minimum level of aggregation may be decided based on uncertainties in the data: High spatially resolute data may be imprecise at fine scales, leading to false conclusion and potential miss-use [@lyytimäki2013], but these values seen over a larger area may be assumed to converge on some true value, making spatially aggregated data better suited for communication to users.
+A minimum level of aggregation may be decided based on uncertainties in the data: high spatially resolute data may be imprecise at fine scales, leading to false sense of certainty and potential misinterpretation [@lyytimäki2013], but these same data seen over a larger area may be assumed to converge on some true value, making spatially aggregated data better suited for communication to users.
 Other common arguments for performing spatial aggregation are to ease visualisation and to allow relevant comparisons [@allain2018].
 For example, ecosystem condition metrics can be visualised on a map, but when both the spatial resolution and spatial variation is high, maps become too complex, even unintelligible.
-It is therefore common to partially aggregate the data and provide a coarser map where different colours, representing values on the map, become distinguishable from each other.
+It is therefore common to partially aggregate the data and provide a coarser map where different colours are used to distinguish different values.
 Similarly, one may choose to aggregate such that the spatial resolution matches some meaningful delineation, notably ecosystem assets in SEEA EA, which makes it possible to easily compare the ecosystem condition between these areas.
+
+Spatial aggregation of ecosystem condition metrics may also be applied to alleviate issues of spatial auto-correlation.
+For example, if data is collected opportunistically with sampling sizes or sampling effort varying between areas, then performing partial spatial aggregation to some intermediate-sized, homogeneous areas, will give equal weight or importance to all areas regardless of how many data points are included in each.
+This approach represents a form of post-stratification that is also relevant to process data from hierarchical (e.g. blocked) sampling designs.
 
@@
 This smoothing out makes the data and its inherent information less suitable for some use cases, such as depicting spatial variation that can be used to support local land-use planning, for aiding a debate about spatial conflicts or identification of high risk areas [@vanbeurden1999].
-
-Spatial aggregating of ecosystem condition metrics is applied to alleviate issues of spatial auto-correlation.
-For example if data is collected opportunistically with sampling sizes or sampling effort varying between areas, then performing partial spatial aggregation to some intermediate sized, homogeneous areas, will give equal weight or importance to all areas regardless of how many data points are included in each.
-This approach represents a form of post-stratification that is also relevant to process data from hierarchical (e.g. blocked) sampling designs.
 
@@
 
-*Variable*: Horizontal coverage (vertical projection) of bilberry (*Vaccinum myrtillus*) recorded in permanent vegetation plots.\
+*Variable*: Horizontal coverage (vertical projection) of European blueberry or bilberry (*Vaccinum myrtillus*) recorded in permanent vegetation plots.\
 
@@
 |------------------|--------|--------|--------------|
-| Variable values  | 2      | 36     | 18           |
+| Variable values  | 2      | 36     | 19           |
 | X~100~           | 20\*   | 20\*   | 20           |
-| Indicator values | 0.1\*  | 1.0\*  | 0.9 (0.55\*) |
+| Indicator values | 0.1\*  | 1.0\*  | 0.95 (0.55\*) |
 
 : Dummy data for the bilberry example.
-Variable values (% cover of bilberry) is recorded for each vegetation plot, and aggregated (arithmetic mean) to the ecosystem asset (EA) level.
+Variable values (% cover of bilberry) are recorded for each vegetation plot, and aggregated (arithmetic mean) to the ecosystem asset (EA) level.
 The X~100~ value is defined for the EA level, but in this hypothetical case it is also dis-aggregated to each vegetation plot and used for normalising variable values also at the plot level.
@@
 Given that X~100~ is defined the way it is, this indicator should follow pathway 3 (@fig-agg1) by spatially aggregating the variable from plot scale to ecosystem asset to allow spatial compensation between plots.
-At the scale of ecosystem assets, the variable and the reference levels are at a commensurable scale, and the variable can be normalised and potentially aggregating it further.
+At the scale of ecosystem assets, the variable and the reference levels are at a commensurable scale, and the variable can be normalised and potentially aggregated further.
 :::
@@
 The type of manipulation depends on the spatial resolution of the underlying data, as well as the size and configuration of the ecosystem assets.
-To elaborate on this issue we first make a distinction between two main types of data: sparse (i.e. spatially discontinuous) data and spatially exhaustive data.
+To elaborate on this issue we first make a distinction between two main types of data: spatially exhaustive data and sparse (i.e. spatially discontinuous) data..
 
-Spatially exhaustive, or wall-to-wall data, include remote sensing products, but also other types of pre-aggregated data, where sparse data are assigned to areal units.
+*Spatially exhaustive*, or wall-to-wall data, include remote sensing products, but also other types of pre-aggregated data, where sparse data are assigned to areal units.
 This last type of data is very common.
 A lot of demographic information for example is made available at the scale of administrative units, rather than at the level of individuals (which would be very ineffective and also problematic in terms of protecting sensitive information).
-Similar pre-aggregation is, however, ubiquitous and true for most complete datasets.
-Consider as an example data from LiDAR (Light Detection And Ranging).
-The raw data is a point cloud, yet the most commonly available datasets are rasters, such as surface or canopy models; wall-to-wall generalisations of the point cloud.
-Also data such as satellite images ,or maps derived from such, have some inherent pre-aggregation where values are averaged for each raster cell.
-Often there is little we can do, as indicator developers, to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (**Box 1**).
-This issue will for example affect our ability to detect true maximum or minimum values (used in *worst case* aggregation, for example in the WFD) for the phenomena our variable is describing, since these extremes are already smoothed out in the pre-aggregation step [@vanbeurden1999].
+Similar pre-aggregation is rather common and present in most complete datasets.
+Consider, as an example, data from LiDAR (Light Detection And Ranging).
+The raw data is a point cloud, yet the most commonly available datasets are rasters, such as surface or canopy models, i.e. wall-to-wall generalisations of the point cloud.
+Also data such as satellite images, or maps derived from such, have some inherent pre-aggregation where values are averaged for each raster cell.
+Often there is little indicator developers can do to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (**Box 1**).
+This issue will, for example, hamper the detection of true maximum or minimum values (used in *worst rule* (i.e.
+one out al out) aggregation, for example in the WFD) for the phenomena described by the variable, since these extremes are already smoothed out in the pre-aggregation step [@vanbeurden1999].
 This will clearly affect the suitability of such data for identifying areas of high risk (very degraded areas, for example).
-The problem is the same when we deliberately aggregate data later in the indicator development process, and we always need to ensure that the spatial aggregation function we use does not have any unintentional effect on the type of inference that we can draw from the metric.
+Similar problems can also occur in subsequent aggregation steps of the indicator calculation workflow, and it is important to ensure that the spatial aggregation function used does not have any unintentional effects on the meaning of the metric.
 
-For sparse data we are considering data where variable values are assigned to singular points or small areas.
+*Sparse data* refers to datasets where variable values are assigned to singular points or isolated small areas.
 In these cases, there will be a lot of area inside each ecosystem asset for which we have no direct information, and some ecosystem assets may have no data at all.
@@
 Both design-based and model-based approaches are valid under SEEA EA. 
-However, transparency, validity and simplicity are all very desirable properties from ECAs, which favours a design-based approach that makes no modeling assumptions [@brus2021].
-We will therefore consider design-based approaches here.
 Model-based approaches (e.g.
-interpolation) also generally means that the variables become simpler to aggregate spatially since the spatial structure of the variables will already be accounted for via the prediction process.
+interpolation) generally make variables easier to aggregate spatially since the spatial structure of the variables will already be accounted for via the prediction process.
 For example, if using kriging to create local predictions for a variable across an ecosystem asset, then the variable is considered random, and the sampling locations may be considered as fixed.
 Thus a simple systematic sampling of the field of values across the ecosystem asset will be valid, and give representative grounds for statistical inference [@brus2021].
+However, transparency, validity and simplicity are all very desirable properties from ECAs, which favours a design-based approach that makes no modeling assumptions [@brus2021].
+From here on, we will focus on design-based approaches.
 
 For design-based approaches, one can assign values to areas via simple aggregation functions used on data values that fall within a given ecosystem asset.
-Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means[^2] [@unitednations2024 §5.54].
-In addition, non-paramteric bootstrapping [@kolstad_area_2025] and bayesian updating [@kolstad_making_2024] can be used to possibly obtain better descriptions of the variation or the uncertainty in the sample.
-If the inclusion probabilities are known, or they can be estimated, then the Horvitz-Thomson estimator for the sum or the mean (and variance) of a spatial sample is appropriate and useful, especially when the data originated from a stratified sample.
-For balanced samples, or samples from a simple random sampling process, we can apply these functions directly, but depending on the random process used to select the sampling points we may need to account for the spatial structure (spatial autocorrelation) of the data as well, when doing this aggregation.
+Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric, or harmonic means[^2] [@unitednations2024 §5.54].
+In addition, non-paramteric bootstrapping [@kolstad_area_2025] and Bayesian updating [@kolstad_making_2024] can be used to possibly obtain better descriptions of the variation or the uncertainty in the sample.
+If the inclusion probabilities are known, or they can be estimated, then the Horvitz-Thomson estimator for the sum or the mean (and variance) of a spatial sample is appropriate and useful, especially when the data originated from a stratified sample (ADD REF).
+Depending on the random process used to select the sampling points we may need to account for the spatial structure (spatial autocorrelation) of the data when doing this aggregation.
 This can be addressed in several ways, such as model-assisted approaches [@brus2021], e.g.
-using regression models @brus2000.
+using regression models [@brus2000].
 One can also add an intermediate aggregation step via homogeneous areas (domains), ecosystem assets or some smaller basal spatial units, to control for this spatial non-independence, and thus reduce the sampling bias and variance.
@@
 In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
-However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to at the modifiable area unit problem (MAUP; **Box 1**).
+However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to as the modifiable area unit problem (MAUP; **Box 1**).
 The MAUP can lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
@@
 
+**Temporal aggregation**
+
+… Should we add some sentences on temporal aggregation, i.e. the operations involved in representing data with a temporal resolution that is finer (shorter) than the accounting frequency?\
+
 **Modifiable area unit problem**
@@
 These units can be arbitrary (*modifiable*), yet different sizes and configuration of units can have large effects on the outcome of aggregation.
-Administrative units are often used as the ecosystem assets in SEEA EAs, yet they are in an ecological sense almost always arbitrary, and can also be subject to change over time.
+Administrative units are often used as the ecosystem assets in SEEA EA, yet they are in an ecological sense almost always arbitrary, and can also be subject to change over time.
 The original spatial units that data are collected at may be a pragmatic choice in much the same way.
 In addition, data may not be available in the original spatial grain, but may only be downloadable in pre aggregated form [@vanbeurden1999].
-All these cases lead to some form of aggregation error[@rastetter1992] where the aggregated information has a non-linear relationship with the true variable value.
+All these cases lead to some form of *aggregation error* [@rastetter1992] where the aggregated information has a non-linear relationship with the true variable value.
 However, the issue of MAUP in SEEA EA deserves an in-depth analyses that we cannot provide here.\
@@
 
-The ecological fallacy [@robinson2009] explains the common misinterpretation of correlations between groups (*i.e.* aggregated information) and some phenomena.
-Simply said, the mean of a group (or an area) cannot accurately describe the qualities of individuals in that group(or pin-points inside that area), yet it is common for people to interpret it that way.
+The ecological fallacy [@robinson2009] is the error of assuming that relationships or patterns observed for groups (at the aggregate level) must also hold true for the individuals within those groups.
+In EA, aggregated data will mask differences within regions and lead some people to interpret this as regions being more homogenous then they really are.
 :::
 
-# Why do we normalise variables? {#sec-norm}
+# Reasons for normalising variables? {#sec-norm}
 
@@
 Firstly, it gives a normative interpretation of a variable, defining a good and a bad state, and simultaneously a directionality to say when something is getting better or worse over time [@czucz_selection_2021].
-SEEA EA encourages normalisation (which they refer to as rescaling), but other environmental assessment or natural resource accounting frameworks, refrains from these kinds of value judgments, or leaves more of this task up to the end users.
-<!--# ref?
-Not sure this is true, and whether to drop the last part of this sentence or not  --> Secondly, normalisation sets a limit to how much a high variable value in one place can compensate for a low value somewhere else, and vice versa [@pedersen2016].
-This is because truncation effectively means that when we spatially aggregate an indicator, we are always aggregating the negative deviations from the upper reference level X~100~, and ignoring any positive deviations which would otherwise compensate for the negative ones (see **Example 2**).
+Secondly, normalisation sets a limit to how much a high variable value in one place can compensate for a low value somewhere else, and vice versa [@pedersen2016].
+This is because truncation effectively means that when spatially aggregating an indicator, one is always aggregating the negative deviations from the upper reference level X~100~, and ignoring any positive deviations which would otherwise compensate for the negative ones (see **Example 2**).
 Transformation can sometimes also have this effect, e.g. sigmoid transformations.
@@
 vertical) aggregation [@jakobsson2021; @unitednations2024 §5.81].
-This is commonly referred to as *the* reason for normalising variables, but as we have shown, it is but one of three main reasons, and also perhaps the least confusing part for many.
+This is commonly referred to as *the* reason for normalising variables, not least because it is perhaps the least confusing one for many; nonetheless, as we have shown, it is the only reason.
 
-<!--# Try add using !expr to get hat and bar sympols into caption:  -->
+<!--# Try add using !expr to get hat and bar symbols into caption:  -->
 
@@
 
-In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2024; -@unitednations2022], there is no mention of aggregation displacement, and there seems to be little awareness in general of the ramification of choosing different spatial aggregation pathways for variables.
-We now describe two ways that aggregation displacement affects ECAs: by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators (@sec-discrep), and secondly, how it can cause confusion about the effects of choosing different aggregation pathways during indicator development (@sec-develop).
+In the statistical standard and recommended guidelines for ecosystem accounts [@unitednations2024; -@unitednations2022], there is no mention of aggregation displacement, and there seems to be little awareness in general of the ramification of choosing different spatial aggregation pathways for variables.
+We now describe two ways that aggregation displacement affects ECAs: firstly, by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators (@sec-discrep), and secondly, how it can cause confusion about the effects of choosing different aggregation pathways during indicator development (@sec-develop).
 
@@
 
-In ecosystem condition accounting the aggregation of both variables and indicators is common practice and raises the issue of aggregation displacement (@fig-agg1) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
+In ECA the aggregation of both variables and indicators is common practice and raises the issue of aggregation displacement (@fig-agg1) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
 This means that the spatial aggregation of metrics leads to a divergence in the type of information held in variables and in their associated indicators, to a point where they become qualitatively very different (Example 2).
@@
 While this may initially seem like a democratic and transparent approach to reporting, this practice places a significant burden on both the end user and the indicator developer.
-Miss-use can ensue, for example by taking the variable value and assigning it the same interpretation as the indicator value.
+Misuse can ensue, for example by taking the variable value and assigning it the same interpretation as the indicator value.
 At the very least, the developer must clarify to the user how the two metrics may have diverged during spatial aggregation and normalisation, complicating the comparison of their respective information.
@@
 Then we see that the first ecosystem asset has more wolves than we would expect under the reference condition, whereas the other two have less.
-Because variable values above X~100~ are truncated, the high wolf numbers in ecosystem asset one does not compensate for the low indicator value in assets two and three, and hence the mean indicator value is less then the sum of the variable values divided by the sum of the X~100~values.
+Because variable values above X~100~ are truncated, the high wolf numbers in ecosystem asset one does not compensate for the low indicator value in assets two and three, and hence the mean indicator value is less than the sum of the variable values divided by the sum of the X~100~values.
 But the high wolf numbers in asset 1 *does compensate* when looking at the aggregated variable value.
@@
 
-Generally, because of the normalisation process, indicators are more aligned with the purpose of ECA with describing ecosystem condition in normative terms.
-Although variable accounts are allowed under the SEEA EA framework, and is indeed a central feature, we believe they should be seen as interim accounts for metrics that cannot or have not yet been normalised.
-We would therefore argue that when an indicator has been developed, the original variable is not presented.
-And when only variable estimates are given, extra care is given to describing the spatial aggregation pathway, since unfavourable local variable values (in terms of ecosystem condition) could have become smoothed out during aggregation, hiding potential environmental problems .
-Furthermore, Eurostat [@eurostat2022] encourages reporting of variable values.
+Generally, because of the normalisation process, indicators are more aligned with the purpose of ECA with describing ecosystem condition in normative terms, than variables.
+Although variable accounts are allowed under the SEEA EA framework, and are indeed a central feature, we believe they should be seen as interim accounts for metrics that cannot be, or have not yet been, normalised.
+We would therefore argue for not presenting the original variable once an indicator has been developed.
+And when only variable estimates are given, extra care is given to describing the spatial aggregation pathway, since unfavourable local variable values (in terms of ecosystem condition) could have become smoothed out during aggregation, hiding potential environmental problems.
 Sometimes, however, variable values cannot be aggregated spatially in any meaningful way (see **Example 3**).
@@
 In addition, unavoidable aggregation displacement (for good and bad) will also affect the indicator values and the inferences that can be drawn from these.
-If not careful, aggregation displacement may also cause errors to sneak into the indicator logic.
-A major determinant of the resulting aggregation displacement from the spatial aggregation of ecosystem condition indicators (@fig-agg1), is the choice about the order of the different steps in the normalisation and aggregation process, i.e.
+A major determinant of the resulting aggregation displacement from the spatial aggregation of ecosystem condition indicators (@fig-agg1) is the choice about the order of the different steps in the normalisation and aggregation process, i.e.
 the aggregation pathway (@fig-pathways).
@@
 Otherwise we get a negative displacement, making it practically impossible to reach an indicator value of 1.
-This means that neither EQR values, nor nEQR values, can be uncritically aggregated spatially to be used in a SEEA EA compliant in ECA.
+This means that neither EQR values nor nEQR values can be uncritically aggregated spatially to be used in a SEEA EA compliant in ECA.
 One solution, to enable the use of WFD indicators in ECAs, would be to change to pathway 6, with (1) scaling, (2) transforming, (3) aggregating and (4) truncating the values.
@@
 
-We believe it is a very desirable property that variable and indicator values are comparable across repeated ECAs and harmonised across assessment scales so that accounts from sub-areas should be able to recreate the account of the larger area.
-This can, however, fail to happen when ECAs use different aggregation pathways for the same indicator.
-Cases when two accounts or assessments use what is considered the same metric, but find different values for the same areas, is cause for both confusion, and could result in reduced general credibility for ECAs.
+We believe it is a very desirable property that variable and indicator values are comparable across repeated ECAs and harmonised across assessment scales so that accounts from sub-areas can be used to recreate the account of the larger area.
+This can, however, be difficult or even impossible when ECAs use different aggregation pathways for the same indicator.
+Cases when two accounts or assessments use what is considered the same metric, but find different values for the same areas, cause confusion and can potentially result in reduced general credibility for ECAs.
 
@@
 This is indeed a desirable property, although not always possible.
-**Example 4** shows a case where the spatial aggregation of an indicator uses area weighs that are different from the total extent of the EA (which is what is recommended by SEEA EA).
-Such alternative weights can be appropriate when the indicator only covers part of the ecosystem covered in the account, for example if it covers only one of many habitat types.
-In these cases, each indicator in an index might require different weights when spatially aggregated, but this is of coarse not possible after the indicators are combined into a single index.
+**Example 4** shows a case where the spatial aggregation of an indicator uses area weights that are different from the total extent of the EA (which is what is recommended by SEEA EA).
+Such alternative weights can be appropriate when the indicator only covers part of an ecosystem, for example if it covers only one of many habitat types.
+In these cases, each indicator in an index might require different weights when spatially aggregated, but this is of course not possible after the indicators are combined into a single index.
 
-A similar aggregation commutativity can be envisioned for step-wise aggregation of a single indicator, where it does not matter if the original data is aggregated directly from the original measurement scale to the EAA level, or via some intermittent scale, such as ecosystem assets or larger regions.
+A similar aggregation commutativity can be envisioned for step-wise aggregation of a single indicator, where it does not matter if the original data is aggregated directly from the original measurement scale to the Ecosystem Accounting Area (EAA) level, or via some intermittent scale, such as ecosystem assets or larger regions.
 This is possible when aggregating metrics using area weighting, and when one is aware of the need for post-stratification when using stratified data sampling.
-However, **Example 4** demonstrates two factors that lead to this indicator-level commutativity is not being fulfilled in this case.
+However, **Example 4** demonstrates two factors entailing that indicator-level commutativity is not being fulfilled in this case.
 
-First, some indicators are possible normalise at the measurement scale, i.e. the finest resolution of the raw data.
-But others may need to be (pre-) aggregated, for example if the variable is designed to depict a density or a proportion that retains to a spatial unit.
+Firstly, some indicators are possible to normalise at the measurement scale, i.e. the finest resolution of the raw data.
+But others may need to be (pre-) aggregated, for example if the variable is designed to depict a density or a proportion that refers to a spatial unit.
 In Example 4, there is no obvious definition of the ecosystem assets, and so one must first define those before calculating the variable values for each of them.
-It turn out that in this case the indicator value at the EAA level is not insensitive to the spatial delineation of the EAs.
+It turns out that in this case the indicator value at the EAA level is sensitive to the spatial delineation of the EAs.
 
@@
 **Example 4 -- Glacier area**\
-*Variable*: Extent of glacier area *anno* 2018-2019 (@fig-mapsGlaciers; A).
+*Variable*: Average extent of glacier area in the years 2018-2019 (@fig-mapsGlaciers; A).
 
-*Reference levels*: X~100~ is defined as the extent of glacier area *anno* 1947-1985.
+*Reference levels*: X~100~ is defined as the average extent of glacier area in the period 1947-1985.
 X~0~ is an absence of glaciers, and X~60~ is a 90% loss of glaciers.
@@
 
-The point estimates for the variable and the reference levels for each EA was then turned into probability distributions by sampling from a normal distribution with a standard deviation equal to 3% of the mean (the uncertainty reported for the maps).
+The point estimates for the variable and the reference levels for each EA were then turned into probability distributions by sampling from a normal distribution with a standard deviation equal to 3% of the mean (the uncertainty reported for the maps).
 To further aggregate the data to geographical regions (@fig-mapsGlaciers; C), these distributions were resampled with replacement and with an *n* corresponding to the extent of glaciers in the historical map, i.e.
@@
 104].
-However, in this case the weighs were not the total area (i.e. the alpine area, since that is the ecosystem that this indicator is designed for).
+However, in this case the weights were not the total area (i.e. the alpine area, since that is the ecosystem that this indicator is designed for).
 Instead the weights were defined by the extent of glaciers in the historic map.
 This implied that EAs with no glaciers in the historic map were coded as *NA* and did not contribute to the EAA estimate, with the argument that the indicator is not valid or relevant for those areas.
-It also meant that areas inside the EA with no initial glacier cover was seen as similarly non-relevant to the indicator.
+It also meant that areas inside the EA with no initial glacier cover were seen as similarly non-relevant to the indicator.
 Alternatively, using alpine extent as the weight during the spatial aggregation caused the indicator values to become considerably lower (@fig-ridge).
-This was partly due to cases where EAs with large alpine extent, but marginal glacier extent in the historic map, had lost all its glaciers in recent time.
-These EA would then get a very low indicator value and also weight very heavy towards the EAA estimate, even though the actual change in absolute glacier extent for that EA was very little.
-It was therefore argued that the area weight should be defined based on the area within the alpine ecosystem for which the indicator was considered relevant, and not defined base on the total EA extent.
+This was partly due to cases where EAs with large alpine extent, but marginal glacier extent in the historic map, had lost all their glaciers in recent times.
+These EAs would then get a very low indicator value and also weigh very heavily in the EAA estimate, even though the actual change in absolute glacier extent for that EA was very little.
+It was therefore argued that the area weight should be defined based on the area within the alpine ecosystem for which the indicator was considered relevant, and not defined based on the total EA extent.
 
@@
 In @fig-pathways we show some aggregation pathways that we have come across, and some that we see as potential new pathways.
-The examples are non-exhaustive, and serve mainly as illustration.
+The examples are non-exhaustive, and serve mainly as illustrations.
 We go on to describe how the aggregation pathways can formally, and succinctly, be described using the names of these operations.
@@
 
-Pathway 5 is a common pathway in the WFD, which does not include spatial aggregation of indicator values.
+Pathway 5 is a common pathway in the WFD, which does not practice spatial aggregation of indicator values.
 Instead one reports the number of water bodies in each condition class (spatial statistics).
-Pathway 6 is a suggested pathway for handling WFD indicator in ECAs (see **Example 3**), where truncation is postponed until after the spatial aggregation for in that way to allow overshooting values to compensate for low indicator values elsewhere (what we refer to as compensation).
+Pathway 6 is a suggested pathway for handling WFD indicator in ECAs (see **Example 3**), where truncation is postponed until after the spatial aggregation in order to allow overshooting values to compensate for low indicator values elsewhere (what we refer to as compensation).
 
-Pathway 7 describe an alternative to pathway 1 for a model-assisted aggregation pathway, where the original sample is the result of a defined sampling strategy, but where a model is used to project (extrapolate or interpolate, depending on the case) values to other spatial points or locations that were not sampled.
+Pathway 7 describes an alternative to pathway 1 for a model-assisted aggregation pathway, where the original sample is the result of a defined sampling strategy, but where a model is used to project (extrapolate or interpolate, depending on the case) values to other spatial points or locations that were not sampled.
 The resulting map (field of values) is then normalised, (re)sampled (which could imply taking all the discrete population units (e.g. all rivers), and finally aggregated.
@@
 
-# Recommendation {#sec-rec}
+# Recommendations {#sec-rec}
 
@@
 ECAs should be accompanied by detailed indicator documentation where a precise description of the steps in the aggregation process is presented.
-Verbose descriptions can be supplemented or replaced by a standardised short-hand notation that we shall present now.
+Verbose descriptions can be supplemented or replaced by a standardised short-hand notation that we present in the following:
 We suggest using the terms as described in @fig-term and elsewhere in the paper: scale, truncate (abbreviated to *trunc.*), transform (*trans.*), spatial aggregation (*sp.
@@
 Note also that in the last example, normalisation only involved a scaling step.
-The description of indicator pathways can be made even easier by referencing specific aggregation pathways by number, as we have done in this paper using @fig-pathways in this paper.Methods used to change either the cover or the extent of the indicator data (see @sec-term) can also be included in the shorthand notation scheme.
+The description of indicator pathways can be made even easier by referencing specific aggregation pathways by number, as we have done in this paper using @fig-pathways in this paper.
+Methods used to change either the cover or the extent of the indicator data (see @sec-term) can also be included in the shorthand notation scheme.
 The three operations we conceive as possible options are extrapolation (extr.), interpolation (inter.), and (sub)sampling.
@@
 Besides making the indicator workflow itself more transparent and reproducible, we believe this recommendation would help raise the general awareness about aggregation displacement in EAs, both among users and developers.
-It will make the interpretation of ECAs less prone to misunderstanding and miss-use, as well as making troubleshooting easier, for example when trying to identify the reason why two ECAs produce dissimilar indicator values from the same underlying data.
+It will make the interpretation of ECAs less prone to misunderstanding and misuse, as well as making troubleshooting easier, for example when trying to identify the reason why two ECAs produce dissimilar indicator values from the same underlying data.
 
@@
 Normalising variables early in the aggregation pathway means that you are aggregating normative measures of condition, which is generally what you want in ECAs (@sec-norm).
-But same as for variables and indicators, reference levels also have a spatial resolution, and they should only be used to normalise variables when the variable is at a scale which is relevant to the way the reference levels are defined, i.e.
+But just like variables and indicators, reference levels also have a spatial resolution, and they should only be used to normalise variables that are at a scale which is relevant to the way the reference levels are defined, i.e.
 they are at a commensurable scale (**Example 1**).
@@
 Yet different versions are often referred to by the same common name.
-This causes confusion about which indicator version is being used, and thus making it difficult to make out which aggregation pathway that has been used.
+This causes confusion about which indicator version is being used, and thus making it difficult to make out which aggregation pathway has been used.
 We recommend making use of stable and unique indicator IDs.
@@
 When spatially aggregating indices it is no longer possible to use different area weights to each single indicator.
-If all indicators making up n index are spatially aggregated using the same weights, typically the extent of the spatial units, then the aggregation operations will be commutative and it will not matter if you aggregate vertically then horisontally , or the other way around (\@fig-pathways; Pathway 8 and 9, respectively).
+If all indicators making up an index are spatially aggregated using the same weights, typically the extent of the spatial units, then the aggregation operations will be commutative and it will not matter if you aggregate vertically then horizontally, or the other way around (@fig-pathways; Pathway 8 and 9, respectively).
 However, indicators are not always valid for the entire ecosystem type that the account is made over.
-They can for example retain to only certain habitat or nature types within the ecosystem type, in which case one might want to use different weights when spatially aggregating the indicator that reflect the extent of only the relevant habitat or nature types, and not of the entire ecosystem (see **Example 4**).
+They can, for example, retain to only certain habitat or nature types within the ecosystem type, in which case one might want to use different weights when spatially aggregating the indicator that reflect the extent of only the relevant habitat or nature types, and not of the entire ecosystem (see **Example 4**).
 But when combined with other indicators into a common index, any subsequent aggregation of that index will give equal spatial weight to all indicators (see next paragraph for an exception to this).
-We therefore recommend caution, and to default to the solution of instead spatially aggregating indicators separately, and then re-calculating indices when needed (\@fig-pathways; Pathway 9).
-This will ensure that nested accounts will be comparable, i.e. that they will present the same index values independent of which spatial units that the indices first are calculated for.
+We therefore recommend caution, and to default to the solution of instead spatially aggregating indicators separately, and then re-calculating indices when needed (@fig-pathways; Pathway 9).
+This will ensure that nested accounts will be comparable, i.e. that they will present the same index values independent of which spatial units the indices first are calculated for.
 
-An alternative solution to ensure agreement across nested accounts is to use weights also when performing the vertical or thematic aggregation of indicators into indicators, where indicators are weighted respective of the area that they are valid over.
-With this approach one can down-weigh indicators that only cover only smaller areas (specific nature types etc.), or which have incomplete spatial coverage.
-When this weighting is already performed at the vertical aggregation stage, any subsequent horisontal aggregation does not need to account for the unique indicator weights again.
+An alternative solution to ensure agreement across nested accounts is to use weights also when performing the vertical or thematic aggregation of indicators into indicators, where indicators are weighted by the area that they are valid over.
+With this approach one can down-weigh indicators that only cover smaller areas (specific nature types etc.), or which have incomplete spatial coverage.
+When this weighting is already performed at the vertical aggregation stage, any subsequent horizontal aggregation does not need to account for the unique indicator weights again.
 However, weighting indicators in vertical aggregation opens a whole other can of worms, and therefore we do not offer this as a generic recommendation.

### 2025-09-12 manuscript/manuscript.qmd
edit abstract

@@
       - ref: nina
-  - name: BC!lint CzC:cz
+  - name: Bálint Czúcz
     email: balint.czucz@nina.no
@@
       - ref: nina
-  - name: ChloC) R. Nater
+  - name: Chloé R. Nater
     email: chloe.nater@nina.no
@@
 abstract: |
-  Ecosystem condition accounts (ECA) make use of variables and indicators to describe key ecosystem characteristics, reflecting their condition and deviations from a reference condition.
-These metrics are often spatially aggregated to represent larger areas, such as countries.
+  As the international community set targets and decide on actions to combate the current biodiversity crises they need tools for accurately describing the current situation ann to monitor trends over time.
+Ecosystem condition accounts (ECA) is one such tools that use variables and indicators to describe key ecosystem characteristics, reflecting their condition and deviations from a reference condition.
+Because the purpose is to inform decisions at relatively high political levels, these metrics are often spatially aggregated to represent larger areas, such as countries.
 However, spatial aggregation of information has the potential to alter the descriptive and normative interpretations one can make from these metrics.
 For example, aggregation displacement causes the information held in variables and indicators to diverge when these are aggregated spatially.
-This process is influenced also by the order of steps involved in normalising and aggregating variables.
+This process is influenced also by the order of steps involved in normalising and aggregating variables, i.e.
+the aggregation pathway.
+Although aggregation displacement and the type of aggregation pathway chosen for the indicator clearly impact both the indicator values and their interpretation, there are no clear guidelines or deliberation on these topics in the SEEA EA standard for ecosystem accounting.
 This paper outlines the consequences of different aggregation pathways, emphasising their impact on the credibility of ECAs, and how these are interpreted by users.
@@
 However, these operations are in practice often intertwined with spatial aggregation.
-Intertwinement can be explicit, for example when using model-based spatial estimation methods (e.g.
-geostatistics) and subsequent spatial sampling of this field of random values [@brus1997].
+Intertwinement can be explicit, for example when using model-based spatial estimation methods (e.g. geostatistics) and subsequent spatial sampling of this field of random values [@brus1997].
 In other cases, inter- or extrapolation can also be combined with spatial aggregation more inadvertently, for example by averaging point data and attributing this value to entire polygons, thus relying on a design-based estimation process [@gruijter1990].
@@
 |   | Example | Main illustrating point |
-|--|--------------------------|---------------------------------------------|
+|-------------|---------------------|--------------------------------------|
 | 1 | Bilberry | Commensurable spatial scales for variable and reference levels (@sec-match) |
@@
 | \-               | Plot 1 | Plot 2 | EA 1          |
-|------------------|--------|--------|--------------|
+|------------------|--------|--------|---------------|
 | Variable values  | 2      | 36     | 19            |
@@
 Depending on the random process used to select the sampling points we may need to account for the spatial structure (spatial autocorrelation) of the data when doing this aggregation.
-This can be addressed in several ways, such as model-assisted approaches [@brus2021], e.g.
-using regression models [@brus2000].
+This can be addressed in several ways, such as model-assisted approaches [@brus2021], e.g. using regression models [@brus2000].
 One can also add an intermediate aggregation step via homogeneous areas (domains), ecosystem assets or some smaller basal spatial units, to control for this spatial non-independence, and thus reduce the sampling bias and variance.
@@
 ECAs should be accompanied by detailed indicator documentation where a precise description of the steps in the aggregation process is presented.
-Verbose descriptions can be supplemented or replaced by a standardised short-hand notation that we present in the following:
-We suggest using the terms as described in @fig-term and elsewhere in the paper: scale, truncate (abbreviated to *trunc.*), transform (*trans.*), spatial aggregation (*sp.
+Verbose descriptions can be supplemented or replaced by a standardised short-hand notation that we present in the following: We suggest using the terms as described in @fig-term and elsewhere in the paper: scale, truncate (abbreviated to *trunc.*), transform (*trans.*), spatial aggregation (*sp.
 agg.*), thematic aggregation (*th.

### 2025-09-12 manuscript/manuscript.qmd
start table

@@
 library(kableExtra)
+library(flextable)
+library(ftExtra)
 ```
@@
 |   | Example | Main illustrating point |
-|-------------|---------------------|--------------------------------------|
+|------------|------------|------------------------------------------------|
 | 1 | Bilberry | Commensurable spatial scales for variable and reference levels (@sec-match) |

### 2025-09-12 manuscript/terms_table.qmd
start table

---
title: "Table"
format: pdf
bibliography: bibliography.bib

---

```{r setup}
#| include: false
library(knitr)
library(tidyverse)
library(flextable)
library(ftExtra)

```

```{r tbl-terms}
#| echo: false
terms <- tibble(
  X = as.character(NULL),
  Y = as.character(NULL)
) |>
  add_row(
    X = "Variables, indicators and metrics",
    Y = "We use the SEEA EA definitions of the terms variables, indicators (normalised variables) and metrics (referring to either a variable or an indicator) [@unitednations2024]. Variables are typically either in percentages or fractions, or as raw biophysical units, such as biomass or density. Indicators, on the other hand, are always on a unitless scale between 0 and 1, where 0 represents very poor (or the worst possible) condition, and 1 represents very good (or the best possible) condition. Sometimes indicators are calculated directly [e.g. @åström], but usually indicators are made by normalising variables based on two or more reference levels [@unitednations2024 §5.60]."
  ) |>

    add_row(
    X = "Reference condition (RC)",
    Y = "Reference condition is the state or condition of the ecosystem that we want to compare the current state against. It can be a purely theoretical construct or a real historical state. It does not need to be defined with any real data."
  ) |>
     add_row(
    X = " Reference levels (RL)",
    Y = "Reference levels (also called reference values) are variable values that act as anchoring points to specific values on the indicator scale, and are used to convert variables into indicators. This definition is different from that in SEEA EA, but at least conceptually it is the same. There can be from two to many RLs per indicator and they are distinguished with a special nomeclature X~n~ where _n_ is the corresponding indicator value times 100. For example, X~0~ and X~100~ refer to variable values coded as 0 or 1 on the indicator scale, respectively (sometimes called the lower and upper RLs). The X~60~ can hold a special meaning if, like in the WFD and in some applications of SEEA EA, the value 0.6 on the indicator scale represents the boundary or threshold between good and reduced condition. Defining X~60~, or any other RL other than X~0~and X~100~, is one way to quantitatively describe a non-linear relationship between the variable and ecosystem condition (referred to as _Transformation_)."
  ) |>
    add_row(
    X = "Spatial aggregation",
    Y = "Spatial (sometimes called horizontal) aggregation is the process of synthesising (e.g. averaging) data from different sub-areas and taking this new value as the value for all sub-areas. In related field of study this operation may be described as increasing the support or reducing the grain [@bierkens2002]. The term _upscaling_ is sometimes used synonymously to our definition of spatial aggregation, but it also has a wider tied to the ambiguous use of the terms scale to mean either grain or extent. It's common to lump spatial aggregation together with some types of extra- or interpolation operation, such as averaging point data and attributing this value to entire polygons [@gruijter1990], or spatial estimation methods (e.g. geostatistics) and subsequent spatial sampling [@brus1997]. However, for the detailed deliberation in this study the distinction between these is important to note."
  ) |>
    add_row(
    X = "Temporal aggregation",
    Y = "Refers to operations involved in combining data from different time points or periods to represenet a single temporal period (e.g. an accounting period or a growing season)."
  ) |>
    add_row(
    X = "Thematic agrgegation",
    Y = "Also called vertical aggregation. In the case of ECA, this refers to an aggregation of two or more indicators into a composite index or indices."
  ) |>
    add_row(
    X = "",
    Y = ""
  ) |>

  rename(Term = X, "Definition and deliberation" = Y)

flextable(terms, cwidth = c(1, 5)) |>
  #autofit() |>
  ftExtra::colformat_md()
```

### 2025-09-15 manuscript/manuscript.qmd
working on table

@@
 abstract: |
-  As the international community set targets and decide on actions to combate the current biodiversity crises they need tools for accurately describing the current situation ann to monitor trends over time.
+  Ecosystem integrity is declining world-wide, and the international community is setting targets and deciding on actions to combate the current biodiversity crises.
+For this to be effective, they need tools for accurately describing the current situation ann to monitor trends over time.
 Ecosystem condition accounts (ECA) is one such tools that use variables and indicators to describe key ecosystem characteristics, reflecting their condition and deviations from a reference condition.
@@
 
-# Reasons for normalising variables? {#sec-norm}
+# Reasons for normalising variables {#sec-norm}
 

### 2025-09-15 manuscript/terms_table.qmd
working on table

@@
 averaging) data from different sub-areas and taking this new value as the value for all sub-areas.
-In related field of study this operation may be described as increasing the support or reducing the grain [@bierkens2002].
-The term _upscaling_ is sometimes used synonymously to our definition of spatial aggregation, but it also has a wider tied to the ambiguous use of the terms scale to mean either grain or extent.
+In related fields of study this operation may be described as _increasing the support_ or _reducing the grain_ [@bierkens2002].
+The term _upscaling_ is sometimes used synonymously to our definition of spatial aggregation, but it also has a wider definition tied to the ambiguous use of the terms scale to mean either grain or extent.
 It's common to lump spatial aggregation together with some types of extra- or interpolation operation, such as averaging point data and attributing this value to entire polygons [@gruijter1990], or spatial estimation methods (e.g.
@@
     add_row(
+    X = "Normalisation",
+    Y = "This is the mathematical process of turning a condition variable into an indicator _sensu_ SEEA EA. It consists of three distinct steps: scaling, truncation, and transformation.
+     **Scaling** a variable means to recode a variable to a unitless unbound scale based on two reference levels, a lower (X~0~) and an upper (X~100~), that define the variable values to be coded as zero and one on the indicator scale, respectively.
+    **Truncation** means to recode scaled variable above 1 to become 1, and values below 0 to become 0.**Transformation** means to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
+This is commonly done by anchoring specific variable values to predefined class boundaries (sensu WFD) which we call reference levels [@czúcz; @jakobsson_setting_2020], but may also be done without explicit reference levels, for example by using a sigmoid or exponential transformation [@mienna2024].
+    Normalisation is frequently referred to as **rescaling** in the SEEA EA literature, but we avoid  this terms because of the need to provide better differentiation between the three steps of the normalisation process, where scaling is one of them."
+  ) |>
+    add_row(
+    X = "Categorisation",
+    Y = "Means to divide the indicator scale into discrete fractions, either based on reference levels, or some other rule (e.g. equal range for each class)."
+  ) |>
+     add_row(
+    X = "Inter- and extrapolation",
+    Y = "Interpolation means to _fill in the gaps_, or to increase the density of the data.
+Extrapolation means to expand the spatial extent of the data or beyond the current data range.
+There is often no clear boundary between these two operations when applied in ECA as an area may for example be within the geographical range of the current data, but outside the environmental range.
+Many different methods are employed to do inter- or extrapolation within ECAs, such as statistical models (e.g.
+regression models) or geostatistical model (e.g.
+kriging).
+These methods are more or less sophisticaled ways to essentially _spread_ or learn from the available data to give values to areas where there is no data.
+This is distinct from the process and purpose of spatial aggregation, which seeks to simplify data."
+  ) |>
+      add_row(
+    X = "The ecological fallacy",
+    Y = "The ecological fallacy [@robinson2009] is the error of assuming that relationships or patterns observed for groups (at the aggregate level) must also hold true for the individuals within those groups.
+In EA, aggregated data will mask differences within regions and lead some people to interpret this as regions being more homogenous then they really are."
+  ) |>
+  add_row(
+    X = "The Modifiable Area Unit Problem (MAUP)",
+    Y = "MAUP describes the sensitivity of spatially aggregated information to the choice of spatial units used for data collection or aggregation [@openshaw1984, @wong2009].
+These units can be arbitrary (*modifiable*), and different sizes and configuration of units can have large effects on the outcome of aggregation."
+  ) |>
+ add_row(
+    X = "Spatial compensation",
+    Y = "In this context, spatial compensation refers specifically to the ability of variable values above X~100~ in some areas to compensate for values below X~100~ elsewhere (assuming here that X~0~ < X~100~).
+Whether or not to allow spatial compensation in spatial aggregation has a large impact on the value and interprettaion of indicator values."
+  ) |>
+     add_row(
+    X = "Aggregation dissplacement",
+    Y = ""
+  ) |>
+     add_row(
+    X = "Ecosystem assets and ecosystem accounting areas",
+    Y = ""
+  ) |>
+    add_row(
     X = "",
@@
 flextable(terms, cwidth = c(1, 5)) |>
-  #autofit() |>
-  ftExtra::colformat_md()
+  ftExtra::colformat_md() |>
+  valign(j = 1, valign = "top", part = "all")
 ```

### 2025-09-15 manuscript/manuscript.qmd
table of terms

@@
 abstract: |
-  Ecosystem integrity is declining world-wide, and the international community is setting targets and deciding on actions to combate the current biodiversity crises.
-For this to be effective, they need tools for accurately describing the current situation ann to monitor trends over time.
+  In face of the ongoing nature crisis, the international community is setting targets and deciding on actions to combate the current biodiversity crises.
+For this to be effective they need tools to accurately describe the current situation and to monitor trends in ecosystems over time.
 Ecosystem condition accounts (ECA) is one such tools that use variables and indicators to describe key ecosystem characteristics, reflecting their condition and deviations from a reference condition.

### 2025-09-15 manuscript/terms_table.qmd
table of terms

@@
   ) |>
-
   add_row(
@@
 The X~60~ can hold a special meaning if, like in the WFD and in some applications of SEEA EA, the value 0.6 on the indicator scale represents the boundary or threshold between good and reduced condition.
-Defining X~60~, or any other RL other than X~0~and X~100~, is one way to quantitatively describe a non-linear relationship between the variable and ecosystem condition (referred to as _Transformation_)."
-  ) |>
-    add_row(
-    X = "Spatial aggregation",
-    Y = "Spatial (sometimes called horizontal) aggregation is the process of synthesising (e.g.
-averaging) data from different sub-areas and taking this new value as the value for all sub-areas.
-In related fields of study this operation may be described as _increasing the support_ or _reducing the grain_ [@bierkens2002].
-The term _upscaling_ is sometimes used synonymously to our definition of spatial aggregation, but it also has a wider definition tied to the ambiguous use of the terms scale to mean either grain or extent.
-It's common to lump spatial aggregation together with some types of extra- or interpolation operation, such as averaging point data and attributing this value to entire polygons [@gruijter1990], or spatial estimation methods (e.g.
-geostatistics) and subsequent spatial sampling [@brus1997].
-However, for the detailed deliberation in this study the distinction between these is important to note."
-  ) |>
-    add_row(
-    X = "Temporal aggregation",
-    Y = "Refers to operations involved in combining data from different time points or periods to represenet a single temporal period (e.g. an accounting period or a growing season)."
-  ) |>
-    add_row(
-    X = "Thematic agrgegation",
-    Y = "Also called vertical aggregation. In the case of ECA, this refers to an aggregation of two or more indicators into a composite index or indices."
+Defining X~60~, or any RL other than X~0~ and X~100~, is one way to quantitatively describe a non-linear relationship between the variable and ecosystem condition (referred to as _Transformation_)."
   ) |>
@@
   add_row(
+    X = "Aggregation displacement",
+    Y = "Referring to the systematic shift that occurs to the relative values of variables and their corresponding indicators when they are spatially aggregated.
+When we take the average of two variables, each of which have attribute values (i.e.
+indicator values) associated with them via some non-linear function (e.g.
+truncation or transformation), then the attribute of the average variable value differs from the average of the attributes of the original variable.
+@rastetter1992 referred to this as aggregation error and used it to describe some of the dangers associated with upscaling in the sense of using precise fine-scale information to explain or predict large scale phenomena.
+However, in ECAs, the difference between variables and indicators is an intentional one, and we therefore choose to refer to what is essentially the same mathematical phenomenon with the less derogatory term aggregation displacement."
+  ) |>
+  add_row(
+    X = "Spatial aggregation",
+    Y = "Spatial (sometimes called horizontal) aggregation is the process of summarising data from different sub-areas and taking this new value as the value for all constituent sub-areas.
+In means going from fine-resolution data to less fine or coarse resolution data.
+In related fields of study this operation may be described as _increasing the support_ or _reducing the grain_ [@bierkens2002].
+Different summary functions can be used, including the area-weighted arithmetic mean, which is required under when aggregation indicator values in SEEA EA.
+The term _upscaling_ is sometimes used synonymously to our definition of spatial aggregation, but it also has a wider definition tied to the ambiguous use of the terms scale to mean either grain or extent.
+It's common to lump spatial aggregation together with some types of extra- or interpolation operation, such as averaging point data and attributing this value to entire polygons [@gruijter1990], or spatial estimation methods (e.g.
+geostatistics) and subsequent spatial sampling [@brus1997].
+However, for the detailed deliberation in this study it is important to note the distinction between these."
+  ) |>
+  add_row(
+    X = "Temporal aggregation",
+    Y = "Refers to operations involved in combining data from different time points or periods to represent a single temporal period (e.g. an accounting period or a growing season)."
+  ) |>
+  add_row(
+    X = "Thematic aggregation",
+    Y = "Also called vertical aggregation. In the case of ECA, this refers to an aggregation of two or more indicators into a composite index (_plural_: indices)."
+  ) |>
+  add_row(
     X = "Inter- and extrapolation",
@@
 Many different methods are employed to do inter- or extrapolation within ECAs, such as statistical models (e.g.
-regression models) or geostatistical model (e.g.
+regression models) or geostatistical models (e.g.
 kriging).
@@
 This is distinct from the process and purpose of spatial aggregation, which seeks to simplify data."
+  ) |>
+  add_row(
+    X = "Ecosystem assets and ecosystem accounting areas",
+    Y = "Ecosystem assets in SEEA EA are the smallest spatial units for which information is shought, made up of contiguous spaces of a specific ecosystem type.
+The ecosystem accounting area is the _'geographical territory for which an ecosystem account is compiled'_ [@unitednations2024]"
   ) |>
@@
   ) |>
-     add_row(
-    X = "Aggregation dissplacement",
-    Y = ""
-  ) |>
-     add_row(
-    X = "Ecosystem assets and ecosystem accounting areas",
-    Y = ""
-  ) |>
-    add_row(
-    X = "",
-    Y = ""
-  ) |>
 

### 2025-09-16 manuscript/terms_table.qmd
child qmd

@@
----
-title: "Table"
-format: pdf
-bibliography: bibliography.bib
-
----
-
-```{r setup}
+```{r setup2}
 #| include: false
@@
 #| echo: false
+#| tbl-cap: "Key terms and concepts as they are used in this paper."
 terms <- tibble(
@@
 
-  rename(Term = X, "Definition and deliberation" = Y)
+  rename("Term or concept" = X, "Definition and deliberation" = Y)
 
-flextable(terms, cwidth = c(1, 5)) |>
-  ftExtra::colformat_md() |>
-  valign(j = 1, valign = "top", part = "all")
+flextable(terms) |>
+  width(j = 1, width = 1) |>     # Relative width
+  width(j = 2, width = 5) |>     # Relative width  
+  valign(j = 1:2, valign = "top", part = "all") |>
+  ftExtra::colformat_md(j = 2)
+
 ```

### 2025-09-17 manuscript/manuscript.qmd
child qmd
further updates

@@
 abstract: |
-  In face of the ongoing nature crisis, the international community is setting targets and deciding on actions to combate the current biodiversity crises.
+  In face of the ongoing nature crisis, the international community is setting targets and deciding on actions to combat the current biodiversity crises.
 For this to be effective they need tools to accurately describe the current situation and to monitor trends in ecosystems over time.
@@
 The world is facing a nature crisis of a magnitude that requires urgent political decision making and international coordination [@ipbes_global_2019].
-Ecosystem accounting (EA) has been fronted as a framework for providing a knowledge base for this.
+Ecosystem accounting has been fronted as a framework for providing a knowledge base for this.
 EAs are environmental assessments aimed at communicating on the status and trends in the extent and condition of ecosystems at a scale and in a format that is relevant for decision makers [@edens2022, @comte2022].
-For ecosystem condition accounting (ECA) and assessments, this often requires that metrics[^1] used to describe the condition of ecosystems are aggregated spatially to produce a single value for a larger area.
+For ecosystem condition accounting (ECA) and assessments, this often requires that **metrics** (terms in bold are explained in @tbl-terms) used to describe the condition of ecosystems are **aggregated spatially** to produce a single value for a larger area.
 This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something that is relevant to decision makers and potentially actionable at a political level [@king2024].
@@
 
-[^1]: We use the SEEA EA definitions of the terms variables, indicators (normalised variables) and metrics (referring to either a variable or an indicator) [@unitednations2024]
-
-Although much attention has been devoted to the vertical aggregation of environmental indicators into composite indices [e.g.
-@langhans2014; @maes2020; @oecd_2008], as well as to the spatial aggregation of relatively unmodified or raw variables [e.g.
-@allain2018], there is a lack of explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting [but see @vanbeurden1999].
+Although much attention has been devoted to the **vertical aggregation** of environmental indicators into composite **indices** [e.g.
+@langhans2014; @maes2020; @oecd_2008], as well as to the spatial aggregation of relatively unmodified or raw **variables** [e.g.
+@allain2018], there is a lack of explicit deliberation on the special case of spatially aggregating highly modified **indicators** used for ecosystem condition accounting [but see @vanbeurden1999].
 This paper therefore focuses on explaining and providing guidelines for the spatial aggregation of ecosystem indicators in the context of the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) [@unitednations2024] - the most used ecosystem accounting framework in the world [@comte2022].
+There is is a jungle of terms for the topics of indicators and aggregation, which are used inconsistently across scientific fields and different frameworks for ecosystem assessments, and there is a clear need for a standardised terminology, at least with the SEEA EA field of practice.
+This study goes some way towards this by including a list of terms and their definitions in @tbl-terms and by suggesting a new reporting terminology for the different steps in spatial aggregation, i.e.
+the **aggregation pathway**.
 
@@
 We will focus on the SEEA EA condition accounts in this paper, although spatial aggregation is also a relevant topic for the other parts of the account, and for other types of indicator-based ecosystem assessments.
-In 2026, EU and EEA countries will be required to report ecosystem condition accounts to Eurostat and the European Commission (?), and these accounts will have to be spatially aggregated to national levels.
+In 2026, EU and EEA countries will be required to report ecosystem condition accounts to Eurostat and the European Commission [@legalReporting], and these accounts will have to be spatially aggregated to national levels.
 This underlines the importance of having a robust methodological framework for the spatial aggregation of data and information.
 
-This paper is aimed primarily at indicator developers, providing them with recommendations for how to choose the right spatial aggregation approach for variables and indicators, and how to report these choices in an accessible and transparent way.
+This paper is aimed primarily at indicator developers to provide them with recommendations for how to choose the best aggregation pathway for variables and indicators, and how to report these choices in an accessible and transparent way.
 Indicator developer subjectivity affects indicator interpretation [@allain2018] and thus ultimately the conclusion from ECAs.
@@
 It is therefore important to be aware of the role that indicator developers play in shaping the type of information and the narratives that are presented to decision makers.
-This paper will make it easier for indicator developers to avoid some common pitfalls that are associated with choosing aggregation pathways.
-We use four examples to illustrate some of our main points (@tbl-examples).
-In @sec-pathways we describe a new reporting terminology for spatial aggregation pathways, and finally, in @sec-rec we give five concrete recommendations aimed at indicator developers that will help ensure a sensible and informed approach to spatial scaling of ecosystem condition metrics.
+We use four real-world examples to illustrate some of the effects arising from choosing different aggregation pathways (@tbl-examples).
 
-## Terminology {#sec-term}
+{{< include terms_table.qmd >}}
 
-There is a jungle of terms used for similar metrics or operations across relevant theoretical frameworks for ecosystem assessments, such as the SEEA EA, WFD, MSFD, IBECA, OSPAR, PAEC, HD, CFP, and others.
-We see a need for a common, more standardised terminology across these fields of practice in order to ease communication and avoid misunderstandings.
-In the following, we define the terms as we use them in this paper.
+# Reasons for normalising variables {#sec-norm}
 
-We follow the definition in @bierkens2002 and treat spatial aggregation as the process of increasing the support (i.e. reducing the grain).
-We see this as a separate process to models that create spatially continuous data from point data by either increasing the extent (extrapolation) or cover (interpolation).
-However, these operations are in practice often intertwined with spatial aggregation.
-Intertwinement can be explicit, for example when using model-based spatial estimation methods (e.g. geostatistics) and subsequent spatial sampling of this field of random values [@brus1997].
-In other cases, inter- or extrapolation can also be combined with spatial aggregation more inadvertently, for example by averaging point data and attributing this value to entire polygons, thus relying on a design-based estimation process [@gruijter1990].
-The term “upscaling” is sometimes used synonymously to our definition of spatial aggregation, but it also has a wider use, related especially to the ambiguous use of the terms scale to mean either grain or extent.
+<!-- The information held by the indicator is also different from that of the variable since additional ecological knowledge can be introduced via the reference levels. -->
 
-SEEA EA distinguishes between indicators and variables, as do we.
-Variables are metrics describing ecosystem characteristics [@unitednations2024 §5.41].
-They are typically either in percentages or fractions, or as raw biophysical units, such as biomass or density.
-Indicators, on the other hand, are always on a unitless scale between 0 and 1, where 0 represents very poor (or the worst possible) condition, and 1 represents very good (or the best possible) condition.
-Sometimes indicators are calculated directly [e.g. @åström], but usually indicators are made by normalising variables based on two or more reference levels [@unitednations2024 §5.60].
-Although some variables may have normative properties (due to decisions about how they are chosen, defined, aggregated or presented [@allain2018]), the normalisation of variables into indicators makes them explicitly normative by introducing a clear directional interpretation, where increasing values always refer to an increase in condition in the direction of a reference condition.
-The information held by the indicator is also different from that of the variable since additional ecological knowledge can be introduced via the reference levels.
+Normalisation of ecosystem condition indicators serves at least three purposes.
+Firstly, it gives a normative interpretation of a variable, defining a good and a bad state, and simultaneously a directionality to say when something is getting better or worse over time [@czucz_selection_2021].
+Although some variables may have normative properties (due to decisions about how they are chosen, defined, aggregated or presented [@allain2018]), the **normalisation** of variables into indicators makes them explicitly normative by introducing a clear directional interpretation, where increasing values always refer to an increase in condition in the direction of a **reference condition**.
 The mathematical part of this normalisation contains several steps (@fig-term).
-These can be performed simultaneously, but for clarity we discuss them as separate steps that may be implemented in different order.
+Two or more **reference levels** are needed to convert variables into indicators.
 
-Variables can be scaled using a minimum of two reference levels, a lower and an upper, that define which variable values are to be coded as zero (X~0~) and as one (X~100~) on the indicator scale.
-Variables can, if needed, be truncated to produce a bound indicator scale between 0 and 1.
-Sometimes variables are transformed to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
-This is commonly done by anchoring specific variable values to predefined class boundaries (sensu WFD) which we call reference levels [@czúcz; @jakobsson_setting_2020], but may also be done without explicit reference levels, for example by using a sigmoid or exponential transformation [@mienna2024].
-Finally, the indicator scale can also be divided into discrete fractions, either based on additional reference levels, or some other rule (e.g. equal range for each class).
-We refer to this step as categorisation.
+Secondly, normalisation sets a limit to how much a high variable value in one place can **compensate** for a low value somewhere else, and vice versa [@pedersen2016].
+This is because **truncation** effectively means that when spatially aggregating an indicator, one is always aggregating the negative deviations from the upper reference level X~100~, and ignoring any positive deviations which would otherwise compensate for the negative ones (see **Example 2**).
+**Transformation** with asymptotes have the same effect as truncation, e.g. sigmoid transformations.
+It is important to note that this is by design, i.e.
+that X~100~ is set (or should be set) so that values above this limit (assuming X~100~ \> X~0~) do not represent any further increase in ecosystem condition (for an exception, see **Example 3**).
+Therefore, aggregation of truncated indicators summarises the estimated ecosystem condition over an area, and not the variable itself.
+The flip-side of this is that spatially aggregating the original variable values does not directly summarise information about ecosystem condition for that area.
+
+Thirdly, we normalise in order to standardise the indicator on the same unitless scale, making them comparable, so that we can perform thematic (i.e.
+vertical) aggregation [@jakobsson2021; @unitednations2024 §5.81].
+This is commonly referred to as *the* reason for normalising variables, not least because it is perhaps the least confusing one for many; nonetheless, as we have shown, it is not the only reason.
 
@@
 For example, if a truncated variable is assumed to have a linear relationship with the indicator scale, then this variable has been normalised even though it has not been transformed.
-The term rescale is a synonym to normalise, but we will only use the latter from here on.
+The term rescale is a synonym to normalise, but we only use the latter to avoid confusion with the specific operation of scaling.
 #| echo: false
@@
 
-Generally, spatial aggregation increases the clarity and accessibility of a metric because it simplifies or compresses the information, sometimes into just a single number (total aggregation).
+Generally, spatial aggregation increases the clarity and accessibility of a metric because it simplifies or compresses the information, sometimes into just a single number (**total aggregation**).
 This may be necessary in some cases of ecosystem accounting when information is required at a relatively high administrative level (e.g.
@@
 
-Sometimes total aggregation of a metric is required for it to actually fit its purpose and become relevant to what it is designed to describe.
+Sometimes aggregation of a metric is required for it to actually fit its purpose and become relevant to what it is designed to describe.
 For example, a hypothetical variable "*proportion of total land area above critical nitrogen load",* informed from a spatial map of nitrogen loads, would only make sense if data was aggregated across some defined area.
@@
 
-Data may also be only party aggregated, or the same metric may be presented at different levels of aggregation [@allain2018].
+Data may also be only **party aggregated**, or the same metric may be presented at different levels of aggregation [@allain2018].
 A minimum level of aggregation may be decided based on uncertainties in the data: high spatially resolute data may be imprecise at fine scales, leading to false sense of certainty and potential misinterpretation [@lyytimäki2013], but these same data seen over a larger area may be assumed to converge on some true value, making spatially aggregated data better suited for communication to users.
@@
 It is therefore common to partially aggregate the data and provide a coarser map where different colours are used to distinguish different values.
-Similarly, one may choose to aggregate such that the spatial resolution matches some meaningful delineation, notably ecosystem assets in SEEA EA, which makes it possible to easily compare the ecosystem condition between these areas.
+Similarly, one may choose to aggregate such that the spatial resolution matches some meaningful delineation, notably **ecosystem assets** in SEEA EA, which makes it possible to easily compare the ecosystem condition between these areas.
 
@@
 For example, if data is collected opportunistically with sampling sizes or sampling effort varying between areas, then performing partial spatial aggregation to some intermediate-sized, homogeneous areas, will give equal weight or importance to all areas regardless of how many data points are included in each.
-This approach represents a form of post-stratification that is also relevant to process data from hierarchical (e.g. blocked) sampling designs.
+This approach represents a form of **post-stratification** that is also relevant to process data from hierarchical (e.g. blocked) sampling designs.
 
 Spatial aggregation can have several consequences that may or may not be desirable, such as smoothing out spatial variability, which can mask local patterns and trends.
-This smoothing out makes the data and its inherent information less suitable for some use cases, such as depicting spatial variation that can be used to support local land-use planning, for aiding a debate about spatial conflicts or identification of high risk areas [@vanbeurden1999].
+This smoothing out makes the data and its inherent information less suitable for some use cases, such as depicting spatial variation that can be used to support local land-use planning, for aiding a debate about spatial conflicts, or for identification of high risk areas [@vanbeurden1999].
 
 |   | Example | Main illustrating point |
-|------------|------------|------------------------------------------------|
+|------------|------------|-------------------------------------------------|
 | 1 | Bilberry | Commensurable spatial scales for variable and reference levels (@sec-match) |
@@
 Similar to variables, the reference levels used for normalising variables also have a spatial assignment, which can be different from that of the variable.
-For example, variable values may exist for unique 10 × 10 m grid cells, but the reference levels may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire accounting area (e.g.
+For example, variable values may exist for unique 10 × 10 m grid cells, but the reference levels may be created with a different spatial scale in mind, for example municipalities, or they can be uniform across the entire **ecosystem accounting area** (e.g.
 a natural zero).
@@
 
-| \-               | Plot 1 | Plot 2 | EA 1          |
-|------------------|--------|--------|---------------|
-| Variable values  | 2      | 36     | 19            |
-| X~100~           | 20\*   | 20\*   | 20            |
-| Indicator values | 0.1\*  | 1.0\*  | 0.95 (0.55\*) |
+{{< include table_bilberry.qmd >}}
 
-: Dummy data for the bilberry example.
-Variable values (% cover of bilberry) are recorded for each vegetation plot, and aggregated (arithmetic mean) to the ecosystem asset (EA) level.
-The X~100~ value is defined for the EA level, but in this hypothetical case it is also dis-aggregated to each vegetation plot and used for normalising variable values also at the plot level.
-The dis-aggregated X~100~ values, and subsequent values that rely on these, are marked with an asterisk (\*).
-{#tbl-billbery}
-
-The bilberry indicator exemplifies different spatial resolutions for the variable and for the reference levels (@tbl-billbery).
+The bilberry indicator exemplifies different spatial resolutions for the variable and for the reference levels (@tbl-bilberry).
 The variable is recorded at the scale of vegetation plots.
@@
 When normalising at the plot scale in this case, one is essentially aggregating only the negative deviations from the upper reference level X~100~, and the aggregated indicator value becomes negatively displaced (0.55) relative to when aggregating variable values (0.9).
-Given that X~100~ is defined the way it is, this indicator should follow pathway 3 (@fig-agg1) by spatially aggregating the variable from plot scale to ecosystem asset to allow spatial compensation between plots.
+Given that X~100~ is defined the way it is, this indicator should follow pathway 3 (@fig-pathways) by spatially aggregating the variable from plot scale to ecosystem asset to allow spatial compensation between plots.
 At the scale of ecosystem assets, the variable and the reference levels are at a commensurable scale, and the variable can be normalised and potentially aggregated further.
@@
 The type of manipulation depends on the spatial resolution of the underlying data, as well as the size and configuration of the ecosystem assets.
-To elaborate on this issue we first make a distinction between two main types of data: spatially exhaustive data and sparse (i.e. spatially discontinuous) data..
+To elaborate on this issue we first make a distinction between two main types of data: spatially exhaustive data and sparse (i.e. spatially discontinuous) data.
 
-*Spatially exhaustive*, or wall-to-wall data, include remote sensing products, but also other types of pre-aggregated data, where sparse data are assigned to areal units.
+Spatially exhaustive, or wall-to-wall data, include remote sensing products, but also other types of pre-aggregated data, where sparse data are assigned to areal units.
 This last type of data is very common.
@@
 Also data such as satellite images, or maps derived from such, have some inherent pre-aggregation where values are averaged for each raster cell.
-Often there is little indicator developers can do to affect this pre-aggregation step, but it is important to be aware of it in light of the ecological fallacy (**Box 1**).
+Often there is little indicator developers can do to affect this pre-aggregation step, but it is important to be aware of it in light of the **ecological fallacy**.
 This issue will, for example, hamper the detection of true maximum or minimum values (used in *worst rule* (i.e.
-one out al out) aggregation, for example in the WFD) for the phenomena described by the variable, since these extremes are already smoothed out in the pre-aggregation step [@vanbeurden1999].
+one out al out) aggregation, for example in the Water Frame Directive - WFD), since these extremes are already smoothed out in the pre-aggregation step [@vanbeurden1999].
 This will clearly affect the suitability of such data for identifying areas of high risk (very degraded areas, for example).
@@
 
-*Sparse data* refers to datasets where variable values are assigned to singular points or isolated small areas.
+Sparse data refers to datasets where variable values are assigned to singular points or isolated small areas.
 In these cases, there will be a lot of area inside each ecosystem asset for which we have no direct information, and some ecosystem assets may have no data at all.
-To assign variable value to all areas (as is required in the SEEA EA framework), we need to do some statistical inference.
-Both design-based and model-based approaches are valid under SEEA EA.
+Spatial prediction can be used assign variable or indicator values to all areas (as is required in the SEEA EA framework).
+Both **design-based** and **model-based** approaches are valid under SEEA EA.
 Model-based approaches (e.g.
 interpolation) generally make variables easier to aggregate spatially since the spatial structure of the variables will already be accounted for via the prediction process.
-For example, if using kriging to create local predictions for a variable across an ecosystem asset, then the variable is considered random, and the sampling locations may be considered as fixed.
-Thus a simple systematic sampling of the field of values across the ecosystem asset will be valid, and give representative grounds for statistical inference [@brus2021].
+For example, if using kriging to create local predictions for a variable across an ecosystem asset, then a simple systematic sampling of the field of values across the ecosystem asset will be valid, and give representative grounds for statistical inference [@brus2021].
 However, transparency, validity and simplicity are all very desirable properties from ECAs, which favours a design-based approach that makes no modeling assumptions [@brus2021].
@@
 For design-based approaches, one can assign values to areas via simple aggregation functions used on data values that fall within a given ecosystem asset.
-Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric, or harmonic means[^2] [@unitednations2024 §5.54].
-In addition, non-paramteric bootstrapping [@kolstad_area_2025] and Bayesian updating [@kolstad_making_2024] can be used to possibly obtain better descriptions of the variation or the uncertainty in the sample.
-If the inclusion probabilities are known, or they can be estimated, then the Horvitz-Thomson estimator for the sum or the mean (and variance) of a spatial sample is appropriate and useful, especially when the data originated from a stratified sample (ADD REF).
+Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric, or harmonic means [@unitednations2024 §5.54].
+In addition, non-paramteric bootstrapping [@kolstad_area_2025] and Bayesian updating [@kolstad_making_2024] can be used to possibly obtain better descriptions of the uncertainty in the estimates.
+If the inclusion probabilities are known, or they can be estimated, then the Horvitz-Thomson estimator for the sum or the mean (and variance) of a spatial sample is appropriate and useful, especially when the data originated from a stratified sample [@horvitz1952].
 Depending on the random process used to select the sampling points we may need to account for the spatial structure (spatial autocorrelation) of the data when doing this aggregation.
-This can be addressed in several ways, such as model-assisted approaches [@brus2021], e.g. using regression models [@brus2000].
+This can be addressed in several ways, such as model-assisted approaches [@brus2021], e.g.
+using regression models [@brus2000].
 One can also add an intermediate aggregation step via homogeneous areas (domains), ecosystem assets or some smaller basal spatial units, to control for this spatial non-independence, and thus reduce the sampling bias and variance.
-
-[^2]: When using means across areas, area weighting is required when following SEEA EA.
 
@@
 In some cases the areal units are even decided politically, for example to administrative units, with little ecological justification for this choice.
-However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to as the modifiable area unit problem (MAUP; **Box 1**).
+However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to as the **modifiable area unit problem (MAUP)**.
 The MAUP can lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
@@
 
-::: callout-note
-**Box 1 -- Related concepts**\
+# Aggregation displacement and spatial compensation {#sec-agg}
 
-**Vertical aggregation**
+@rastetter1992 showed that when taking the average of two variables, each of which have attribute values (e.g.
+indicator values) associated with them via some non-linear function, then the attribute of the average variable value differs from the average of the attributes of the original variables.
+They referred to this as aggregation error and used it to describe some of the dangers associated with **upscaling** in the sense of using precise fine-scale information to explain or predict large scale phenomena.
+However, in our case, the difference between variables and indicators is an intentional one, and we therefore choose to refer to what is essentially the same mathematical effect with the less derogatory term **aggregation displacement** (@fig-agg1).
 
-It is common in the ECA community of practice to use the term aggregation about the process of aggregating normalised indicators into indices (i.e.
-thematic or vertical aggregation; @langhans2014), but in this paper we are primarily discussing spatial (i.e.
-horizontal) aggregation.\
-
-**Temporal aggregation**
-
-… Should we add some sentences on temporal aggregation, i.e. the operations involved in representing data with a temporal resolution that is finer (shorter) than the accounting frequency?\
-
-**Modifiable area unit problem**
-
-The modifiable area unit problem (MAUP) describes the sensitivity of spatially aggregated information to the choice of spatial units used for data collection or aggregation [@openshaw1984, @wong2009].
-These units can be arbitrary (*modifiable*), yet different sizes and configuration of units can have large effects on the outcome of aggregation.
-Administrative units are often used as the ecosystem assets in SEEA EA, yet they are in an ecological sense almost always arbitrary, and can also be subject to change over time.
-The original spatial units that data are collected at may be a pragmatic choice in much the same way.
-In addition, data may not be available in the original spatial grain, but may only be downloadable in pre aggregated form [@vanbeurden1999].
-All these cases lead to some form of *aggregation error* [@rastetter1992] where the aggregated information has a non-linear relationship with the true variable value.
-However, the issue of MAUP in SEEA EA deserves an in-depth analyses that we cannot provide here.\
-
-**Ecological fallacy**
-
-The ecological fallacy [@robinson2009] is the error of assuming that relationships or patterns observed for groups (at the aggregate level) must also hold true for the individuals within those groups.
-In EA, aggregated data will mask differences within regions and lead some people to interpret this as regions being more homogenous then they really are.
-:::
-
-# Reasons for normalising variables {#sec-norm}
-
-Normalisation of ecosystem condition indicators, as defined in @fig-term, serves at least three purposes.
-Firstly, it gives a normative interpretation of a variable, defining a good and a bad state, and simultaneously a directionality to say when something is getting better or worse over time [@czucz_selection_2021].
-Secondly, normalisation sets a limit to how much a high variable value in one place can compensate for a low value somewhere else, and vice versa [@pedersen2016].
-This is because truncation effectively means that when spatially aggregating an indicator, one is always aggregating the negative deviations from the upper reference level X~100~, and ignoring any positive deviations which would otherwise compensate for the negative ones (see **Example 2**).
-Transformation can sometimes also have this effect, e.g. sigmoid transformations.
-One reason to want to aggregate the negative deviations only, is because the X~100~ is set (or should be set) so that values above this limit (assuming X~100~ \> X~0~) do not represent any further increase in ecosystem condition (for an exception, see **Example 3**).
-Therefore, this way of aggregation summarises the estimated ecosystem condition over an area, and not the variable itself.
-In other words, normalising variables facilitate the spatial aggregation of our ecosystem condition estimates (i.e. the indicator values).
-The flip-side of this is that spatially aggregating variable values does not directly summarise information about ecosystem condition for that area.
-<!--# ref?
-Is this list of three exhaustive?
---> Thirdly, we normalise in order to standardise the indicator on the same unitless scale so that we can perform thematic (i.e.
-vertical) aggregation [@jakobsson2021; @unitednations2024 §5.81].
-This is commonly referred to as *the* reason for normalising variables, not least because it is perhaps the least confusing one for many; nonetheless, as we have shown, it is the only reason.
+In the statistical standard and recommended guidelines for ecosystem accounts [@unitednations2024; -@unitednations2022], there is no mention of aggregation displacement, and there seems to be little awareness in general of the ramification of choosing different spatial aggregation pathways for variables.
+We now describe two ways that aggregation displacement affects ECAs: firstly, by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators (@sec-discrep), and secondly, how it can cause confusion about the effects of choosing different aggregation pathways during indicator development (@sec-develop).
 
@@
 ```
-
-# Aggregation displacement and spatial compensation {#sec-agg}
-
-The spatial aggregation of data can lead to biases arising from a number of sources.
-For example, direct extrapolation based on uneven or spatially biased sampling leads to biased estimates [@kolstad_making_2024].
-We will, however, focus now on what we consider an extended case of aggregation error, *sensu* @rastetter1992.
-They showed that when we take the average of two variables, each of which have attribute values (e.g.
-indicator values) associated with them via some non-linear function, then the attribute of the average variable value differs from the average of the attributes of the original variables.
-@rastetter1992 referred to this as aggregation error and used it to describe some of the dangers associated with upscaling in the sense of using precise fine-scale information to explain or predict large scale phenomena.
-However, in our case, the difference between variables and indicators is an intentional one, and we therefore choose to refer to what is essentially the same mathematical effect with the less derogatory term aggregation displacement.
-
-In the statistical standard and recommended guidelines for ecosystem accounts [@unitednations2024; -@unitednations2022], there is no mention of aggregation displacement, and there seems to be little awareness in general of the ramification of choosing different spatial aggregation pathways for variables.
-We now describe two ways that aggregation displacement affects ECAs: firstly, by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators (@sec-discrep), and secondly, how it can cause confusion about the effects of choosing different aggregation pathways during indicator development (@sec-develop).
 
@@
 This hypothetical example (@tbl-wolf) shows how different the interpretation of a variable and an indicator can become when the information is spatially aggregated.
-When species-based indicators are aggregated to the EAA level, it is common for the general public to interpret this value (0.5) as the status for this species, for example the conservation status[^3].
+When species-based indicators are aggregated to the EAA level, it is common for the general public to interpret this value (0.5) as the status for this species, for example the conservation status[^1].
 However, for the EAA seen overall, the population size is actually equal to the expected population size in the reference condition.
@@
 
-[^3]:
+[^1]:
     ::: callout-note
@@
 agg.
-= spatial aggregation (arithemtic mean); trunc.
+= spatial aggregation (arithmetic mean); trunc.
 truncation; stat.
@@
 
-Pathway 2 involves aggregating the variable before normalising[^4].
+Pathway 2 involves aggregating the variable before normalising.
 This preserves the ability for very high or very low variable values to compensate for very low or very high values elsewhere, respectively.
-
-[^4]: In a parallel terminology, normalisation can be considered modeling, and spatial aggregation before and after normalisation can be referred to as pre or post processing, respectively [@vanbeurden1999].
 

### 2025-09-17 manuscript/table_bilberry.qmd
further updates

```{r}
#| include: false
library(tidyverse)
library(knitr)
library(kableExtra)
```

```{r dat}
#| echo: false
dat <- tibble(
    " " = c(
        "Variable values",
        "$X_{100}$",
        "Indicator values"),
    "Plot 1" = c(
        "2",
        "20*",
        "0.1*"),
    "Plot 2" = c(
        "36",
        "20*",
        "1.0*"),
    "EA" = c(
        "19",
        "20",
        "0.95 (0.55*)"))
```

```{r tbl-bilberry}
#| echo: false
#| tbl-cap: "Dummy data for the bilberry example. Variable values (% cover of bilberry) are recorded for each vegetation plot, and aggregated (arithmetic mean) to the ecosystem asset (EA) level. The X~100~ value is defined for the EA level, but in this hypothetical case it is also dis-aggregated to each vegetation plot and used for normalising variable values also at the plot level. The dis-aggregated X~100~ values, and subsequent values that rely on these, are marked with an asterisk (*)."

kbl(dat,
    format = "latex",
    booktabs = TRUE,
    escape = F
    ) 
```

### 2025-09-17 manuscript/terms_table.qmd
further updates

@@
     X = "Normalisation",
-    Y = "This is the mathematical process of turning a condition variable into an indicator _sensu_ SEEA EA. It consists of three distinct steps: scaling, truncation, and transformation.
+    Y = "This is the mathematical process of turning a condition variable into an indicator _sensu_ SEEA EA.
+It consists of three distinct steps: scaling, truncation, and transformation (see also @fig-term).
      **Scaling** a variable means to recode a variable to a unitless unbound scale based on two reference levels, a lower (X~0~) and an upper (X~100~), that define the variable values to be coded as zero and one on the indicator scale, respectively.
@@
 This is commonly done by anchoring specific variable values to predefined class boundaries (sensu WFD) which we call reference levels [@czúcz; @jakobsson_setting_2020], but may also be done without explicit reference levels, for example by using a sigmoid or exponential transformation [@mienna2024].
-    Normalisation is frequently referred to as **rescaling** in the SEEA EA literature, but we avoid  this terms because of the need to provide better differentiation between the three steps of the normalisation process, where scaling is one of them."
+    Normalisation is frequently referred to as **rescaling** in the SEEA EA literature, but we avoid  this terms because of the need to provide better differentiation between the three steps of the normalisation process, where scaling is one of them.
+Also, in a parallel terminology, normalisation can be considered modeling, and spatial aggregation before and after normalisation can be referred to as pre or post processing, respectively [@vanbeurden1999]."
   ) |>
@@
 In related fields of study this operation may be described as _increasing the support_ or _reducing the grain_ [@bierkens2002].
+Aggregation can be partial, resulting in two or more spatial areas with unique values, or total, if it results in just a single number summarising all the data.
 Different summary functions can be used, including the area-weighted arithmetic mean, which is required under when aggregation indicator values in SEEA EA.
@@
   add_row(
-    X = "Inter- and extrapolation",
-    Y = "Interpolation means to _fill in the gaps_, or to increase the density of the data.
-Extrapolation means to expand the spatial extent of the data or beyond the current data range.
+    X = "Spatial prediction",
+    Y = "It refers to any estimation of the value of a variable at an unobserved location, and therefore include both inter- and extraolation.
+**Interpolation** means to _fill in the gaps_, or to increase the density of the data.
+**Extrapolation** means to expand the spatial extent of the data or beyond the current data range.
 There is often no clear boundary between these two operations when applied in ECA as an area may for example be within the geographical range of the current data, but outside the environmental range.
-Many different methods are employed to do inter- or extrapolation within ECAs, such as statistical models (e.g.
+Many different methods are employed for spatial predictions within ECAs, such as statistical models (e.g.
 regression models) or geostatistical models (e.g.
@@
 This is distinct from the process and purpose of spatial aggregation, which seeks to simplify data."
+  ) |>
+  add_row(
+    X = "Design-based and model based prediction",
+    Y = "These are two fundamentally different ways of doing and thinking about spatialand statistical prediction.
+Design-based approaches relies on classical sampling theory and uses what is know about the sampling design to make spatial inferences.
+It is generally applied in more simple ways, such as with look-up tables, but model-assisted estimation using statistical models is also possible.
+Design-based approaches  produce things like the sample mean and sample variance and assumes the sampled values are correct (fixed), and uncertainty arises due to where we chose to sample (the locations are random).
+Model-based prediction uses geostatistical models, such as kriging and produces estimates such as the model mean or model variance.
+In this framework, the locations are considered fixed and the uncertainty arises due in the statistical model.
+Readers are referred to @brus2021 for more details."
   ) |>
@@
   ) |>
+  add_row(
+    X = "Post-stratfication",
+    Y = "A statistical technique used to correct for a biased sample by forcing the sample to match the main population in some key characteristcis.
+In ECAs it is common to have data that lack a clear probability based sampling design, and therefore end up with more samples from certain strata (e.g.
+some geographic region or within some shared environmental condition).
+Post-stratification seeks to remove this unevenness in the _post_-processing, e.g.
+by doing a step-wise spatial aggregation via the same strata."
+  )|>
 

### 2025-09-18 manuscript/manuscript.qmd
down to pathways

@@
 It is therefore important to be aware of the role that indicator developers play in shaping the type of information and the narratives that are presented to decision makers.
-We use four real-world examples to illustrate some of the effects arising from choosing different aggregation pathways (@tbl-examples).
+We use three examples to illustrate some of the effects arising from choosing different aggregation pathways.
 
@@
 # Reasons for normalising variables {#sec-norm}
-
-<!-- The information held by the indicator is also different from that of the variable since additional ecological knowledge can be introduced via the reference levels. -->
 
@@
 
-# Spatial aggregation of data in ecosystem condition accounting
+# The challenges and rationale for spatial aggregation in ecosystem condition accounting
 
@@
 
-|   | Example | Main illustrating point |
-|------------|------------|-------------------------------------------------|
-| 1 | Bilberry | Commensurable spatial scales for variable and reference levels (@sec-match) |
-| 2 | Wolves | Spatial compensation and discrepancy between variable and indicator (@sec-discrep) |
-| 3 | PTI | Compensation and the order of operations (@sec-pathways) |
-| 4 | Glaciers | Aggregation commutativity and alternative indicator weights (@sec-comp) |
-
-: Examples used in this paper, and how they illustrate specific points raised {#tbl-examples}
-
-## Matching the scale of variables and reference levels {#sec-match}
+<!-- ## Matching the scale of variables and reference levels -->
 
@@
 a natural zero).
-Variables should not be normalised based on reference levels unless they are at a commensurable spatial scale.
-The bilberry example (**Example 1**) shows that sometimes variables need to be spatially aggregated before they can be normalised.
+The bilberry example (**Example 1**) shows that, to ensure variables and reference levels are at a commensurable spatial scale, sometimes variables need to be partly spatially aggregated before they can be normalised.
+
+<!-- Some indicators are possible to normalise at the measurement scale, 
+i.e. the finest resolution of the raw data.
+Others, however, need to be (pre-) aggregated, 
+for example if the variable is designed to depict a density or a proportion that refers to a spatial unit. 
+Generally, variables that describe "the share of X in Y" 
+(e.g. the proportion of one type of "surface cover" within a larger area) 
+are particularly prone to issues regarding scale dependence of the reference levels.
+If you look at such a variable for a large area, then you can extract some meaningful and precise etimate about the proportion of X in Y, 
+but for increasingly smaller areas the variable become more binary, until you have either only Y or no Y present.
+At this zoom level the reference levels is unlikely to make much sense. 
+This stresses the need to define ecosystem assets that are appropriate in size. -->
 
@@
 **Example 1 -- Bilberry**\
+
+This example is to show what it means that the variable and the reference levels need to be at a comensurable spatial scale before doing the normalisition.
 
@@
 
-## Fitting data to ecosystem assets
+# Spatial prediction: fitting data to ecosystem assets
 
@@
 They referred to this as aggregation error and used it to describe some of the dangers associated with **upscaling** in the sense of using precise fine-scale information to explain or predict large scale phenomena.
+In ECA, the aggregation of both variables and indicators is common practice and raises this issue of aggregation error (@fig-agg1) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
 However, in our case, the difference between variables and indicators is an intentional one, and we therefore choose to refer to what is essentially the same mathematical effect with the less derogatory term **aggregation displacement** (@fig-agg1).
 
-In the statistical standard and recommended guidelines for ecosystem accounts [@unitednations2024; -@unitednations2022], there is no mention of aggregation displacement, and there seems to be little awareness in general of the ramification of choosing different spatial aggregation pathways for variables.
-We now describe two ways that aggregation displacement affects ECAs: firstly, by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators (@sec-discrep), and secondly, how it can cause confusion about the effects of choosing different aggregation pathways during indicator development (@sec-develop).
+Spatial aggregation leads to a divergence in the type of information held in variables and in their associated indicators, to a point where they become qualitatively very different (**Example 2**).
+Aggregated indicators reflect the average ecosystem condition, where areas in very good or very poor condition have limited ability to compensate for opposite extreme values elsewhere.
+The spatially aggregated variable on the other hand, may reflect something like a sum of individuals, an average of a population, or some other aspect where values above and below the reference levels are able to compensate for each other.
+The information held by the indicator is also different from that of the variable since additional ecological knowledge can be introduced via the reference levels.
+
+In the statistical standard and recommended guidelines for ecosystem accounts [@unitednations2024; -@unitednations2022], there is no mention of aggregation displacement, and there seems to be little awareness in general of its meening and the ramification of choosing different spatial aggregation pathways for variables.
+Indicators and variables are sometimes presented side-by-side in ecosystem condition assessments or accounts.
+This is also recommended practice following the SEEA EA [@unitednations2024; ch. 5.3.4].
 
@@
 
-## Discrepancies between variables and indicators {#sec-discrep}
-
-In ECA the aggregation of both variables and indicators is common practice and raises the issue of aggregation displacement (@fig-agg1) when variables have nonlinear relationships to the normalised indicators [@rastetter1992], which they typically do.
-This means that the spatial aggregation of metrics leads to a divergence in the type of information held in variables and in their associated indicators, to a point where they become qualitatively very different (Example 2).
-
-Indicators and variables are sometimes presented side-by-side in ecosystem condition assessments or accounts.
-This is also recommended practice following the SEEA EA [@unitednations2024; ch. 5.3.4].
-However, due to aggregation displacement, spatially aggregated indicators should not be interpreted as normalised versions of aggregated variables.
-Aggregated indicators reflect the average ecosystem condition, where areas in very good or very poor condition have limited ability to compensate for opposite extreme values elsewhere.
-The spatially aggregated variable on the other hand, may reflect something like a sum of individuals, an average of a population, or some other aspect where values above and below the reference levels are able to compensate for each other.
-
-Aggregated variables and the associated aggregated indicators cannot act as alternative evidence or interpretations for the state of nature.
-While this may initially seem like a democratic and transparent approach to reporting, this practice places a significant burden on both the end user and the indicator developer.
-Misuse can ensue, for example by taking the variable value and assigning it the same interpretation as the indicator value.
-At the very least, the developer must clarify to the user how the two metrics may have diverged during spatial aggregation and normalisation, complicating the comparison of their respective information.
-
 \clearpage
@@
 **Example 2 -- Wolves**\
+
+This hypothetical example (@tbl-wolf) shows how different the interpretation of a variable and an indicator can become when the information is spatially aggregated.
+More specifically, it shows shows how spatial compensation in the spatial aggregation of variables leads to aggregation displacement and a discrepancy between the variable and the indicator.
 
@@
 
-|                  | EA 1 | EA 2 | EA 3 | EAA aggregate |
-|------------------|------|------|------|---------------|
-| Variable values  | 20   | 5    | 0    | Sum = 25      |
-| X~100~           | 10   | 10   | 5    | Sum = 25      |
-| Indicator values | 1.0  | 0.5  | 0.0  | Mean = 0.5    |
+{{< include table_wolves.qmd >}}
 
-: Dummy data for the wolves example. EA = ecosystem asset (e.g. a municipality). EAA = ecosystem accounting area (e.g. country). {#tbl-wolf}
-
-This hypothetical example (@tbl-wolf) shows how different the interpretation of a variable and an indicator can become when the information is spatially aggregated.
-When species-based indicators are aggregated to the EAA level, it is common for the general public to interpret this value (0.5) as the status for this species, for example the conservation status[^1].
+When species-based indicators are aggregated to the EAA level, it is common for the general public to interpret this value (0.5) as the status for this species, for example the conservation status.
 However, for the EAA seen overall, the population size is actually equal to the expected population size in the reference condition.
@@
 
-[^1]:
-    ::: callout-note
-    It is sometimes worth going back to the variable, and see if it can be defined differently to avoid this kind of confusion.
-    Wolf numbers could in this example be converted to wolf density.
-    :::
-
-Generally, because of the normalisation process, indicators are more aligned with the purpose of ECA with describing ecosystem condition in normative terms, than variables.
-Although variable accounts are allowed under the SEEA EA framework, and are indeed a central feature, we believe they should be seen as interim accounts for metrics that cannot be, or have not yet been, normalised.
-We would therefore argue for not presenting the original variable once an indicator has been developed.
-And when only variable estimates are given, extra care is given to describing the spatial aggregation pathway, since unfavourable local variable values (in terms of ecosystem condition) could have become smoothed out during aggregation, hiding potential environmental problems.
-Sometimes, however, variable values cannot be aggregated spatially in any meaningful way (see **Example 3**).
-
-## Aggregation displacement in indicator development {#sec-develop}
+<!-- It is sometimes worth going back to the variable, and see if it can be defined differently to avoid this kind of confusion. Wolf numbers could in this example be converted to wolf density. -->
 
@@
 the aggregation pathway (@fig-pathways).
-If normalisation occurs early in the aggregation process, the amount of compensation, or off-setting, is limited (**Example 2**).
+If normalisation occurs early in the aggregation process, the amount of spatial compensation is limited (**Example 2**).
 However, sometimes complete or partial compensation is justified (**Example 1** and **3**).
@@
 
-# Agreement across nested ECAs {#sec-comp}
+<!-- SEEA EA uses the term aggregation commutativity for the property that the order of vertical and horizontal aggregation is irrelevant to the end result.
+A similar aggregation commutativity can be envisioned for step-wise aggregation of a single indicator, where it does not matter if the original data is aggregated directly from the original measurement scale to the Ecosystem Accounting Area (EAA) level, or via some intermittent scale, such as ecosystem assets or larger regions.
+This is possible when aggregating metrics using area weighting, and when one is aware of the need for post-stratification when using stratified sampling. -->
 
-We believe it is a very desirable property that variable and indicator values are comparable across repeated ECAs and harmonised across assessment scales so that accounts from sub-areas can be used to recreate the account of the larger area.
-This can, however, be difficult or even impossible when ECAs use different aggregation pathways for the same indicator.
-Cases when two accounts or assessments use what is considered the same metric, but find different values for the same areas, cause confusion and can potentially result in reduced general credibility for ECAs.
-
-We believe one solution to this is to acknowledge that the aggregation pathway can affect the indicators (and indices) so much, sometimes as much as variation in the variable itself, that the use of different pathways imply that metrics are named and treated uniquely, for example using different common names, but perhaps better still, different version numbers or indicator IDs.
-For each unique indicator the precise aggregation pathway should be specified as part of the indicator metadata.
-This would show ECA users if two accounts are using different metrics that cannot be assumed to behave similarly or report the same thing, even if they have the same common name and data source.
-
-SEEA EA uses the term aggregation commutativity for the property that the order of vertical and horizontal aggregation is irrelevant to the end result.
-This is indeed a desirable property, although not always possible.
-**Example 4** shows a case where the spatial aggregation of an indicator uses area weights that are different from the total extent of the EA (which is what is recommended by SEEA EA).
-Such alternative weights can be appropriate when the indicator only covers part of an ecosystem, for example if it covers only one of many habitat types.
-In these cases, each indicator in an index might require different weights when spatially aggregated, but this is of course not possible after the indicators are combined into a single index.
-
-A similar aggregation commutativity can be envisioned for step-wise aggregation of a single indicator, where it does not matter if the original data is aggregated directly from the original measurement scale to the Ecosystem Accounting Area (EAA) level, or via some intermittent scale, such as ecosystem assets or larger regions.
-This is possible when aggregating metrics using area weighting, and when one is aware of the need for post-stratification when using stratified data sampling.
-However, **Example 4** demonstrates two factors entailing that indicator-level commutativity is not being fulfilled in this case.
-
-Firstly, some indicators are possible to normalise at the measurement scale, i.e. the finest resolution of the raw data.
-But others may need to be (pre-) aggregated, for example if the variable is designed to depict a density or a proportion that refers to a spatial unit.
-In Example 4, there is no obvious definition of the ecosystem assets, and so one must first define those before calculating the variable values for each of them.
-It turns out that in this case the indicator value at the EAA level is sensitive to the spatial delineation of the EAs.
-
-Secondly, SEEA EA mentions several arithmetic or parametric aggregation functions, such as different means, median, min or max values.
+<!-- Secondly, SEEA EA mentions several arithmetic or parametric aggregation functions, such as different means, median, min or max values.
 It does not mention the possibility of bootstrap resampling [@jakobsson2021; @kolstad_making_2024].
 This method allows for using probability distributions to represent indicator values and uncertainties, and to use weighted resampling to create new joint probability distributions for the spatially aggregated indicator.
-In Example 4, point estimates for the variable and reference levels are replaced with probability distributions where a 3% uncertainty is introduced based on the reported mapping uncertainty of the raw data.
-The shape of the aggregated joint distributions depend on the spatial size and configuration of the spatial units where this uncertainty is introduced.
-
-::: callout-note
-**Example 4 -- Glacier area**\
-*Variable*: Average extent of glacier area in the years 2018-2019 (@fig-mapsGlaciers; A).
-
-*Reference levels*: X~100~ is defined as the average extent of glacier area in the period 1947-1985.
-X~0~ is an absence of glaciers, and X~60~ is a 90% loss of glaciers.
-
-This indicator [@kolstad_area_2025] uses historical data as the reference levels against which to normalise the variable.
-Both the variable and the reference levels are vector maps, but the metric itself needs to be numbers assigned to distinct ecosystem assets (EA).
-Seventy EAs (@fig-mapsGlaciers; B) were defined as unique combinations of geographical regions (@fig-mapsGlaciers; C) and bioclimatic regions.
-There is then some spatial aggregation done to get the summed glacier area for each EA.
-Of the 70 EAs, 29 had glaciers in the historical glacier map.
-The remaining EAs did not get assigned indicator values.
-
-The point estimates for the variable and the reference levels for each EA were then turned into probability distributions by sampling from a normal distribution with a standard deviation equal to 3% of the mean (the uncertainty reported for the maps).
-To further aggregate the data to geographical regions (@fig-mapsGlaciers; C), these distributions were resampled with replacement and with an *n* corresponding to the extent of glaciers in the historical map, i.e.
-using historical glacier extent as the area weights.
-The same bootstrapping procedure was done for aggregation to the country (EAA) level.
-
-In SEEA EA it says that "\[...\] the values recorded in an ecosystem condition variable account should be calculated as the area-weighted arithmetic mean of ecosystem assets belonging to the particular ecosystem type within the EAA" [@unitednations2024, p.
-104].
-However, in this case the weights were not the total area (i.e. the alpine area, since that is the ecosystem that this indicator is designed for).
-Instead the weights were defined by the extent of glaciers in the historic map.
-This implied that EAs with no glaciers in the historic map were coded as *NA* and did not contribute to the EAA estimate, with the argument that the indicator is not valid or relevant for those areas.
-It also meant that areas inside the EA with no initial glacier cover were seen as similarly non-relevant to the indicator.
-Alternatively, using alpine extent as the weight during the spatial aggregation caused the indicator values to become considerably lower (@fig-ridge).
-This was partly due to cases where EAs with large alpine extent, but marginal glacier extent in the historic map, had lost all their glaciers in recent times.
-These EAs would then get a very low indicator value and also weigh very heavily in the EAA estimate, even though the actual change in absolute glacier extent for that EA was very little.
-It was therefore argued that the area weight should be defined based on the area within the alpine ecosystem for which the indicator was considered relevant, and not defined based on the total EA extent.
-
-The spatial delineation of EAs can affect the indicator estimate at the EAA level.
-In this case the EAA estimates were not identical, but qualitatively similar (@fig-ridge).
-However, using larger EAs generally led to more narrow probability distributions which we believe is an artifact, and not a real reduction in uncertainty.
-In any case, this exemplifies how the MAUP and the choice of spatial units can affect the aggregated values.
-
-```{r fig-mapsGlaciers}
-#| fig-cap: 'Extent of glaciers in Norway can be visualised either as actuall occurences (A) or as the summed area for fifferent spatial units: B) ecosystem asset; C) gegrapical regions; D) the whole country.
-The legend is omitted for simplicity.
-The different maps exemplifies the modifiable area unit problem.
-From Kolstad (2025)'
-#| echo: false
-knitr::include_graphics("../img/maps.png")
-```
-
-```{r fig-ridge}
-#| fig-cap: 'Indicator values (probability distributions; vertical lines are quartiles) for the indicator _Glacier extent_ aggregated to EAA (country) level.
-The *indicator* is first normalised at the scale of the original EAs; *indicator_level2* is pre-aggregated to geographical regions before it it normalised, and; *indicator_level3* is normalised at the EAA scale with no further spatial aggregation.
-The difference between these three distributions show that the definition of the spatial untis and and especially the sice of the ecosystem assets does matter to the EAA estimate.
-*indicator_mtArea* is an indicator normalised at the EA scale (similar to *indicator*), but which is aggregted using the total alpine extent for each EA rather than the intial glacer extent as the weights.
-This approach puts a lot more emphasis on areas especially in the south of the country where there is a lot of alpine area but little intial glacer extent.
-Any loss of glacers, even if quite small in absolute terms, had a great effect on lowering the indicator estimates which where then propagated to the EAA level.'
-#| echo: false
-#| out-width: 500
-knitr::include_graphics("../img/ridge.png")
-```
-:::
+-->
 
@@
 
-<!--# a possible fifth recommendation is to not present aggregated variables and indicators side-by-side as alternative to choose from -->
-
 Based on the above discussion and examples, we here summarise five recommendations for developers of ecosystem condition indicators and assessments for avoiding some of the pitfalls from having a too casual approach to the choice of aggregation pathways.
@@
 
-Indicators (and variables) often exist in multiple versions, varying slightly in the raw data or in the methods used to produce the data or the metrics.
-Yet different versions are often referred to by the same common name.
+Indicators (and variables) often exist in multiple versions, varying slightly in the raw data or in the methods or aggregation pathway used to produce the data or the metrics.
+Yet different versions are often referred to by the same common name, for example in nested or repeated ECAs.
 This causes confusion about which indicator version is being used, and thus making it difficult to make out which aggregation pathway has been used.
-We recommend making use of stable and unique indicator IDs.
-This has for example already been implemented in ecRxiv, a GitHub-based publishing platform for ecosystem condition indicators [@Kolstad_ecRxiv].
+We need to acknowledge that the aggregation pathway can affect the indicators (and indices) so much that the use of different pathways imply that metrics are named and treated uniquely.
+We recommend making use of stable and unique indicator IDs and version numbers.
+This has for example already been implemented in ecRxiv, a GitHub-based publishing platform for ecosystem condition indicators used in the Norwegian national pilot ECA [@Kolstad_ecRxiv].
 On ecRxiv, indicator documentation follows a structured approach to both reporting and peer review (including code review).
@@
 
+### 6. Don't present variables and their respective indicators as alternative evidences to choose from
+
+Aggregation displacement causes a divergence in the type of information held in ecosystem condition variables versus indicators, 
+and spatially aggregated indicators should not be interpreted as normalised versions of aggregated variables.
+Therefore, aggregated variables and the associated aggregated indicators cannot act as alternative evidence or interpretations for the state of nature.
+While this may initially seem like a democratic and transparent approach to reporting, this practice places a significant burden on both the end user and the indicator developer.
+Misuse can ensue, for example by taking the variable value and assigning it the same interpretation as the indicator value.
+At the very least, the developer must clarify to the user how the two metrics may have diverged during spatial aggregation and normalisation, complicating the comparison of their respective information.
+Generally, because of the normalisation process, indicators are more aligned with the purpose of ECA with describing ecosystem condition in normative terms, than variables.
+Although variable accounts are allowed under the SEEA EA framework, and are indeed a central feature, we believe they should be seen as interim accounts for metrics that cannot be, or have not yet been, normalised.
+
 # Conclusion

### 2025-09-22 manuscript/manuscript.qmd
Hannos new example

@@
 
+```{=html}
 <!-- Some indicators are possible to normalise at the measurement scale, 
@@
 This stresses the need to define ecosystem assets that are appropriate in size. -->
+```
 
@@
 ::: callout-note
-**Example 3 -- Phytoplankton trophic index (PTI)**\
+**Example 3 -- Water Framework Directive, indicator ASPT**\
 
-*Variable*: Mean score of algal species present, based on a set of indicator species scored for phosphorus requirements/tolerance.
-The variable is recorded in water bodies (lakes of 0.5 km^2^ or more).
+*Variable*: ASPT (*average score per taxon*) is the mean eutrophication tolerance score of invertebrate species present in a river water body.\
 
-*Reference levels*: X~100~ = median variable value for water bodies in reference condition.
-X~0~, X~20~, X~40~, X~60~ and X~80~ = intercalibrated threshold values, based on dose–response curves.\
+*Reference levels*: X~100~ = median variable value recorded in reference water bodies.
+X~80~ = lower boundary of variable values in reference water bodies.
+X~0~ = natural zero (no species present).
+X~20~, X~40~, and X~60~ = intercalibrated threshold values, based on dose–response curves.\
 
-```{r read}
-#| include: false
-#| cache: false
-dat <- readxl::read_xlsx(here::here("data/PTI_data.xlsx")) |>
-  mutate(X = X |> str_replace_all("%", "\\\\%"))
-cn <- colnames(dat) |>
-    str_replace_all("2", "") |>
-    str_replace_all("X", "") |>
-    str_replace_all("RL", "X\\\\textsubscript{100}")
-```
+The Water Framework Directive (WFD) indicators use aggregation pathway 5 (@fig-pathways), 
+with truncation, scaling, and transformation, in that order.
+The value obtained after the first two steps is called an EQR (ecological quality ratio), 
+and the value obtained after the third step is called nEQR 
+(normalised EQR, and true indicator values *sensu* SEEA EA).
+Spatial aggregation, the way that we use the term here, is not done for the WFD.
+Instead, the aggregated reporting metric is the proportion of water bodies in each of the five condition classes 
+("bad", "poor", "moderate", "good", "high"), 
+separated by the non-terminal reference levels. \
 
-```{r tbl-pti}
-#| echo: false
-#| tbl-cap: "Dummy data for the PTI example.
-value = value for a hypothetical variable.
-agg.
-= spatial aggregation (arithmetic mean); trunc.
-truncation; stat.
-= spatial statistics.
-The transformation step is left out for simplicity.
-Pathways refer to figure 3."
-#| cache: false
-kbl(dat, format = "latex",
-    booktabs = TRUE,
-    col.names = cn,
-    escape = F
-    ) |>
-  kable_styling() %>%
-  add_header_above(c(
-    " " = 3, 
-    "Pathway 5" = 4,
-     "Pathway 6" = 3)) |>
-  collapse_rows(
-    columns = c(9,10),
-    latex_hline = "major"
-  )
-```
+When WFD data is put into use in other contexts, however, aggregation becomes important, 
+and it is crucial that aggregation is done after the (scaling and) transformation step, 
+because the transformation (using a stepwise-linear function) adds ecological meaning to the data.
+As the example in @fig-aspt shows, 
+the ecologically meaningful interpretation is that the EAA has "poor" condition at time 1, improving to "moderate" condition at time 2.
+When aggregation is erroneously done before transformation, however, 
+the same measurements suggest that condition is "bad" and deteriorating.
+This illustrates that an ecosystem condition indicator account cannot be derived directly from an ecosystem condition variable account in conjunction with reference levels, 
+as suggested by SEEA EA (@unitednations2024: 5.76 and table 5.3).
+The reason in the case of ASPT is that the transformation function is non-linear,
+and aggregation displacement therefore occurs in the spatial aggregation of variables and indicators.
+Reference levels are the same throughout the entire EAA, and variation in reference levels is thus not part of the problem here. \
 
-The Water Framework Directive (WFD) indicators uses aggregation pathway 5 (@fig-pathways), with truncation, scaling, and transformation, in that order (see worked example in @tbl-pti).
-The value obtained after the first two steps is called an EQR (ecological quality ratio), and the value obtained after the third step nEQR (normalised EQR).
-Spatial aggregation, the way that we use the term here, is not done for the WFD.
-Instead, the proportion of water bodies in each condition class is the aggregated reporting metric.
-Therefore aggregation displacement is not an issue.
-When WFD data is put into use in other contexts, however, aggregation becomes important.
+![Different aggregation pathways can result in different indicator values, exemplified by the ASPT indicator from the WFD.
+Two ecosystem assets (Rivers A and B) are measured at two points in time.
+Regular font numbers are EQR values; boldface numbers are nEQR values (indicator values *sensu* SEEA EA).
+When EQR values are first aggregated and then transformed ("transf."), ASPT incorrectly indicates a deteriorating and "bad" condition.
+When EQR values are first transformed and then aggregated, ASPT correctly indicates an improvement from "poor" to "moderate" condition.
+](images/ASPT.jpg){#fig-aspt}
 
-\
-Because of how X~100~ is defined based on the median value across reference lakes, overshooting values (\> X~100~) should be preserved in the spatial aggregation.
-Otherwise we get a negative displacement, making it practically impossible to reach an indicator value of 1.
-This means that neither EQR values nor nEQR values can be uncritically aggregated spatially to be used in a SEEA EA compliant in ECA.
-One solution, to enable the use of WFD indicators in ECAs, would be to change to pathway 6, with (1) scaling, (2) transforming, (3) aggregating and (4) truncating the values.
-In our hypothetical example (@tbl-pti) this gives an aggregated value of 0.85, whereas the mean value for the nEQRs would be 0.75 (i.e. negatively displaced).
+Another issue that can be illustrated with ASPT (or other WFD indicators) is the effect of truncation.
+Aggregation pathway 5 uses truncation as the first step. 
+When WFD data is used in ecosystem condition accounting, however, 
+truncation needs to be the final step, 
+applied only after aggregation. 
+This is because X~100~ is defined as the median of variable values measured in reference water bodies. 
+In other words, half of the values measured in reference water bodies is expected to be above X~100~. 
+When truncation is done as the first step, 
+the average indicator value of reference water bodies will be around 0.95, 
+i.e. it would be prevented from actually reaching 1.0. 
+Overshooting values therefore need to be preserved until after aggregation (spatial compensation),
+and only after aggregation should overshooting values be truncated to 1.0. \
+
+Likewise, in the case where model-based extrapolation of a WFD indicator is used, 
+it is important to postpone truncation until after the modelling and extrapolation. 
+Otherwise the distribution of the extrapolated values may be severely distorted, 
+at least if many of the original measurements are in or close to the X~100~ value. 
 :::
 
+```{=html}
 <!-- SEEA EA uses the term aggregation commutativity for the property that the order of vertical and horizontal aggregation is irrelevant to the end result.
@@
 This is possible when aggregating metrics using area weighting, and when one is aware of the need for post-stratification when using stratified sampling. -->
+```
 
+```{=html}
 <!-- Secondly, SEEA EA mentions several arithmetic or parametric aggregation functions, such as different means, median, min or max values.
@@
 -->
+```
 
@@
 #|   Flowchart showing pathways for the spatial aggregation of variables in ecosystem condition accounting.
-Normalisation refers to the steps that make a variable conform to the definition of an indicator in SEEA EA, which may include some, but not necesarily all the steps scale, truncate and transform.
+Normalisation refers to the steps that make a variable conform to the definition of an indicator in SEEA EA, which may include some, but not necessarily all, of the steps scale, truncate and transform.
 #| out-width: 400
@@
 
-Aggregation displacement causes a divergence in the type of information held in ecosystem condition variables versus indicators, 
-and spatially aggregated indicators should not be interpreted as normalised versions of aggregated variables.
+Aggregation displacement causes a divergence in the type of information held in ecosystem condition variables versus indicators, and spatially aggregated indicators should not be interpreted as normalised versions of aggregated variables.
 Therefore, aggregated variables and the associated aggregated indicators cannot act as alternative evidence or interpretations for the state of nature.

### 2025-09-23 manuscript/manuscript.qmd
table text size

@@
 and the value obtained after the third step is called nEQR 
-(normalised EQR, and true indicator values *sensu* SEEA EA).
+[normalised EQR, and true indicator values *sensu* SEEA EA; @van_de_bund_ecological_2007].
 Spatial aggregation, the way that we use the term here, is not done for the WFD.

### 2025-09-23 manuscript/terms_table.qmd
table text size

@@
     **Truncation** means to recode scaled variable above 1 to become 1, and values below 0 to become 0.**Transformation** means to adjust the indicator scale to reflect potentially nonlinear relationships between the variable and ecosystem condition.
-This is commonly done by anchoring specific variable values to predefined class boundaries (sensu WFD) which we call reference levels [@czúcz; @jakobsson_setting_2020], but may also be done without explicit reference levels, for example by using a sigmoid or exponential transformation [@mienna2024].
+This is commonly done by anchoring specific variable values to predefined class boundaries [sensu WFD; @directiv2000] which we call reference levels [@czúcz; @jakobsson_setting_2020], but may also be done without explicit reference levels, for example by using a sigmoid or exponential transformation [@mienna2024].
     Normalisation is frequently referred to as **rescaling** in the SEEA EA literature, but we avoid  this terms because of the need to provide better differentiation between the three steps of the normalisation process, where scaling is one of them.
@@
     X = "Thematic aggregation",
-    Y = "Also called vertical aggregation. In the case of ECA, this refers to an aggregation of two or more indicators into a composite index (_plural_: indices)."
+    Y = "Also called vertical aggregation. In the case of ECA, this refers to an aggregation of two or more indicators into a composite **index** (_plural_: **indices**)."
   ) |>
@@
 flextable(terms) |>
-  width(j = 1, width = 1) |>     # Relative width
-  width(j = 2, width = 5) |>     # Relative width  
+  width(j = 1, width = 1) |>     
+  width(j = 2, width = 5) |>     
   valign(j = 1:2, valign = "top", part = "all") |>
-  ftExtra::colformat_md(j = 2)
+  ftExtra::colformat_md(j = 2) |>
+  fontsize(size = 9, part = "all")
 

### 2025-12-01 manuscript/manuscript.qmd
change title

@@
 ---
-title: On the spatial aggregation of ecosystem condition indicators
+title: On the spatial aggregation of condition metrics for ecosystem accounting
 author:
@@
 However, this decision clearly affects the results of the spatial aggregation via a well-described (but often forgotten?) statistical problem referred to as the **modifiable area unit problem (MAUP)**.
-The MAUP can lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust (@sec-comp).
+The MAUP can lead to inconsistencies between repeated or nested ECAs, leading to confusion and mistrust.
 This is unfortunate, as it is important for decision makers to see that the information they are basing their decision on is robust and insensitive to aggregation methods [@vanbeurden1999].
@@
 For this, and all other aggregation pathways, the picture can be made more complicated by adding additional aggregation levels (i.e.
-a step-wise aggregation, @sec-comp)
+a step-wise aggregation).
 
@@
 If perfectly commutative then the two approaches will yield the same result.
-However, if the indicators that make up the indices use different area weights as part of their spatial aggregation (**Example 4**), then pathway 9 is the only one that gives the same index values across nested accounts (@sec-comp).
+However, if the indicators that make up the indices use different area weights as part of their spatial aggregation (**Example 4**), then pathway 9 is the only one that gives the same index values across nested accounts.
 
@@
 The description of indicator pathways can be made even easier by referencing specific aggregation pathways by number, as we have done in this paper using @fig-pathways in this paper.
-Methods used to change either the cover or the extent of the indicator data (see @sec-term) can also be included in the shorthand notation scheme.
+Methods used to change either the cover or the extent of the indicator data (see @tbl-terms) can also be included in the shorthand notation scheme.
 The three operations we conceive as possible options are extrapolation (extr.), interpolation (inter.), and (sub)sampling.
 We use the term spatial projection (projection for short) to encompass both extra- and interpolation.
-Thus, pathway 7 in Fig.
-3 can be written as $project - scale - sample - sp.agg. - trunc$.
+Thus, pathway 7 in @fig-pathways can be written as $project - scale - sample - sp.agg. - trunc$.
 The type of model used for the spatial projection can be added as a footnote.
@@
 However, internal consistency within assessments should be possible, and this can make it easier to communicate and interpret the information conveyed through the indicator values.
-This can be achieved by for example always scaling early (see recommendation number 2 above) and using the same aggregation levels (same spatial units) and same approach to step-wise aggregation across levels (@sec-comp), for all indicators.
+This can be achieved by for example always scaling early (see recommendation number 2 above) and using the same aggregation levels (same spatial units) and same approach to step-wise aggregation across levels, for all indicators.
 
@@
 
-...
-
 # References {.unnumbered}

### 2025-12-01 manuscript/manuscript.qmd
conclusion

@@
 #| fig-cap: |
-#|   Flowchart showing pathways for the spatial aggregation of variables in ecosystem condition accounting.
+#|   Flowchart showing  a selection of nine possible pathways for the spatial aggregation of variables in ecosystem condition accounting.
 Normalisation refers to the steps that make a variable conform to the definition of an indicator in SEEA EA, which may include some, but not necessarily all, of the steps scale, truncate and transform.
+The pathway description used the standardised shorthand aggregation athway terminology, where trunc.
+= truncation, trans.
+= transformation, sp.
+agg.
+= spatial aggregation, proj.
+= project (extrapolation or interpolation), sample = sub-sampling from a larger dataset, th.
+agg.
+= thematic aggregation.
 #| out-width: 400
@@
 # Conclusion
+This paper has described the reasons why we need to do spatial generalisation in ecosystem accounting, and 
+clarified several misconscptions about this process. 
+We have showed that indeitical raw data, when spatially aggregated through different pathways, can yield fundamentally different types of information about ecosystems condition. 
+The recommendations offered here is intended to help researchers 
+design ecosystem condition metrics that better align 
+with the type of information they aim to communicate from the data, 
+while also improving transparency through the use of a standardised terminology.
+With a clearer understanding of the challenges inherrent to spatial generalisation,
+we that hope future ecosystem condition accounts will be more internally consistent, more comparable between each other, and less susceptible to unintended mathematical artefacts.
 

### 2025-12-01 manuscript/manuscript.qmd

@@
 #| fig-cap: |
-#|   Examples showing the difference in indicator values (y-axis) obtained from taking the mean (V bar) of two variable values (V1 and V1) that are on the original scale (i hat) or on the normalised indicator scale (i bar).
+#|   Examples showing the difference in indicator values (y-axis) obtained from taking the mean of two variable values V1 and V1 ($\bar{V}$) that are on the original scale ($\hat{i}$) or on the normalised indicator scale ($\bar{i}$).
 Pane A: Due to truncation at X~100~, the latter results in a comparatively lower indicator value.
 Pane B: Example where the normalisation function includes a transformation step (see Figure 1 for definition) that anchors X~60~ to 0.6 on the indicator scale.
-This reverses the order of i hat and i bar compared to example in pane A.
+This reverses the order of $\hat{i}$ and $\bar{i}$ compared to example in pane A.
 The solid coloured line represents the normalisation function f(V).
@@
 Normalisation refers to the steps that make a variable conform to the definition of an indicator in SEEA EA, which may include some, but not necessarily all, of the steps scale, truncate and transform.
-The pathway description used the standardised shorthand aggregation athway terminology, where trunc.
+The pathway description uses the standardised shorthand aggregation pathway terminology, where trunc.
 = truncation, trans.
@@
 = thematic aggregation.
-#| out-width: 400
+#| out-width: 600
 #| echo: false

### 2025-12-02 manuscript/manuscript.qmd
final tweak

@@
 For this to be effective they need tools to accurately describe the current situation and to monitor trends in ecosystems over time.
-Ecosystem condition accounts (ECA) is one such tools that use variables and indicators to describe key ecosystem characteristics, reflecting their condition and deviations from a reference condition.
+Ecosystem condition accounts (ECA) is one such tool that use variables and indicators to describe key ecosystem characteristics, reflecting their condition and deviations from a reference condition.
 Because the purpose is to inform decisions at relatively high political levels, these metrics are often spatially aggregated to represent larger areas, such as countries.
@@
 The world is facing a nature crisis of a magnitude that requires urgent political decision making and international coordination [@ipbes_global_2019].
-Ecosystem accounting has been fronted as a framework for providing a knowledge base for this.
-EAs are environmental assessments aimed at communicating on the status and trends in the extent and condition of ecosystems at a scale and in a format that is relevant for decision makers [@edens2022, @comte2022].
+Ecosystem accounting has been fronted as a framework for providing a knowledge base for this [@bagstad_lessons_2021; @grondard_ecosystem_2021].
+Ecosystem accounts are environmental assessments aimed at communicating on the status and trends in the extent and condition of ecosystems at a scale and in a format that is relevant for decision makers [@edens2022; @comte2022].
 For ecosystem condition accounting (ECA) and assessments, this often requires that **metrics** (terms in bold are explained in @tbl-terms) used to describe the condition of ecosystems are **aggregated spatially** to produce a single value for a larger area.
@@
 When normalising at the plot scale in this case, one is essentially aggregating only the negative deviations from the upper reference level X~100~, and the aggregated indicator value becomes negatively displaced (0.55) relative to when aggregating variable values (0.9).
-Given that X~100~ is defined the way it is, this indicator should follow pathway 3 (@fig-pathways) by spatially aggregating the variable from plot scale to ecosystem asset to allow spatial compensation between plots.
+Given that X~100~ is defined the way it is, this indicator should follow pathway 3 (@tbl-pathways) by spatially aggregating the variable from plot scale to ecosystem asset to allow spatial compensation between plots.
 At the scale of ecosystem assets, the variable and the reference levels are at a commensurable scale, and the variable can be normalised and potentially aggregated further.
@@
 Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric, or harmonic means [@unitednations2024 §5.54].
-In addition, non-paramteric bootstrapping [@kolstad_area_2025] and Bayesian updating [@kolstad_making_2024] can be used to possibly obtain better descriptions of the uncertainty in the estimates.
+In addition, non-parametric bootstrapping [@kolstad_area_2025] and Bayesian updating [@kolstad_making_2024] can be used to possibly obtain better descriptions of the uncertainty in the estimates.
 If the inclusion probabilities are known, or they can be estimated, then the Horvitz-Thomson estimator for the sum or the mean (and variance) of a spatial sample is appropriate and useful, especially when the data originated from a stratified sample [@horvitz1952].
@@
 
-In the statistical standard and recommended guidelines for ecosystem accounts [@unitednations2024; -@unitednations2022], there is no mention of aggregation displacement, and there seems to be little awareness in general of its meening and the ramification of choosing different spatial aggregation pathways for variables.
+In the statistical standard and recommended guidelines for ecosystem accounts [@unitednations2024; -@unitednations2022], there is no mention of aggregation displacement, and there seems to be little awareness in general of its meaning and the ramification of choosing different spatial aggregation pathways for variables.
 Indicators and variables are sometimes presented side-by-side in ecosystem condition assessments or accounts.
-This is also recommended practice following the SEEA EA [@unitednations2024; ch. 5.3.4].
-
-<!--# Try add using !expr to get hat and bar symbols into caption:  -->
+This is also recommended practice following the SEEA EA [@unitednations2024, ch. 5.3.4].
 
@@
 A major determinant of the resulting aggregation displacement from the spatial aggregation of ecosystem condition indicators (@fig-agg1) is the choice about the order of the different steps in the normalisation and aggregation process, i.e.
-the aggregation pathway (@fig-pathways).
+the aggregation pathway (@tbl-pathways).
 If normalisation occurs early in the aggregation process, the amount of spatial compensation is limited (**Example 2**).
@@
 
-The Water Framework Directive (WFD) indicators use aggregation pathway 5 (@fig-pathways), 
+The Water Framework Directive (WFD) indicators use aggregation pathway 5 (@tbl-pathways), 
 with truncation, scaling, and transformation, in that order.
@@
 
-```{=html}
-<!-- SEEA EA uses the term aggregation commutativity for the property that the order of vertical and horizontal aggregation is irrelevant to the end result.
-A similar aggregation commutativity can be envisioned for step-wise aggregation of a single indicator, where it does not matter if the original data is aggregated directly from the original measurement scale to the Ecosystem Accounting Area (EAA) level, or via some intermittent scale, such as ecosystem assets or larger regions.
-This is possible when aggregating metrics using area weighting, and when one is aware of the need for post-stratification when using stratified sampling. -->
-```
-
-```{=html}
-<!-- Secondly, SEEA EA mentions several arithmetic or parametric aggregation functions, such as different means, median, min or max values.
-It does not mention the possibility of bootstrap resampling [@jakobsson2021; @kolstad_making_2024].
-This method allows for using probability distributions to represent indicator values and uncertainties, and to use weighted resampling to create new joint probability distributions for the spatially aggregated indicator.
--->
-```
-
 # Aggregation pathways {#sec-pathways}
@@
 We introduce the term aggregation pathway to describe the order of the steps used to go from a variable at the measurement scale, to a spatially aggregated metric, usually an indicator, that describes the ecosystem condition for a larger area.
-In @fig-pathways we show some aggregation pathways that we have come across, and some that we see as potential new pathways.
+In @tbl-pathways we show some aggregation pathways that we have come across, and some that we see as potential new pathways.
 The examples are non-exhaustive, and serve mainly as illustrations.
@@
 
-Pathways 8 and 9 describe alternative ways to create indices at different spatial scales, either by spatially aggregating the indices (pathway 8), or by aggregating the indicators separately and recalculating the indices at different spatial scales (pathway 9).
-If perfectly commutative then the two approaches will yield the same result.
-However, if the indicators that make up the indices use different area weights as part of their spatial aggregation (**Example 4**), then pathway 9 is the only one that gives the same index values across nested accounts.
+Pathways 8 and 9 describe alternative ways to create indices at different spatial scales, either by performing the spatial aggregation on the indices (i.e.
+after thematic aggregation; pathway 8) or on the indicators (i.e.
+before thematic aggregation; pathway 9).
+If all indicators making up an index are spatially aggregated using the same weights, typically the extent of the spatial units, then the aggregation operations will be commutative and it will not matter if you aggregate vertically then horizontally, or the other way around (@tbl-pathways; Pathway 8 and 9, respectively).
+However, indicators are not always valid for the entire ecosystem type that the account is made over.
+They can, for example, retain to only certain habitats or nature types within the ecosystem type, in which case one might want to use different weights when spatially aggregating the indicator that reflect the extent of only the relevant habitat or nature types, and not of the entire ecosystem.
+This can be achieved using pathway 9.
 
-```{r fig-pathways}
-#| fig-cap: |
-#|   Flowchart showing  a selection of nine possible pathways for the spatial aggregation of variables in ecosystem condition accounting.
+```{r tbl-pathways}
+#| tbl-cap: |
+#|   Table showing  a selection of nine possible pathways for the spatial aggregation of variables in ecosystem condition accounting.
 Normalisation refers to the steps that make a variable conform to the definition of an indicator in SEEA EA, which may include some, but not necessarily all, of the steps scale, truncate and transform.
@@
 These terms can be abbreviated as shown in the parentheses, and placed together into a single string, where operations performed simultaneously are enclosed in square brackets.
-For example, pathway 1 in @fig-pathways can be annotated as $scale - trunc. - trans. - sp. agg.$.
+For example, pathway 1 in @tbl-pathways can be annotated as $scale - trunc. - trans. - sp. agg.$.
 If the three normalisation steps are performed simultaneously this can be written as $[scale - trunc. - trans.] - sp. agg.$.
@@
 Note also that in the last example, normalisation only involved a scaling step.
-The description of indicator pathways can be made even easier by referencing specific aggregation pathways by number, as we have done in this paper using @fig-pathways in this paper.
+The description of indicator pathways can be made even easier by referencing specific aggregation pathways by number, as we have done in this paper using @tbl-pathways in this paper.
 Methods used to change either the cover or the extent of the indicator data (see @tbl-terms) can also be included in the shorthand notation scheme.
@@
 We use the term spatial projection (projection for short) to encompass both extra- and interpolation.
-Thus, pathway 7 in @fig-pathways can be written as $project - scale - sample - sp.agg. - trunc$.
+Thus, pathway 7 in @tbl-pathways can be written as $project - scale - sample - sp.agg. - trunc$.
 The type of model used for the spatial projection can be added as a footnote.
 
-Besides making the indicator workflow itself more transparent and reproducible, we believe this recommendation would help raise the general awareness about aggregation displacement in EAs, both among users and developers.
+Besides making the indicator workflow itself more transparent and reproducible, we believe this recommendation would help raise the general awareness about aggregation displacement in ecosystem accounting, both among users and developers.
 It will make the interpretation of ECAs less prone to misunderstanding and misuse, as well as making troubleshooting easier, for example when trying to identify the reason why two ECAs produce dissimilar indicator values from the same underlying data.
@@
 When spatially aggregating indices it is no longer possible to use different area weights to each single indicator.
-If all indicators making up an index are spatially aggregated using the same weights, typically the extent of the spatial units, then the aggregation operations will be commutative and it will not matter if you aggregate vertically then horizontally, or the other way around (@fig-pathways; Pathway 8 and 9, respectively).
-However, indicators are not always valid for the entire ecosystem type that the account is made over.
-They can, for example, retain to only certain habitat or nature types within the ecosystem type, in which case one might want to use different weights when spatially aggregating the indicator that reflect the extent of only the relevant habitat or nature types, and not of the entire ecosystem (see **Example 4**).
-But when combined with other indicators into a common index, any subsequent aggregation of that index will give equal spatial weight to all indicators (see next paragraph for an exception to this).
-We therefore recommend caution, and to default to the solution of instead spatially aggregating indicators separately, and then re-calculating indices when needed (@fig-pathways; Pathway 9).
+
+When spatially aggregating indices (@tbl-pathways; pathway 8) you loose this ability to give different weights to different indicators.
+The result can be that the aggregated index value will not be determinable from the aggregated indices, 
+and that  ECAs using the same indicator set, 
+but focusing on different spatial scales 
+(e.g. one one presneting regional values only, and anther also presenting local values), 
+will not get equal index values for the same areas, even though they use the same indicator set.
+We therefore recommend caution, and to default to the solution of instead of spatially aggregating indicators separately, and then re-calculating indices when needed (@tbl-pathways; Pathway 9).
 This will ensure that nested accounts will be comparable, i.e. that they will present the same index values independent of which spatial units the indices first are calculated for.
-
-An alternative solution to ensure agreement across nested accounts is to use weights also when performing the vertical or thematic aggregation of indicators into indicators, where indicators are weighted by the area that they are valid over.
-With this approach one can down-weigh indicators that only cover smaller areas (specific nature types etc.), or which have incomplete spatial coverage.
-When this weighting is already performed at the vertical aggregation stage, any subsequent horizontal aggregation does not need to account for the unique indicator weights again.
-However, weighting indicators in vertical aggregation opens a whole other can of worms, and therefore we do not offer this as a generic recommendation.
 
@@
 
+# CRediT statement
+
+**Anders Kolstad**: Conceptualization, Data Curation, Writing - Original Draft, Visualization, Funding acquisition.
+**Hanno Sandvik**: Conceptualization, Writing - Review & Editing, Visualization.
+**Bálint Czúcz**: Conceptualization, Writing - Review & Editing.
+**Chloé R. Nater**: Conceptualization, Writing - Review & Editing.
+
+# Funding
+
+This work received partial founding by the Norwegian Environment Agency (nr. 22047007).
+
 # References {.unnumbered}
